---
title: "markeR Based Signature Usage"
subtitle: "Prognosis Potential of SGScore using TCGA Data"
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---


# Introduction

In this document, I assess the potential of SGScore as a cancer prognosis tool. The SGScore was calculated using the markeR package. For more information, please refer to the original package developed by **Martins-Silva, R.** ([GitHub Link](https://github.com/DiseaseTranscriptomicsLab/markeR))

## Methodology


```{r, warning=F, message=F}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)
library(scater)
set.seed(1906)
```


# Preparation of TCGA data

First, I load the TCGA data. Please refer to the original resource in case of need, at the **GDC Portal** ([TCGA Link](https://portal.gdc.cancer.gov/)).

```{r}
#This object contains both counts and metadata. For preprocessing, please refer to markdown file "SIGNATURE_TCGA_pre.Rmd". 
TCGAobj_tissue <- readRDS("~/Projects/TCGA/TCGAobj_tissue.RDS")

#Not all TCGA tissues have enough number of tumour and normal samples to warrant a direct comparsion. Herein, I select only those that do (at least 5 samples of each)
tissuestocheck <- c("BLCA", "BRCA", "CHOL", "COAD", "ESCA", "HNSC", "READ", "STAD", "UCEC", "KICH", "KIRC", "KIRP", "LIHC", "LUAD", "LUSC", "PRAD", "THCA")

TCGAobj_tissue_TvsN <- TCGAobj_tissue[names(TCGAobj_tissue) %in% tissuestocheck]


#Preprocessing of the TCGA data. For simplicity, I keep only either tumour or normal samples, discarding only sample types. 
for (tissue in names(TCGAobj_tissue_TvsN)) {

  # Retrieve metadata and logCPM for the tissue
  metadata <- as.character(TCGAobj_tissue_TvsN[[tissue]]$metadata)
  logCPM <- TCGAobj_tissue_TvsN[[tissue]]$logCPM
  
  # Replace "01" with "Tumor" and "11" with "Normal"
  metadata[metadata == "01"] <- "Tumor"
  #metadata[metadata == "11"] <- "Normal"

  # Find the two most common tumour types (now updated to "Tumor" and "Normal")
  top_events <- names(sort(table(metadata), decreasing = TRUE))[1:2]

  # Filter metadata and logCPM to include only the two most common events
  keep_indices <- which(metadata %in% top_events)
  TCGAobj_tissue_TvsN[[tissue]]$metadata <- metadata[keep_indices]
  TCGAobj_tissue_TvsN[[tissue]]$logCPM <- logCPM[, keep_indices, drop = FALSE]
}



```

# Opening Signatures

In here, I load up the object with simulated signature scores. For information on how these score were obtained, please refer to the markdown file "Signature_markeR_bi.Rmd".

```{r}
Signatureobj <- readRDS("~/Projects/StressGranules/SGPAPER/Simulations/simulationlist.RDS")

# Signatureobj <- lapply(Signatureobj, function(tissue_list) {
#   tissue_list[!grepl("^SG2", names(tissue_list))]
# })
```

# Opening Metadata

Here, I load the corresponding metadata to my TCGA samples. Please refer to the original resource in case of need, at the **GDC Portal** ([TCGA Link](https://portal.gdc.cancer.gov/)).

```{r}
# Define the base folder path (update this to your actual path)
base_folder <- "/mnt/scratch/home/jacek/data/TCGA_tumours_only/Counts/"

# Get all files in the base folder
tissue_files <- list.files(base_folder, full.names = TRUE)

# Filter only files that contain "clinical_info" in their name
clinical_files <- grep("clinical_info", tissue_files, value = TRUE)

# Create an empty list to store data
Metadata <- list()

# Loop through each clinical file
for (file_path in clinical_files) {
  # Extract filename
  filename <- basename(file_path)
  
  # Extract tissue name (assuming format: "TCGA_TISSUE_Counts_clinical_info")
  tissue_name <- sub("^TCGA_(.*?)_Counts_clinical_info.*$", "\\1", filename)
  
  print(paste0("Loading file for tissue: ", tissue_name))
  
  # Read the file and store in the list
  Metadata[[tissue_name]] <- fread(file_path, sep = "\t", header = TRUE)
}

# Check the structure of the list
print(names(Metadata))  # Lists cleaned tissue names


```

### Extracting only relevant columnns

For simplicity, I will maintain only relevant columns. I opted to select direct columns that can help decipher how certain tumours react to therapy, or directly those that allow me to calculate survival.

```{r}
metacols <- c("barcode", "person_neoplasm_cancer_status", "primary_therapy_outcome_success", "ajcc_pathologic_stage","days_to_death", "days_to_last_followup", "vital_status")

i <- 1
for (tissue in names(Metadata)) {
  print(paste0("Filtering Metacols ", tissue, " (", i, "/", length(names(Metadata)), ")"))
  
  # Ensure Metadata[[tissue]] is a data.table
  if (!is.null(Metadata[[tissue]]) && is.data.table(Metadata[[tissue]])) {
    # Select only columns that exist
    available_cols <- intersect(metacols, colnames(Metadata[[tissue]]))
    Metadata[[tissue]] <- Metadata[[tissue]][, ..available_cols]
  } else {
    warning(paste("Skipping", tissue, "- Metadata is missing or not a data.table"))
  }
  
  i <- i + 1
}

```

# Therapy Response and Tumour Stage *Not Included

Starting with the therapy response. I also check the pathological state, as it could be that more advanced tumours could have more (or less) stress granules.

```{r fig.height=30, fig.width=20}
custom_palette <- qualitative_hcl(20, palette = "Pastel")
custom_palette <- sample(custom_palette,20)


library(ggpubr)
library(ggplot2)
library(dplyr)
library(rstatix)

# Function to filter out small or empty categories
# ✅ Fix: Remove both NA and empty ("") categories
filter_categories <- function(df, col, min_size = 5) {
  df <- df %>% 
    filter(!is.na(.data[[col]]), .data[[col]] != "")  # Remove NAs & empty strings
  
  category_counts <- table(df[[col]])  # Count occurrences per category
  valid_categories <- names(category_counts[category_counts >= min_size])  # Keep only large enough groups
  df <- df %>% filter(.data[[col]] %in% valid_categories)  # Filter out small categories
  return(df)
}

compute_metrics <- function(score_list, metadata) {
  results <- list()

  for (sig in names(score_list)) {
    score_df <- score_list[[sig]]

    # Ensure 'barcode' is in both before merging
    if (!"barcode" %in% colnames(score_df) || !"barcode" %in% colnames(metadata)) {
      warning("Skipping ", sig, ": 'barcode' missing in score data or metadata.")
      next
    }

    score_long <- reshape_scores(score_df, sig) %>%
      left_join(metadata, by = "barcode")

    for (col in setdiff(names(metadata), "barcode")) {
      test_data <- score_long %>%
        select(score, all_of(col)) %>%
        drop_na()

      if (n_distinct(test_data[[col]]) > 1) {  
        anova_result <- anova_test(data = test_data, score ~ .data[[col]])
        p_value <- anova_result$p
        cohen_f <- anova_result$cohen_f
      } else {
        p_value <- NA
        cohen_f <- NA
      }

      results[[sig]][[col]] <- list(p_value = p_value, cohen_f = cohen_f)
    }
  }

  return(results)
}

# Function to Plot Violin Plots for SG1 & SG2
plot_violin <- function(score_list, metadata, tissue, pointSize = 2, ColorValues = NULL) {
  plots <- list()  # Store plots for combining later
  
  # Ensure metadata contains 'barcode' before proceeding
  if (!"barcode" %in% colnames(metadata)) {
    stop("Error: 'barcode' column missing from metadata!")
  }

  # Keep only the two relevant metadata columns (check if they exist first)
  relevant_cols <- c("primary_therapy_outcome_success", "ajcc_pathologic_stage")
  relevant_cols <- relevant_cols[relevant_cols %in% colnames(metadata)]  # Ensure they exist
  if (length(relevant_cols) == 0) {
    warning("No relevant metadata columns found. Skipping tissue: ", tissue)
    return(NULL)
  }
  
  selected_metadata <- metadata %>% select(barcode, all_of(relevant_cols))

  for (sig in c("SG1")) {
    if (!sig %in% names(score_list)) next  # Skip if SG1 or SG2 is missing

    score_df <- score_list[[sig]]
    colnames(score_df)[1] <- "barcode"
    
    # Ensure score_df has a barcode column before proceeding
    if (!"barcode" %in% colnames(score_df)) {
      stop("Error: 'barcode' column missing from score data for signature: ", sig)
    }

    score_long <- merge(score_df, selected_metadata, by = "barcode")

    for (col in relevant_cols) {
      test_data <- score_long %>% 
        select(score, all_of(col)) %>%
        filter_categories(col)  # ✅ Remove small categories and blanks
      
      if (n_distinct(test_data[[col]]) > 1) {
        # Compute ANOVA and Cohen’s F
        formula <- reformulate(col, response = "score")
        anova_result <- anova_test(data = test_data, formula)
        
        p_value <- round(anova_result$p, 4)
        cohen_f <- round(sqrt(anova_result$ges / (1 - anova_result$ges)), 4)
        p_label <- paste0("\np = ", p_value, "\nCohen’s F = ", cohen_f)
        
        # Rename x-axis labels
        x_label <- ifelse(col == "primary_therapy_outcome_success", 
                          "Primary Therapy Outcome", "Pathologic Stage")

        # Create violin plot
        plot <- ggplot(test_data, aes(x = .data[[col]], y = score, fill = .data[[col]], color = .data[[col]])) +
          geom_violin(alpha = 0.5, scale = "width") +  # Violin plot
          geom_jitter(size = pointSize, alpha = 0.5, width = 0.2) +  # Jitter points
          ggtitle(paste0(sig, " - ", tissue, " (", x_label, ")")) +
          xlab(x_label) +
          theme_minimal() +
          annotate("text", x = 1.5, y = max(test_data$score, na.rm = TRUE), label = p_label, hjust = 0) +
          scale_fill_manual(values = ColorValues) +
          scale_color_manual(values = ColorValues) +  # Ensure points and violins use ColorValues
          theme(legend.position = "none")  # Hide legend for cleaner plots
        
        plots[[paste0(tissue, "_", sig, "_", col)]] <- plot  # Store plot
      }
    }
  }
  return(plots)  # Return list of plots
}

# Store all plots
all_plots <- list()
results_list <- list()
# Loop through tissues
for (tissue in names(Metadata)) {
  metadata <- Metadata[[tissue]]
  scores <- Signatureobj[[tissue]]
  
  # Generate violin plots and store them
  tissue_plots <- plot_violin(scores, metadata, tissue, ColorValues = custom_palette)
  
  if (!is.null(tissue_plots)) {
    all_plots <- c(all_plots, tissue_plots)  # Merge into master list
  }
  
  # ✅ FIX compute_metrics function call (remove "barcode")
  results_list[[tissue]] <- compute_metrics(scores, metadata)
}

# Arrange all plots in a panel (if there are any valid plots)
if (length(all_plots) > 0) {
  final_plot <- ggarrange(plotlist = all_plots, ncol = 2, nrow = ceiling(length(all_plots) / 2))
  print(final_plot)
} else {
  print("No valid plots generated.")
}

# tiff("~/Projects/StressGranules/SGPAPER/TherapyResponsePanel.tiff", width = 14, height=60, units = "in", res=300)
# final_plot
# dev.off()

```

# Survival Analysis

## Creating a joint "days" column

```{r}
i <- 1
for (tissue in names(Metadata)) {
  print(paste0("Merging Metadata ", tissue, " (", i, "/", length(names(Metadata)), ")"))
  
  # Ensure Metadata_Obj[[tissue]] is a dataframe
  if (!is.null(Metadata[[tissue]]) && is.data.frame(Metadata[[tissue]])) {
    # Replace NAs with 0 properly
    Metadata[[tissue]][is.na(Metadata[[tissue]])] <- 0

    # Add FollowTime column
    Metadata[[tissue]]$FollowTime <- Metadata[[tissue]]$days_to_death + Metadata[[tissue]]$days_to_last_followup
  } else {
    warning(paste("Skipping", tissue, "- Metadata_Obj is missing or not a dataframe"))
  }
  
  i <- i + 1
}

```
## TOP 25 vs Bottom 25

For the survival analysis I opted to always compare top 25% of signature score samples (SGScore and simulated) vs the bottom 25% of signature score samples.

```{r fig.height=70, fig.width=12, warning=FALSE}
# ✅ Initialize empty lists for results & plots
library(survival)
library(survminer)
Survival_Results <- list()
all_survival_plots <- list()

i <- 1
for (tissue in names(Metadata)) {
  print(paste0("Processing Tissue: ", tissue, " (", i, "/", length(names(Metadata)), ")"))

  data       <- Metadata[[tissue]]
  signatures <- names(Signatureobj[[tissue]])

  for (sig in signatures) {
    score_df <- Signatureobj[[tissue]][[sig]]
    colnames(score_df)[1] <- "barcode"

    if (!"barcode" %in% colnames(score_df) || !"barcode" %in% colnames(data)) {
      warning("Skipping ", sig, " in ", tissue, ": 'barcode' missing.")
      next
    }

    data_merged <- left_join(data, score_df, by = "barcode") %>%
      mutate(
        SGScore_quantile = ntile(score, 4),
        Group = case_when(
          SGScore_quantile == 1 ~ "Low",
          SGScore_quantile == 4 ~ "High",
          TRUE                  ~ NA_character_
        )
      ) %>%
      filter(!is.na(Group)) %>%
      mutate(
        FollowTime   = as.numeric(FollowTime),
        vital_status = ifelse(vital_status == "Dead", 1, 0),
        Group        = factor(Group, levels = c("Low", "High"))
      )

    if (nrow(data_merged) < 10) {
      warning("Skipping ", sig, " in ", tissue, ": Not enough survival data.")
      next
    }

    surv_object   <- Surv(time = data_merged$FollowTime, event = data_merged$vital_status)
    fit           <- survfit(surv_object ~ Group, data = data_merged)
    surv_diff     <- survdiff(surv_object ~ Group, data = data_merged)
    log_rank_pval <- 1 - pchisq(surv_diff$chisq, df = length(surv_diff$n) - 1)

    # ── HR from Cox model (Low as reference) ─────────────────────────────────
    cox_fit  <- coxph(surv_object ~ Group, data = data_merged)
    cox_sum  <- summary(cox_fit)
    HR       <- cox_sum$conf.int[1, "exp(coef)"]
    HR_low   <- cox_sum$conf.int[1, "lower .95"]
    HR_high  <- cox_sum$conf.int[1, "upper .95"]

    KM_plot <- ggsurvplot(
      fit, data = data_merged,
      conf.int     = TRUE,
      pval         = FALSE,
      risk.table   = TRUE,
      legend.title = paste(sig, "Group"),
      legend.labs  = c("Low", "High"),
      xlab         = "",
      ylab         = "",
      ggtheme      = theme_minimal()
    )

    KM_plot_final <- KM_plot$plot +
      ggtitle(paste0(tissue)) +
      scale_color_manual(values = c("High" = "#BA3B46", "Low" = "#53A2BE")) +
      scale_fill_manual( values = c("High" = "#BA3B46", "Low" = "#53A2BE")) +
      theme(
        axis.line        = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border     = element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        text             = element_text(size = 25, family = "Arial MS"),
        plot.title       = element_text(hjust = 0.5),
        plot.subtitle    = element_text(hjust = 0.5, face = "italic"),
        legend.position  = "none"
      )

    if (!tissue %in% names(Survival_Results)) Survival_Results[[tissue]] <- list()

    Survival_Results[[tissue]][[sig]] <- list(
      fit           = fit,
      log_rank_pval = log_rank_pval,
      HR            = HR,
      HR_low        = HR_low,
      HR_high       = HR_high,
      KM_plot       = KM_plot_final
    )

    all_survival_plots[[paste0(tissue, "_", sig)]] <- KM_plot_final
  }
  i <- i + 1
}

# ============================================================
# Empirical FPR for SG1 against SG1_Sim signatures
# ============================================================

fpr_results <- list()

for (tissue in names(Survival_Results)) {
  if (!"SG1" %in% names(Survival_Results[[tissue]])) next

  real_p <- Survival_Results[[tissue]][["SG1"]]$log_rank_pval

  random_sigs <- names(Survival_Results[[tissue]])[
    grepl("^SG1_[Ss]im", names(Survival_Results[[tissue]]))
  ]

  if (length(random_sigs) == 0) {
    message("No SG1_Sim signatures found for ", tissue)
    next
  }

  random_pvals <- sapply(random_sigs, function(sig) {
    Survival_Results[[tissue]][[sig]]$log_rank_pval
  })

  fpr <- mean(random_pvals <= real_p, na.rm = TRUE)

  fpr_results[[tissue]] <- data.frame(
    tissue         = tissue,
    real_p         = real_p,
    n_random       = length(random_pvals),
    n_more_extreme = sum(random_pvals <= real_p, na.rm = TRUE),
    FPR            = fpr
  )
}

fpr_df <- bind_rows(fpr_results)
print("Empirical FPR for SG1 vs simulated signatures:")
print(fpr_df)
saveRDS(fpr_df, "~/Projects/StressGranules/SGPAPER/Simulations/SG1_FPR.RDS")

# ============================================================
# BH correction + HR + FPR annotation for SG1
# ============================================================

sg1_pvals <- list()
for (tissue in names(Survival_Results)) {
  if ("SG1" %in% names(Survival_Results[[tissue]])) {
    sg1_pvals[[tissue]] <- Survival_Results[[tissue]][["SG1"]]$log_rank_pval
  }
}

if (length(sg1_pvals) > 0) {

  # ── Base R assignment to avoid dplyr scoping issues ──────────────────────
  sg1_pval_df <- data.frame(
    tissue = names(sg1_pvals),
    p_raw  = unname(unlist(sg1_pvals)),
    stringsAsFactors = FALSE
  )
  sg1_pval_df$p_adj_BH <- p.adjust(sg1_pval_df$p_raw, method = "BH")
  sg1_pval_df$sig      <- sg1_pval_df$p_adj_BH < 0.05

  # Merge FPR
  sg1_pval_df <- merge(sg1_pval_df,
                       fpr_df[, c("tissue", "FPR")],
                       by = "tissue", all.x = TRUE)

  print(sg1_pval_df)

  # ── Add annotation to each SG1 plot ──────────────────────────────────────
  for (tissue in sg1_pval_df$tissue) {

    adj_p   <- sg1_pval_df$p_adj_BH[sg1_pval_df$tissue == tissue]
    fpr_val <- sg1_pval_df$FPR[sg1_pval_df$tissue == tissue]
    HR      <- Survival_Results[[tissue]][["SG1"]]$HR
    HR_low  <- Survival_Results[[tissue]][["SG1"]]$HR_low
    HR_high <- Survival_Results[[tissue]][["SG1"]]$HR_high

    p_label <- paste0(
      "adj.p = ", formatC(adj_p,  format = "e", digits = 2), "\n",
      "FPR = ",   ifelse(is.na(fpr_val), "NA",
                         formatC(fpr_val, format = "f", digits = 2)), "\n",
      "HR = ",    round(HR, 2),
      " [", round(HR_low, 2), "-", round(HR_high, 2), "]"
    )

    Survival_Results[[tissue]][["SG1"]]$KM_plot <-
      Survival_Results[[tissue]][["SG1"]]$KM_plot +
      annotate(
        "text",
        x      = Inf, y = 0,
        label  = p_label,
        hjust  = 1.1,
        size   = 8,
        colour = "black"
      )

    key <- paste0(tissue, "_SG1")
    if (key %in% names(all_survival_plots)) {
      all_survival_plots[[key]] <- Survival_Results[[tissue]][["SG1"]]$KM_plot
    }
  }

  saveRDS(sg1_pval_df,
          "~/Projects/StressGranules/SGPAPER/Simulations/SG1_pval_corrected.RDS")

} else {
  message("SG1 not found in any tissue — no correction performed.")
}

# ── Combine and save ──────────────────────────────────────────────────────────
if (length(all_survival_plots) > 0) {
  final_survival_plot <- ggarrange(
    plotlist = all_survival_plots,
    ncol = 2, nrow = ceiling(length(all_survival_plots) / 2)
  )
  print(final_survival_plot)
}

for (tissue in names(Survival_Results)) {
  for (sig in names(Survival_Results[[tissue]])) {
    print(paste0(
      "Tissue: ", tissue, " | Signature: ", sig,
      " - Log-rank p-value: ", round(Survival_Results[[tissue]][[sig]]$log_rank_pval, 4)
    ))
  }
}

# ── Save only the SG1 plots (one per tissue, no Sim plots) ───────────────────
plots_to_keep <- all_survival_plots[grepl("_SG1$", names(all_survival_plots))]
saveRDS(plots_to_keep,
        "~/Projects/StressGranules/SGPAPER/Simulations/SurvivalResults_SG1.RDS")
```

## Separating Results

```{r}
# ✅ Initialize empty lists for split results
Survival_Results_SG1 <- list()
Survival_Results_SG2 <- list()

# ✅ Loop through each tissue
for (tissue in names(Survival_Results)) {
  
  # ✅ Split SG1-related signatures
  SG1_sigs <- names(Survival_Results[[tissue]])[grepl("^TNF", names(Survival_Results[[tissue]]))]
  if (length(SG1_sigs) > 0) {
    Survival_Results_SG1[[tissue]] <- Survival_Results[[tissue]][SG1_sigs]
  }
  
  # ✅ Split SG2-related signatures
  SG2_sigs <- names(Survival_Results[[tissue]])[grepl("^EMT", names(Survival_Results[[tissue]]))]
  if (length(SG2_sigs) > 0) {
    Survival_Results_SG2[[tissue]] <- Survival_Results[[tissue]][SG2_sigs]
  }
}

# ✅ Check if split worked correctly
print(paste0("Tissues in SG1 list: ", length(Survival_Results_SG1)))
print(paste0("Tissues in SG2 list: ", length(Survival_Results_SG2)))

# ✅ Print a summary
for (tissue in names(Survival_Results_SG1)) {
  print(paste0("Tissue: ", tissue, " - SG1 signatures: ", length(Survival_Results_SG1[[tissue]])))
}
for (tissue in names(Survival_Results_SG2)) {
  print(paste0("Tissue: ", tissue, " - SG2 signatures: ", length(Survival_Results_SG2[[tissue]])))
}

```

### Functions to plot

The biggest issue with cancer data, is that almost every gene set has prognostic value. Therefore, I opted to compare my signatures, with random simulated ones. I then calculate the False Positive Rate (FPR), the percentage of simulated signatures which obtain equal or more extreme values (significance) than my signatures. For more information on how these simulated signatures were created, please refer to the markdown file "Signature_markeR_Bi.Rmd".

```{r}
plot_survival_pvalues <- function(Survival_Results_Subset, title_for_plot, pointSize = 2, bigPointSize = 4, titlesize = 12, labsize = 10) {
  
  library(ggplot2)
  library(dplyr)
  
  # ✅ Extract p-values into a data frame
  pval_data <- do.call(rbind, lapply(names(Survival_Results_Subset), function(tissue) {
    sigs <- Survival_Results_Subset[[tissue]]
    data.frame(
      Tissue = tissue,
      Signature = names(sigs),
      PValue = sapply(sigs, function(x) x$log_rank_pval)
    )
  }))

  # ✅ Mark the original SG1/SG2 signatures
  pval_data$Type <- ifelse(pval_data$Signature %in% c("TNF", "EMT"), "Original", "Simulated")

  # ✅ Count how many simulated signatures have a lower p-value than each original signature in each tissue
  comparison_data <- pval_data %>%
    filter(Type == "Original") %>%
    left_join(
      pval_data %>% filter(Type == "Simulated"),
      by = "Tissue", suffix = c("_Original", "_Simulated")
    ) %>%
    filter(PValue_Simulated < PValue_Original) %>%
    group_by(Tissue, Signature_Original) %>%
    summarise(Simulations_Lower = n(), .groups = 'drop')

  # ✅ Print the result: How many simulations are lower than the original p-values
  if (nrow(comparison_data) > 0) {
    cat("Number of simulations with lower p-value than the original signature:\n")
    print(comparison_data)
  } else {
    cat("No simulations have a lower p-value than the original signatures.\n")
  }

  # ✅ Ensure tissues are plotted in correct alphabetical order
  pval_data$Tissue <- factor(pval_data$Tissue, levels = rev(sort(unique(pval_data$Tissue))))

  # ✅ Generate the plot **without coord_flip()**
  p <- ggplot(pval_data, aes(y = Tissue, x = PValue)) +  # Flip axes here manually
    
    # Violin plot for distribution
    geom_violin(fill = "lightblue", alpha = 0.5) +
    
    # Jitter points for simulated signatures (smaller)
    geom_jitter(data = pval_data %>% filter(Type == "Simulated"), 
                aes(color = Type), height = 0.2, size = pointSize, alpha = 0.7) +
    
    # Jitter points for original signatures (bigger, plotted on top)
    geom_jitter(data = pval_data %>% filter(Type == "Original"), 
                aes(color = Type), height = 0.2, size = bigPointSize) +
    
    # Color scheme for points
    scale_color_manual(values = c("Original" = "#9B1D20", "Simulated" = "#AAAAAA")) +
    
    # Labels and styling
    labs(title = title_for_plot, y = "Tissue", x = "Log-rank p-value") +  # Adjusted labels
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = labsize),  # Adjust y-axis text since it's now categorical
      plot.title = element_text(hjust = 0.5, size = titlesize),
      legend.title = element_blank()
    )+theme(legend.position="none") 

  print(p)
}


```

# P Value Plots

Finally, we plot all of this information on a graph.

```{r fig.height=16, fig.width=16}
# Plot for SG1-related signatures
plot_survival_pvalues(Survival_Results_SG1, "Survival Analysis - SG1")

plot_survival_pvalues(Survival_Results_SG2, "Survival Analysis - SG2")

# Arrange all plots in a panel
SG_Plots <- ggarrange(SG1, SG2, ncol = 2)  # Adjust layout as needed

# Display the panel
print(SG_Plots)

# tiff("~/Projects/StressGranules/SGPAPER/Images/SurvivalPlots.tiff", width = 16, height=16, units = "in", res=300)
# SG_Plots
# dev.off()
```

#NEW THESIS FIGURE

```{r}
ExpreTable <- TCGAobj_tissue_TvsN$SARC$logCPM
genes <- c("AGO2", "BRF1", "CAPRIN1", "CPEB1", "DDX6", "DISC1", "EIF2A", "FASTK", "FMR1", "FXR1", "G3BP1", "G3BP2", "L1TD1", "SAMD4A", "SMN1","TIA1", "TIAL1", "ZFP36")
library(survival)
library(survminer)
library(dplyr)
library(ggpubr)

# ============================================================
# 1. Setup — only 25/75 split
# ============================================================

splits <- list(
  "25/75" = list(low_q = 0.25, high_q = 0.75)
)

all_plots  <- list()
pval_store <- list()

# ============================================================
# 2. Loop — add coxph for HR
# ============================================================

for (gene in genes) {

  if (!gene %in% rownames(ExpreTable)) {
    warning("Skipping ", gene, ": not found in ExpreTable.")
    next
  }

  gene_expr <- as.numeric(ExpreTable[gene, ])
  names(gene_expr) <- colnames(ExpreTable)

  common_samples <- intersect(names(gene_expr), surv_meta$barcode)
  if (length(common_samples) < 10) {
    warning("Skipping ", gene, ": fewer than 10 matched samples.")
    next
  }

  expr_matched <- as.numeric(gene_expr[common_samples])
  meta_matched <- surv_meta[surv_meta$barcode %in% common_samples, ]
  meta_matched <- meta_matched[match(common_samples, meta_matched$barcode), ]

  low_cut  <- quantile(expr_matched, 0.25, na.rm = TRUE)
  high_cut <- quantile(expr_matched, 0.75, na.rm = TRUE)

  meta_matched$Group <- case_when(
    expr_matched <= low_cut  ~ "Low",
    expr_matched >= high_cut ~ "High",
    TRUE                     ~ NA_character_
  )

  plot_df <- meta_matched %>%
    mutate(
      FollowTime   = as.numeric(FollowTime),
      vital_status = ifelse(vital_status == "Dead", 1, 0),
      Group        = factor(Group, levels = c("Low", "High"))  # Low = reference
    ) %>%
    filter(!is.na(Group))

  if (nrow(plot_df) < 10) {
    warning("Skipping ", gene, " (25/75): not enough samples after split.")
    next
  }

  surv_obj   <- Surv(time = plot_df$FollowTime, event = plot_df$vital_status)
  fit        <- survfit(surv_obj ~ Group, data = plot_df)
  surv_diff  <- survdiff(surv_obj ~ Group, data = plot_df)
  log_rank_p <- 1 - pchisq(surv_diff$chisq, df = length(surv_diff$n) - 1)

  # HR from Cox model (Low as reference)
  cox_fit   <- coxph(surv_obj ~ Group, data = plot_df)
  cox_sum   <- summary(cox_fit)
  HR        <- cox_sum$conf.int[1, "exp(coef)"]
  HR_low    <- cox_sum$conf.int[1, "lower .95"]
  HR_high   <- cox_sum$conf.int[1, "upper .95"]

  pval_store[[gene]] <- log_rank_p
  all_plots[[gene]]  <- list(
    plot    = ggsurvplot(
      fit, data = plot_df,
      conf.int     = TRUE,
      pval         = FALSE,
      risk.table   = FALSE,
      legend.title = gene,
      legend.labs  = c("High", "Low"),
      xlab         = "",
      ylab         = "",
      ggtheme      = theme_minimal()
    )$plot,
    HR      = HR,
    HR_low  = HR_low,
    HR_high = HR_high
  )
}

# ============================================================
# 3. Multiple testing correction
# ============================================================

pval_df <- data.frame(
  gene  = names(pval_store),
  p_raw = unlist(pval_store),
  stringsAsFactors = FALSE
) %>%
  mutate(
    p_adj_BH = p.adjust(p_raw, method = "BH"),
    sig       = p_adj_BH < 0.05
  )

print(pval_df)

# ============================================================
# 4. Finalise plots — BH p + HR annotation, no raw p
# ============================================================

final_plots <- list()

for (gene in names(all_plots)) {

  adj_p   <- pval_df$p_adj_BH[pval_df$gene == gene]
  HR      <- all_plots[[gene]]$HR
  HR_low  <- all_plots[[gene]]$HR_low
  HR_high <- all_plots[[gene]]$HR_high
  p_col   <- ifelse(adj_p < 0.05, "#BA3B46", "grey40")

  p_label <- paste0(
    "adj.logrank p = ", formatC(adj_p, format = "e", digits = 2), "\n",
    "HR = ",     round(HR, 2),
    " [", round(HR_low, 2), "-", round(HR_high, 2), "]"
  )

  KM_final <- all_plots[[gene]]$plot +
    ggtitle(gene) +
    scale_color_manual(values = c("High" = "#BA3B46", "Low" = "#53A2BE")) +
    scale_fill_manual( values = c("High" = "#BA3B46", "Low" = "#53A2BE")) +
    annotate(
      "text",
      x      = Inf, y = Inf,
      label  = p_label,
      hjust  = 1.1, vjust = 1.4,
      size   = 8,
      colour = p_col
    ) +
    theme(
      axis.line        = element_line(colour = "black"),
      panel.grid       = element_blank(),
      panel.border     = element_blank(),
      panel.background = element_blank(),
      text             = element_text(size = 25, family = "Arial MS"),
      plot.title       = element_text(hjust = 0.5),
      plot.subtitle    = element_text(hjust = 0.5, face = "italic"),
      legend.position  = "none"
    )

  final_plots[[gene]] <- KM_final
}

# ============================================================
# 5. Arrange and save
# ============================================================

final_panel <- ggarrange(
  plotlist = final_plots,
  ncol     = 4,
  nrow     = ceiling(length(final_plots) / 4)
)

print(final_panel)


# ============================================================
# 6. Save
# ============================================================

png("~/Projects/StressGranules/SGPAPER/Images/PNG/SurvivalGenes.png",
    width  = 28,
    height = 12,
    units  = "in", res = 300)
print(final_panel)
dev.off()
```



# Session Info

```{r}
sessionInfo()
```
