---
title: "Analyses of GSE119977"
subtitle: "RNA partitioning into stress granules is based on the summation of multiple interactions"
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---


# Introduction

In this document, I present all analyses pertaining to the dataset **GSE119977**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE119977)),  
originally published by **T. Matheny, 2021**.

## Methodology

```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)


set.seed(1906)
```


## First Steps

```{r}

#This section here is to load up the genelenght file I will be using throughout the analyses.
genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)

genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")

genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()

genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")



# Get the table of read counts by indicating the path to the file. This file is what we obtained from the running featureCounts.

sg.read.counts <- read.table("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/featureCounts_results.txt", header=TRUE)

# Afterwards, I discard the columns which don't have read counts. These columns are only informations about the gene.
sg.read.counts <- sg.read.counts[,-c(2:6)]

biomart <- fread("~/biomart.txt")[,-1]


colnames(sg.read.counts)[1] <- "ensembl_gene_id_version"
colnames(biomart) <- c("ensembl_gene_id_version", "hgnc_symbol")

eltable <- merge(sg.read.counts,biomart,by = "ensembl_gene_id_version", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="") #this removes reads that dont have a corresponding hgnc symbol.

eltable$ensembl_gene_id_version <- NULL

read.counts <- eltable %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

read.counts <- as.data.frame(read.counts)

row.names(read.counts) <- read.counts$hgnc_symbol #changes rownames into hgnc symbols
read.counts$hgnc_symbol <- NULL #We no longer need this column, as we have the rownames.

```

## Section for SPINSIM Preparation

This serves as an example

```{r}
mof <- fread("~/Projects/StressGranules/Simulator/MolecularWeights.csv")
colnames(mof)[2] <- "GeneID" 
mof <- mof[!duplicated(mof$GeneID), ]


stress.tpm <- read.counts[,28]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/SG2021/SG2021SOGF_3.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)
```


## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, warning=FALSE, message=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(10, palette = "Pastel")
custom_palette <- sample(custom_palette, 10)

# Use SampleColouring from metadata
SampleColouring <- c("WT.SG.AR.PA", "WT.SG.AR.PA", "WT.SG.AR.PA", "WT.BU.AR", "WT.BU.AR", "WT.BU.AR", "WT.SG.SO.PA", "WT.SG.SO.PA", "WT.BU.SO", "WT.BU.SO", "KO.SG.SO.PA", "KO.SG.SO.PA", "KO.SG.SO.PA", "KO.BU.SO", "KO.BU.SO", "KO.BU.SO", "GF.BU.UN", "GF.BU.UN", "GF.BU.UN", "KO.BU.UN", "KO.BU.UN", "KO.BU.UN", "GF.SG.SO.G3", "GF.SG.SO.G3", "GF.SG.SO.G3", "GF.BU.SO", "GF.BU.SO", "GF.BU.SO")

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(4, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )
```


As we can see in these plots, in this case there are no individual sample outliers in terms of complexity or read depth. Excellent!

# Differential Expression Analysis

## Filtration

### ***Before Filtration***

Let us begin with the filtration. Personally, I find that its easier to do visually. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bi-modal distributions, with the first mode corresponding to lowly expressed genes. I remove these by cutting just before the second mode.

```{r, warning=FALSE}
alternew.WT.SG.Arsenite.PA <- log10(rowSums(read.counts[,1:3]))
alternew.WT.SG.Arsenite.PA <- as.data.frame(alternew.WT.SG.Arsenite.PA)

alternew.WT.Bulk.Arsenite <- log10(rowSums(read.counts[,4:6]))
alternew.WT.Bulk.Arsenite <- as.data.frame(alternew.WT.Bulk.Arsenite)

alternew.WT.SG.Sorbitol.PA <- log10(rowSums(read.counts[,7:8]))
alternew.WT.SG.Sorbitol.PA <- as.data.frame(alternew.WT.SG.Sorbitol.PA)

alternew.WT.Bulk.Sorbitol <- log10(rowSums(read.counts[,9:10]))
alternew.WT.Bulk.Sorbitol <- as.data.frame(alternew.WT.Bulk.Sorbitol)

alternew.KO.SG.Sorbitol.PA <- log10(rowSums(read.counts[,11:13]))
alternew.KO.SG.Sorbitol.PA <- as.data.frame(alternew.KO.SG.Sorbitol.PA)

alternew.KO.Bulk.Sorbitol <- log10(rowSums(read.counts[,14:16]))
alternew.KO.Bulk.Sorbitol <- as.data.frame(alternew.KO.Bulk.Sorbitol)

alternew.GFP.Unstressed <- log10(rowSums(read.counts[,17:19]))
alternew.GFP.Unstressed <- as.data.frame(alternew.GFP.Unstressed)

alternew.KO.Unstressed <- log10(rowSums(read.counts[,20:22]))
alternew.KO.Unstressed <- as.data.frame(alternew.KO.Unstressed)

alternew.GFP.SG.Sorbitol.G3 <- log10(rowSums(read.counts[,23:25]))
alternew.GFP.SG.Sorbitol.G3 <- as.data.frame(alternew.GFP.SG.Sorbitol.G3)

alternew.GFP.Bulk.Sorbitol <- log10(rowSums(read.counts[,26:28]))
alternew.GFP.Bulk.Sorbitol <- as.data.frame(alternew.GFP.Bulk.Sorbitol)

alternew <- log10(rowSums(read.counts))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 4, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.WT.SG.Arsenite.PA, aes(x=alternew.WT.SG.Arsenite.PA)) + geom_density(fill="#a2d1e6", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("WT Stress Granules - Arsenite - PABPC1")

ggplot(alternew.WT.Bulk.Arsenite, aes(x=alternew.WT.Bulk.Arsenite)) + geom_density(fill="#769fca", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("WT Whole Cell - Arsenite")

ggplot(alternew.WT.SG.Sorbitol.PA, aes(x=alternew.WT.SG.Sorbitol.PA)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("WT Stress Granules - Sorbitol - PABPC1")

ggplot(alternew.WT.Bulk.Sorbitol, aes(x=alternew.WT.Bulk.Sorbitol)) + geom_density(fill="#de425b", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("WT Whole Cell - Sorbitol")

ggplot(alternew.KO.SG.Sorbitol.PA, aes(x=alternew.KO.SG.Sorbitol.PA)) + geom_density(fill="#4e9c81", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("G3BP1 KO Stress Granules - Sorbitol - PABPC1")

ggplot(alternew.KO.Bulk.Sorbitol, aes(x=alternew.KO.Bulk.Sorbitol)) + geom_density(fill="#207567", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("G3BP1 KO Whole Cell - Sorbitol")

ggplot(alternew.GFP.Unstressed, aes(x=alternew.GFP.Unstressed)) + geom_density(fill="#878787", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("GFP-G3BP1 Whole Cell - Unstressed")

ggplot(alternew.KO.Unstressed, aes(x=alternew.KO.Unstressed)) + geom_density(fill="#555555", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("G3BP1 KO Whole Cell - Unstressed")

ggplot(alternew.GFP.SG.Sorbitol.G3, aes(x=alternew.GFP.SG.Sorbitol.G3)) + geom_density(fill="yellow", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("GFP-G3BP1 Stress Granules - Sorbitol - G3BP1")

ggplot(alternew.GFP.Bulk.Sorbitol, aes(x=alternew.GFP.Bulk.Sorbitol)) + geom_density(fill="orange", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("GFP-G3BP1 Whole Cell - Sorbitol")
```


By seeing the shape of the graph, we can see where we should cut. We have a **binomial distribution** (low reads and noise vs actual biological meaningful reads). The ideal place to cut them is right before the wedge.

### ***After Filtration***

I didn't do it this time, but I've maintained the code to filter by sample type. This time, I considered that it was more correct to filter all samples together.


```{r}
filterednew <- subset(read.counts, alternew>4)
filterednew.WTSGARPA <- subset(read.counts, alternew.WT.SG.Arsenite.PA>2)
filterednew.WTBUAR <- subset(read.counts, alternew.WT.Bulk.Arsenite>2.5)
filterednew.WTSGSOPA <- subset(read.counts, alternew.WT.SG.Sorbitol.PA>2)
filterednew.WTBUSO <- subset(read.counts, alternew.WT.Bulk.Sorbitol>2)
filterednew.KOSGSOPA <- subset(read.counts, alternew.KO.SG.Sorbitol.PA>2.5)
filterednew.KOBUSO <- subset(read.counts, alternew.KO.Bulk.Sorbitol>2.5)
filterednew.GFBUUN <- subset(read.counts, alternew.GFP.Unstressed>2.5)
filterednew.KOBUUN <- subset(read.counts, alternew.KO.Unstressed>2.5)
filterednew.GFSGSOG3 <- subset(read.counts, alternew.GFP.SG.Sorbitol.G3>2.75)
filterednew.GFBUSO <- subset(read.counts, alternew.GFP.Bulk.Sorbitol>2.5)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))
impgenes.WTSGARPA <- c(row.names(filterednew.WTSGARPA))
impgenes.WTBUAR <- c(row.names(filterednew.WTBUAR))
impgenes.WTSGSOPA <- c(row.names(filterednew.WTSGSOPA))
impgenes.WTBUSO <- c(row.names(filterednew.WTBUSO))
impgenes.KOSGSOPA <- c(row.names(filterednew.KOSGSOPA))
impgenes.KOBUSO <- c(row.names(filterednew.KOBUSO))
impgenes.GFBUUN <- c(row.names(filterednew.GFBUUN))
impgenes.KOBUUN <- c(row.names(filterednew.KOBUUN))
impgenes.GFSGSOG3 <- c(row.names(filterednew.GFSGSOG3))
impgenes.GFBUSO <- c(row.names(filterednew.GFBUSO))

#Now we need to select only the rows which have those genes.
finaltable <- read.counts[row.names(read.counts)%in%impgenes,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WTSGARPA,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WTBUAR,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WTSGSOPA,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WTBUSO,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.KOSGSOPA,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.KOBUSO,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.GFBUUN,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.KOBUUN,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.GFSGSOG3,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.GFBUSO,]

write.csv(finaltable, "~/Projects/StressGranules/Simulator/OriginalSamples/SG2021/Filtcounts.txt", row.names = T)
```
## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
forstress <- c("Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Unstressed", "Unstressed", "Unstressed", "Unstressed", "Unstressed", "Unstressed", "Sorbitol", "Sorbitol","Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol")

forstress.simple <- c("Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "Unstressed", "Unstressed", "Unstressed", "Unstressed", "Unstressed", "Unstressed", "Stress", "Stress","Stress", "Stress", "Stress", "Stress")

forsample <- c("SG", "SG", "SG", "Bulk", "Bulk", "Bulk", "SG", "SG", "Bulk", "Bulk", "SG", "SG", "SG", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "SG", "SG","SG", "Bulk", "Bulk", "Bulk")

forcelltype <- c("WT", "WT", "WT", "WT", "WT", "WT", "WT", "WT", "WT", "WT", "KO", "KO", "KO", "KO", "KO", "KO", "GFP", "GFP", "GFP", "KO", "KO", "KO", "GFP", "GFP","GFP", "GFP", "GFP", "GFP")

#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)


#for the others, we need to select only relevant columns. 

#Then we calculate the normalization factors.
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)


#Now we do our matrix itself
mm <- model.matrix(~0+SampleColouring)


#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```

As we can see, the **mean-variance trend** shows a good curve, showing that the filtration was well done.

## Normalisation

For normalization I convert the read counts into log2CPMs. Its a good way to visualize how the reads are. 

```{r}

dfforboxy <- finaltable

orderforplots <- c("WT.SG.Arsenite.PA", "WT.SG.Arsenite.PA", "WT.SG.Arsenite.PA", "WT.Bulk.Arsenite", "WT.Bulk.Arsenite", "WT.Bulk.Arsenite", "WT.SG.Sorbitol.PA", "WT.SG.Sorbitol.PA", "WT.Bulk.Sorbitol", "WT.Bulk.Sorbitol", "KO.SG.Sorbitol.PA", "KO.SG.Sorbitol.PA", "KO.SG.Sorbitol.PA", "KO.Bulk.Sorbitol", "KO.Bulk.Sorbitol", "KO.Bulk.Sorbitol", "GFP.Unstressed", "GFP.Unstressed", "GFP.Unstressed", "KO.Unstressed", "KO.Unstressed", "KO.Unstressed", "GFP.SG.Sorbitol.G3", "GFP.SG.Sorbitol.G3","GFP.SG.Sorbitol.G3", "GFP.Bulk.Sorbitol", "GFP.Bulk.Sorbitol", "GFP.Bulk.Sorbitol")
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#a2d1e6", "#769fca", "#e88184", "#de425b", "#4e9c81", "#207567", "#878787", "#555555", "yellow", "orange"), labels=c("WT.SG.Arsenite.PA", "WT.Bulk.Arsenite", "WT.SG.Sorbitol.PA", "WT.Bulk.Sorbitol", "KO.SG.Sorbitol.PA", "KO.Bulk.Sorbitol", "GFP.Unstressed", "KO.Unstressed", "GFP.SG.Sorbitol.G3", "GFP.Bulk.Sorbitol"), breaks =c("WT.SG.Arsenite.PA", "WT.Bulk.Arsenite", "WT.SG.Sorbitol.PA", "WT.Bulk.Sorbitol", "KO.SG.Sorbitol.PA", "KO.Bulk.Sorbitol", "GFP.Unstressed", "KO.Unstressed", "GFP.SG.Sorbitol.G3", "GFP.Bulk.Sorbitol")) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

#Using the Standard Matrix

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#a2d1e6", "#769fca", "#e88184", "#de425b", "#4e9c81", "#207567", "#878787", "#555555", "yellow", "orange"),labels=c("WT.SG.Arsenite.PA", "WT.Bulk.Arsenite", "WT.SG.Sorbitol.PA", "WT.Bulk.Sorbitol", "KO.SG.Sorbitol.PA", "KO.Bulk.Sorbitol", "GFP.Unstressed", "KO.Unstressed", "GFP.SG.Sorbitol.G3", "GFP.Bulk.Sorbitol"), breaks =c("WT.SG.Arsenite.PA", "WT.Bulk.Arsenite", "WT.SG.Sorbitol.PA", "WT.Bulk.Sorbitol", "KO.SG.Sorbitol.PA", "KO.Bulk.Sorbitol", "GFP.Unstressed", "KO.Unstressed", "GFP.SG.Sorbitol.G3", "GFP.Bulk.Sorbitol")) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```

## PCA

Before we move on, lets do a **Principal Component analysis**. It is a good exploratory tool to find out how our samples relate to each other and if there are any obvious outliers. We would expect our replicates to be nearer to one another than to other samples, and more different samples should be further away. 

```{r, fig.height=7, fig.width=9}
# Define color palette
custom_palette <- qualitative_hcl(4, palette = "Pastel")
custom_palette <- sample(custom_palette, 4)

#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- forsample
dfforpca$Stress <- forstress
dfforpca$Cell <- forcelltype

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-3)], scale = FALSE)

p_Sample <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Sample") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Stress <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Stress", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Stress", values = custom_palette) +
  ggtitle("PCA - Coloured by Stress") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Cell <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Cell", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Cell", values = custom_palette) +
  ggtitle("PCA - Coloured by Cell") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Sample
p_Stress
p_Cell

#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])


```
As one can observe from the PCAs, our PC1 is mostly explained by whether a sample is SG or WC.

# Simple Linear Models

In these linear models, I simply compare samples directly.

```{r}

mm <- model.matrix(~0+SampleColouring)

D0 <- DGEList(finaltable)

#Then we calculate the normalization factors for these new generated data frames
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)

y <- voom(D0, mm, plot = T)

fitted <- lmFit(y, mm)

head(coef(fitted))

WT.AR.PA <- makeContrasts(SampleColouringWT.SG.AR.PA  - SampleColouringWT.BU.AR , levels = colnames(coef(fitted)))
WT.SO.PA <- makeContrasts(SampleColouringWT.SG.SO.PA  - SampleColouringWT.BU.SO , levels = colnames(coef(fitted)))
KO.SO.PA <- makeContrasts(SampleColouringKO.SG.SO.PA  - SampleColouringKO.BU.SO , levels = colnames(coef(fitted)))
GF.SO.G3 <- makeContrasts(SampleColouringGF.SG.SO.G3  - SampleColouringGF.BU.SO , levels = colnames(coef(fitted)))

tmp_WT.AR.PA <- contrasts.fit(fitted, WT.AR.PA)
tmp_WT.SO.PA <- contrasts.fit(fitted, WT.SO.PA)
tmp_KO.SO.PA <- contrasts.fit(fitted, KO.SO.PA)
tmp_GF.SO.G3 <- contrasts.fit(fitted, GF.SO.G3)

tmp_WT.AR.PA <- eBayes(tmp_WT.AR.PA)
tmp_WT.SO.PA <- eBayes(tmp_WT.SO.PA)
tmp_KO.SO.PA <- eBayes(tmp_KO.SO.PA)
tmp_GF.SO.G3 <- eBayes(tmp_GF.SO.G3)

top_WT.AR.PA <- topTable(tmp_WT.AR.PA, sort.by = "logFC", n=Inf)
top_WT.SO.PA <- topTable(tmp_WT.SO.PA, sort.by = "logFC", n=Inf)
top_KO.SO.PA <- topTable(tmp_KO.SO.PA, sort.by = "logFC", n=Inf)
top_GF.SO.G3 <- topTable(tmp_GF.SO.G3, sort.by = "logFC", n=Inf)

DEGenes_WT.AR.PA <- top_WT.AR.PA
DEGenes_WT.SO.PA <- top_WT.SO.PA
DEGenes_KO.SO.PA <- top_KO.SO.PA
DEGenes_GF.SO.G3 <- top_GF.SO.G3

ndegs <- 20
```

# SG vs WC (WT Cells, Arsenite, PABPC1-IP)

## Volcano Plot

```{r}
write.table(x = DEGenes_WT.AR.PA,
            file = "~/Projects/StressGranules/SGPAPER/SGvsWC/TMatheny2021_WTARPA.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes_WT.AR.PA[DEGenes_WT.AR.PA$logFC>3 & DEGenes_WT.AR.PA$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_WT.AR.PA[DEGenes_WT.AR.PA$logFC< -3 & DEGenes_WT.AR.PA$adj.P.Val<0.05,] )

ggplot(data=DEGenes_WT.AR.PA,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_WT.AR.PA$Gene <- row.names(DEGenes_WT.AR.PA)
colnames(genelength) <- c("Length","Gene")
DEGenes_WT.AR.PA <- merge(DEGenes_WT.AR.PA, genelength, by="Gene")

ggplot(data=DEGenes_WT.AR.PA,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_WT.AR.PA)
liner.SG <- rstandard(smooth.SG)
DEGenes_WT.AR.PA <-cbind(DEGenes_WT.AR.PA, liner.SG)
colnames(DEGenes_WT.AR.PA)[9] <- "Residuals"
row.names(DEGenes_WT.AR.PA) <- DEGenes_WT.AR.PA$Gene
linemarkers <- as.data.frame(DEGenes_WT.AR.PA[abs(DEGenes_WT.AR.PA$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_WT.AR.PA$t, log10(DEGenes_WT.AR.PA$Length), method = "spearman")



ggplot(DEGenes_WT.AR.PA, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs WC") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

A massive lenght bias is seen.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_WT.AR.PA <- DEGenes_WT.AR.PA[order(DEGenes_WT.AR.PA[,4]),]
gseatool <- DEGenes_WT.AR.PA[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_WT.AR.PA$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# SG vs WC (WT Cells, Sorbitol, PABPC1-IP)

## Volcano Plot

```{r}
write.table(x = DEGenes_WT.SO.PA,
            file = "~/Projects/StressGranules/SGPAPER/SGvsWC/TMatheny2021_WTSOPA.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes_WT.SO.PA[DEGenes_WT.SO.PA$logFC>3 & DEGenes_WT.SO.PA$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_WT.SO.PA[DEGenes_WT.SO.PA$logFC< -3 & DEGenes_WT.SO.PA$adj.P.Val<0.05,] )

ggplot(data=DEGenes_WT.SO.PA,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_WT.SO.PA$Gene <- row.names(DEGenes_WT.SO.PA)
colnames(genelength) <- c("Length","Gene")
DEGenes_WT.SO.PA <- merge(DEGenes_WT.SO.PA, genelength, by="Gene")

ggplot(data=DEGenes_WT.SO.PA,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_WT.SO.PA)
liner.SG <- rstandard(smooth.SG)
DEGenes_WT.SO.PA <-cbind(DEGenes_WT.SO.PA, liner.SG)
colnames(DEGenes_WT.SO.PA)[9] <- "Residuals"
row.names(DEGenes_WT.SO.PA) <- DEGenes_WT.SO.PA$Gene
linemarkers <- as.data.frame(DEGenes_WT.SO.PA[abs(DEGenes_WT.SO.PA$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_WT.SO.PA$t, log10(DEGenes_WT.SO.PA$Length), method = "spearman")



ggplot(DEGenes_WT.SO.PA, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs WC") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

A massive lenght bias is seen.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_WT.SO.PA <- DEGenes_WT.SO.PA[order(DEGenes_WT.SO.PA[,4]),]
gseatool <- DEGenes_WT.SO.PA[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_WT.SO.PA$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


# SG vs WC (G3BP KO Cells, Sorbitol, PABPC1-IP)

## Volcano Plot

```{r}
write.table(x = DEGenes_KO.SO.PA,
            file = "~/Projects/StressGranules/SGPAPER/SGvsWC/TMatheny2021_KOSOPA.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes_KO.SO.PA[DEGenes_KO.SO.PA$logFC>3 & DEGenes_KO.SO.PA$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_KO.SO.PA[DEGenes_KO.SO.PA$logFC< -3 & DEGenes_KO.SO.PA$adj.P.Val<0.05,] )

ggplot(data=DEGenes_KO.SO.PA,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_KO.SO.PA$Gene <- row.names(DEGenes_KO.SO.PA)
colnames(genelength) <- c("Length","Gene")
DEGenes_KO.SO.PA <- merge(DEGenes_KO.SO.PA, genelength, by="Gene")

ggplot(data=DEGenes_KO.SO.PA,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_KO.SO.PA)
liner.SG <- rstandard(smooth.SG)
DEGenes_KO.SO.PA <-cbind(DEGenes_KO.SO.PA, liner.SG)
colnames(DEGenes_KO.SO.PA)[9] <- "Residuals"
row.names(DEGenes_KO.SO.PA) <- DEGenes_KO.SO.PA$Gene
linemarkers <- as.data.frame(DEGenes_KO.SO.PA[abs(DEGenes_KO.SO.PA$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_KO.SO.PA$t, log10(DEGenes_KO.SO.PA$Length), method = "spearman")



ggplot(DEGenes_KO.SO.PA, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs WC") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

A massive lenght bias is seen.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_KO.SO.PA <- DEGenes_KO.SO.PA[order(DEGenes_KO.SO.PA[,4]),]
gseatool <- DEGenes_KO.SO.PA[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_KO.SO.PA$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```



# SG vs WC (G3BP-GFP Cells, Sorbitol, GFP-IP)

## Volcano Plot

```{r}
write.table(x = DEGenes_GF.SO.G3,
            file = "~/Projects/StressGranules/SGPAPER/SGvsWC/TMatheny2021_GFSOG3.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes_GF.SO.G3[DEGenes_GF.SO.G3$logFC>3 & DEGenes_GF.SO.G3$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_GF.SO.G3[DEGenes_GF.SO.G3$logFC< -3 & DEGenes_GF.SO.G3$adj.P.Val<0.05,] )

ggplot(data=DEGenes_GF.SO.G3,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_GF.SO.G3$Gene <- row.names(DEGenes_GF.SO.G3)
colnames(genelength) <- c("Length","Gene")
DEGenes_GF.SO.G3 <- merge(DEGenes_GF.SO.G3, genelength, by="Gene")

ggplot(data=DEGenes_GF.SO.G3,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_GF.SO.G3)
liner.SG <- rstandard(smooth.SG)
DEGenes_GF.SO.G3 <-cbind(DEGenes_GF.SO.G3, liner.SG)
colnames(DEGenes_GF.SO.G3)[9] <- "Residuals"
row.names(DEGenes_GF.SO.G3) <- DEGenes_GF.SO.G3$Gene
linemarkers <- as.data.frame(DEGenes_GF.SO.G3[abs(DEGenes_GF.SO.G3$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_GF.SO.G3$t, log10(DEGenes_GF.SO.G3$Length), method = "spearman")



ggplot(DEGenes_GF.SO.G3, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs WC") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

A massive lenght bias is seen.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_GF.SO.G3 <- DEGenes_GF.SO.G3[order(DEGenes_GF.SO.G3[,4]),]
gseatool <- DEGenes_GF.SO.G3[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_GF.SO.G3$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# Complex Linear Models

In these type of linear models, I take into account the different variables, such as stress and genotype.

```{r}

metadata <- data.frame(Sample=forsample, Stress=forstress, Cell=forcelltype)
metadata$Stress <- factor(metadata$Stress, levels = c("Unstressed", "Arsenite", "Sorbitol"))
metadata$Sample <- factor(metadata$Sample, levels = c("Bulk", "SG"))
metadata$Cell <- factor(metadata$Cell, levels = c("WT", "KO", "GFP"))

mm <- model.matrix(~0+Sample+Stress+Cell,metadata)

D0 <- DGEList(finaltable)

#Then we calculate the normalization factors for these new generated data frames
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)

y <- voom(D0, mm, plot = T)

fitted <- lmFit(y, mm)

head(coef(fitted))

SG.vs.WC <- makeContrasts(SampleSG  - SampleBulk , levels = colnames(coef(fitted)))

tmp_SG.vs.WC <- contrasts.fit(fitted, SG.vs.WC)
tmp_SG.vs.WC <- eBayes(tmp_SG.vs.WC)
tmp <- eBayes(fitted)

top_SG.vs.WC <- topTable(tmp_SG.vs.WC, sort.by = "logFC", n=Inf)
DEGenes_SG.vs.WC <- top_SG.vs.WC
top_Arse <- topTable(tmp,3, sort.by = "logFC", n=Inf)
DEGenes_Arse <- top_Arse
top_Sorb <- topTable(tmp,4, sort.by = "logFC", n=Inf)
DEGenes_Sorb <- top_Sorb
top_KO <- topTable(tmp,5, sort.by = "logFC", n=Inf)
DEGenes_KO <- top_KO
top_GFP <- topTable(tmp,6, sort.by = "logFC", n=Inf)
DEGenes_GFP <- top_GFP
ndegs=20
```
# SG vs WC

## Volcano Plot

```{r}
write.table(x = DEGenes_SG.vs.WC,
            file = "~/Projects/StressGranules/SGPAPER/SGvsWC/TMatheny2021.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes_SG.vs.WC[DEGenes_SG.vs.WC$logFC>3 & DEGenes_SG.vs.WC$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_SG.vs.WC[DEGenes_SG.vs.WC$logFC< -3 & DEGenes_SG.vs.WC$adj.P.Val<0.05,] )

ggplot(data=DEGenes_SG.vs.WC,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_SG.vs.WC$Gene <- row.names(DEGenes_SG.vs.WC)
colnames(genelength) <- c("Length","Gene")
DEGenes_SG.vs.WC <- merge(DEGenes_SG.vs.WC, genelength, by="Gene")

ggplot(data=DEGenes_SG.vs.WC,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_SG.vs.WC)
liner.SG <- rstandard(smooth.SG)
DEGenes_SG.vs.WC <-cbind(DEGenes_SG.vs.WC, liner.SG)
colnames(DEGenes_SG.vs.WC)[9] <- "Residuals"
row.names(DEGenes_SG.vs.WC) <- DEGenes_SG.vs.WC$Gene
linemarkers <- as.data.frame(DEGenes_SG.vs.WC[abs(DEGenes_SG.vs.WC$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_SG.vs.WC$t, log10(DEGenes_SG.vs.WC$Length), method = "spearman")



ggplot(DEGenes_SG.vs.WC, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs WC") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
A massive lenght bias is seen.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_SG.vs.WC <- DEGenes_SG.vs.WC[order(DEGenes_SG.vs.WC[,4]),]
gseatool <- DEGenes_SG.vs.WC[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_SG.vs.WC$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# Arsenite Stress

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
upmarkers <- as.data.frame(DEGenes_Arse[DEGenes_Arse$logFC>5 & DEGenes_Arse$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_Arse[DEGenes_Arse$logFC< -3 & DEGenes_Arse$adj.P.Val<0.05,] )

ggplot(data=DEGenes_Arse,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Arsenite")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_Arse$Gene <- row.names(DEGenes_Arse)
colnames(genelength) <- c("Length","Gene")
DEGenes_Arse <- merge(DEGenes_Arse, genelength, by="Gene")

ggplot(data=DEGenes_Arse,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Arsenite")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_Arse)
liner.SG <- rstandard(smooth.SG)
DEGenes_Arse <-cbind(DEGenes_Arse, liner.SG)
colnames(DEGenes_Arse)[9] <- "Residuals"
row.names(DEGenes_Arse) <- DEGenes_Arse$Gene
linemarkers <- as.data.frame(DEGenes_Arse[abs(DEGenes_Arse$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_Arse$t, log10(DEGenes_Arse$Length), method = "spearman")



ggplot(DEGenes_Arse, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("Arsenite") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
A minor negative lenght bias is seen.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_Arse <- DEGenes_Arse[order(DEGenes_Arse[,4]),]
gseatool <- DEGenes_Arse[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_Arse$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# Sorbitol Stress

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
upmarkers <- as.data.frame(DEGenes_Sorb[DEGenes_Sorb$logFC>4 & DEGenes_Sorb$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_Sorb[DEGenes_Sorb$logFC< -3 & DEGenes_Sorb$adj.P.Val<0.05,] )

ggplot(data=DEGenes_Sorb,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Sorbitol")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_Sorb$Gene <- row.names(DEGenes_Sorb)
colnames(genelength) <- c("Length","Gene")
DEGenes_Sorb <- merge(DEGenes_Sorb, genelength, by="Gene")

ggplot(data=DEGenes_Sorb,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Sorbitol")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_Sorb)
liner.SG <- rstandard(smooth.SG)
DEGenes_Sorb <-cbind(DEGenes_Sorb, liner.SG)
colnames(DEGenes_Sorb)[9] <- "Residuals"
row.names(DEGenes_Sorb) <- DEGenes_Sorb$Gene
linemarkers <- as.data.frame(DEGenes_Sorb[abs(DEGenes_Sorb$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_Sorb$t, log10(DEGenes_Sorb$Length), method = "spearman")



ggplot(DEGenes_Sorb, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("Sorbitol") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
A major negative lenght bias is seen.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_Sorb <- DEGenes_Sorb[order(DEGenes_Sorb[,4]),]
gseatool <- DEGenes_Sorb[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_Sorb$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# KO Effect

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
upmarkers <- as.data.frame(DEGenes_KO[DEGenes_KO$logFC>3 & DEGenes_KO$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_KO[DEGenes_KO$logFC< -3 & DEGenes_KO$adj.P.Val<0.05,] )

ggplot(data=DEGenes_KO,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("KO Effect")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_KO$Gene <- row.names(DEGenes_KO)
colnames(genelength) <- c("Length","Gene")
DEGenes_KO <- merge(DEGenes_KO, genelength, by="Gene")

ggplot(data=DEGenes_KO,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("KO")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_KO)
liner.SG <- rstandard(smooth.SG)
DEGenes_KO <-cbind(DEGenes_KO, liner.SG)
colnames(DEGenes_KO)[9] <- "Residuals"
row.names(DEGenes_KO) <- DEGenes_KO$Gene
linemarkers <- as.data.frame(DEGenes_KO[abs(DEGenes_KO$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_KO$t, log10(DEGenes_KO$Length), method = "spearman")



ggplot(DEGenes_KO, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("KO") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_KO <- DEGenes_KO[order(DEGenes_KO[,4]),]
gseatool <- DEGenes_KO[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_KO$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# GFP Effect

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
upmarkers <- as.data.frame(DEGenes_GFP[DEGenes_GFP$logFC>5 & DEGenes_GFP$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_GFP[DEGenes_GFP$logFC< -3 & DEGenes_GFP$adj.P.Val<0.05,] )

ggplot(data=DEGenes_GFP,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("GFP")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_GFP$Gene <- row.names(DEGenes_GFP)
colnames(genelength) <- c("Length","Gene")
DEGenes_GFP <- merge(DEGenes_GFP, genelength, by="Gene")

ggplot(data=DEGenes_GFP,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("GFP")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_GFP)
liner.SG <- rstandard(smooth.SG)
DEGenes_GFP <-cbind(DEGenes_GFP, liner.SG)
colnames(DEGenes_GFP)[9] <- "Residuals"
row.names(DEGenes_GFP) <- DEGenes_GFP$Gene
linemarkers <- as.data.frame(DEGenes_GFP[abs(DEGenes_GFP$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_GFP$t, log10(DEGenes_GFP$Length), method = "spearman")



ggplot(DEGenes_GFP, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("GFP") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_GFP <- DEGenes_GFP[order(DEGenes_GFP[,4]),]
gseatool <- DEGenes_GFP[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_GFP$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# SPINSIM v.2

After my SpinSim pipeline, this is how I compared these simulated SGs with bulk and their original counterparts. For more details on this pipeline, please refer to its corresponding files in the Simulations/ folder in the GitHub repository.

```{r}
read.counts$Gene <- row.names(read.counts)

#To maintain equitable read depth, I base my final total count number per sample on the original samples
goalsum_SGWTAR1 <- sum(read.counts[,1])
goalsum_SGWTAR2 <- sum(read.counts[,2])
goalsum_SGWTAR3 <- sum(read.counts[,3])
goalsum_SGWTSO1 <- sum(read.counts[,7])
goalsum_SGWTSO2 <- sum(read.counts[,8])
goalsum_SGWTSO3 <- sum(read.counts[,11])
goalsum_SGKOSO1 <- sum(read.counts[,12])
goalsum_SGKOSO2 <- sum(read.counts[,13])
goalsum_SGGFSO1 <- sum(read.counts[,23])
goalsum_SGGFSO2 <- sum(read.counts[,24])
goalsum_SGGFSO3 <- sum(read.counts[,25])

colnames(genelength) <- c("Length","Gene")

#For the simulated samples
SGWTAR1 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/SG2021/ARWT_1.txt", sep = ",")
SGWTAR1 <- na.omit(SGWTAR1)
colnames(SGWTAR1) <- c("Gene", "WTAR1")
SGWTAR2 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/SG2021/ARWT_2.txt", sep = ",")
SGWTAR2 <- na.omit(SGWTAR2)
colnames(SGWTAR2) <- c("Gene", "WTAR2")
SGWTAR3 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/SG2021/ARWT_3.txt", sep = ",")
SGWTAR3 <- na.omit(SGWTAR3)
colnames(SGWTAR3) <- c("Gene", "WTAR3")
SGWTSO1 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/SG2021/SOWT_1.txt", sep = ",")
SGWTSO1 <- na.omit(SGWTSO1)
colnames(SGWTSO1) <- c("Gene", "WTSO1")
SGWTSO2 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/SG2021/SOWT_2.txt", sep = ",")
SGWTSO2 <- na.omit(SGWTSO2)
colnames(SGWTSO2) <- c("Gene", "WTSO2")
SGWTSO3 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/SG2021/SOWT_3.txt", sep = ",")
SGWTSO3 <- na.omit(SGWTSO3)
colnames(SGWTSO3) <- c("Gene", "WTSO3")
SGKOSO1 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/SG2021/SOKO_1.txt", sep = ",")
SGKOSO1 <- na.omit(SGKOSO1)
colnames(SGKOSO1) <- c("Gene", "KOSO1")
SGKOSO2 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/SG2021/SOKO_2.txt", sep = ",")
SGKOSO2 <- na.omit(SGKOSO2)
colnames(SGKOSO2) <- c("Gene", "KOSO2")
SGGFSO1 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/SG2021/SOGF_1.txt", sep = ",")
SGGFSO1 <- na.omit(SGGFSO1)
colnames(SGGFSO1) <- c("Gene", "GFSO1")
SGGFSO2 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/SG2021/SOGF_2.txt", sep = ",")
SGGFSO2 <- na.omit(SGGFSO2)
colnames(SGGFSO2) <- c("Gene", "GFSO2")
SGGFSO3 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/SG2021/SOGF_3.txt", sep = ",")
SGGFSO3 <- na.omit(SGGFSO3)
colnames(SGGFSO3) <- c("Gene", "GFSO3")


#merging simulated samples up
simsamples <- merge(SGWTAR1, SGWTAR2, by="Gene")
simsamples <- merge(simsamples, SGWTAR3, by="Gene")
simsamples <- merge(simsamples, SGWTSO1, by="Gene")
simsamples <- merge(simsamples, SGWTSO2, by="Gene")
simsamples <- merge(simsamples, SGWTSO3, by="Gene")
simsamples <- merge(simsamples, SGKOSO1, by="Gene")
simsamples <- merge(simsamples, SGKOSO2, by="Gene")
simsamples <- merge(simsamples, SGGFSO1, by="Gene")
simsamples <- merge(simsamples, SGGFSO2, by="Gene")
simsamples <- merge(simsamples, SGGFSO3, by="Gene")
simsamples <- merge(simsamples, genelength, by="Gene")

#De-Tpming old samples
simsamples$Counts_WTAR1 <- simsamples$WTAR1*simsamples$Length
simsamples$Counts_WTAR1 <- simsamples$Counts_WTAR1/sum(simsamples$Counts_WTAR1)*goalsum_SGWTAR1
simsamples$Counts_WTAR2 <- simsamples$WTAR2*simsamples$Length
simsamples$Counts_WTAR2 <- simsamples$Counts_WTAR2/sum(simsamples$Counts_WTAR2)*goalsum_SGWTAR2
simsamples$Counts_WTAR3 <- simsamples$WTAR3*simsamples$Length
simsamples$Counts_WTAR3 <- simsamples$Counts_WTAR3/sum(simsamples$Counts_WTAR3)*goalsum_SGWTAR3
simsamples$Counts_WTSO1 <- simsamples$WTSO1*simsamples$Length
simsamples$Counts_WTSO1 <- simsamples$Counts_WTSO1/sum(simsamples$Counts_WTSO1)*goalsum_SGWTSO1
simsamples$Counts_WTSO2 <- simsamples$WTSO2*simsamples$Length
simsamples$Counts_WTSO2 <- simsamples$Counts_WTSO2/sum(simsamples$Counts_WTSO2)*goalsum_SGWTSO2
simsamples$Counts_WTSO3 <- simsamples$WTSO3*simsamples$Length
simsamples$Counts_WTSO3 <- simsamples$Counts_WTSO3/sum(simsamples$Counts_WTSO3)*goalsum_SGWTSO3
simsamples$Counts_KOSO1 <- simsamples$KOSO1*simsamples$Length
simsamples$Counts_KOSO1 <- simsamples$Counts_KOSO1/sum(simsamples$Counts_KOSO1)*goalsum_SGKOSO1
simsamples$Counts_KOSO2 <- simsamples$KOSO2*simsamples$Length
simsamples$Counts_KOSO2 <- simsamples$Counts_KOSO2/sum(simsamples$Counts_KOSO2)*goalsum_SGKOSO2
simsamples$Counts_GFSO1 <- simsamples$GFSO1*simsamples$Length
simsamples$Counts_GFSO1 <- simsamples$Counts_GFSO1/sum(simsamples$Counts_GFSO1)*goalsum_SGGFSO1
simsamples$Counts_GFSO2 <- simsamples$GFSO2*simsamples$Length
simsamples$Counts_GFSO2 <- simsamples$Counts_GFSO2/sum(simsamples$Counts_GFSO2)*goalsum_SGGFSO2
simsamples$Counts_GFSO3 <- simsamples$GFSO3*simsamples$Length
simsamples$Counts_GFSO3 <- simsamples$Counts_GFSO3/sum(simsamples$Counts_GFSO3)*goalsum_SGGFSO3

simsamples <- simsamples[,c(1,14:24)] 


OGsamples <- read.counts

SG_City <- merge(OGsamples, simsamples, by="Gene")
SG_City <- SG_City %>% distinct(Gene, .keep_all = TRUE)
row.names(SG_City) <- SG_City$Gene
SG_City$Gene <- NULL
```


## Filtration

### ***Before Filtration***

Let us begin with the filtration. 

```{r, warning=FALSE}
alternew <- log10(rowSums(SG_City))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 4, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")
```

### ***After Filtration***

```{r filtration 2}
MitoGenes <- fread("~/MitoGenes.txt", header = FALSE)[[1]]
filterednew <- subset(SG_City, alternew>4)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew), MitoGenes)

#Now we need to select only the rows which have those genes.
finaltable <- SG_City[row.names(SG_City)%in%impgenes,]
```
## Matrices

Good! Now we can proceed by designing our matrices. 

```{r matrixes and normalization}
#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)

#Then we calculate the normalization factors.
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)

forall <- c("SG", "SG", "SG", "Bulk", "Bulk", "Bulk", "SG", "SG", "Bulk", "Bulk", "SG", "SG", "SG", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "SG", "SG","SG", "Bulk", "Bulk", "Bulk", rep("SimSG",11))

#Now we do our matrix itself
mm <- model.matrix(~0+forall)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```


## Normalisation

For normalisation I convert the read counts into log2CPMs. Its a good way to visualize how the reads are. 

```{r}

dfforboxy <- finaltable
custom_palette <- qualitative_hcl(10, palette = "Pastel")
custom_palette <- sample(custom_palette,10)


orderforplots <- forall
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = (custom_palette), labels=unique(forall), breaks =unique(forall)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = unique(custom_palette), labels=unique(forall), breaks =unique(forall)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```


## PCA

```{r PCA}
#this is done to be able to do traditional PCA
dfforpca <- y$E
dfforpca <- rbind(dfforpca, forall)
row.names(dfforpca)[nrow(dfforpca)] <- "forall"
dfforpca <- t(dfforpca)
numcol <- c(1:(ncol(dfforpca)-1))
dfforpca <- as.data.frame(dfforpca) #for some reason I can only change colum type if its a dataframe

dfforpca[ ,numcol] <- apply(dfforpca[ , numcol], 2,            # to turn our columns into numeric
                            function(x) as.numeric(as.character(x)))

#sapply(dfforpca, class) #to check class

# Or specify the factor levels in the order you want
dfforpca$forall <- factor(dfforpca$forall, levels = unique(forall))

pca_res <- prcomp(dfforpca[,numcol], scale=FALSE) #prepare the pca

#to make everything pretty
p <- autoplot(pca_res, data = dfforpca, x=1, y=2, colour = "forall", size=8, alpha=0.7)
p <- p + theme_bw() + scale_colour_manual((name = "Samples"), values = custom_palette, labels=unique(forall))
p <- p + ggtitle("PCA Plot") + theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())
p


#in case you want to see them yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])
```


## Linear Models

Now we must perform linear model fitting.

```{r}

forstress <- c("Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Unstressed", "Unstressed", "Unstressed", "Unstressed", "Unstressed", "Unstressed", "Sorbitol", "Sorbitol","Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol","Arsenite", "Arsenite", "Arsenite","Sorbitol", "Sorbitol", "Sorbitol","Sorbitol", "Sorbitol","Sorbitol", "Sorbitol", "Sorbitol")

forsample <- c("SG", "SG", "SG", "Bulk", "Bulk", "Bulk", "SG", "SG", "Bulk", "Bulk", "SG", "SG", "SG", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "SG", "SG","SG", "Bulk", "Bulk", "Bulk", rep("SimSG",11))

forcelltype <- c("WT", "WT", "WT", "WT", "WT", "WT", "WT", "WT", "WT", "WT", "KO", "KO", "KO", "KO", "KO", "KO", "GFP", "GFP", "GFP", "KO", "KO", "KO", "GFP", "GFP","GFP", "GFP", "GFP", "GFP", rep("WT",6), "KO", "KO", rep("GFP",3))

metadata <- data.frame(Stress=forstress, Cell=forcelltype, Sample=forsample) #Once again, I remove the KOKO samples


mm <- model.matrix(~0+forsample+forcelltype+forstress)

fitted <- lmFit(y, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr <- makeContrasts(forsampleSG - forsampleSimSG, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)

top.table <- topTable(tmp, sort.by = "logFC", n = Inf)

DEGenes <- top.table

ndegs <- 20
```



## Batch Effect Correction


```{r}
batch_cols <- c("Cell", "Stress")
metadata_cols <- c("Sample")
mmbatch <- model.matrix(reformulate(paste0("~0 + ",paste0( batch_cols,collapse = " + "))), data=metadata)
mmkeep <- model.matrix(reformulate(paste0("~0 + ",paste0( metadata_cols,collapse =  "+" ))), data=metadata)
mm <- cbind(mmkeep,mmbatch)
D0 <- DGEList(2^y$E)
D0 <- calcNormFactors(D0)
y <- voom(D0, mm, plot = F)
fit <- lmFit(y, mm)
beta <- fit$coefficients[,-(1:ncol(mmkeep)),drop=FALSE]
beta[is.na(beta)] <- 0
corrcounts_v1 <- as.matrix(y$E) - beta %*% t(mmbatch)
offset <- apply(corrcounts_v1,1,min) - apply(y$E,1,min)
corrcounts_v1 <- as.data.frame(corrcounts_v1-offset)
corrcounts_v1 <- 2^corrcounts_v1

mm <- model.matrix(~0+Sample, metadata)
y <- voom(corrcounts_v1, mm, plot = T)

fitted <- lmFit(y$E, mm)

head(coef(fitted))

contr <- makeContrasts(SampleSG  - SampleSimSG , levels = colnames(coef(fitted)))

tmp_contr <- contrasts.fit(fitted, contr)

tmp <- eBayes(tmp_contr)
top.table <- topTable(tmp, sort.by = "logFC", n=Inf)
DEGenes <- top.table
ndegs=20
```

## Stressed SG vs Unstressed SG ("Normal" Volcano Plot)

```{r sgvsgs-vulcan, results='hide', warning=FALSE, message=FALSE}

write.table(x = DEGenes,
            file = "~/Projects/StressGranules/SGPAPER/RealSGs/RealtMat21.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2.75 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2.75 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Real SG vs Simulated SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```


## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Real SG vs Sim SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("Real SG vs Sim SG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```
## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


# Session Info

```{r}
sessionInfo()
```