---
title: "Analyses of GSE138058"
subtitle: "G3BP1-linked mRNA partitioning supports selective protein synthesis in response to oxidative stress"
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

In this document, I present all analyses pertaining to the dataset **GSE138058**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE138058)),  
originally published by **SP. Somasekharan, 2020**.

## Methodology

```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)


set.seed(1906)

```


## First Steps

```{r}
#This section here is to load up the genelenght file I will be using throughout the analyses.
genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)

genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")

genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()

genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")



#Get the table of read counts by indicating the path to the file. This file is what we obtained from the running featureCounts.

sg.read.counts <- read.table("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2020/TrimmedSamples/alignment_STAR/featureCounts_results.txt", header=TRUE)

# Afterwards, I discard the columns which don't have read counts. These columns are only informations about the gene.
sg.read.counts <- sg.read.counts[,-c(2:6)]


biomart <- fread("~/biomart.txt")[,-1]


colnames(sg.read.counts)[1] <- "ensembl_gene_id_version"
colnames(biomart) <- c("ensembl_gene_id_version", "hgnc_symbol")

eltable <- merge(sg.read.counts,biomart,by = "ensembl_gene_id_version", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="") #this removes reads that dont have a corresponding hgnc symbol.

eltable$ensembl_gene_id_version <- NULL

read.counts <- eltable %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

read.counts <- as.data.frame(read.counts)

row.names(read.counts) <- read.counts$hgnc_symbol #changes rownames into hgnc symbols
read.counts$hgnc_symbol <- NULL #We no longer need this column, as we have the rownames.

```

## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, message=FALSE, warning=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(8, palette = "Pastel")
custom_palette <- sample(custom_palette, 8)

# Use SampleColouring from metadata
SampleColouring <- c(rep("CMV_Unstressed", 12),rep("G3BP1_Unstressed", 12),rep("CMV_Stressed", 12),rep("G3BP1_Stressed", 12), rep("Total_Unstressed", 12), rep("Total_Stressed", 12), rep("Polysomal_Unstressed", 12), rep("Polysomal_Stressed", 12) )

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(4, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )
```

As we can see in these plots, in this case there are no individual sample outliers in terms of complexity or read depth. Excellent!

# Differential Expression Analysis

## Filtration

### ***Before Filtration***

Let us begin with the filtration. Personally, I find that its easier to do visually. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bi-modal distributions, with the first mode corresponding to lowly expressed genes. I remove these by cutting just before the second mode.

```{r, warning=FALSE}
alternew.CMVu <- log10(rowSums(read.counts[,c(1:12)]))
alternew.CMVu <- as.data.frame(alternew.CMVu)

alternew.G3u <- log10(rowSums(read.counts[,c(13:24)]))
alternew.G3u <- as.data.frame(alternew.G3u)

alternew.CMVs <- log10(rowSums(read.counts[,c(25:36)]))
alternew.CMVs <- as.data.frame(alternew.CMVs)

alternew.G3s <- log10(rowSums(read.counts[,c(37:48)]))
alternew.G3s <- as.data.frame(alternew.G3s)

alternew.TOTu <- log10(rowSums(read.counts[,c(49:60)]))
alternew.TOTu <- as.data.frame(alternew.TOTu)

alternew.TOTs <- log10(rowSums(read.counts[,c(61:72)]))
alternew.TOTs <- as.data.frame(alternew.TOTs)

alternew.POLu <- log10(rowSums(read.counts[,c(73:84)]))
alternew.POLu <- as.data.frame(alternew.POLu)

alternew.POLs <- log10(rowSums(read.counts[,c(85:96)]))
alternew.POLs <- as.data.frame(alternew.POLs)

alternew <- log10(rowSums(read.counts))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.CMVu, aes(x=alternew.CMVu)) + geom_density(fill="#207567", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - CMV Control Unstressed")


ggplot(alternew.CMVs, aes(x=alternew.CMVs)) + geom_density(fill="#124038", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - CMV Control Stressed")

ggplot(alternew.G3u, aes(x=alternew.G3u)) + geom_density(fill="#E44E5A", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - G3BP1 Pulldown Unstressed")


ggplot(alternew.G3s, aes(x=alternew.G3s)) + geom_density(fill="#931621", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - G3BP1 Pulldown Stressed")

ggplot(alternew.TOTu, aes(x=alternew.TOTu)) + geom_density(fill="#769fca", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Total Unstressed RNA")

ggplot(alternew.TOTs, aes(x=alternew.TOTs)) + geom_density(fill="#396593", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Total Stressed RNA")

ggplot(alternew.POLu, aes(x=alternew.POLu)) + geom_density(fill="#B88B4A", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Polysomal Pulldown Unstressed")

ggplot(alternew.POLs, aes(x=alternew.POLs)) + geom_density(fill="#533E2D", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Polysomal Pulldown Stressed")
```


By seeing the shape of the graph, we can see where we should cut. We have a **binomial distribution** (low reads and noise vs actual biological meaningful reads). The ideal place to cut them is right before the wedge.

### ***After Filtration***

I didn't do it this time, but I've maintained the code to filter by sample type. This time, I considered that it was more correct to filter all samples together.


```{r}

MitoGenes <- fread("~/MitoGenes.txt", header = FALSE)[[1]]

filterednew <- subset(read.counts, alternew> 3.75)
filterednew.CMVu <- subset(read.counts, alternew.CMVu>3)
filterednew.CMVs <- subset(read.counts, alternew.CMVs>3)
filterednew.G3u <- subset(read.counts, alternew.G3u>2.75)
filterednew.G3s <- subset(read.counts, alternew.G3s>2.75)
filterednew.TOTu <- subset(read.counts, alternew.TOTu>2.5)
filterednew.TOTs <- subset(read.counts, alternew.TOTs>2.5)
filterednew.POLu <- subset(read.counts, alternew.POLu>2.5)
filterednew.POLs <- subset(read.counts, alternew.POLs>2.5)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- row.names(falternew)
impgenes.CMVu <- c(row.names(filterednew.CMVu))
impgenes.CMVs <- c(row.names(filterednew.CMVs))
impgenes.G3u <- c(row.names(filterednew.G3u))
impgenes.G3s <- c(row.names(filterednew.G3s))
impgenes.TOTu <- c(row.names(filterednew.TOTu))
impgenes.TOTs <- c(row.names(filterednew.TOTs))
impgenes.POLu <- c(row.names(filterednew.POLu))
impgenes.POLs <- c(row.names(filterednew.POLs))

#Now we need to select only the rows which have those genes.
finaltable <- read.counts[row.names(read.counts)%in%impgenes,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.CMVu,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.CMVs,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.G3u,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.G3s,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.TOTu,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.TOTs,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.POLu,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.POLs,]
```


## Matrices

Good! Now we can proceed by designing our matrices.

```{r}
#Let us prepare metadata to add to our samples
forsample <- c(rep("CMV",12),rep("G3BP1",12),rep("CMV",12),rep("G3BP1",12), rep("Total",24), rep("Polysomal",24))
forstress <- c(rep("CTRL",24), rep("Stressed",24), rep("CTRL",12), rep("Stressed",12),rep("CTRL",12), rep("Stressed",12))
forsg <- c(rep("No", 36), rep("Yes", 12), rep("No",48))

#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)

#Then we calculate the normalisation factors for these new generated data frames
#Bear in mind this step calculates the normalisation factors, but does not normalise itself.
D0 <- calcNormFactors(D0)

#Now we do the matrixes
mm <- model.matrix(~0+SampleColouring)

#and now we normalise. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)

```

## Normalisation

Now, we are almost ready to move on to the differential gene expression analysis. I will first check to see if any issues arose during the normalisation process, by looking at samples before and after this process. 

```{r, fig.height=12, fig.width=20}

dfforboxy <- finaltable

orderforplots <- SampleColouring
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#207567", "#E44E5A", "#124038", "#931621", "#769fca", "#396593", "#B88B4A", "#533E2D"), labels=unique(orderforplots)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

#Using the Standard Matrix

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#207567", "#E44E5A", "#124038", "#931621", "#769fca", "#396593", "#B88B4A", "#533E2D"), labels=unique(orderforplots)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```


I like to start with a Principal Component Analysis (PCA), which helps to identify what drives the largest variation in the data. This time, I also plotted a simple correlation heatmap between samples.

```{r, fig.height=7, fig.width=9}
# Define color palette
custom_palette <- qualitative_hcl(8, palette = "Pastel")
custom_palette <- sample(custom_palette, 8)
#custom_palette <- c("#124038","#AD1F36",  "#DBD56E","#5D8DC0")

#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- forsample
dfforpca$Stress <- forstress
dfforpca$SG <- forsg
dfforpca$All <- SampleColouring

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-4)], scale = FALSE)

p_Sample <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8, alpha=0.7) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + theme_classic() +
  theme(axis.line = element_line(colour = "black")) +
  theme(text=element_text(size=20,  family="Arial MS"))


p_Stress <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Stress", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Stress", values = custom_palette) +
  ggtitle("PCA - Coloured by Stress") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_SG <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "SG", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "SG", values = custom_palette) +
  ggtitle("PCA - Coloured by SG") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_All <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "All", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "All", values = custom_palette) +
  ggtitle("PCA - Coloured by All") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Sample
p_Stress
p_SG
p_All

 png("~/Images/PCA_Example4.png", width = 7, height=4, units = "in", res=300)
 p_Sample
 dev.off()
 
#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])


```

As one can observe from the PCAs, our PC1 is mostly explained by wheter a sample is polyssomal or not.

## Linear Models

Now we must perform linear model fitting. In this section, I will use simple linear models, comparing samples directly. 

```{r}
#The object with full description
fitted_y <- lmFit(y, mm)

#Just to see if everything is ok
head(coef(fitted_y))


#Lets do the comparisons
#Stressed SG vs Unstressed SG
SG.vs.SG <- makeContrasts(SampleColouringG3BP1_Stressed - SampleColouringG3BP1_Unstressed, levels = colnames(coef(fitted_y)))
#Stressed SG vs Stressed WC
SG.vs.WC <- makeContrasts(SampleColouringG3BP1_Stressed - SampleColouringTotal_Stressed, levels = colnames(coef(fitted_y)))
#Stressed WC vs Unstressed WC
WC.vs.WC <- makeContrasts(SampleColouringTotal_Stressed - SampleColouringTotal_Unstressed, levels = colnames(coef(fitted_y)))
#Stressed SG vs Stressed Control
SG.vs.CT <- makeContrasts(SampleColouringG3BP1_Unstressed - SampleColouringTotal_Stressed, levels = colnames(coef(fitted_y)))
#if more is necessary, add more

#ContrastFitting
tmp_SG.vs.SG <- contrasts.fit(fitted_y, SG.vs.SG)
tmp_SG.vs.WC <- contrasts.fit(fitted_y, SG.vs.WC)
tmp_WC.vs.WC <- contrasts.fit(fitted_y, WC.vs.WC)
tmp_SG.vs.CT <- contrasts.fit(fitted_y, SG.vs.CT)

#Correction for multiple testing
tmp_SG.vs.SG <- eBayes(tmp_SG.vs.SG)
tmp_SG.vs.WC <- eBayes(tmp_SG.vs.WC)
tmp_WC.vs.WC <- eBayes(tmp_WC.vs.WC)
tmp_SG.vs.CT <- eBayes(tmp_SG.vs.CT)
```


# Simple Linear Models

Lets move on to the main analyses.

```{r}
#TOP Tables
top_SG.vs.SG <- topTable(tmp_SG.vs.SG, sort.by = "logFC", n=Inf)
top_SG.vs.WC <- topTable(tmp_SG.vs.WC, sort.by = "logFC", n=Inf)
top_WC.vs.WC <- topTable(tmp_WC.vs.WC, sort.by = "logFC", n=Inf)
top_SG.vs.CT <- topTable(tmp_SG.vs.CT, sort.by = "logFC", n=Inf)

#DEGenes
DEGenes_SG.vs.SG <- top_SG.vs.SG
DEGenes_SG.vs.WC <- top_SG.vs.WC
DEGenes_WC.vs.WC <- top_WC.vs.WC
DEGenes_SG.vs.CT <- top_SG.vs.CT
```

# Stressed SG vs Stressed Control

## Volcano Plot 

This interaction should give the G3BP1 interactome. 

```{r, results='hide', warning=FALSE, message=FALSE}
ndegs <- 10
upmarkers <- as.data.frame(DEGenes_SG.vs.CT[DEGenes_SG.vs.CT$logFC>1.5 & DEGenes_SG.vs.CT$B>50,] )
downmarkers <- as.data.frame(DEGenes_SG.vs.CT[DEGenes_SG.vs.CT$logFC< -1.5 & DEGenes_SG.vs.CT$B>50,] )

ggplot(data=DEGenes_SG.vs.CT,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs Vector")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
There is a massive increase in G3BP1 and G3BP1P1. Apex is connected to the protein, not the transcript, so this is a bit surprising.

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_SG.vs.CT$Gene <- row.names(DEGenes_SG.vs.CT)
colnames(genelength) <- c("Length","Gene")
DEGenes_SG.vs.CT <- merge(DEGenes_SG.vs.CT, genelength, by="Gene")

ggplot(data=DEGenes_SG.vs.CT,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs Vector")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_SG.vs.CT)
liner.SG <- rstandard(smooth.SG)
DEGenes_SG.vs.CT <-cbind(DEGenes_SG.vs.CT, liner.SG)
colnames(DEGenes_SG.vs.CT)[9] <- "Residuals"
row.names(DEGenes_SG.vs.CT) <- DEGenes_SG.vs.CT$Gene
linemarkers <- as.data.frame(DEGenes_SG.vs.CT[abs(DEGenes_SG.vs.CT$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_SG.vs.CT$t, log10(DEGenes_SG.vs.CT$Length), method = "spearman")



ggplot(DEGenes_SG.vs.CT, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs Vector") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
No length bias whatsoever. 

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_SG.vs.CT <- DEGenes_SG.vs.CT[order(DEGenes_SG.vs.CT[,4]),]
gseatool <- DEGenes_SG.vs.CT[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_SG.vs.CT$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


# Stressed SG vs Unstressed SG

This gives the changes in G3BP1 interactome in stress.

## Volcano Plot 

```{r, results='hide', warning=FALSE, message=FALSE}
ndegs <- 10
upmarkers <- as.data.frame(DEGenes_SG.vs.SG[DEGenes_SG.vs.SG$logFC>1 & DEGenes_SG.vs.SG$B>20,] )
downmarkers <- as.data.frame(DEGenes_SG.vs.SG[DEGenes_SG.vs.SG$logFC< -2 & DEGenes_SG.vs.SG$B>10,] )

ggplot(data=DEGenes_SG.vs.SG,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
As expected, most are related to stress response, like EGR1, HSPA6 and DNAJB1.
## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_SG.vs.SG$Gene <- row.names(DEGenes_SG.vs.SG)
colnames(genelength) <- c("Length","Gene")
DEGenes_SG.vs.SG <- merge(DEGenes_SG.vs.SG, genelength, by="Gene")

ggplot(data=DEGenes_SG.vs.SG,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_SG.vs.SG)
liner.SG <- rstandard(smooth.SG)
DEGenes_SG.vs.SG <-cbind(DEGenes_SG.vs.SG, liner.SG)
colnames(DEGenes_SG.vs.SG)[9] <- "Residuals"
row.names(DEGenes_SG.vs.SG) <- DEGenes_SG.vs.SG$Gene
linemarkers <- as.data.frame(DEGenes_SG.vs.SG[abs(DEGenes_SG.vs.SG$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_SG.vs.SG$t, log10(DEGenes_SG.vs.SG$Length), method = "spearman")



ggplot(DEGenes_SG.vs.SG, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs SG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
A negative lenght bias is observed.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_SG.vs.SG <- DEGenes_SG.vs.SG[order(DEGenes_SG.vs.SG[,4]),]
gseatool <- DEGenes_SG.vs.SG[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_SG.vs.SG$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# Stressed WC vs Unstressed WC

This gives the stress response in PC3 cells.

## Volcano Plot 

```{r, results='hide', warning=FALSE, message=FALSE}
ndegs <- 10
upmarkers <- as.data.frame(DEGenes_WC.vs.WC[DEGenes_WC.vs.WC$logFC>1 & DEGenes_WC.vs.WC$B>10,] )
downmarkers <- as.data.frame(DEGenes_WC.vs.WC[DEGenes_WC.vs.WC$logFC< -1 & DEGenes_WC.vs.WC$B>10,] )

ggplot(data=DEGenes_WC.vs.WC,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("WC vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

stressresponse_soma <- DEGenes_WC.vs.WC
write.table(stressresponse_soma, "~/Projects/StressGranules/SGPAPER/PC3BULKStress.txt", row.names = T, col.names = T)

```
As before, most genes are related to stress response: DNAJB1, DUSP1, EGR1, HSPA6.

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_WC.vs.WC$Gene <- row.names(DEGenes_WC.vs.WC)
colnames(genelength) <- c("Length","Gene")
DEGenes_WC.vs.WC <- merge(DEGenes_WC.vs.WC, genelength, by="Gene")

ggplot(data=DEGenes_WC.vs.WC,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_WC.vs.WC)
liner.SG <- rstandard(smooth.SG)
DEGenes_WC.vs.WC <-cbind(DEGenes_WC.vs.WC, liner.SG)
colnames(DEGenes_WC.vs.WC)[9] <- "Residuals"
row.names(DEGenes_WC.vs.WC) <- DEGenes_WC.vs.WC$Gene
linemarkers <- as.data.frame(DEGenes_WC.vs.WC[abs(DEGenes_WC.vs.WC$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_WC.vs.WC$t, log10(DEGenes_WC.vs.WC$Length), method = "spearman")



ggplot(DEGenes_WC.vs.WC, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs SG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
A negative lenght bias is observed.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_WC.vs.WC <- DEGenes_WC.vs.WC[order(DEGenes_WC.vs.WC[,4]),]
gseatool <- DEGenes_WC.vs.WC[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_WC.vs.WC$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# Stressed SG vs Stressed WC

## Volcano Plot 

```{r, results='hide', warning=FALSE, message=FALSE}
write.table(x = DEGenes_SG.vs.WC,
            file = "~/Projects/StressGranules/SGPAPER/SGvsWC/Somasekharan.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)



ndegs <- 10
upmarkers <- as.data.frame(DEGenes_SG.vs.WC[DEGenes_SG.vs.WC$logFC>2 & DEGenes_SG.vs.WC$B>50,] )
downmarkers <- as.data.frame(DEGenes_SG.vs.WC[DEGenes_SG.vs.WC$logFC< -2 & DEGenes_SG.vs.WC$B>50,] )

ggplot(data=DEGenes_SG.vs.WC,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

stressoma_t <- data.frame(Gene=row.names(DEGenes_SG.vs.WC), so_t=DEGenes_SG.vs.WC$t, so_logFC=DEGenes_SG.vs.WC$logFC)

```

Similar results to comparison with the vector. 

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_SG.vs.WC$Gene <- row.names(DEGenes_SG.vs.WC)
colnames(genelength) <- c("Length","Gene")
DEGenes_SG.vs.WC <- merge(DEGenes_SG.vs.WC, genelength, by="Gene")

ggplot(data=DEGenes_SG.vs.WC,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_SG.vs.WC)
liner.SG <- rstandard(smooth.SG)
DEGenes_SG.vs.WC <-cbind(DEGenes_SG.vs.WC, liner.SG)
colnames(DEGenes_SG.vs.WC)[9] <- "Residuals"
row.names(DEGenes_SG.vs.WC) <- DEGenes_SG.vs.WC$Gene
linemarkers <- as.data.frame(DEGenes_SG.vs.WC[abs(DEGenes_SG.vs.WC$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_SG.vs.WC$t, log10(DEGenes_SG.vs.WC$Length), method = "spearman")



ggplot(DEGenes_SG.vs.WC, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs SG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
A negative lenght bias is seen.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_SG.vs.WC <- DEGenes_SG.vs.WC[order(DEGenes_SG.vs.WC[,4]),]
gseatool <- DEGenes_SG.vs.WC[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_SG.vs.WC$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# Complex Linear Models

In here, I use more complex linear models, which in theory allows me to isolate the actual SG transcriptome. Since the polyssome samples are so different, I dont use them in this analysis.

```{r}
forsample <- c(rep("CMV",12),rep("G3BP1",12),rep("CMV",12),rep("G3BP1",12), rep("Total",24), rep("Polysomal",24))[1:72]
forstress <- c(rep("CTRL",24), rep("Stressed",24), rep("CTRL",12), rep("Stressed",12),rep("CTRL",12), rep("Stressed",12))[1:72]
forsg <- c(rep("No", 36), rep("Yes", 12), rep("No",48))[1:72]

metadata <- data.frame(Sample=forsample, Stress=forstress, SG=forsg)

mm <- model.matrix(~1+Sample*Stress,metadata)

print(mm)

#Here, I center my model matrix
mm[,2] <- mm[,2] - mean(mm[,2])
mm[,3] <- mm[,3] - mean(mm[,3])
mm[,4] <- mm[,4] - mean(mm[,4])
mm[,5] <- mm[,4] * mm[,2]
mm[,6] <- mm[,4] * mm[,3]



D0 <- DGEList(finaltable[,1:72])

#Then we calculate the normalization factors for these new generated data frames
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)

y <- voom(D0, mm, plot = T)

fitted <- lmFit(y$E, mm)


head(coef(fitted))

tmp <- eBayes(fitted)

top.table <- topTable(tmp, 5, sort.by = "logFC", n=Inf)
DEGenes <- top.table
ndegs=20
```

```{r, fig.height=7, fig.width=9}
# Define color palette
custom_palette <- qualitative_hcl(8, palette = "Pastel")
custom_palette <- sample(custom_palette, 8)
#custom_palette <- c("#124038","#AD1F36",  "#DBD56E","#5D8DC0")

#This prepares my data for the PCA
dfforpca <- y$E[,c(1:72)]
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- forsample[1:72]
dfforpca$Stress <- forstress[1:72]
dfforpca$SG <- forsg[1:72]
dfforpca$All <- SampleColouring[1:72]

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-4)], scale = FALSE)

p_Sample <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8, alpha=0.7) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + theme_classic() +
  theme(axis.line = element_line(colour = "black")) +
  theme(text=element_text(size=20,  family="Arial MS"))


p_Stress <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Stress", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Stress", values = custom_palette) +
  ggtitle("PCA - Coloured by Stress") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_SG <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "SG", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "SG", values = custom_palette) +
  ggtitle("PCA - Coloured by SG") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_All <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "All", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "All", values = custom_palette) +
  ggtitle("PCA - Coloured by All") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Sample
p_Stress
p_SG
p_All

 png("~/Images/PCA_Example4.png", width = 7, height=4, units = "in", res=300)
 p_Sample
 dev.off()
 
#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])


```


```{r}
# Pull t-stats for the interaction and its two main-effect components
top_interact <- topTable(tmp, coef=5, sort.by="logFC", n=Inf)   # SampleG3BP1:StressStressed
top_sample   <- topTable(tmp, coef=2, sort.by="logFC", n=Inf)   # SampleG3BP1 (main effect)
top_stress   <- topTable(tmp, coef=4, sort.by="logFC", n=Inf)   # StressStressed (main effect)

interact_t <- data.frame(Gene=row.names(top_interact), interact_t=top_interact$t)
sample_t   <- data.frame(Gene=row.names(top_sample), sample_t=top_sample$t)
stress_t   <- data.frame(Gene=row.names(top_stress), stress_t=top_stress$t)

MitoGenes <- fread("~/MitoGenes.txt")[[1]]
NucGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]

## ---- Interaction vs Sample (G3BP1) component ----

comp1_df <- merge(interact_t, sample_t, by="Gene")
row.names(comp1_df) <- comp1_df$Gene
MitoData_c1 <- comp1_df[comp1_df$Gene %in% MitoGenes, ]
NucData_c1  <- comp1_df[comp1_df$Gene %in% NucGenes, ]

spearman_c1 <- cor(comp1_df$interact_t, comp1_df$sample_t, method = "spearman")

(p_c1 <- ggplot(comp1_df, aes(x=sample_t, y=interact_t)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData_c1, aes(x=sample_t, y=interact_t), color="#73937E", size=4) +
  geom_point(data=NucData_c1, aes(x=sample_t, y=interact_t), color="#CB643F", size=4) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color="grey40") +
  theme_bw() + ylab("Interaction (t-stat)") + xlab("Sample Component (t-stat)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Interaction vs Sample Component") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=15, family="Arial MS")) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size=5))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/InteractionVsSample_Soma.png", width = 7, height=6, units = "in", res=300)
p_c1
dev.off()

## ---- Interaction vs Stress component ----

comp2_df <- merge(interact_t, stress_t, by="Gene")
row.names(comp2_df) <- comp2_df$Gene
MitoData_c2 <- comp2_df[comp2_df$Gene %in% MitoGenes, ]
NucData_c2  <- comp2_df[comp2_df$Gene %in% NucGenes, ]

spearman_c2 <- cor(comp2_df$interact_t, comp2_df$stress_t, method = "spearman")

(p_c2 <- ggplot(comp2_df, aes(x=stress_t, y=interact_t)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData_c2, aes(x=stress_t, y=interact_t), color="#73937E", size=4) +
  geom_point(data=NucData_c2, aes(x=stress_t, y=interact_t), color="#CB643F", size=4) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color="grey40") +
  theme_bw() + ylab("Interaction (t-stat)") + xlab("Stress Component (t-stat)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Interaction vs Stress Component") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=15, family="Arial MS")) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size=5))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/InteractionVsStress_Soma.png", width = 7, height=6, units = "in", res=300)
p_c2
dev.off()
```


## Volcano Plot

```{r}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>1.5 & DEGenes$B>20,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -1.5 & DEGenes$B>30,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

write.table(DEGenes, "~/Projects/StressGranules/SGPAPER/RealSGs/RealSoma.txt", row.names=T, col.names = T)


MitoGenes <- fread("~/MitoGenes.txt")[[1]]
MitoData <- DEGenes[rownames(DEGenes) %in% MitoGenes, ]
NucGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]
NucData <- DEGenes[rownames(DEGenes) %in% NucGenes, ]
(m <- ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=B)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=NucData, aes(x=logFC, y=B), color="#CB643F", size=4) +
  geom_point(data=MitoData, aes(x=logFC, y=B), color="#73937E", size=4) +
  ylab("B-Statistic") +
  xlab(expression(Log[2]~"Fold-Change")) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),   # remove major gridlines
    panel.grid.minor = element_blank(),   # remove minor gridlines
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/SomaMito.png", width = 7, height=6, units = "in", res=300)
m
dev.off()

```

# MA Plot

```{r}
# MA plot: mean expression (x) vs logFC (y), same highlighting scheme as the volcano

ggplot(data=DEGenes,
                 aes(x=AveExpr,
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=AveExpr, y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=AveExpr, y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  theme_bw() + ylab(expression(Log[2]~"Fold-Change")) + xlab("Average Expression (Log2 CPM)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("T. Matheny Both - MA Plot") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,], mapping=aes(x=AveExpr, y=logFC, label=row.names(upmarkers)[1:ndegs]), size=4, point.padding=-0.5, color="#A10702", max.overlaps=20, fontface="italic") +
  geom_label_repel(downmarkers[1:ndegs,], mapping=aes(x=AveExpr, y=logFC, label=row.names(downmarkers)[1:ndegs]), size=4, point.padding=-0.5, color="#03256C", max.overlaps=20, fontface="italic") +
  geom_point(data=upmarkers, aes(x=AveExpr, y=logFC), color="#A10702", size=4) +
  geom_point(data=downmarkers, aes(x=AveExpr, y=logFC), color="#03256C", size=4) +
  theme(text=element_text(size=15, family="Arial MS"))

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=AveExpr,
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=AveExpr, y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=AveExpr, y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) + geom_density2d(color = "#800020", size = 1.2) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+
  stat_cor(data = MitoData, aes(x=AveExpr, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#73937E", label.y.npc = 0.1) +
  stat_cor(data = NucData, aes(x=AveExpr, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#CB643F", label.y.npc = 0.2) +
  xlab("Average Expression (Log2 CPM)") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/SomaMito_MA.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()
```
## MA Plot

```{r}
MitoGenes <- fread("~/MitoGenes.txt")[[1]]

genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)
genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")
genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()
genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")

genelength$Cromo <- ifelse(genelength$Gene %in% MitoGenes, "Mito", "Other")

# Apply the length-based weight formula
genelength$S <- ifelse(
  genelength$Cromo == "Mito",
  (genelength$Length)^(1/3) * 3.23e-11 + 1.17e-9,   # Formula for mitochondrial RNAs
  (genelength$Length)^(1/3) * 6.09e-11 + 3.88e-10    # Formula for other RNAs
)
DEGenes$Gene <- row.names(DEGenes)
DEGenes <- merge(DEGenes, genelength, by="Gene")
MitoGenes <- fread("~/MitoGenes.txt")[[1]]
MitoData <- DEGenes[DEGenes$Gene %in% MitoGenes, ]
NucGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]
NucData <- DEGenes[DEGenes$Gene %in% NucGenes, ]

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=log10(Length),
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=log10(Length), y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=log10(Length), y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) + geom_density2d(color = "#800020", size = 1.2) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+
  stat_cor(data = MitoData, aes(x=log10(Length), y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#73937E", label.y.npc = 0.1) +
  stat_cor(data = NucData, aes(x=log10(Length), y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#CB643F", label.y.npc = 0.2) +
  scale_x_continuous(
    breaks = log10(c(10, 100, 1000, 10000)),  # tick positions (in log10 scale)
    labels = c("10", "100", "1000", "10000") # what you want to show
  )+
  xlab("RNA Length (log10)") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/SomaMito_MA_length.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=(S),
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=(S), y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=(S), y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) + geom_density2d(color = "#800020", size = 1.2) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+
  stat_cor(data = MitoData, aes(x=S, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#73937E", label.y.npc = 0.1) +
  stat_cor(data = NucData, aes(x=S, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#CB643F", label.y.npc = 0.2) +
  xlab("S") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/SomaMito_MA_S.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()
```
## Mito / Nuc Signature Violin

```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(rstatix)

# ── 0. Subset to the samples that take part in the SampleG3BP1:StressStressed interaction ──
# These are the CMV and G3BP1 arms only (indices 1:48); the Total arm is not part of this coefficient
interact_idx <- which(metadata$Sample %in% c("CMV", "G3BP1"))

expression.df <- y$E[, interact_idx]
sample_grp <- metadata$Sample[interact_idx]
stress_grp <- metadata$Stress[interact_idx]

# Combined 4-level grouping for the x-axis, keeps Sample and Stress crossed and visible
Condition <- interaction(sample_grp, stress_grp, sep=".")

custom_palette <- c(
  "CMV.CTRL"       = "#a2d1e6",
  "CMV.Stressed"   = "#769fca",
  "G3BP1.CTRL"     = "#e88184",
  "G3BP1.Stressed" = "#de425b"
)

# ── 1. Gene sets already loaded earlier in the script (MitoGenes, NucGenes) ──
mito_genes_present <- MitoGenes[MitoGenes %in% rownames(expression.df)]
nuc_genes_present  <- NucGenes[NucGenes %in% rownames(expression.df)]
cat(length(mito_genes_present), "Mito genes found in expression matrix\n")
cat(length(nuc_genes_present), "Nuc genes found in expression matrix\n")

# ── 2. Rank all genes per sample (1 = least expressed) ───────────────────────
rank_matrix <- apply(expression.df, 2, function(x) rank(x, ties.method = "average"))
rank_matrix <- as.data.frame(rank_matrix)

# ── 3. Compute a rank-sum score per sample, one gene set at a time ───────────
compute_score <- function(genes, rank_matrix, n_total) {
  colSums(rank_matrix[genes, , drop = FALSE]) / n_total
}

mito_score <- compute_score(mito_genes_present, rank_matrix, nrow(expression.df))
nuc_score  <- compute_score(nuc_genes_present, rank_matrix, nrow(expression.df))

# ── 4. Assemble plot dataframe, long format across both gene sets ────────────
score_df <- rbind(
  data.frame(Sample = colnames(expression.df), Score = as.numeric(mito_score),
             SampleGrp = sample_grp, StressGrp = stress_grp,
             Condition = Condition, GeneSet = "Mito"),
  data.frame(Sample = colnames(expression.df), Score = as.numeric(nuc_score),
             SampleGrp = sample_grp, StressGrp = stress_grp,
             Condition = Condition, GeneSet = "Nuc")
)

# ── 5. Statistics per gene set: two-way ANOVA (Sample*Stress), interaction term ──
# This checks if the gene set's score is significantly driven by the interaction,
# same logic as the t-stat scatter plots - is it more than the sum of its components?
stat_labels <- score_df %>%
  group_by(GeneSet) %>%
  group_modify(~{
    aov_res  <- aov(Score ~ SampleGrp * StressGrp, data = .x)
    aov_tidy <- summary(aov_res)

    # The interaction term is the third row of the ANOVA table
    p_raw  <- aov_tidy[[1]][["Pr(>F)"]][3]
    p_adj  <- p.adjust(p_raw, method = "BH")

    ss_interact <- aov_tidy[[1]][["Sum Sq"]][3]
    ss_total    <- sum(aov_tidy[[1]][["Sum Sq"]])
    eta_sq      <- ss_interact / ss_total
    cohens_f    <- sqrt(eta_sq / (1 - eta_sq))

    data.frame(
      label = paste0("Interaction Cohen's f = ", round(cohens_f, 3),
                      "\nadj. p = ", formatC(p_adj, format = "e", digits = 2))
    )
  })

# ── 6. Plot ───────────────────────────────────────────────────────────────────
vilsigplot <- ggplot(score_df, aes(x = Condition, y = Score, fill = Condition)) +
  geom_violin(colour = "black", alpha = 0.6, trim = FALSE) +
  geom_point(
    position = position_jitter(width = 0.1),
    alpha    = 0.7,
    size     = 2
  ) +
  stat_summary(fun = mean,   geom = "point", shape = 18,
               size = 4, colour = "blue") +
  stat_summary(fun = median, geom = "point", shape = 17,
               size = 4, colour = "blue") +
  geom_text(
    data  = stat_labels,
    aes(x = -Inf, y = Inf, label = label),
    hjust = -0.05, vjust = 1.2,
    size  = 4, fontface = "bold", lineheight = 1.1,
    inherit.aes = FALSE
  ) +
  facet_wrap(~GeneSet, scales = "free_y") +
  scale_fill_manual(values = custom_palette) +
  labs(
    x     = "Sample : Stress",
    y     = "Signature Score (Rank Sum)",
    title = "Mitochondrial vs Nuclear-Localised Gene Scores"
  ) +
  theme_bw() +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x     = element_text(size = 12, angle=20, hjust=1),
    strip.text      = element_text(size = 12, face = "bold"),
    legend.position = "none",    
    text = element_text(size = 20, family = "Arial MS", face = "bold")
  )

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Soma_InteractionMitoSigs.png", width = 14, height=6, units = "in", res=300)
vilsigplot
dev.off()

vilsigplot
```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}


ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[11] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs SG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

Massive negative lenght bias.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=10, fig.width=14}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>%
  filter(padj < thresh) %>%
  mutate(pathway = gsub("^HALLMARK_", "", pathway),
         pathway = gsub("_", " ", pathway))

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
(SomaGSEA <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = Status), alpha = 0.75) + 
  scale_fill_manual(values = c("#53A2BE", "#BA3B46"), 
                    name = "Regulation Status", 
                    labels = c("Downregulated", "Upregulated"), 
                    breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x = "Gene Set", 
       y = "Normalized Enrichment Score",
       title = "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) +
  theme_bw() +
  theme(legend.position = "none",
        text = element_text(size = 25, family = "Arial MS", face = "bold"),
        plot.title = element_text(hjust = 0.25, face = "bold"),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "bold")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/SomaGSEA.png", width = 14, height=10, units = "in", res=300)
SomaGSEA
dev.off()
```

### For Mito

```{r}
# --- Rank genes by t-statistic as usual ---
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]
vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol

# --- Define MitoGenes as a custom gene set ---
MitoGenes <- fread("~/MitoGenes.txt")[[1]]
custom.gsets <- list(MitoGenes = toupper(MitoGenes))

# --- Run fgsea ---
mito.res <- fgsea(pathways = custom.gsets,
                  stats = vectorranks)

mito.plot.base <- plotEnrichment(custom.gsets[["MitoGenes"]], vectorranks)
curve.data <- mito.plot.base$data

# --- key fix: match the curve's internal descending order ---
ord <- order(vectorranks, decreasing = TRUE)
ranked.names <- names(vectorranks)[ord]
gene.positions <- which(ranked.names %in% custom.gsets[["MitoGenes"]])

rug.top    <- min(curve.data$y) - 0.03
rug.bottom <- rug.top - 0.04

y.min <- rug.bottom - 0.03
y.max <- max(curve.data$y) + 0.05

(mito.plot <- ggplot(curve.data, aes(x = x, y = y)) +
  geom_area(fill = "#73937E", alpha = 0.4) +
  geom_line(color = "#73937E", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_segment(data = data.frame(x = gene.positions),
               aes(x = x, xend = x, y = rug.top, yend = rug.bottom),
               color = "#73937E", alpha = 0.6, linewidth = 0.9,
               inherit.aes = FALSE) +
  annotate("text", x = 9000, y = 0.5,
           label = paste0("NES = ", round(mito.res$NES, 2),
                          "\nadj.p ≈ ", round(mito.res$padj, 4)),
           size = 5, family = "Arial MS", fontface = "bold") +
  coord_cartesian(ylim = c(y.min, y.max), clip = "off") +
  theme_bw() +
  theme(text = element_text(size = 25, family = "Arial MS"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/MitoGSEA_Soma.png", width=7, height=6, units="in", res=300)
mito.plot
dev.off()
```
### For Nuc

```{r}
# --- Rank genes by t-statistic as usual ---
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]
vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol

# --- Define MitoGenes as a custom gene set ---
MitoGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]
custom.gsets <- list(MitoGenes = toupper(MitoGenes))

# --- Run fgsea ---
mito.res <- fgsea(pathways = custom.gsets,
                  stats = vectorranks)

mito.plot.base <- plotEnrichment(custom.gsets[["MitoGenes"]], vectorranks)
curve.data <- mito.plot.base$data

# --- key fix: match the curve's internal descending order ---
ord <- order(vectorranks, decreasing = TRUE)
ranked.names <- names(vectorranks)[ord]
gene.positions <- which(ranked.names %in% custom.gsets[["MitoGenes"]])

rug.top    <- min(curve.data$y) - 0.03
rug.bottom <- rug.top - 0.04

y.min <- rug.bottom - 0.03
y.max <- max(curve.data$y) + 0.05

(mito.plot <- ggplot(curve.data, aes(x = x, y = y)) +
  geom_area(fill = "#CB643F", alpha = 0.4) +
  geom_line(color = "#CB643F", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_segment(data = data.frame(x = gene.positions),
               aes(x = x, xend = x, y = rug.top, yend = rug.bottom),
               color = "#CB643F", alpha = 0.6, linewidth = 0.9,
               inherit.aes = FALSE) +
  annotate("text", x = 4500, y = -0.1,
           label = paste0("NES = ", round(mito.res$NES, 2),
                          "\nadj.p = ", round(mito.res$padj, 4)),
           size = 5, family = "Arial MS", fontface = "bold") +
  coord_cartesian(ylim = c(y.min, y.max), clip = "off") +
  theme_bw() +
  theme(text = element_text(size = 25, family = "Arial MS"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/NucGSEA_Soma.png", width=7, height=6, units="in", res=300)
mito.plot
dev.off()
```


# Session Info

```{r}
sessionInfo()
```