---
title: "Analyses of GSE171792"
subtitle: "Sexually dimorphic RNA helicases DDX3X and DDX3Y differentially regulate RNA metabolism through phase separation."
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

In this document, I present all analyses pertaining to the dataset **GSE171792**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE171792)),  
originally published by **H. Shen, 2022**.


```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)


set.seed(1906)
```


## First Steps

```{r}
#This section here is to load up the genelenght file I will be using throughout the analyses.
genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)

genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")

genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()

genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")


#Get the table of read counts by indicating the path to the file. This file is what we obtained from the running featureCounts.

sg.read.counts <- read.table("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2022/TrimmedSamples/alignment_STAR/featureCounts_result.txt", header=TRUE)

# Afterwards, I discard the columns which don't have read counts. These columns are only informations about the gene.
sg.read.counts <- sg.read.counts[,-c(2:6)]

biomart <- fread("~/biomart.txt")[,-1]


colnames(sg.read.counts)[1] <- "ensembl_gene_id_version"
colnames(biomart) <- c("ensembl_gene_id_version", "hgnc_symbol")

eltable <- merge(sg.read.counts,biomart,by = "ensembl_gene_id_version", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="") #this removes reads that dont have a corresponding hgnc symbol.

eltable$ensembl_gene_id_version <- NULL

read.counts <- eltable %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

read.counts <- as.data.frame(read.counts)

row.names(read.counts) <- read.counts$hgnc_symbol #changes rownames into hgnc symbols
read.counts$hgnc_symbol <- NULL #We no longer need this column, as we have the rownames.

```


## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, warning=FALSE, message=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(7, palette = "Pastel")
custom_palette <- sample(custom_palette, 7)

# Use SampleColouring from metadata
SampleColouring <- c("GFP_Stressed","GFP_Stressed", "DDX3X_Stressed", "DDX3X_Stressed", "DDX3Y_Stressed", "DDX3Y_Stressed", "DDX3Y_Unstressed", "DDX3Y_Unstressed", "DDX3X_Unstressed", "DDX3X_Unstressed", "GFP_Unstressed", "GFP_Unstressed")

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(4, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )
```

As we can see in these plots, even thought complexity and read depth changes according to sample type, there are no individual sample outliers in terms of complexity or read depth. Excellent!

## Filtration

### ***Before Filtration***

Let us begin with the filtration.

```{r, warning=FALSE}
#Lets first try without separating sample types
#All samples had vastly different distributions, so I opted to do the filtration by sample type instead


alternew.GS <- log10(rowSums(read.counts[,c(1:2)]))
alternew.GS <- as.data.frame(alternew.GS)

alternew.XS <- log10(rowSums(read.counts[,c(3:4)]))
alternew.XS <- as.data.frame(alternew.XS)

alternew.YS <- log10(rowSums(read.counts[,c(5:6)]))
alternew.YS <- as.data.frame(alternew.YS)

alternew.GU <- log10(rowSums(read.counts[,c(11:12)]))
alternew.GU <- as.data.frame(alternew.GU)

alternew.XU <- log10(rowSums(read.counts[,c(9:10)]))
alternew.XU <- as.data.frame(alternew.XU)

alternew.YU <- log10(rowSums(read.counts[,c(7:8)]))
alternew.YU <- as.data.frame(alternew.YU)

alternew <- log10(rowSums(read.counts))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.GS, aes(x=alternew.GS)) + geom_density(fill="#207567", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - GFP Stressed Samples")


ggplot(alternew.GU, aes(x=alternew.GU)) + geom_density(fill="#769fca", alpha=0.5) + theme_bw() + geom_vline(xintercept = 1.25, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - GFP Unstressed Samples")

ggplot(alternew.XS, aes(x=alternew.XS)) + geom_density(fill="#207567", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - DDX3X Stressed Samples")


ggplot(alternew.XU, aes(x=alternew.XU)) + geom_density(fill="#769fca", alpha=0.5) + theme_bw() + geom_vline(xintercept = 1.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - DDX3X Unstressed Samples")

ggplot(alternew.YS, aes(x=alternew.YS)) + geom_density(fill="#207567", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - DDX3Y Stressed Samples")

ggplot(alternew.YU, aes(x=alternew.YU)) + geom_density(fill="#769fca", alpha=0.5) + theme_bw() + geom_vline(xintercept = 1.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - DDX3Y Unstressed Samples")
```


By seeing the shape of the graph, we can see where we should cut. We have a **binomial distribution** (low reads and noise vs actual biological meaningful reads). The ideal place to cut them is right before the wedge.

### ***After Filtration***

While I usually filter all samples together, due to the problems that you will see on the PCA, I filtered by sample type this time. This attenuated at least in part the problem, even though it didn't fix it. 

```{r}
filterednew <- subset(read.counts, alternew> 3)
filterednew.GS <- subset(read.counts, alternew.GS>2.75)
filterednew.GU <- subset(read.counts, alternew.GU>1.25)
filterednew.XS <- subset(read.counts, alternew.XS>2.5)
filterednew.XU <- subset(read.counts, alternew.XU>1.75)
filterednew.YS <- subset(read.counts, alternew.YS>2.5)
filterednew.YU <- subset(read.counts, alternew.YU>1.5)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))
impgenes.GS <- c(row.names(filterednew.GS))
impgenes.GU <- c(row.names(filterednew.GU))
impgenes.XS <- c(row.names(filterednew.XS))
impgenes.XU <- c(row.names(filterednew.XU))
impgenes.YS <- c(row.names(filterednew.YS))
impgenes.YU <- c(row.names(filterednew.YU))

#Now we need to select only the rows which have those genes.
finaltable <- read.counts[row.names(read.counts)%in%impgenes,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.GS,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.GU,]
inaltable <- finaltable[row.names(finaltable)%in%impgenes.XS,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.XU,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.YS,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.YU,]
```


## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#Let us prepare metadata to add to our samples
fordescriptive <- c("GFP_Stressed_1","GFP_Stressed_2", "DDX3X_Stressed", "DDX3X_Stressed", "DDX3Y_Stressed", "DDX3Y_Stressed", "DDX3Y_Unstressed", "DDX3Y_Unstressed", "DDX3X_Unstressed", "DDX3X_Unstressed", "GFP_Unstressed", "GFP_Unstressed")

forstress <- c("Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "CT", "CT", "CT", "CT", "CT", "CT")

forgenome <- c("GFP", "GFP", "DDX3X","DDX3X", "DDX3Y", "DDX3Y", "DDX3Y", "DDX3Y", "DDX3X", "DDX3X", "GFP", "GFP")

#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)


#Then we calculate the normalization factors for these new generated data frames
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)

#Now we do the matrixes
mm <- model.matrix(~0+fordescriptive)


#and now we normalize. This step also turns our counts into log2-counts per million.
#The object with all simple
y <- voom(D0, mm, plot = T)

```

As we can see, the **mean-variance trend** does not show a good curve, showing some issues with the filtration.


## Normalisation

For Normalisation I convert the read counts into log2CPMs. Its a good way to visualise how the reads are. 

```{r}

dfforboxy <- finaltable

orderforplots <- fordescriptive
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#B6CBE2", "#AD1F36", "#124038","#2CA08D","#5D8DC0", "#F0A8B4", "grey"), labels=unique(orderforplots)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

#Using the Standard Matrix

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#B6CBE2", "#AD1F36", "#124038","#2CA08D","#5D8DC0", "#F0A8B4", "grey"), labels=unique(orderforplots)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```
Nevertheless, the normalisation process appears mostly alright.

## PCA

```{r}
#this is done to be able to do traditional PCA
dfforpca <- y$E
dfforpca <- rbind(dfforpca, fordescriptive)
dfforpca <- t(dfforpca)
numcol <- c(1:(ncol(dfforpca)-1))
dfforpca <- as.data.frame(dfforpca) #for some reason I can only change colum type if its a dataframe

dfforpca[ ,numcol] <- apply(dfforpca[ , numcol], 2,            # to turn our columns into numeric
                            function(x) as.numeric(as.character(x)))

#sapply(dfforpca, class) #to check class

# Or specify the factor levels in the order you want
dfforpca$fordescriptive <- factor(dfforpca$fordescriptive, levels = unique(fordescriptive))

pca_res <- prcomp(dfforpca[,numcol], scale=FALSE) #prepare the pca

#to make everything pretty
p <- autoplot(pca_res, data = dfforpca, x=1, y=2, colour = "fordescriptive", size=8, alpha=0.7)
p <- p + theme_bw() + scale_colour_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey"), labels=c(unique(fordescriptive)))
p <- p + ggtitle("PCA Plot") + theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())
p


#in case you want to see them yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])
```
From the PCA we can see that the biggest issue in on the GFP_Stressed samples. The two of them are completely different. From the plot, it seems that the problem is the rightmost GFP_Stressed sample. 

While on the original study they show that these two samples are good replicates, I believe that the issue is linked to improper filtration. If 50% of a sample genes are a "0", you can have similar samples, even if the remainder of the genes are vastly different. 

I opted to still use them together on the "Simple Linear Model" Section. On the "Complex Linear Model", I use them both together and singularly. 

# Simple Linear Models

Now we must perform linear model fitting. Once more, we will have 4, one for each matrix we previously designed.

```{r}
#Let us prepare metadata to add to our samples
fordescriptive <- c("GFP_Stressed","GFP_Stressed", "DDX3X_Stressed", "DDX3X_Stressed", "DDX3Y_Stressed", "DDX3Y_Stressed", "DDX3Y_Unstressed", "DDX3Y_Unstressed", "DDX3X_Unstressed", "DDX3X_Unstressed", "GFP_Unstressed", "GFP_Unstressed")

#The object with full description
mm <- model.matrix(~0+fordescriptive)

y <- voom(D0, mm, plot = T)

fitted_y <- lmFit(y, mm)

#Just to see if everything is ok

head(coef(fitted_y))

#Lets do the comparisons
GFP.vs.SGX <- makeContrasts(fordescriptiveDDX3X_Stressed   - fordescriptiveGFP_Stressed , levels = colnames(coef(fitted_y)))
GFP.vs.SGY <- makeContrasts(fordescriptiveDDX3Y_Stressed - fordescriptiveGFP_Stressed , levels = colnames(coef(fitted_y)))
SGY.CTY <- makeContrasts(fordescriptiveDDX3Y_Stressed - fordescriptiveDDX3Y_Unstressed , levels = colnames(coef(fitted_y)))
SGX.CTX <- makeContrasts(fordescriptiveDDX3X_Stressed  - fordescriptiveDDX3X_Unstressed , levels = colnames(coef(fitted_y)))
SGY.SGX <- makeContrasts(fordescriptiveDDX3Y_Stressed - fordescriptiveDDX3X_Stressed , levels = colnames(coef(fitted_y)))

#ContrastFitting
tmp_GFP.vs.SGX <- contrasts.fit(fitted_y, GFP.vs.SGX)
tmp_GFP.vs.SGY <- contrasts.fit(fitted_y, GFP.vs.SGY)
tmp_SGY.CTY <- contrasts.fit(fitted_y, SGY.CTY)
tmp_SGX.CTX <- contrasts.fit(fitted_y, SGX.CTX)
tmp_SGY.SGX <- contrasts.fit(fitted_y, SGY.SGX)

#Correction for multiple testing
tmp_GFP.vs.SGX <- eBayes(tmp_GFP.vs.SGX)
tmp_GFP.vs.SGY <- eBayes(tmp_GFP.vs.SGY)
tmp_SGY.CTY <- eBayes(tmp_SGY.CTY)
tmp_SGX.CTX <- eBayes(tmp_SGX.CTX)
tmp_SGY.SGX <- eBayes(tmp_SGY.SGX)
```

Lets move on to the main analyses.

```{r}
#TOP Tables
top_GFP.vs.SGX <- topTable(tmp_GFP.vs.SGX,sort.by = "logFC", n=Inf)
top_GFP.vs.SGY <- topTable(tmp_GFP.vs.SGY, sort.by = "logFC", n=Inf)
top_SGY.CTY <- topTable(tmp_SGY.CTY, sort.by = "logFC", n=Inf)
top_SGX.CTX <- topTable(tmp_SGX.CTX, sort.by = "logFC", n=Inf)
top_SGY.SGX <- topTable(tmp_SGY.SGX, sort.by = "logFC", n=Inf)

#DEGenes

DEGenes_GFP.vs.SGX <- top_GFP.vs.SGX
DEGenes_GFP.vs.SGY <- top_GFP.vs.SGY
DEGenes_SGY.CTY <- top_SGY.CTY
DEGenes_SGX.CTX <- top_SGX.CTX
DEGenes_SGY.SGX <- top_SGY.SGX

```


# DDX3X vs GFP

## Volcano Plot 

```{r, results='hide', warning=FALSE, message=FALSE}
  write.table(x = DEGenes_GFP.vs.SGX,
             file = "~/Projects/StressGranules/SGPAPER/SGvsWC/Shen_X.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)
 #


ndegs <- 10
upmarkers <- as.data.frame(DEGenes_GFP.vs.SGX[DEGenes_GFP.vs.SGX$logFC>2 & DEGenes_GFP.vs.SGX$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_GFP.vs.SGX[DEGenes_GFP.vs.SGX$logFC< -2 & DEGenes_GFP.vs.SGX$adj.P.Val<0.05,] )

ggplot(data=DEGenes_GFP.vs.SGX,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("DDX3X vs GFP")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
Even the volcano shape is a bit weird

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_GFP.vs.SGX$Gene <- row.names(DEGenes_GFP.vs.SGX)
colnames(genelength) <- c("Length","Gene")
DEGenes_GFP.vs.SGX <- merge(DEGenes_GFP.vs.SGX, genelength, by="Gene")

ggplot(data=DEGenes_GFP.vs.SGX,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("DDX3X vs GFP")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_GFP.vs.SGX)
liner.SG <- rstandard(smooth.SG)
DEGenes_GFP.vs.SGX <-cbind(DEGenes_GFP.vs.SGX, liner.SG)
colnames(DEGenes_GFP.vs.SGX)[9] <- "Residuals"
row.names(DEGenes_GFP.vs.SGX) <- DEGenes_GFP.vs.SGX$Gene
linemarkers <- as.data.frame(DEGenes_GFP.vs.SGX[abs(DEGenes_GFP.vs.SGX$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_GFP.vs.SGX$t, log10(DEGenes_GFP.vs.SGX$Length), method = "spearman")



ggplot(DEGenes_GFP.vs.SGX, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("DDX3X vs GFP") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```

No evidence of lenght bias.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression, and contamination.

```{r, fig.height=12, fig.width=12}
upmarkers <- as.data.frame(DEGenes_GFP.vs.SGX[DEGenes_GFP.vs.SGX$logFC>2 & DEGenes_GFP.vs.SGX$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_GFP.vs.SGX[DEGenes_GFP.vs.SGX$logFC< -2 & DEGenes_GFP.vs.SGX$adj.P.Val<0.05,] )
finagenes <- rbind(upmarkers, downmarkers)

dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:12))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=fordescriptive)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) + scale_colour_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_GFP.vs.SGX <- DEGenes_GFP.vs.SGX[order(DEGenes_GFP.vs.SGX[,4]),]
gseatool <- DEGenes_GFP.vs.SGX[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_GFP.vs.SGX$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# DDX3Y vs GFP

## Volcano Plot 

```{r, results='hide', warning=FALSE, message=FALSE}
  write.table(x = DEGenes_GFP.vs.SGY,
             file = "~/Projects/StressGranules/SGPAPER/SGvsWC/Shen_Y.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)
 #


ndegs <- 10
upmarkers <- as.data.frame(DEGenes_GFP.vs.SGY[DEGenes_GFP.vs.SGY$logFC>2 & DEGenes_GFP.vs.SGY$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_GFP.vs.SGY[DEGenes_GFP.vs.SGY$logFC< -2 & DEGenes_GFP.vs.SGY$adj.P.Val<0.05,] )

ggplot(data=DEGenes_GFP.vs.SGY,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("DDX3Y vs GFP")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
Even the volcano shape is a bit weird

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_GFP.vs.SGY$Gene <- row.names(DEGenes_GFP.vs.SGY)
colnames(genelength) <- c("Length","Gene")
DEGenes_GFP.vs.SGY <- merge(DEGenes_GFP.vs.SGY, genelength, by="Gene")

ggplot(data=DEGenes_GFP.vs.SGY,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("DDX3Y vs GFP")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_GFP.vs.SGY)
liner.SG <- rstandard(smooth.SG)
DEGenes_GFP.vs.SGY <-cbind(DEGenes_GFP.vs.SGY, liner.SG)
colnames(DEGenes_GFP.vs.SGY)[9] <- "Residuals"
row.names(DEGenes_GFP.vs.SGY) <- DEGenes_GFP.vs.SGY$Gene
linemarkers <- as.data.frame(DEGenes_GFP.vs.SGY[abs(DEGenes_GFP.vs.SGY$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_GFP.vs.SGY$t, log10(DEGenes_GFP.vs.SGY$Length), method = "spearman")



ggplot(DEGenes_GFP.vs.SGY, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("DDX3Y vs GFP") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```

No evidence of lenght bias.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression, and contamination.

```{r, fig.height=12, fig.width=12}
upmarkers <- as.data.frame(DEGenes_GFP.vs.SGY[DEGenes_GFP.vs.SGY$logFC>2 & DEGenes_GFP.vs.SGY$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_GFP.vs.SGY[DEGenes_GFP.vs.SGY$logFC< -2 & DEGenes_GFP.vs.SGY$adj.P.Val<0.05,] )
finagenes <- rbind(upmarkers, downmarkers)

dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:12))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=fordescriptive)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) + scale_colour_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_GFP.vs.SGY <- DEGenes_GFP.vs.SGY[order(DEGenes_GFP.vs.SGY[,4]),]
gseatool <- DEGenes_GFP.vs.SGY[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_GFP.vs.SGY$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# DDX3X vs Ct Unstressed

## Volcano Plot 

```{r, results='hide', warning=FALSE, message=FALSE}

ndegs <- 10
upmarkers <- as.data.frame(DEGenes_SGX.CTX[DEGenes_SGX.CTX$logFC>2 & DEGenes_SGX.CTX$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_SGX.CTX[DEGenes_SGX.CTX$logFC< -2 & DEGenes_SGX.CTX$adj.P.Val<0.05,] )

ggplot(data=DEGenes_SGX.CTX,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("DDX3X vs Ct Unstressed")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
Even the volcano shape is a bit weird

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_SGX.CTX$Gene <- row.names(DEGenes_SGX.CTX)
colnames(genelength) <- c("Length","Gene")
DEGenes_SGX.CTX <- merge(DEGenes_SGX.CTX, genelength, by="Gene")

ggplot(data=DEGenes_SGX.CTX,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("DDX3X vs Ct Unstressed")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_SGX.CTX)
liner.SG <- rstandard(smooth.SG)
DEGenes_SGX.CTX <-cbind(DEGenes_SGX.CTX, liner.SG)
colnames(DEGenes_SGX.CTX)[9] <- "Residuals"
row.names(DEGenes_SGX.CTX) <- DEGenes_SGX.CTX$Gene
linemarkers <- as.data.frame(DEGenes_SGX.CTX[abs(DEGenes_SGX.CTX$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_SGX.CTX$t, log10(DEGenes_SGX.CTX$Length), method = "spearman")



ggplot(DEGenes_SGX.CTX, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("DDX3X vs Ct Unstressed") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```

Negative lenght bias.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression, and contamination.

```{r, fig.height=12, fig.width=12}
upmarkers <- as.data.frame(DEGenes_SGX.CTX[DEGenes_SGX.CTX$logFC>2 & DEGenes_SGX.CTX$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_SGX.CTX[DEGenes_SGX.CTX$logFC< -2 & DEGenes_SGX.CTX$adj.P.Val<0.05,] )
finagenes <- rbind(upmarkers, downmarkers)

dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:12))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=fordescriptive)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) + scale_colour_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_SGX.CTX <- DEGenes_SGX.CTX[order(DEGenes_SGX.CTX[,4]),]
gseatool <- DEGenes_SGX.CTX[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_SGX.CTX$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```



# DDX3Y vs Ct Unstressed

## Volcano Plot 

```{r, results='hide', warning=FALSE, message=FALSE}


ndegs <- 10
upmarkers <- as.data.frame(DEGenes_SGY.CTY[DEGenes_SGY.CTY$logFC>2 & DEGenes_SGY.CTY$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_SGY.CTY[DEGenes_SGY.CTY$logFC< -2 & DEGenes_SGY.CTY$adj.P.Val<0.05,] )

ggplot(data=DEGenes_SGY.CTY,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("DDX3Y vs Ct Unstressed")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
Even the volcano shape is a bit weird

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_SGY.CTY$Gene <- row.names(DEGenes_SGY.CTY)
colnames(genelength) <- c("Length","Gene")
DEGenes_SGY.CTY <- merge(DEGenes_SGY.CTY, genelength, by="Gene")

ggplot(data=DEGenes_SGY.CTY,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("DDX3Y vs Ct Unstressed")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_SGY.CTY)
liner.SG <- rstandard(smooth.SG)
DEGenes_SGY.CTY <-cbind(DEGenes_SGY.CTY, liner.SG)
colnames(DEGenes_SGY.CTY)[9] <- "Residuals"
row.names(DEGenes_SGY.CTY) <- DEGenes_SGY.CTY$Gene
linemarkers <- as.data.frame(DEGenes_SGY.CTY[abs(DEGenes_SGY.CTY$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_SGY.CTY$t, log10(DEGenes_SGY.CTY$Length), method = "spearman")



ggplot(DEGenes_SGY.CTY, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("DDX3Y vs Ct Unstressed") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```

Negative lenght bias.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression, and contamination.

```{r, fig.height=12, fig.width=12}
upmarkers <- as.data.frame(DEGenes_SGY.CTY[DEGenes_SGY.CTY$logFC>2 & DEGenes_SGY.CTY$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_SGY.CTY[DEGenes_SGY.CTY$logFC< -2 & DEGenes_SGY.CTY$adj.P.Val<0.05,] )
finagenes <- rbind(upmarkers, downmarkers)

dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:12))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=fordescriptive)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) + scale_colour_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes_SGY.CTY <- DEGenes_SGY.CTY[order(DEGenes_SGY.CTY[,4]),]
gseatool <- DEGenes_SGY.CTY[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_SGY.CTY$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# DDX3Y vs DDX3X

## Volcano Plot 

```{r, results='hide', warning=FALSE, message=FALSE}

ndegs <- 10
upmarkers <- as.data.frame(DEGenes_SGY.SGX[DEGenes_SGY.SGX$logFC>1 & DEGenes_SGY.SGX$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes_SGY.SGX[DEGenes_SGY.SGX$logFC< -1 & DEGenes_SGY.SGX$adj.P.Val<0.05,] )

ggplot(data=DEGenes_SGY.SGX,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("DDX3Y vs DDX3X")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
Even the volcano shape is a bit weird. Not many differences.

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes_SGY.SGX$Gene <- row.names(DEGenes_SGY.SGX)
colnames(genelength) <- c("Length","Gene")
DEGenes_SGY.SGX <- merge(DEGenes_SGY.SGX, genelength, by="Gene")

ggplot(data=DEGenes_SGY.SGX,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("DDX3Y vs Ct Unstressed")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes_SGY.SGX)
liner.SG <- rstandard(smooth.SG)
DEGenes_SGY.SGX <-cbind(DEGenes_SGY.SGX, liner.SG)
colnames(DEGenes_SGY.SGX)[9] <- "Residuals"
row.names(DEGenes_SGY.SGX) <- DEGenes_SGY.SGX$Gene
linemarkers <- as.data.frame(DEGenes_SGY.SGX[abs(DEGenes_SGY.SGX$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes_SGY.SGX$t, log10(DEGenes_SGY.SGX$Length), method = "spearman")



ggplot(DEGenes_SGY.SGX, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("DDX3Y vs Ct Unstressed") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```

Negative lenght bias.

# Complex Linear Models

# SG (With Both GFP Samples)

Now we must perform linear model fitting. Once more, we will have 4, one for each matrix we previously designed.

```{r}
#Let us prepare metadata to add to our samples
forstress <- c("Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "CT", "CT", "CT", "CT", "CT", "CT")

forgenome <- c("GFP", "GFP", "DDX3X","DDX3X", "DDX3Y", "DDX3Y", "DDX3Y", "DDX3Y", "DDX3X", "DDX3X", "GFP", "GFP")

forsg <- c("NO","NO", "YES", "YES", "YES", "YES", "NO", "NO", "NO", "NO", "NO", "NO")

#The object with full description
mm <- model.matrix(~0+forstress+forgenome+forsg)

y <- voom(D0, mm, plot = T)

fitted_y <- lmFit(y, mm)

#Just to see if everything is ok

head(coef(fitted_y))


#Correction for multiple testing
tmp <- eBayes(fitted_y)

```
Lets move on to the main analyses.

```{r}
#TOP Tables
top <- topTable(tmp,5,sort.by = "logFC", n=Inf)


#DEGenes

DEGenes <- top


```


## Volcano Plot 

```{r, results='hide', warning=FALSE, message=FALSE}


ndegs <- 10
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>1 & DEGenes$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -1 & DEGenes$adj.P.Val<0.05,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
Even the volcano shape is a bit weird

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```

# SG (With GFP 2)

Now we must perform linear model fitting. Once more, we will have 4, one for each matrix we previously designed.

```{r}
#Let us prepare metadata to add to our samples
forstress <- c("Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "CT", "CT", "CT", "CT", "CT", "CT")[-1]

forgenome <- c("GFP", "GFP", "DDX3X","DDX3X", "DDX3Y", "DDX3Y", "DDX3Y", "DDX3Y", "DDX3X", "DDX3X", "GFP", "GFP")[-1]

forsg <- c("NO","NO", "YES", "YES", "YES", "YES", "NO", "NO", "NO", "NO", "NO", "NO")[-2]

#The object with full description
mm <- model.matrix(~0+forstress+forgenome+forsg)

y <- voom(D0[,-1], mm, plot = T)

fitted_y <- lmFit(y, mm)

#Just to see if everything is ok

head(coef(fitted_y))


#Correction for multiple testing
tmp <- eBayes(fitted_y)

```
Lets move on to the main analyses.

```{r}
#TOP Tables
top <- topTable(tmp,5,sort.by = "logFC", n=Inf)


#DEGenes

DEGenes <- top


```


## Volcano Plot 

```{r, results='hide', warning=FALSE, message=FALSE}

ndegs <- 10
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>10 & DEGenes$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2 & DEGenes$adj.P.Val<0.05,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
Even the volcano shape is a bit weird

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```
No lenght bias

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression, and contamination.

```{r, fig.height=12, fig.width=12}
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>10 & DEGenes$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2 & DEGenes$adj.P.Val<0.05,] )
finagenes <- rbind(upmarkers, downmarkers)

dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:12))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=fordescriptive)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) + scale_colour_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# SG (With GFP 1)

Now we must perform linear model fitting. Once more, we will have 4, one for each matrix we previously designed.

```{r}
#Let us prepare metadata to add to our samples
forstress <- c("Stress", "Stress", "Stress", "Stress", "Stress", "Stress", "CT", "CT", "CT", "CT", "CT", "CT")[-2]

forgenome <- c("GFP", "GFP", "DDX3X","DDX3X", "DDX3Y", "DDX3Y", "DDX3Y", "DDX3Y", "DDX3X", "DDX3X", "GFP", "GFP")[-2]

forsg <- c("NO","NO", "YES", "YES", "YES", "YES", "NO", "NO", "NO", "NO", "NO", "NO")[-2]

#The object with full description
mm <- model.matrix(~0+forstress+forgenome+forsg)

y <- voom(D0[,-2], mm, plot = T)

fitted_y <- lmFit(y, mm)

#Just to see if everything is ok

head(coef(fitted_y))


#Correction for multiple testing
tmp <- eBayes(fitted_y)

```
Lets move on to the main analyses.

```{r}
#TOP Tables
top <- topTable(tmp,5,sort.by = "logFC", n=Inf)


#DEGenes

DEGenes <- top


```


## Volcano Plot 

```{r, results='hide', warning=FALSE, message=FALSE}


ndegs <- 10
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>8 & DEGenes$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2 & DEGenes$adj.P.Val<0.05,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
Even the volcano shape is a bit weird

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```
No lenght bias

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression, and contamination.

```{r, fig.height=12, fig.width=12}
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>8 & DEGenes$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2 & DEGenes$adj.P.Val<0.05,] )
finagenes <- rbind(upmarkers, downmarkers)

dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:12))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=fordescriptive)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) + scale_colour_manual(name = "Sample Type", values = c("#FF0000", "#FFA500", "#FFFF00", "#00FF00", "#0000FF", "#800080", "grey")) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


# Session Info

```{r}
sessionInfo()
```