---
title: "Analyses of GSE138988"
subtitle: "Transcriptome-Wide Comparison of Stress Granules and P-Bodies Reveals that Translation Plays a Major Role in RNA Partitioning"
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---


# Introduction

In this document, I present all analyses pertaining to the dataset **GSE138988**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE138988)),  
originally published by **T. Matheny, 2019**.

## Methodology

```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)

set.seed(1906)
```


## Data Loading
```{r}
#This section here is to load up the genelenght file I will be using throughout the analyses.
genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)

genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")

genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()

genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")




# Get the table of read counts by indicating the path to the file. This file is what we obtained from the running featureCounts.
sg.read.counts <- read.table("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2019/TrimmedSamples/alignment_STAR/featureCounts_results.txt", header=TRUE)

# Afterwards, I discard the columns which don't have read counts. These columns are only informations about the gene.
sg.read.counts <- sg.read.counts[,-c(2:6)] #use 6 if its going to be without length

biomart <- fread("~/biomart.txt")[,-1]


colnames(sg.read.counts)[1] <- "ensembl_gene_id_version"
colnames(biomart) <- c("ensembl_gene_id_version", "hgnc_symbol")

eltable <- merge(sg.read.counts,biomart,by = "ensembl_gene_id_version", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="") 
eltable$ensembl_gene_id_version <- NULL
stress.read.counts <- eltable %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE)) #this removes duplicate reads


stress.read.counts <- as.data.frame(stress.read.counts)

row.names(stress.read.counts) <- stress.read.counts$hgnc_symbol #changes rownames into hgnc symbols

stress.read.counts$hgnc_symbol <- NULL
```


### Bulk TPM for Centrifugation

In this section I prepare the bulk samples that I then use to simulate SGs from.
```{r}
read.counts <- subset(stress.read.counts, !(SRR10294577Aligned.sortedByCoord.out.bam == 0 | SRR10294578Aligned.sortedByCoord.out.bam == 0 | SRR10294579Aligned.sortedByCoord.out.bam == 0 | SRR10294580Aligned.sortedByCoord.out.bam == 0 | SRR10294581Aligned.sortedByCoord.out.bam == 0 | SRR10294582Aligned.sortedByCoord.out.bam == 0))


mof <- fread("~/Projects/StressGranules/Simulator/MolecularWeights.csv")
colnames(mof)[2] <- "GeneID" 
mof <- mof[!duplicated(mof$GeneID), ]



#Stressed Samp1
stress.tpm <- read.counts[,7]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/Bulk_Uns_1.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)

#Stressed Samp2
stress.tpm <- read.counts[,8]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/Bulk_Uns_2.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)

#Stressed Samp3
stress.tpm <- read.counts[,9]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/Bulk_Uns_3.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)


#Stressed Samp1
stress.tpm <- read.counts[,10]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/Bulk_Str_1.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)

#Stressed Samp2
stress.tpm <- read.counts[,11]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/Bulk_Str_2.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)

#Stressed Samp3
stress.tpm <- read.counts[,12]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/Bulk_Str_3.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)
```


## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, warning=FALSE, message=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(4, palette = "Pastel")
custom_palette <- sample(custom_palette, 4)

# Use SampleColouring from metadata
SampleColouring <- c("SG_U", "SG_U", "SG_U", "SG_S", "SG_S", "SG_S","BU_U", "BU_U", "BU_U", "BU_S", "BU_S", "BU_S")

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(6, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )
```

As we can see in these plots, the same sample types tend to have the same complexity, although one SG_Stressed sample does appear to have lower complexity. It is important to keep this in mind, in case some problems later arise (for example with normalisation).

# Differential Expression Analysis

## Filtration

### ***Before Filtration***

Let us begin with the filtration. Personally, I find that its easier to do visually. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bi-modal distributions, with the first mode corresponding to lowly expressed genes. I remove these by cutting just before the second mode. This time, I opted to also observe every sample type by itself.


```{r, warning=FALSE}
alternew.sg.unstressed <- log10(rowSums(stress.read.counts[,1:3]))
alternew.sg.unstressed <- as.data.frame(alternew.sg.unstressed)

alternew.sg.stressed <- log10(rowSums(stress.read.counts[,4:6]))
alternew.sg.stressed <- as.data.frame(alternew.sg.stressed)

alternew.bu.unstressed <- log10(rowSums(stress.read.counts[,7:9]))
alternew.bu.unstressed <- as.data.frame(alternew.bu.unstressed)

alternew.bu.stressed <- log10(rowSums(stress.read.counts[,10:12]))
alternew.bu.stressed <- as.data.frame(alternew.bu.stressed)

alternew <- log10(rowSums(stress.read.counts))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.sg.unstressed, aes(x=alternew.sg.unstressed)) + geom_density(fill="#a2d1e6", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Stress Granules Samples - Unstressed")

ggplot(alternew.sg.stressed, aes(x=alternew.sg.stressed)) + geom_density(fill="#769fca", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Stress Granules Samples - Stressed")

ggplot(alternew.bu.unstressed, aes(x=alternew.bu.unstressed)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Bulk Samples - Unstressed")

ggplot(alternew.bu.stressed, aes(x=alternew.bu.stressed)) + geom_density(fill="#de425b", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Bulk Samples - Stressed")


dffordensy <- stress.read.counts
champdata2 <- c("SG_U", "SG_U", "SG_U", "SG_S", "SG_S", "SG_S", "Bulk_U", "Bulk_U", "Bulk_U", "Bulk_S", "Bulk_S", "Bulk_S")
names(champdata2) <- colnames(dffordensy)
dffordensy <- reshape2::melt(t(dffordensy))
colnames(dffordensy) <- c("Sample", "Gene", "Expression")

densyplot <- ggplot(dffordensy, aes(x=log10(Expression), group=Sample, fill=Sample)) + geom_density(alpha=0.5) + facet_wrap(.~Sample, ncol=3, labeller = labeller(Sample=champdata2))
densyplot <- densyplot + scale_fill_manual(name = "Sample Type", values = c("#a2d1e6", "#a2d1e6", "#a2d1e6", "#769fca", "#769fca", "#769fca", "#e88184", "#e88184", "#e88184", "#de425b", "#de425b", "#de425b"))
densyplot <- densyplot + theme_bw() + xlab("Number of Reads (log10)") + ylab("Events") + theme(legend.position = "none") 
densyplot + geom_vline(xintercept = 2, linetype = "dashed") + theme(
      strip.text.x = element_text(
        size = 12
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   )
```


By seeing the shape of the graph, we can see where we should cut. We have a **binomial distribution** (low reads and noise vs actual biological meaningful reads). The ideal place to cut them is right before the wedge.

### ***After Filtration***

I didn't do it this time, but I've maintained the code to filter by sample type. This time, I considered that it was more correct to filter all samples together.

```{r}
filterednew <- subset(stress.read.counts, alternew>3.5)
filterednew.sgu <- subset(stress.read.counts, alternew.sg.unstressed>3)
filterednew.sgs <- subset(stress.read.counts, alternew.sg.stressed>2.75)
filterednew.buu <- subset(stress.read.counts, alternew.bu.unstressed>3)
filterednew.bus <- subset(stress.read.counts, alternew.bu.stressed>3)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))
impgenes.sgu <- c(row.names(filterednew.sgu))
impgenes.sgs <- c(row.names(filterednew.sgs))
impgenes.buu <- c(row.names(filterednew.buu))
impgenes.bus <- c(row.names(filterednew.bus))

#Now we need to select only the rows which have those genes.
finaltable <- stress.read.counts[row.names(stress.read.counts)%in%impgenes,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.sgu,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.sgs,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.buu,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.bus,]
```
## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#Let us prepare metadata to add to our samples
#Generic full sample name
forall <- c("SG_U", "SG_U", "SG_U", "SG_S", "SG_S", "SG_S", "Bulk_U", "Bulk_U", "Bulk_U", "Bulk_S", "Bulk_S", "Bulk_S")

#Sample Type Information
forsamp <- c("SG", "SG", "SG", "SG", "SG", "SG", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk", "Bulk")

#Stress Information
forstress <- c("U", "U", "U", "S", "S", "S", "U", "U", "U", "S", "S", "S")
forstress <- factor(forstress, levels = c("U", "S"))

#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)


#Then we calculate the normalization factors.
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)


#I start with the simplest analysis possible
mm <- model.matrix(~0+forall)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)

```

As we can see, the **mean-variance trend** shows a good curve, showing that the filtration was well done.

## Normalisation

Now, we are almost ready to move on to the differential gene expression analysis. I will first check to see if any issues arose during the normalisation process, by looking at samples before and after this process.

```{r normalization}

dfforboxy <- finaltable

orderforplots <- c("SG-U", "SG-U", "SG-U", "SG-S", "SG-S", "SG-S", "Bulk-U", "Bulk-U", "Bulk-U", "Bulk-S", "Bulk-S", "Bulk-S")
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#a2d1e6", "#769fca", "#e88184", "#de425b"), labels=c("SG-U", "SG-S", "Bulk-U", "Bulk-S"), breaks =c("SG-U", "SG-S", "Bulk-U", "Bulk-S")) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#a2d1e6", "#769fca", "#e88184", "#de425b"), labels=c("SG-U", "SG-S", "Bulk-U", "Bulk-S"), breaks =c("SG-U", "SG-S", "Bulk-U", "Bulk-S")) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```

In this case, from the boxplots, we can see that there no samples which can be considered outliers.

## PCA

I like to start with a Principal Component Analysis (PCA), which helps to identify what drives the largest variation in the data.

```{r}

#In here, I am just defining some colours to use
custom_palette <- c("#43F3D5", "#769fca", "#e88184", "#de425b")

#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- forsamp
dfforpca$Stress <- forstress

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-2)], scale = FALSE)

p_Sample <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Sample") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Stress <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Stress", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Stress", values = custom_palette) +
  ggtitle("PCA - Coloured by Stress") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Sample
p_Stress

#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])

```



As one can observe from the PCAs, our PC1 is mostly explained by stress state, while our PC2 is explained by sample type.

## Linear Models

Finally, we move on to differential gene expression! In this chunk, I define the linear model I will actually use for my analyses. In this case, since I have 4 comparisons I want to check, I will have 4 contrast matrixes.

```{r}
fitted <- lmFit(y, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr.sgvsgs <- makeContrasts(forallSG_S - forallSG_U, levels = colnames(coef(fitted)))

contr.buvsbu <- makeContrasts(forallBulk_S - forallBulk_U, levels = colnames(coef(fitted)))

contr.unstress <- makeContrasts(forallSG_U - forallBulk_U, levels = colnames(coef(fitted)))

contr.stress <- makeContrasts(forallSG_S - forallBulk_S , levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp.sgvsgs <- contrasts.fit(fitted, contr.sgvsgs)
tmp.buvsbu <- contrasts.fit(fitted, contr.buvsbu)
tmp.unstress <- contrasts.fit(fitted, contr.unstress)
tmp.stress <- contrasts.fit(fitted, contr.stress)

#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp.sgvsgs <- eBayes(tmp.sgvsgs)
tmp.buvsbu <- eBayes(tmp.buvsbu)
tmp.unstress <- eBayes(tmp.unstress)
tmp.stress <- eBayes(tmp.stress)

```

Lets move on to the main analyses.

```{r}
#time to see which genes are differentially expressed!
top.table.sgvsgs <- topTable(tmp.sgvsgs, sort.by = "logFC", n = Inf)
top.table.buvsbu <- topTable(tmp.buvsbu, sort.by = "logFC", n = Inf)
top.table.unstress <- topTable(tmp.unstress, sort.by = "logFC", n = Inf)
top.table.stress <- topTable(tmp.stress, sort.by = "logFC", n = Inf)

DEGenes.sgvsgs <- top.table.sgvsgs
DEGenes.buvsbu <- top.table.buvsbu
DEGenes.unstress <- top.table.unstress
DEGenes.stress <- top.table.stress


ndegs <- 20
```

# Stressed SG vs Unstressed SG ("Normal" Volcano Plot)

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes.sgvsgs[DEGenes.sgvsgs$logFC>2.5 & DEGenes.sgvsgs$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.sgvsgs[DEGenes.sgvsgs$logFC< -2.5 & DEGenes.sgvsgs$adj.P.Val<0.05,] )

ggplot(data=DEGenes.sgvsgs,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Stressed SG vs Unstressed SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

stress_t <- data.frame(Gene=row.names(DEGenes.sgvsgs), st_t=DEGenes.sgvsgs$t)

write.table(x = DEGenes.sgvsgs,
            file = "~/Projects/StressGranules/SGPAPER/Stressresponse_u2os_rnp.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)
```

## Stressed SG vs Unstressed SG (Volcano Plot with Gene Length)

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes.sgvsgs$Gene <- row.names(DEGenes.sgvsgs)
colnames(genelength) <- c("Length","Gene")
DEGenes.sgvsgs <- merge(DEGenes.sgvsgs, genelength, by="Gene")

ggplot(data=DEGenes.sgvsgs,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Stressed SG vs Unstressed SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.sgvsgs)
liner.SG <- rstandard(smooth.SG)
DEGenes.sgvsgs <-cbind(DEGenes.sgvsgs, liner.SG)
colnames(DEGenes.sgvsgs)[9] <- "Residuals"
row.names(DEGenes.sgvsgs) <- DEGenes.sgvsgs$Gene
linemarkers <- as.data.frame(DEGenes.sgvsgs[abs(DEGenes.sgvsgs$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.sgvsgs$t, log10(DEGenes.sgvsgs$Length), method = "spearman")



ggplot(DEGenes.sgvsgs, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SGS vs SGU Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

Some lenght bias is present, although it isn't as drastic as previously thought.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*


```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes.sgvsgs <- DEGenes.sgvsgs[order(DEGenes.sgvsgs[,4]),]
gseatool <- DEGenes.sgvsgs[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.sgvsgs$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```

## GSEA Length

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, results=FALSE}
#for all genes
DEGenes.sgvsgs <- DEGenes.sgvsgs[order(DEGenes.sgvsgs[,8]),]
gseatool <- DEGenes.sgvsgs[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.sgvsgs$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)

#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")
```
Some bias towards "long gene sets" is observed

# Stressed Bulk vs Unstressed Bulk (Normal Volcano Plot)

```{r, results='hide', warning=FALSE, message=FALSE}

write.table(x = DEGenes.buvsbu,
            file = "~/Projects/StressGranules/SGPAPER/Stressresponse_u2os.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)


upmarkers <- as.data.frame(DEGenes.buvsbu[DEGenes.buvsbu$logFC>3 & DEGenes.buvsbu$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.buvsbu[DEGenes.buvsbu$logFC< -3 & DEGenes.buvsbu$adj.P.Val<0.05,] )

ggplot(data=DEGenes.buvsbu,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Stressed WC vs Unstressed WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

stressresponse_u20s <- DEGenes.buvsbu
```
As expected, most extreme genes are typical of a stress response.

## Stressed Bulk vs Unstressed Bulk (Volcano Plot with Gene Length)

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes.buvsbu$Gene <- row.names(DEGenes.buvsbu)
colnames(genelength) <- c("Length","Gene")
DEGenes.buvsbu <- merge(DEGenes.buvsbu, genelength, by="Gene")

ggplot(data=DEGenes.buvsbu,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Stressed SG vs Unstressed SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.buvsbu)
liner.SG <- rstandard(smooth.SG)
DEGenes.buvsbu <-cbind(DEGenes.buvsbu, liner.SG)
colnames(DEGenes.buvsbu)[9] <- "Residuals"
row.names(DEGenes.buvsbu) <- DEGenes.buvsbu$Gene
linemarkers <- as.data.frame(DEGenes.buvsbu[abs(DEGenes.buvsbu$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.buvsbu$t, log10(DEGenes.buvsbu$Length), method = "spearman")



ggplot(DEGenes.buvsbu, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("WCS vs WCU Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=15, fig.width=20, results=FALSE}
#for all genes
DEGenes.buvsbu <- DEGenes.buvsbu[order(DEGenes.buvsbu[,4]),]
gseatool <- DEGenes.buvsbu[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.buvsbu$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```


## GSEA Length

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, results=FALSE}
#for all genes
DEGenes.buvsbu <- DEGenes.buvsbu[order(DEGenes.buvsbu[,8]),]
gseatool <- DEGenes.buvsbu[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.buvsbu$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)

#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")
```

# Unstressed "SG" vs Unstressed Bulk (Normal Volcano Plot)

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes.unstress[DEGenes.unstress$logFC>1.5 & DEGenes.unstress$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.unstress[DEGenes.unstress$logFC< -1.5 & DEGenes.unstress$adj.P.Val<0.05,] )

ggplot(data=DEGenes.unstress,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Unstressed SG vs Unstressed WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

unstress_t <- data.frame(Gene=row.names(DEGenes.unstress), uns_t=DEGenes.unstress$t)
```
## Unstressed SG vs Unstressed Bulk (Volcano Plot with Gene Length)

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes.unstress$Gene <- row.names(DEGenes.unstress)
colnames(genelength) <- c("Length","Gene")
DEGenes.unstress <- merge(DEGenes.unstress, genelength, by="Gene")

ggplot(data=DEGenes.unstress,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Unstressed SG vs Unstressed WC")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.unstress)
liner.SG <- rstandard(smooth.SG)
DEGenes.unstress <-cbind(DEGenes.unstress, liner.SG)
colnames(DEGenes.unstress)[9] <- "Residuals"
row.names(DEGenes.unstress) <- DEGenes.unstress$Gene
linemarkers <- as.data.frame(DEGenes.unstress[abs(DEGenes.unstress$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.unstress$t, log10(DEGenes.unstress$Length), method = "spearman")



ggplot(DEGenes.unstress, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SGU vs WCU Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```
Some minor lenght bias is present.

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes.unstress <- DEGenes.unstress[order(DEGenes.unstress[,4]),]
gseatool <- DEGenes.unstress[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.unstress$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```


## GSEA Length

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, results=FALSE}
#for all genes
DEGenes.unstress <- DEGenes.unstress[order(DEGenes.unstress[,8]),]
gseatool <- DEGenes.unstress[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.unstress$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)

#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")
```

# Stressed "SG" vs Stressed Bulk (Normal Volcano Plot)

```{r, results='hide', warning=FALSE, message=FALSE}

write.table(x = DEGenes.stress,
            file = "~/Projects/StressGranules/SGPAPER/SGvsWC/TMatheny2019.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)


upmarkers <- as.data.frame(DEGenes.stress[DEGenes.stress$logFC>1.15 & DEGenes.stress$B>0,] )
downmarkers <- as.data.frame(DEGenes.stress[DEGenes.stress$logFC< -1.5 & DEGenes.stress$B>0,] )

ggplot(data=DEGenes.stress,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Stressed SG vs Stressed WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

stress_rnpt <- data.frame(Gene=row.names(DEGenes.stress), srnp_t=DEGenes.stress$t, srnp_logFC=DEGenes.stress$logFC)
```

## Stressed SG vs Stressed Bulk (Volcano Plot with Gene Length)

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes.stress$Gene <- row.names(DEGenes.stress)
colnames(genelength) <- c("Length","Gene")
DEGenes.stress <- merge(DEGenes.stress, genelength, by="Gene")

ggplot(data=DEGenes.stress,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Stressed SG vs Stressed WC")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.stress)
liner.SG <- rstandard(smooth.SG)
DEGenes.stress <-cbind(DEGenes.stress, liner.SG)
colnames(DEGenes.stress)[9] <- "Residuals"
row.names(DEGenes.stress) <- DEGenes.stress$Gene
linemarkers <- as.data.frame(DEGenes.stress[abs(DEGenes.stress$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.stress$t, log10(DEGenes.stress$Length), method = "spearman")



ggplot(DEGenes.stress, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SGS vs WCS Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```

This time, we have a larger lenght bias. 

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes.stress <- DEGenes.stress[order(DEGenes.stress[,4]),]
gseatool <- DEGenes.stress[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.stress$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```



## GSEA Length

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, results=FALSE}
#for all genes
DEGenes.stress <- DEGenes.stress[order(DEGenes.stress[,8]),]
gseatool <- DEGenes.stress[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.stress$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)

#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")
```

# More Complex Linear Models

Finally, we move on to differential gene expression! In this chunk, I define the linear model I will actually use for my analyses. In this case, since I have 4 comparisons I want to check, I will have 4 contrast matrixes.

```{r}

#I start with the simplest analysis possible
mm <- model.matrix(~0+forsamp*forstress)

fitted <- lmFit(y, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
colnames(fitted$coefficients) <- make.names(colnames(fitted$coefficients))
contr.sgvsbu <- makeContrasts(forsampSG - forsampBulk, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp.sgvsbu <- contrasts.fit(fitted, contr.sgvsbu)

#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp.sgvsbu <- eBayes(tmp.sgvsbu)
tmp <- eBayes(fitted)

top.table.sgvsbu <- topTable(tmp.sgvsbu, sort.by = "logFC", n = Inf)
top.table.uvss <- topTable(tmp,3, sort.by = "logFC", n = Inf)
top.table.int <- topTable(tmp, 4, sort.by = "logFC", n = Inf)

DEGenes.sgvsbu <- top.table.sgvsbu
DEGenes.uvss <- top.table.uvss
DEGenes.int <- top.table.int

```

# SG vs Bulk

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes.sgvsbu[DEGenes.sgvsbu$logFC>1.15 & DEGenes.sgvsbu$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.sgvsbu[DEGenes.sgvsbu$logFC< -1.5 & DEGenes.sgvsbu$adj.P.Val<0.05,] )

ggplot(data=DEGenes.sgvsbu,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs Bulk")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```

## SG vs Bulk (Volcano Plot with Gene Length)


```{r}

DEGenes.sgvsbu$Gene <- row.names(DEGenes.sgvsbu)
colnames(genelength) <- c("Length","Gene")
DEGenes.sgvsbu <- merge(DEGenes.sgvsbu, genelength, by="Gene")

ggplot(data=DEGenes.sgvsbu,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs Bulk")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.sgvsbu)
liner.SG <- rstandard(smooth.SG)
DEGenes.sgvsbu <-cbind(DEGenes.sgvsbu, liner.SG)
colnames(DEGenes.sgvsbu)[9] <- "Residuals"
row.names(DEGenes.sgvsbu) <- DEGenes.sgvsbu$Gene
linemarkers <- as.data.frame(DEGenes.sgvsbu[abs(DEGenes.sgvsbu$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.sgvsbu$t, log10(DEGenes.sgvsbu$Length), method = "spearman")

ggplot(DEGenes.sgvsbu, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs Bu Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```


## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes.sgvsbu <- DEGenes.sgvsbu[order(DEGenes.sgvsbu[,4]),]
gseatool <- DEGenes.sgvsbu[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.sgvsbu$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```


## GSEA Length

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, results=FALSE}
#for all genes
DEGenes.sgvsbu <- DEGenes.sgvsbu[order(DEGenes.sgvsbu[,8]),]
gseatool <- DEGenes.sgvsbu[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.sgvsbu$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)

#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")
```

# Stress vs Control

```{r uvss-vulcan, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes.uvss[DEGenes.uvss$logFC>2.5 & DEGenes.uvss$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.uvss[DEGenes.uvss$logFC< -2.5 & DEGenes.uvss$adj.P.Val<0.05,] )

ggplot(data=DEGenes.uvss,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Stress vs Control")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```

## Stress vs Control (Volcano Plot with Gene Length)

```{r}

DEGenes.uvss$Gene <- row.names(DEGenes.uvss)
colnames(genelength) <- c("Length","Gene")
DEGenes.uvss <- merge(DEGenes.uvss, genelength, by="Gene")

ggplot(data=DEGenes.uvss,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Stress vs Control")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.uvss)
liner.SG <- rstandard(smooth.SG)
DEGenes.uvss <-cbind(DEGenes.uvss, liner.SG)
colnames(DEGenes.uvss)[9] <- "Residuals"
row.names(DEGenes.uvss) <- DEGenes.uvss$Gene
linemarkers <- as.data.frame(DEGenes.uvss[abs(DEGenes.uvss$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.uvss$t, log10(DEGenes.uvss$Length), method = "spearman")



ggplot(DEGenes.uvss, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("Stress vs Control Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```


## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes.uvss <- DEGenes.uvss[order(DEGenes.uvss[,4]),]
gseatool <- DEGenes.uvss[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.uvss$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```


## GSEA Length

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, results=FALSE}
#for all genes
DEGenes.uvss <- DEGenes.uvss[order(DEGenes.uvss[,8]),]
gseatool <- DEGenes.uvss[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.uvss$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)

#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")
```

# Interaction (SG)

```{r int-vulcan, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes.int[DEGenes.int$logFC>1.25 & DEGenes.int$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.int[DEGenes.int$logFC< -1.2 & DEGenes.int$adj.P.Val<0.05,] )


ggplot(data=DEGenes.int,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Interaction (SG)")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
## Interaction (SG) (Volcano Plot with Gene Length)

```{r}

DEGenes.int$Gene <- row.names(DEGenes.int)
colnames(genelength) <- c("Length","Gene")
DEGenes.int <- merge(DEGenes.int, genelength, by="Gene")

ggplot(data=DEGenes.int,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Interaction (SG)")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.int)
liner.SG <- rstandard(smooth.SG)
DEGenes.int <-cbind(DEGenes.int, liner.SG)
colnames(DEGenes.int)[9] <- "Residuals"
row.names(DEGenes.int) <- DEGenes.int$Gene
linemarkers <- as.data.frame(DEGenes.int[abs(DEGenes.int$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.int$t, log10(DEGenes.int$Length), method = "spearman")



ggplot(DEGenes.int, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

## GSEA

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes.int <- DEGenes.int[order(DEGenes.int[,4]),]
gseatool <- DEGenes.int[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.int$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```


## GSEA Length

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, results=FALSE}
#for all genes
DEGenes.int <- DEGenes.int[order(DEGenes.int[,8]),]
gseatool <- DEGenes.int[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.int$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)

#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")
```

# Correlation SGS vs SGU Enrichment

```{r}
sgs_t <- merge(stress_rnpt, unstress_t, by="Gene")
sgs_t <- merge(sgs_t, genelength, by="Gene")



smooth.SG <- lm(uns_t~(srnp_t), sgs_t)
liner.SG <- rstandard(smooth.SG)
sgs_t <-cbind(sgs_t, liner.SG)
colnames(sgs_t)[5] <- "Residuals"
row.names(sgs_t) <- sgs_t$Gene
linemarkers <- as.data.frame(sgs_t[(sgs_t$st)>20,])

cor.test(sgs_t$uns_t, (sgs_t$srnp_t), method="spearman")

ggplot(sgs_t, aes(y=uns_t, x=(srnp_t))) +geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+
 xlab("Stressed SG Enrichment") + ylab("Unstressed SG Enrichment") + geom_density2d(colour="red")+ scale_colour_gradient(low="black", high="red")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("SGS vs SGU Enrichment")+stat_cor(method = "spearman", cor.coef.name = "rho")+
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed") +
  geom_label_repel(linemarkers[1:ndegs,] , mapping=aes(label=rownames(linemarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=linemarkers, aes(x=srnp_t, y= uns_t), color="#399e5a")
```

# Alternative Methods (after centrifugation findings)
 
## Repetition of Interaction but with matrix centering

```{r}

#Lets create a dataframe with this information
centrifugation <- c(rep("Yes",6), rep("No",6))
stresstype <- c(rep("Ct",3), rep("Stressed",3),rep("Ct",3), rep("Stressed",3))

metadata <- data.frame(Centrifugation=centrifugation, Stress=stresstype)

mm <- model.matrix(~1+Centrifugation*Stress, metadata)
fitted <- lmFit(y, mm)



mm[,2] <- c(rep(0.5,6), rep(-0.5,6))
mm[,3] <- c(rep(-0.5,3), rep(0.5,3),rep(-0.5,3), rep(0.5,3))
mm[,4] <- mm[,2]*mm[,3]

head(coef(fitted))

fitted <- lmFit(y$E, mm)

top.table <- topTable(tmp,4, sort.by = "logFC", n=Inf)
DEGenes_sg.pres <- top.table

ndegs=20
```

```{r}
tmp <- eBayes(fitted)

top_interact <- topTable(tmp, coef="CentrifugationYes:StressStressed", sort.by="logFC", n=Inf)
top_centri   <- topTable(tmp, coef="CentrifugationYes", sort.by="logFC", n=Inf)
top_stress   <- topTable(tmp, coef="StressStressed", sort.by="logFC", n=Inf)

interact_t <- data.frame(Gene=row.names(top_interact), interact_t=top_interact$t)
centri_t   <- data.frame(Gene=row.names(top_centri), centri_t=top_centri$t)
stress_t   <- data.frame(Gene=row.names(top_stress), stress_t=top_stress$t)

MitoGenes <- fread("~/MitoGenes.txt")[[1]]
NucGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]

## ---- Interaction vs Centrifugation component ----

comp1_df <- merge(interact_t, centri_t, by="Gene")
row.names(comp1_df) <- comp1_df$Gene
MitoData_c1 <- comp1_df[comp1_df$Gene %in% MitoGenes, ]
NucData_c1  <- comp1_df[comp1_df$Gene %in% NucGenes, ]

spearman_c1 <- cor(comp1_df$interact_t, comp1_df$centri_t, method = "spearman")

(p_c1 <- ggplot(comp1_df, aes(x=centri_t, y=interact_t)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData_c1, aes(x=centri_t, y=interact_t), color="#73937E", size=4) +
  geom_point(data=NucData_c1, aes(x=centri_t, y=interact_t), color="#CB643F", size=4) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color="grey40") +
  theme_bw() + ylab("Interaction (t-stat)") + xlab("Centrifugation Component (t-stat)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Interaction vs Centrifugation Component") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=15, family="Arial MS")) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size=5))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Tmat19_InteractionVsCentri.png", width = 7, height=6, units = "in", res=300)
p_c1
dev.off()

## ---- Interaction vs Stress component ----

comp2_df <- merge(interact_t, stress_t, by="Gene")
row.names(comp2_df) <- comp2_df$Gene
MitoData_c2 <- comp2_df[comp2_df$Gene %in% MitoGenes, ]
NucData_c2  <- comp2_df[comp2_df$Gene %in% NucGenes, ]

spearman_c2 <- cor(comp2_df$interact_t, comp2_df$stress_t, method = "spearman")

(p_c2 <- ggplot(comp2_df, aes(x=stress_t, y=interact_t)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData_c2, aes(x=stress_t, y=interact_t), color="#73937E", size=4) +
  geom_point(data=NucData_c2, aes(x=stress_t, y=interact_t), color="#CB643F", size=4) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color="grey40") +
  theme_bw() + ylab("Interaction (t-stat)") + xlab("Stress Component (t-stat)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Interaction vs Stress Component") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=15, family="Arial MS")) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size=5))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Tmat19_InteractionVsStress.png", width = 7, height=6, units = "in", res=300)
p_c2
dev.off()
```


### Volcano Plot
```{r}

write.table(x = DEGenes_sg.pres,
            file = "~/Projects/StressGranules/SGPAPER/RealSGs/RealMatheny2019.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)


upmarkers <- as.data.frame(DEGenes_sg.pres[DEGenes_sg.pres$logFC>1.25 & DEGenes_sg.pres$B>0,] )
downmarkers <- as.data.frame(DEGenes_sg.pres[DEGenes_sg.pres$logFC< -1.25 & DEGenes_sg.pres$B>0,] )

ggplot(data=DEGenes_sg.pres,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#769fca",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#de425b",max.overlaps = 20)+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#769fca")+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#de425b")

DEGenes <- DEGenes_sg.pres
MitoGenes <- fread("~/MitoGenes.txt")[[1]]
MitoData <- DEGenes[rownames(DEGenes) %in% MitoGenes, ]
NucGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]
NucData <- DEGenes[rownames(DEGenes) %in% NucGenes, ]
(m <- ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=B)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=NucData, aes(x=logFC, y=B), color="#CB643F", size=4) +
  geom_point(data=MitoData, aes(x=logFC, y=B), color="#73937E", size=4) +
  ylab("B-Statistic") +
  xlab(expression(Log[2]~"Fold-Change")) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),   # remove major gridlines
    panel.grid.minor = element_blank(),   # remove minor gridlines
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/TMat19Mito.png", width = 7, height=6, units = "in", res=300)
m
dev.off()

```
# MA Plot

```{r}
# MA plot: mean expression (x) vs logFC (y), same highlighting scheme as the volcano

ggplot(data=DEGenes,
                 aes(x=AveExpr,
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=AveExpr, y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=AveExpr, y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  theme_bw() + ylab(expression(Log[2]~"Fold-Change")) + xlab("Average Expression (Log2 CPM)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("T. Matheny Both - MA Plot") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,], mapping=aes(x=AveExpr, y=logFC, label=row.names(upmarkers)[1:ndegs]), size=4, point.padding=-0.5, color="#A10702", max.overlaps=20, fontface="italic") +
  geom_label_repel(downmarkers[1:ndegs,], mapping=aes(x=AveExpr, y=logFC, label=row.names(downmarkers)[1:ndegs]), size=4, point.padding=-0.5, color="#03256C", max.overlaps=20, fontface="italic") +
  geom_point(data=upmarkers, aes(x=AveExpr, y=logFC), color="#A10702", size=4) +
  geom_point(data=downmarkers, aes(x=AveExpr, y=logFC), color="#03256C", size=4) +
  theme(text=element_text(size=15, family="Arial MS"))

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=AveExpr,
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=AveExpr, y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=AveExpr, y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) +
  xlab("Average Expression (Log2 CPM)") + geom_density2d(color = "#800020", size = 1.2) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+
  stat_cor(data = MitoData, aes(x=AveExpr, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#73937E", label.y.npc = 0.1) +
  stat_cor(data = NucData, aes(x=AveExpr, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#CB643F", label.y.npc = 0.2) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Tmat19Mito_MA.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()
```
## MA Plot

```{r}
MitoGenes <- fread("~/MitoGenes.txt")[[1]]

genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)
genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")
genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()
genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")

genelength$Cromo <- ifelse(genelength$Gene %in% MitoGenes, "Mito", "Other")

# Apply the length-based weight formula
genelength$S <- ifelse(
  genelength$Cromo == "Mito",
  (genelength$Length)^(1/3) * 3.23e-11 + 1.17e-9,   # Formula for mitochondrial RNAs
  (genelength$Length)^(1/3) * 6.09e-11 + 3.88e-10    # Formula for other RNAs
)
DEGenes$Gene <- row.names(DEGenes)
DEGenes <- merge(DEGenes, genelength, by="Gene")
MitoGenes <- fread("~/MitoGenes.txt")[[1]]
MitoData <- DEGenes[DEGenes$Gene %in% MitoGenes, ]
NucGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]
NucData <- DEGenes[DEGenes$Gene %in% NucGenes, ]

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=log10(Length),
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=log10(Length), y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=log10(Length), y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) +
  xlab("RNA Length (log10)") + geom_density2d(color = "#800020", size = 1.2) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+
  stat_cor(data = MitoData, aes(x=log10(Length), y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#73937E", label.y.npc = 0.1) +
  stat_cor(data = NucData, aes(x=log10(Length), y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#CB643F", label.y.npc = 0.2) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/TmatMito_MA_length.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=(S),
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=(S), y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=(S), y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) + geom_density2d(color = "#800020", size = 1.2) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+
  stat_cor(data = MitoData, aes(x=S, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#73937E", label.y.npc = 0.1) +
  stat_cor(data = NucData, aes(x=S, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#CB643F", label.y.npc = 0.2) +
  xlab("S") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/TmatMito_MA_S.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()
```

### Volcano Plot (Length)

```{r}

DEGenes_sg.pres$Gene <- row.names(DEGenes_sg.pres)
DEGenes_sg.pres <- merge(DEGenes_sg.pres, genelength, by="Gene")

ggplot(data=DEGenes_sg.pres,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ ylab("B stat")+xlab("Log 2 Fold Change") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "blue", high = "red")

smooth.SG <- lm(t~log10(Length), DEGenes_sg.pres)
liner.SG <- rstandard(smooth.SG)
DEGenes_sg.pres <-cbind(DEGenes_sg.pres, liner.SG)
colnames(DEGenes_sg.pres)[11] <- "Residuals"
row.names(DEGenes_sg.pres) <- DEGenes_sg.pres$Gene
linemarkers <- as.data.frame(DEGenes_sg.pres[abs(DEGenes_sg.pres$Residuals)>4,])

cor.test(DEGenes_sg.pres$t, log10(DEGenes_sg.pres$Length), method="spearman")

ggplot(DEGenes_sg.pres, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("t-stat") + xlab("Length") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("SG")+
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho")
```
## Violins

```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(rstatix)

expression.df <- y$E
centri_grp <- metadata$Centrifugation
stress_grp <- metadata$Stress
Condition  <- interaction(centri_grp, stress_grp, sep=".")

custom_palette <- c(
  "No.Ct"        = "#a2d1e6",
  "No.Stressed"  = "#769fca",
  "Yes.Ct"       = "#e88184",
  "Yes.Stressed" = "#de425b"
)

# ── 1. Gene sets already loaded earlier in the script (MitoGenes, NucGenes) ──
mito_genes_present <- MitoGenes[MitoGenes %in% rownames(expression.df)]
nuc_genes_present  <- NucGenes[NucGenes %in% rownames(expression.df)]
cat(length(mito_genes_present), "Mito genes found in expression matrix\n")
cat(length(nuc_genes_present), "Nuc genes found in expression matrix\n")

# ── 2. Rank all genes per sample (1 = least expressed) ───────────────────────
rank_matrix <- apply(expression.df, 2, function(x) rank(x, ties.method = "average"))
rank_matrix <- as.data.frame(rank_matrix)

# ── 3. Compute a rank-sum score per sample, one gene set at a time ───────────
compute_score <- function(genes, rank_matrix, n_total) {
  colSums(rank_matrix[genes, , drop = FALSE]) / n_total
}

mito_score <- compute_score(mito_genes_present, rank_matrix, nrow(expression.df))
nuc_score  <- compute_score(nuc_genes_present, rank_matrix, nrow(expression.df))

# ── 4. Assemble plot dataframe, long format across both gene sets ────────────
score_df <- rbind(
  data.frame(Sample = colnames(expression.df), Score = as.numeric(mito_score),
             CentriGrp = centri_grp, StressGrp = stress_grp,
             Condition = Condition, GeneSet = "Mito"),
  data.frame(Sample = colnames(expression.df), Score = as.numeric(nuc_score),
             CentriGrp = centri_grp, StressGrp = stress_grp,
             Condition = Condition, GeneSet = "Nuc")
)

# ── 5. Statistics per gene set: two-way ANOVA (Centrifugation*Stress), interaction term ──
stat_labels <- score_df %>%
  group_by(GeneSet) %>%
  group_modify(~{
    aov_res  <- aov(Score ~ CentriGrp * StressGrp, data = .x)
    aov_tidy <- summary(aov_res)

    # The interaction term is the third row of the ANOVA table
    p_raw  <- aov_tidy[[1]][["Pr(>F)"]][3]
    p_adj  <- p.adjust(p_raw, method = "BH")

    ss_interact <- aov_tidy[[1]][["Sum Sq"]][3]
    ss_total    <- sum(aov_tidy[[1]][["Sum Sq"]])
    eta_sq      <- ss_interact / ss_total
    cohens_f    <- sqrt(eta_sq / (1 - eta_sq))

    data.frame(
      label = paste0("Interaction Cohen's f = ", round(cohens_f, 3),
                      "\nadj. p = ", formatC(p_adj, format = "e", digits = 2))
    )
  })

# ── 6. Plot ───────────────────────────────────────────────────────────────────
vilsigplot <- ggplot(score_df, aes(x = Condition, y = Score, fill = Condition)) +
  geom_violin(colour = "black", alpha = 0.6, trim = FALSE) +
  geom_point(
    position = position_jitter(width = 0.1),
    alpha    = 0.7,
    size     = 1.5
  ) +
  stat_summary(fun = mean,   geom = "point", shape = 18,
               size = 3, colour = "black") +
  stat_summary(fun = median, geom = "point", shape = 17,
               size = 3, colour = "black") +
  geom_text(
    data  = stat_labels,
    aes(x = -Inf, y = Inf, label = label),
    hjust = -0.05, vjust = 1.2,
    size  = 4, fontface = "bold", lineheight = 1.1,
    inherit.aes = FALSE
  ) +
  facet_wrap(~GeneSet, scales = "free_y") +
  scale_fill_manual(values = custom_palette) +
  labs(
    x     = "Centrifugation : Stress",
    y     = "Signature Score (Rank Sum)",
    title = "Mitochondrial vs Nuclear-Localized Gene Scores"
  ) +
  theme_bw() +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x     = element_text(size = 12, angle=20, hjust=1),
    strip.text      = element_text(size = 12, face = "bold"),
    legend.position = "none"
  )

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Tmat19_MitoSigs.png", width = 14, height=6, units = "in", res=300)
vilsigplot
dev.off()

vilsigplot
```


### GSEA

```{r fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes_sg.pres <- DEGenes_sg.pres[order(DEGenes_sg.pres[,4]),]
gseatool <- DEGenes_sg.pres[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes_sg.pres$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```


### For Mito

```{r}
# --- Rank genes by t-statistic as usual ---
DEGenes_sg.pres <- DEGenes_sg.pres[order(DEGenes_sg.pres[,4]),]
gseatool <- DEGenes_sg.pres[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes_sg.pres)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]
vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol

# --- Define MitoGenes as a custom gene set ---
MitoGenes <- fread("~/MitoGenes.txt")[[1]]
custom.gsets <- list(MitoGenes = toupper(MitoGenes))

# --- Run fgsea ---
mito.res <- fgsea(pathways = custom.gsets,
                  stats = vectorranks)

mito.plot.base <- plotEnrichment(custom.gsets[["MitoGenes"]], vectorranks)
curve.data <- mito.plot.base$data

# --- key fix: match the curve's internal descending order ---
ord <- order(vectorranks, decreasing = TRUE)
ranked.names <- names(vectorranks)[ord]
gene.positions <- which(ranked.names %in% custom.gsets[["MitoGenes"]])

rug.top    <- min(curve.data$y) - 0.03
rug.bottom <- rug.top - 0.04

y.min <- rug.bottom - 0.03
y.max <- max(curve.data$y) + 0.05

(mito.plot <- ggplot(curve.data, aes(x = x, y = y)) +
  geom_area(fill = "#73937E", alpha = 0.4) +
  geom_line(color = "#73937E", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_segment(data = data.frame(x = gene.positions),
               aes(x = x, xend = x, y = rug.top, yend = rug.bottom),
               color = "#73937E", alpha = 0.6, linewidth = 0.9,
               inherit.aes = FALSE) +
  annotate("text", x = 1500, y = 0.2,
           label = paste0("NES = ", round(mito.res$NES, 2),
                          "\nadj.p = ", round(mito.res$padj, 4)),
           size = 5, family = "Arial MS", fontface = "bold") +
  coord_cartesian(ylim = c(y.min, y.max), clip = "off") +
  theme_bw() +
  theme(text = element_text(size = 25, family = "Arial MS"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/MitoGSEA_TMat19.png", width=7, height=6, units="in", res=300)
mito.plot
dev.off()
```

The results are very similar with and without centering.

## SG as a Variable (for variance % calc)

```{r}

#Lets create a dataframe with this information
centrifugation <- c(rep("Yes",6), rep("No",6))
stresstype <- c(rep("Ct",3), rep("Stressed",3),rep("Ct",3), rep("Stressed",3))
sgpres <- c(rep("No",3), rep("Yes",3), rep("No",6))

metadata <- data.frame(Centrifugation=centrifugation, Stress=stresstype, SG=sgpres)

mm <- model.matrix(~0+Centrifugation+Stress+SG, metadata)
fitted <- lmFit(y, mm)

head(coef(fitted))

fitted <- lmFit(y$E, mm)

top.table <- topTable(tmp,4, sort.by = "logFC", n=Inf)
DEGenes_sg.pres <- top.table

ndegs=20
```

```{r}

get_variance_explained_manual <- function(mat, metadata, variables) {
  # mat: genes x samples (rows = genes, cols = samples)
  # metadata: dataframe of sample annotations
  # variables: character vector of column names in metadata
  
  message("Calculating variance explained per gene...")
  
  # Ensure sample ordering matches
  stopifnot(ncol(mat) == nrow(metadata))
  
  results <- lapply(variables, function(var) {
    message("Processing ", var, "...")
    sapply(1:nrow(mat), function(i) {
      y <- as.numeric(mat[i, ])
      df <- data.frame(y = y, x = metadata[[var]])
      # handle categorical or numeric
      model <- lm(y ~ x, data = df)
      summary(model)$r.squared
    })
  })
  
  names(results) <- variables
  varMat <- as.data.frame(results)
  rownames(varMat) <- rownames(mat)
  return(varMat)
}
plot_variance_explained_density <- function(varMat, variables, colours, scale = "log") {
  df_plot <- reshape2::melt(varMat, variable.name = "Variable", value.name = "VarianceExplained")
  
  # Convert to percentage variance explained
  df_plot$VarianceExplained <- df_plot$VarianceExplained * 100
  
  # Convert input colours to pastel and name them to ensure correct mapping
  pastel_colours <- colorspace::lighten(colours, amount = 0.5)
  names(pastel_colours) <- variables
  
  # Build x scale based on argument
  x_scale <- if (scale == "log") {
    scale_x_log10(
      name = "% variance explained",
      breaks = c(0.001, 0.01, 0.1, 1, 10, 100),
      labels = c("0.001", "0.01", "0.1", "1", "10", "100")
    )
  } else if (scale == "linear") {
    scale_x_continuous(
      name = "% variance explained",
      breaks = c(0, 20, 40, 60, 80, 100),
      labels = c("0", "20", "40", "60", "80", "100"),
      limits = c(0, 100)
    )
  } else {
    stop("Invalid scale argument. Use 'log' or 'linear'.")
  }
  
  ggplot(df_plot, aes(x = VarianceExplained, color = Variable)) +
    geom_density(size = 1.2, adjust = 1.2) +
    x_scale +
    ylab("Density") +
    theme_bw(base_size = 16) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 20),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank()
    )  +
    scale_color_manual(values = pastel_colours)
}
#Example Usage
   #expression is your normalised expression table, with genes as rows and samples as coloumns (most likely y$E if you used voom)
   #metadata is your metadata frame where each column is a variable, and each row a sample. Must be in the same order as expression
   #variables must be a vector of the variables (columns in metadata) that you want to test. So something like  c("Centrifugation", "Stress", "SG")
   #colours must be a vector of the colours you want to use. Must be the same number as the variables. So something like c("#229954", "#EC7063", "blue")
varMat <- get_variance_explained_manual(y$E, metadata, colnames(metadata))
# Log scale (default — no need to specify)
p <- plot_variance_explained_density(varMat, colnames(metadata), c("red", "blue", "green"))
print(p)
# Linear scale
p <- plot_variance_explained_density(varMat, colnames(metadata), c("red", "blue", "green"), scale = "linear")
print(p)

png("~/Projects/StressGranules/SGPAPER/Images/PNG/SupFigure4.png", width = 10, height=6, units = "in", res=300)
p
dev.off()

```


# Session Info

```{r}
sessionInfo()
```
