---
title: "Analyses of GSE212380"
subtitle: "N4-acetylcytidine (ac4C) promotes mRNA localization to stress granules."
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

In this document, I present all analyses pertaining to the dataset **GSE212380**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE212380)),  
originally published by **P. Kudrin, 2024**.

```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)

set.seed(1906)
```


## Loading up the data

```{r}
#This section here is to load up the genelenght file I will be using throughout the analyses.
genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)

genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")

genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()

genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")

#Get the table of read counts by indicating the path to the file. This file is what we obtained from the running featureCounts.
raw.read.counts <- read.table("~/Projects/StressGranules/SG_Transcriptome/ac4C/TrimmedSamples/alignment_STAR/featureCounts_Results.txt", header = T)[,-c(2:6)]

biomart <- fread("~/biomart.txt")[,-1]


colnames(raw.read.counts)[1] <- "ensembl_gene_id_version"
colnames(biomart) <- c("ensembl_gene_id_version", "hgnc_symbol")

eltable <- merge(raw.read.counts,biomart,by = "ensembl_gene_id_version", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="") #this removes reads that dont have a corresponding hgnc symbol.

eltable$ensembl_gene_id_version <- NULL

read.counts <- eltable %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

read.counts <- as.data.frame(read.counts)

row.names(read.counts) <- read.counts$hgnc_symbol #changes rownames into hgnc symbols
read.counts$hgnc_symbol <- NULL #We no longer need this column, as we have the rownames.

```


## Section for SPINSIM Preparation

This serves as an example

```{r}
mof <- fread("~/Projects/StressGranules/Simulator/MolecularWeights.csv")
colnames(mof)[2] <- "GeneID" 
mof <- mof[!duplicated(mof$GeneID), ]


stress.tpm <- read.counts[,8]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/ac4C/NAT_4.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)
```


## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, warning=FALSE, message=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(4, palette = "Pastel")
custom_palette <- sample(custom_palette, 4)

# Use SampleColouring from metadata
SampleColouring <- sampnamer <- c(rep("BUWT",4), rep("BUNT",4), rep("SGWT",4), rep("SGNT", 4))

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(4, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )
```


As we can see, there is at least one SGWT with lower complexity. It is useful to keep it in mind for downstream steps.

# Differential Expression Analysis

## Filtration

### ***Before Filtration***

Let us begin with the filtration. Personally, I find that its easier to do visually. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bi-modal distributions, with the first mode corresponding to lowly expressed genes. I remove these by cutting just before the second mode.

```{r filtration, warning=FALSE}
alternew.SGKO <- log10(rowSums(read.counts[,1:4]))
alternew.SGKO <- as.data.frame(alternew.SGKO)

alternew.SGWT <- log10(rowSums(read.counts[,5:8]))
alternew.SGWT <- as.data.frame(alternew.SGWT)

alternew.BUKO <- log10(rowSums(read.counts[,9:12]))
alternew.BUKO <- as.data.frame(alternew.BUKO)

alternew.BUWT <- log10(rowSums(read.counts[,13:15]))
alternew.BUWT <- as.data.frame(alternew.BUWT)

alternew <- log10(rowSums(read.counts))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3.25, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.SGKO, aes(x=alternew.SGKO)) + geom_density(fill="#a2d1e6", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - SG NAT10 KO")

ggplot(alternew.SGWT, aes(x=alternew.SGWT)) + geom_density(fill="#769fca", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - SG WT")

ggplot(alternew.BUKO, aes(x=alternew.BUKO)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - BU NAT10 KO")

ggplot(alternew.BUWT, aes(x=alternew.BUWT)) + geom_density(fill="#de425b", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.25, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - BU WT")

```


By seeing the shape of the graph, we can see where we should cut. We have a **binomial distribution** (low reads and noise vs actual biological meaningful reads). The ideal place to cut them is right before the wedge.

### ***After Filtration***

I didn't do it this time, but I've maintained the code to filter by sample type. This time, I considered that it was more correct to filter all samples together.


```{r}
filterednew <- subset(read.counts, alternew>3.25)
filterednew.SGKO <- subset(read.counts, alternew.SGKO>3)
filterednew.SGWT <- subset(read.counts, alternew.SGWT>2.75)
filterednew.BUKO <- subset(read.counts, alternew.BUKO>2)
filterednew.BUWT <- subset(read.counts, alternew.BUWT>2.25)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))
impgenes.SGKO <- c(row.names(filterednew.SGKO))
impgenes.SGWT <- c(row.names(filterednew.SGWT))
impgenes.BUKO <- c(row.names(filterednew.BUKO))
impgenes.BUWT <- c(row.names(filterednew.BUWT))

#Now we need to select only the rows which have those genes.
finaltable <- read.counts[row.names(read.counts)%in%impgenes,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.SGKO,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.SGWT,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.BUKO,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.BUWT,]
```


## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#Let us prepare metadata to add to our samples
forsamp <- c(rep("Bulk",8), rep("SG",8))

forcell <- c(rep("WT",4), rep("NT",4),rep("WT",4), rep("NT",4))

#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)

#for the others, we need to select only relevant columns. 

#Then we calculate the normalization factors.
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)


#Now we do our matrix itself
mm <- model.matrix(~0+forsamp+forcell)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```


As we can see, the **mean-variance trend** shows a good curve, showing that the filtration was well done.

## Normalisation

For normalisation I convert the read counts into log2CPMs. Its a good way to visualise how the reads are. 

```{r}

dfforboxy <- finaltable

orderforplots <- sampnamer
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#43F3D5", "#769fca", "#e88184", "#de425b"), labels=unique(sampnamer), breaks =unique(sampnamer)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#43F3D5", "#769fca", "#e88184", "#de425b"), labels=unique(sampnamer), breaks =unique(sampnamer)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```


## PCA

Before we move on, lets do a **Principal Component analysis**. It is a good exploratory tool to find out how our samples relate to each other and if there are any obvious outliers. We would expect our replicates to be nearer to one another than to other samples, and more different samples should be further away. 

```{r, fig.height=7, fig.width=9}
# Define color palette
custom_palette <- qualitative_hcl(4, palette = "Pastel")
custom_palette <- sample(custom_palette, 4)

#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- forsamp
dfforpca$Colour <- SampleColouring
dfforpca$Cell <- forcell

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-3)], scale = FALSE)

p_Sample <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Sample") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())


p_Cell <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Cell", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Cell", values = custom_palette) +
  ggtitle("PCA - Coloured by Cell") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Colour <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Colour", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Colour", values = custom_palette) +
  ggtitle("PCA - Coloured by Colour") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Sample
p_Cell
p_Colour

#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])


```

Bulk Samples clearly separate from SG samples, although there is one BUWT sample that is very similar to BUNT samples.

## Linear Models

Now we must perform linear model fitting.

```{r}

colnames(mm) <- c("forsampBulk","forsampSG","forcellWT")
fitted <- lmFit(y, mm)
head(coef(fitted))


#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(forsampSG  - forsampBulk, levels = mm)


#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```

# SG vs WC

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
write.table(x = DEGenes,
            file = "~/Projects/StressGranules/SGPAPER/SGvsWC/SGAC.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2.5 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2.5 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "pearson")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

We observe a positive length bias.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}

finagenes <- rbind(upmarkers, downmarkers)

dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:16))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=SampleColouring)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


## GSEA Length

Now, since SGs are enriched for longer transcripts, these pathways may simply be the ones which are tipically enriched in longer transcripts. Therefore, I will repeat the analysis, but using transcript lenght instead of t-statistic. 

```{r, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,8]),]
gseatool <- DEGenes[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)

#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg_pval", "sg", "size_pval", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")

```

Barring some exceptions which I labeled on the plot, most gene set enrichment is driven by transcript size.

# NAT10 WT vs KO

```{r}
head(coef(fitted))


tmp <- eBayes(fitted)


top.table <- topTable(tmp, 3, sort.by = "logFC", n=Inf)
DEGenes <- top.table
```

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2.5 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2.5 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("NAT10 WT vs KO")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```
## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("NAT10 WT vs KO")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "pearson")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("NAT10 WT vs KO") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

WT cells have a "shorter overall transcriptome". That is, they express more of shorter genes.

# WT SG vs BU

```{r}

#Now we do our matrix itself
mm <- model.matrix(~0+sampnamer)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
fitted <- lmFit(y, mm)
head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(sampnamerSGWT  - sampnamerBUWT, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2.5 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2.5 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

wt_t <- data.frame(Gene=row.names(DEGenes), wt_t = DEGenes$t)
```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "pearson")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

We observe a positive length bias, even with the WC WT outlier.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}

finagenes <- rbind(upmarkers, downmarkers)

dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:16))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=SampleColouring)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


## GSEA Length

Now, since SGs are enriched for longer transcripts, these pathways may simply be the ones which are tipically enriched in longer transcripts. Therefore, I will repeat the analysis, but using transcript lenght instead of t-statistic. 

```{r, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,8]),]
gseatool <- DEGenes[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)

#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg_pval", "sg", "size_pval", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")

```

# NT SG vs BU

```{r}

#Now we do our matrix itself
mm <- model.matrix(~0+sampnamer)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
fitted <- lmFit(y, mm)
head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(sampnamerSGNT  - sampnamerBUNT, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2.5 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2.5 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

nt_t <- data.frame(Gene=row.names(DEGenes), nt_t = DEGenes$t)
```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "pearson")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

We observe a positive length bias, even with the WC WT outlier.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}

finagenes <- rbind(upmarkers, downmarkers)

dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:16))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=SampleColouring)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


## GSEA Length

Now, since SGs are enriched for longer transcripts, these pathways may simply be the ones which are tipically enriched in longer transcripts. Therefore, I will repeat the analysis, but using transcript lenght instead of t-statistic. 

```{r, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,8]),]
gseatool <- DEGenes[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)

#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg_pval", "sg", "size_pval", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")

```

# Comparisons

```{r}

ac4c <- fread("~/Projects/StressGranules/Simulator/MolecularWeights_ac4C.csv")[,c(1,3,5)]

wtvsnt <- merge(wt_t, nt_t, by="Gene")
colnames(ac4c) <- c("Gene", "NAC", "Weight")
wtvsnt <- merge(wtvsnt, ac4c, all.y=F, all.x=T)
wtvsnt[is.na(wtvsnt)] <- 0

smooth.SG <- lm(wt_t~(nt_t), wtvsnt)
liner.SG <- rstandard(smooth.SG)
wtvsnt <-cbind(wtvsnt, liner.SG)
colnames(wtvsnt)[6] <- "Residuals"
#row.names(wtvsnt) <- wtvsnt$Gene
linemarkers <- as.data.frame(wtvsnt[abs(wtvsnt$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(wtvsnt$wt_t, (wtvsnt$nt_t), method = "spearman")

ogres <- wtvsnt[,c(1,6)]

ggplot(wtvsnt, aes(y=wt_t, x=(nt_t))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)  +   theme_bw()+xlab("ACKO SG Enrichment")+
  ylab("WT SG Enrichment")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(text=element_text(size=15,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=5)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")


ggplot(wtvsnt, aes(y=NAC, x=(liner.SG))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ 
 ylab("Number of ac4C") + xlab("Residuals") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  + stat_cor(method = "spearman", cor.coef.name = "rho", size=5)


wtvsnt$bins <- ifelse(wtvsnt$NAC <= 8, wtvsnt$NAC,
                      ifelse(wtvsnt$NAC>=9 & wtvsnt$NAC<=12, "[9,12]","[13,["))


wtvsnt$bins <- factor(wtvsnt$bins, levels=c(0:8, "[9,12]","[13,[" ))
ggplot(wtvsnt, aes(x=bins, y=(liner.SG)))  + geom_jitter(alpha=0.2) + geom_boxplot(fill="#43F3D5", alpha=0.4) +theme_bw()+ 
 xlab("Number of ac4C") + ylab("Residuals") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) 

```
Enrichment is very similar between WT and ACKO stress granules. Additionally, it seems most of these differences is indeed linked to number of ac4C modifications.

Just to confirm, I will directly compare WC with WC samples and SG vs SG samples. If the differences are driven by SG specific differences, the enrichment should be different between sample types.

# NT WC vs WT WC

```{r}

#Now we do our matrix itself
mm <- model.matrix(~0+sampnamer)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
fitted <- lmFit(y, mm)
head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(sampnamerBUNT  - sampnamerBUWT, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```

## Volcano Plot

```{r}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2.5 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2.5 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("WC vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```

## Volcano Plot with Transcript Length

```{r}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("WC vs WC")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("KO Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)
```

So, KO samples are "longer" than their wt counterparts.

# NT SG vs WT SG

```{r}

#Now we do our matrix itself
mm <- model.matrix(~0+sampnamer)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
fitted <- lmFit(y, mm)
head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(sampnamerSGNT  - sampnamerSGWT, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```


## Volcano Plot

```{r}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2.5 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2.5 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))
```

## Volcano Plot with Transcript Length

```{r}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("KO Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)
```

But on SG, it is linked with shorter transcripts. This is good! Everything seems to point that the differences in enrichment are indeed driven by ac4C. 

# SPIN-SIM

```{r}
read.counts$Gene <- row.names(read.counts)

goalsum_WT1 <- sum(read.counts$SRR21363237Aligned.sortedByCoord.out.bam)
goalsum_WT2 <- sum(read.counts$SRR21363238Aligned.sortedByCoord.out.bam)
goalsum_WT3 <- sum(read.counts$SRR21363239Aligned.sortedByCoord.out.bam)
goalsum_WT4 <- sum(read.counts$SRR21363240Aligned.sortedByCoord.out.bam)
goalsum_NT1 <- sum(read.counts$SRR21363241Aligned.sortedByCoord.out.bam)
goalsum_NT2 <- sum(read.counts$SRR21363242Aligned.sortedByCoord.out.bam)
goalsum_NT3 <- sum(read.counts$SRR21363243Aligned.sortedByCoord.out.bam)
goalsum_NT4 <- sum(read.counts$SRR21363244Aligned.sortedByCoord.out.bam)

colnames(genelength)[2] <- "Gene"


#For the new samples
simwt1 <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/ac4C/WTSG1.txt", sep = ",")[,c(1,3)]
colnames(simwt1) <- c("Gene", "SGWT1")
simwt2 <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/ac4C/WTSG2.txt", sep = ",")[,c(1,3)]
colnames(simwt2) <- c("Gene", "SGWT2")
simwt3 <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/ac4C/WTSG3.txt", sep = ",")[,c(1,3)]
colnames(simwt3) <- c("Gene", "SGWT3")
simwt4 <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/ac4C/WTSG4.txt", sep = ",")[,c(1,3)]
colnames(simwt4) <- c("Gene", "SGWT4")

simnt1 <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/ac4C/NTSG1.txt", sep = ",")[,c(1,3)]
colnames(simnt1) <- c("Gene", "SGNT1")
simnt2 <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/ac4C/NTSG2.txt", sep = ",")[,c(1,3)]
colnames(simnt2) <- c("Gene", "SGNT2")
simnt3 <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/ac4C/NTSG3.txt", sep = ",")[,c(1,3)]
colnames(simnt3) <- c("Gene", "SGNT3")
simnt4 <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/ac4C/NTSG4.txt", sep = ",")[,c(1,3)]
colnames(simnt4) <- c("Gene", "SGNT4")


#WT Sample 1 is the wt sample that on the PCA appears close to the NTs. I opted to do this part without it.
#simsamples <- merge(simwt1, simwt2, by="Gene")
simsamples <- merge(simwt2, simwt3, by="Gene")
#simsamples <- merge(simsamples, simwt3, by="Gene")
simsamples <- merge(simsamples, simwt4, by="Gene")
simsamples <- merge(simsamples, simnt1, by="Gene")
simsamples <- merge(simsamples, simnt2, by="Gene")
simsamples <- merge(simsamples, simnt3, by="Gene")
simsamples <- merge(simsamples, simnt4, by="Gene")
simsamples <- merge(simsamples, genelength, by="Gene")

simsamples <- simsamples %>%
  group_by(Gene) %>%
  slice(which.max(Length)) %>%
  ungroup()

simsamples <- as.data.frame(simsamples)

colnames(simsamples)[9] <- "Length"

#simsamples$WTCounts_1 <- simsamples$SGWT1*simsamples$Length
#simsamples$WTCounts_1 <- simsamples$WTCounts_1/sum(simsamples$WTCounts_1)*goalsum_WT1
simsamples$WTCounts_2 <- simsamples$SGWT2*simsamples$Length
simsamples$WTCounts_2 <- simsamples$WTCounts_2/sum(simsamples$WTCounts_2)*goalsum_WT2
simsamples$WTCounts_3 <- simsamples$SGWT3*simsamples$Length
simsamples$WTCounts_3 <- simsamples$WTCounts_3/sum(simsamples$WTCounts_3)*goalsum_WT3
simsamples$WTCounts_4 <- simsamples$SGWT4*simsamples$Length
simsamples$WTCounts_4 <- simsamples$WTCounts_4/sum(simsamples$WTCounts_4)*goalsum_WT4

simsamples$NTCounts_1 <- simsamples$SGNT1*simsamples$Length
simsamples$NTCounts_1 <- simsamples$NTCounts_1/sum(simsamples$NTCounts_1)*goalsum_NT1
simsamples$NTCounts_2 <- simsamples$SGNT2*simsamples$Length
simsamples$NTCounts_2 <- simsamples$NTCounts_2/sum(simsamples$NTCounts_2)*goalsum_NT2
simsamples$NTCounts_3 <- simsamples$SGNT3*simsamples$Length
simsamples$NTCounts_3 <- simsamples$NTCounts_3/sum(simsamples$NTCounts_3)*goalsum_NT3
simsamples$NTCounts_4 <- simsamples$SGNT4*simsamples$Length
simsamples$NTCounts_4 <- simsamples$NTCounts_4/sum(simsamples$NTCounts_4)*goalsum_NT4

simsamples <- simsamples[,c(1,10:16)] 

OGsamples <- read.counts[,c(2:8,17)]

simandog <- merge(simsamples, OGsamples, by="Gene")
row.names(simandog) <- simandog$Gene
simandog$Gene <- NULL
```


## Filtration

### ***Before Filtration***

Let us begin with the filtration. 

```{r, warning=FALSE}
alternew.sgwt <- log10(rowSums(simandog[,1:3]))
alternew.sgwt <- as.data.frame(alternew.sgwt)

alternew.sgnt <- log10(rowSums(simandog[,4:7]))
alternew.sgnt <- as.data.frame(alternew.sgnt)

alternew.buwt <- log10(rowSums(simandog[,8:10]))
alternew.buwt <- as.data.frame(alternew.buwt)

alternew.bunt <- log10(rowSums(simandog[,11:14]))
alternew.bunt <- as.data.frame(alternew.bunt)

alternew <- log10(rowSums(simandog))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.sgwt, aes(x=alternew.sgwt)) + geom_density(fill="#769fca", alpha=0.5) + theme_bw() + geom_vline(xintercept = 1.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Stress Granules Samples - SimWT")

ggplot(alternew.sgnt, aes(x=alternew.sgnt)) + geom_density(fill="#de425b", alpha=0.5) + theme_bw() + geom_vline(xintercept = 1.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Stress Granules Samples - SimNT")

ggplot(alternew.buwt, aes(x=alternew.buwt)) + geom_density(fill="#769fca", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - SBulk Samples - WT")

ggplot(alternew.bunt, aes(x=alternew.bunt)) + geom_density(fill="#de425b", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Bulk Samples - NT")


sampy <- c(rep("SGWTSIM",3), rep("SGNTSIM",4), rep("BUWT",3), rep("BUNT",4))

```

### ***After Filtration***

From my spinsim pipeline, a few outliers usually appear so I opted to filter by separated sample types.

```{r}
filterednew <- subset(simandog, alternew>3)
filterednew.buwt <- subset(simandog, alternew.buwt>2.75)
filterednew.bunt <- subset(simandog, alternew.bunt>2.75)
filterednew.sgwt <- subset(simandog, alternew.sgwt>1.5)
filterednew.sgnt <- subset(simandog, alternew.sgnt>1.5)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))
impgenes.buwt <- c(row.names(filterednew.buwt))
impgenes.bunt <- c(row.names(filterednew.bunt))
impgenes.sgwt <- c(row.names(filterednew.sgwt))
impgenes.sgnt <- c(row.names(filterednew.sgnt))

#Now we need to select only the rows which have those genes.
finaltable <- simandog[row.names(simandog)%in%impgenes,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.buwt,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.bunt,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.sgwt,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.sgnt,]
```

## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)

#Then we calculate the normalization factors.
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)


#Now we do our matrix itself
mm <- model.matrix(~0+sampy)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```


## Normalisation

For normalisation I convert the read counts into log2CPMs. Its a good way to visualise how the reads are. 

```{r}

dfforboxy <- finaltable

orderforplots <- sampy
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#a2d1e6", "#769fca", "green", "purple"), labels=c("SGWTSIM", "SGNTSIM", "BUWT", "BUNT"), breaks =c("SGWTSIM", "SGNTSIM", "BUWT", "BUNT")) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#a2d1e6", "#769fca", "green", "purple"), labels=c("SGWTSIM", "SGNTSIM", "BUWT", "BUNT"), breaks =c("SGWTSIM", "SGNTSIM", "BUWT", "BUNT")) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```



## PCA

```{r}
#this is done to be able to do traditional PCA
dfforpca <- y$E
dfforpca <- rbind(dfforpca, sampy)
dfforpca <- t(dfforpca)
numcol <- c(1:(ncol(dfforpca)-1))
dfforpca <- as.data.frame(dfforpca) #for some reason I can only change colum type if its a dataframe

dfforpca[ ,numcol] <- apply(dfforpca[ , numcol], 2,            # to turn our columns into numeric
                            function(x) as.numeric(as.character(x)))

#sapply(dfforpca, class) #to check class

# Or specify the factor levels in the order you want
dfforpca$forall <- factor(dfforpca$sampy, levels = c("SGWTSIM", "SGNTSIM", "BUWT", "BUNT"))

pca_res <- prcomp(dfforpca[,numcol], scale=FALSE) #prepare the pca

#to make everything pretty
p <- autoplot(pca_res, data = dfforpca, x=1, y=2, colour = "forall", size=8, alpha=0.7)
p <- p + theme_bw() + scale_colour_manual((name = "Samples"), values = c("#a2d1e6", "#769fca", "green", "purple"), labels=c("SGWTSIM", "SGNTSIM", "BUWT", "BUNT"))
p <- p + ggtitle("PCA Plot") + theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())
p


#in case you want to see them yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])
```

From the spin-sim, we can see that my simulated SGs "mirror" their bulk counterparts, from which the expression was based on.

## Linear Models

Now we must perform linear model fitting. Once more, we will have 4, one for each matrix we previously designed.

```{r}
fitted <- lmFit(y, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr <- makeContrasts(sampySGWTSIM - sampyBUWT, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)

top.table <- topTable(tmp, sort.by = "logFC", n = Inf)

DEGenes <- top.table

ndegs <- 20
```
# WT Samples

## Volcano Plots

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>1.25 & DEGenes$B>0,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -1.5 & DEGenes$B>5,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Sim SG WT")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#769fca",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#a2d1e6",max.overlaps = 20)+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#769fca")+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#a2d1e6")

simwt <- data.frame(Gene=row.names(DEGenes), simwt_t=DEGenes$t, simwt_logFC=DEGenes$logFC)
```


```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength)[2] <- "Gene"
colnames(genelength)[1] <- "Length"
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SIM WT SG")+ ylab("B stat")+xlab("Log 2 Fold Change") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "blue", high = "red")


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])
cor.test(DEGenes$t, log10(DEGenes$Length), method="spearman")

ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+
 ylab("t") + xlab("Gene Length (Log10 Base Pairs)") + geom_density2d()+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("t vs Gene Length")+
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed") +
  geom_label_repel(linemarkers[1:ndegs,] , mapping=aes(label=rownames(linemarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)+
  geom_point(data=linemarkers, aes(y=t, x= log10(Length)), color="#399e5a")

```


# KO Samples


```{r}
fitted <- lmFit(y, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr <- makeContrasts(sampySGNTSIM  - sampyBUNT , levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)

top.table <- topTable(tmp, sort.by = "logFC", n = Inf)

DEGenes <- top.table

ndegs <- 20
```

## Volcano Plots

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>1.25 & DEGenes$B>0,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -1.5 & DEGenes$B>5,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Sim SG NT")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#769fca",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#a2d1e6",max.overlaps = 20)+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#769fca")+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#a2d1e6")

simnt <- data.frame(Gene=row.names(DEGenes), simnt_t=DEGenes$t, simnt_logFC=DEGenes$logFC)
```


```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength)[2] <- "Gene"
colnames(genelength)[1] <- "Length"
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SIM NT SG")+ ylab("B stat")+xlab("Log 2 Fold Change") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "blue", high = "red")


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])
cor.test(DEGenes$t, log10(DEGenes$Length), method="spearman")

ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+
 ylab("t") + xlab("Gene Length (Log10 Base Pairs)") + geom_density2d()+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("t vs Gene Length")+
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed") +
  geom_label_repel(linemarkers[1:ndegs,] , mapping=aes(label=rownames(linemarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)+
  geom_point(data=linemarkers, aes(y=t, x= log10(Length)), color="#399e5a")

```

# Comparisons

```{r}
wtvsnt <- merge(simwt, simnt, by="Gene")


smooth.SG <- lm(simwt_t~(simnt_t), wtvsnt)
liner.SG <- rstandard(smooth.SG)
wtvsnt <-cbind(wtvsnt, liner.SG)
colnames(wtvsnt)[6] <- "Residuals"
row.names(wtvsnt) <- wtvsnt$Gene
linemarkers <- as.data.frame(wtvsnt[abs(wtvsnt$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(wtvsnt$simwt_t, (wtvsnt$simnt_t), method = "spearman")



ggplot(wtvsnt, aes(y=simwt_t, x=(simnt_t))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("WT Enrichment") + xlab("KO Enrichemnt") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)



wtvsnt <- merge(wtvsnt, ac4c, all.y=F, all.x=T)
wtvsnt[is.na(wtvsnt)] <- 0

wtvsnt$bins <- ifelse(wtvsnt$NAC <= 8, wtvsnt$NAC,
                      ifelse(wtvsnt$NAC>=9 & wtvsnt$NAC<=12, "[9,12]","[13,["))


wtvsnt$bins <- factor(wtvsnt$bins, levels=c(0:8, "[9,12]","[13,[" ))
ggplot(wtvsnt, aes(x=bins, y=(Residuals)))  + geom_jitter(alpha=0.2) + geom_boxplot(fill="#43F3D5", alpha=0.4) +theme_bw()+ 
 xlab("Number of ac4C") + ylab("Residuals") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(wtvsnt, aes(y=NAC, x=(Residuals))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ 
 xlab("Number of ac4C") + ylab("Residuals") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  + stat_cor(method = "spearman", cor.coef.name = "rho", size=5)

simres <- wtvsnt[,c(1,6)]

colnames(ogres)[2] <- "Residuals"

resvres <- merge(ogres, simres, by="Gene")


smooth.SG <- lm(Residuals.y~(Residuals.x), resvres)
liner.SG <- rstandard(smooth.SG)
resvres <-cbind(resvres, liner.SG)
colnames(resvres)[4] <- "Residuals"
#row.names(resvres) <- resvres$Gene
linemarkers <- as.data.frame(resvres[abs(resvres$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(resvres$Residuals.y, (resvres$Residuals.x), method = "spearman")

genebin <- wtvsnt[,c(1,9)]

resvres <- merge(resvres, genebin, by="Gene")
resvres$bins <- factor(resvres$bins, levels=c(0:8, "[9,12]","[13,[" ))

upmarkers <- as.data.frame(resvres[resvres$bins=="[9,12]" | resvres$bins=="[13,[",] )

ggplot(resvres, aes(y=Residuals.y, x=(Residuals.x))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black')  +   theme_bw()+xlab("Original Residuals")+
  ylab("Simulated Residuals")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(text=element_text(size=15,  family="Arial MS")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_point(data=upmarkers, aes(x=Residuals.x, y=  Residuals.y), color="#43F3D5",fill="#43F3D5",alpha=0.3, size=5, shape=21) 
```
It does seems plausible that the differences in enrichment can be caused not by biological properties of ac4C, but instead simply because of its added weight. In teal, we can see that RNAs with more residues do sediment more.

# "Real" SG

After the spin-sim pipeline, we can then compared the original SGs to our simulated ones, which allows us to determine genes which are more or less enriched than what was expected by centrifugation alone.

```{r}
read.counts$Gene <- row.names(read.counts)

goalsum_SGWT1 <- sum(read.counts[,9])
goalsum_SGWT2 <- sum(read.counts[,10])
goalsum_SGWT3 <- sum(read.counts[,11])
goalsum_SGWT4 <- sum(read.counts[,12])
goalsum_SGNAT1 <- sum(read.counts[,13])
goalsum_SGNAT2 <- sum(read.counts[,14])
goalsum_SGNAT3 <- sum(read.counts[,15])
goalsum_SGNAT4 <- sum(read.counts[,16])


colnames(genelength) <- c("Length","Gene")

#For the simulated samples
SGWT1 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/ac4C/SGWT1.txt", sep = ",")
SGWT1 <- na.omit(SGWT1)
colnames(SGWT1) <- c("Gene", "SGWT1")
SGWT2 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/ac4C/SGWT2.txt", sep = ",")
SGWT2 <- na.omit(SGWT2)
colnames(SGWT2) <- c("Gene", "SGWT2")
SGWT3 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/ac4C/SGWT3.txt", sep = ",")
SGWT3 <- na.omit(SGWT3)
colnames(SGWT3) <- c("Gene", "SGWT3")
SGWT4 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/ac4C/SGWT4.txt", sep = ",")
SGWT4 <- na.omit(SGWT4)
colnames(SGWT4) <- c("Gene", "SGWT4")

SGNAT1 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/ac4C/SGNAT1.txt", sep = ",")
SGNAT1 <- na.omit(SGNAT1)
colnames(SGNAT1) <- c("Gene", "SGNAT1")
SGNAT2 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/ac4C/SGNAT2.txt", sep = ",")
SGNAT2 <- na.omit(SGNAT2)
colnames(SGNAT2) <- c("Gene", "SGNAT2")
SGNAT3 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/ac4C/SGNAT3.txt", sep = ",")
SGNAT3 <- na.omit(SGNAT3)
colnames(SGNAT3) <- c("Gene", "SGNAT3")
SGNAT4 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/ac4C/SGNAT4.txt", sep = ",")
SGNAT4 <- na.omit(SGNAT4)
colnames(SGNAT4) <- c("Gene", "SGNAT4")


#merging simulated samples up
simsamples <- merge(SGWT1, SGWT2, by="Gene")
simsamples <- merge(simsamples, SGWT3, by="Gene")
simsamples <- merge(simsamples, SGWT4, by="Gene")
simsamples <- merge(simsamples, SGNAT1, by="Gene")
simsamples <- merge(simsamples, SGNAT2, by="Gene")
simsamples <- merge(simsamples, SGNAT3, by="Gene")
simsamples <- merge(simsamples, SGNAT4, by="Gene")
simsamples <- merge(simsamples, genelength, by="Gene")

#De-Tpming old samples
simsamples$Counts_WT1 <- simsamples$SGWT1*simsamples$Length
simsamples$Counts_WT1 <- simsamples$Counts_WT1/sum(simsamples$Counts_WT1)*goalsum_SGWT1
simsamples$Counts_WT2 <- simsamples$SGWT2*simsamples$Length
simsamples$Counts_WT2 <- simsamples$Counts_WT2/sum(simsamples$Counts_WT2)*goalsum_SGWT2
simsamples$Counts_WT3 <- simsamples$SGWT3*simsamples$Length
simsamples$Counts_WT3 <- simsamples$Counts_WT3/sum(simsamples$Counts_WT3)*goalsum_SGWT3
simsamples$Counts_WT4 <- simsamples$SGWT4*simsamples$Length
simsamples$Counts_WT4 <- simsamples$Counts_WT4/sum(simsamples$Counts_WT4)*goalsum_SGWT4
simsamples$Counts_NAT1 <- simsamples$SGNAT1*simsamples$Length
simsamples$Counts_NAT1 <- simsamples$Counts_NAT1/sum(simsamples$Counts_NAT1)*goalsum_SGNAT1
simsamples$Counts_NAT2 <- simsamples$SGNAT2*simsamples$Length
simsamples$Counts_NAT2 <- simsamples$Counts_NAT2/sum(simsamples$Counts_NAT2)*goalsum_SGNAT2
simsamples$Counts_NAT3 <- simsamples$SGNAT3*simsamples$Length
simsamples$Counts_NAT3 <- simsamples$Counts_NAT3/sum(simsamples$Counts_NAT3)*goalsum_SGNAT3
simsamples$Counts_NAT4 <- simsamples$SGNAT4*simsamples$Length
simsamples$Counts_NAT4 <- simsamples$Counts_NAT4/sum(simsamples$Counts_NAT4)*goalsum_SGNAT4


simsamples <- simsamples[,c(1,11:18)] 


OGsamples <- read.counts

SG_City <- merge(OGsamples, simsamples, by="Gene")
SG_City <- SG_City %>% distinct(Gene, .keep_all = TRUE)
row.names(SG_City) <- SG_City$Gene
SG_City$Gene <- NULL
```


## Filtration

### ***Before Filtration***

Let us begin with the filtration. 

```{r, warning=FALSE}
alternew <- log10(rowSums(SG_City))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3.25, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")
```


### ***After Filtration***

```{r}
filterednew <- subset(SG_City, alternew>3.25)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))

#Now we need to select only the rows which have those genes.
finaltable <- SG_City[row.names(SG_City)%in%impgenes,]
```

## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)

#Then we calculate the normalization factors.
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)

forall <- c(rep("BUWT",4), rep("BUNT",4), rep("SGWT",4), rep("SGNT", 4),rep("SimSGWT",4), rep("SimSGNT", 4))

#Now we do our matrix itself
mm <- model.matrix(~0+forall)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```


## Normalisation

For normalisation I convert the read counts into log2CPMs. Its a good way to visualise how the reads are. 

```{r}

dfforboxy <- finaltable
custom_palette <- qualitative_hcl(6, palette = "Pastel")
custom_palette <- sample(custom_palette,6)


orderforplots <- forall
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = (custom_palette), labels=unique(forall), breaks =unique(forall)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = unique(custom_palette), labels=unique(forall), breaks =unique(forall)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```


## PCA

```{r}
#this is done to be able to do traditional PCA
dfforpca <- y$E
dfforpca <- rbind(dfforpca, forall)
row.names(dfforpca)[nrow(dfforpca)] <- "forall"
dfforpca <- t(dfforpca)
numcol <- c(1:(ncol(dfforpca)-1))
dfforpca <- as.data.frame(dfforpca) #for some reason I can only change colum type if its a dataframe

dfforpca[ ,numcol] <- apply(dfforpca[ , numcol], 2,            # to turn our columns into numeric
                            function(x) as.numeric(as.character(x)))

#sapply(dfforpca, class) #to check class

# Or specify the factor levels in the order you want
dfforpca$forall <- factor(dfforpca$forall, levels = unique(forall))

pca_res <- prcomp(dfforpca[,numcol], scale=FALSE) #prepare the pca

#to make everything pretty
p <- autoplot(pca_res, data = dfforpca, x=1, y=2, colour = "forall", size=8, alpha=0.7)
p <- p + theme_bw() + scale_colour_manual((name = "Samples"), values = custom_palette, labels=unique(forall))
p <- p + ggtitle("PCA Plot") + theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())
p


#in case you want to see them yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])
```


## Linear Models

Now we must perform linear model fitting.

```{r}
forsample <- c(rep("BU",8), rep("SG",8),rep("SimSG",8))

forcelltype <- c(rep("WT",4), rep("NT",4), rep("WT",4), rep("NT", 4),rep("WT",4), rep("NT", 4))

mm <- model.matrix(~0+forsample+forcelltype)

fitted <- lmFit(y, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr <- makeContrasts(forsampleSG - forsampleSimSG, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)

top.table <- topTable(tmp, sort.by = "logFC", n = Inf)

DEGenes <- top.table

ndegs <- 20
```

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}

write.table(x = DEGenes,
            file = "~/Projects/StressGranules/SGPAPER/RealSGs/Realac4c.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)


upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>5 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -5 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Real SG")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#769fca",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#a2d1e6",max.overlaps = 20)+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#769fca")+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#a2d1e6")

#simstress <- data.frame(Gene=row.names(DEGenes), ss_t=DEGenes$t, ss_logFC=DEGenes$logFC)
#oldmethod <- data.frame(Gene=row.names(DEGenes), old_t=DEGenes$t, old_logFC=DEGenes$logFC)
#newmethod <- data.frame(Gene=row.names(DEGenes), new_t=DEGenes$t, new_logFC=DEGenes$logFC)
```


## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}
DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Real SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)

```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score (NES)",
       title= "Hallmark Pathways",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")
```

# Session Info

```{r}
sessionInfo()
```