---
title: "Analyses of GSE242766"
subtitle: "M(6)A reduction relieves FUS-associated ALS granules."
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

In this document, I present all analyses pertaining to the dataset **GSE242766**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE242766)),  
originally published by **G. Di Timoteo, 2024**.


```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)

set.seed(1906)
```



## First Steps

```{r}
#This section here is to load up the genelenght file I will be using throughout the analyses.
genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)

genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")

genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()

genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")

#Get the table of read counts by indicating the path to the file. This file is obtained from the GEO dataset site.
raw.read.counts <- fread("~/Projects/StressGranules/SG_Transcriptome/m6A/GSE242766_counts_table.tsv")

biomart <- fread("~/biomart.txt")[,-2]


colnames(raw.read.counts)[1] <- "ensembl_gene_id"
colnames(biomart) <- c("ensembl_gene_id", "hgnc_symbol")

eltable <- merge(raw.read.counts,biomart,by = "ensembl_gene_id", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="") #this removes reads that dont have a corresponding hgnc symbol.

eltable$ensembl_gene_id <- NULL

read.counts <- eltable %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

read.counts <- as.data.frame(read.counts)

row.names(read.counts) <- read.counts$hgnc_symbol #changes rownames into hgnc symbols
read.counts$hgnc_symbol <- NULL #We no longer need this column, as we have the rownames.

```

## Section for SPINSIM Preparation

This serves as an example.

```{r}
mof <- fread("~/Projects/StressGranules/Simulator/MolecularWeights.csv")
colnames(mof)[2] <- "GeneID" 
mof <- mof[!duplicated(mof$GeneID), ]


stress.tpm <- read.counts[,16]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/m6A/WTEX2.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)
```

## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, warning=FALSE, message=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(8, palette = "Pastel")
custom_palette <- sample(custom_palette, 8)

# Use SampleColouring from metadata
SampleColouring  <- c("ALWCCT","WTWCCT","ALWCEX","WTSGEX","WTSGCT","ALSGEX","ALSGCT","WTWCEX","WTWCCT","ALWCEX","ALWCCT","WTSGEX","WTSGCT","ALSGEX","ALSGCT","WTWCEX")

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(4, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )
```
As we can see in these plots, in this case there are no individual sample outliers in terms of complexity or read depth. Excellent!


# Differential Expression Analysis

## Filtration

### ***Before Filtration***

Let us begin with the filtration. Personally, I find that its easier to do visually. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bi-modal distributions, with the first mode corresponding to lowly expressed genes. I remove these by cutting just before the second mode.

```{r, warning=FALSE}
alternew.ALSWC_ND <- log10(rowSums(read.counts[,c(1,11)]))
alternew.ALSWC_ND <- as.data.frame(alternew.ALSWC_ND)

alternew.ALSSG_ND <- log10(rowSums(read.counts[,c(2,9)]))
alternew.ALSSG_ND <- as.data.frame(alternew.ALSSG_ND)

alternew.WTWC_ND <- log10(rowSums(read.counts[,c(3,10)]))
alternew.WTWC_ND <- as.data.frame(alternew.WTWC_ND)

alternew.WTSG_ND <- log10(rowSums(read.counts[,c(4,12)]))
alternew.WTSG_ND <- as.data.frame(alternew.WTSG_ND)

alternew.ALSWC_D <- log10(rowSums(read.counts[,c(5,13)]))
alternew.ALSWC_D <- as.data.frame(alternew.ALSWC_D)

alternew.ALSSG_D <- log10(rowSums(read.counts[,c(6,14)]))
alternew.ALSSG_D <- as.data.frame(alternew.ALSSG_D)

alternew.WTWC_D <- log10(rowSums(read.counts[,c(7,15)]))
alternew.WTWC_D <- as.data.frame(alternew.WTWC_D)

alternew.WTSG_D <- log10(rowSums(read.counts[,c(8,16)]))
alternew.WTSG_D <- as.data.frame(alternew.WTSG_D)

alternew <- log10(rowSums(read.counts))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.ALSWC_ND, aes(x=alternew.ALSWC_ND)) + geom_density(fill="#FFB6C1", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - ALS WC No Doxo")

ggplot(alternew.ALSSG_ND, aes(x=alternew.ALSSG_ND)) + geom_density(fill="#FFDAB9", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - ALS SG No Doxo")


ggplot(alternew.WTWC_ND, aes(x=alternew.WTWC_ND)) + geom_density(fill="#E6E6FA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - WT WC No Doxo")

ggplot(alternew.WTSG_ND, aes(x=alternew.WTSG_ND)) + geom_density(fill="#98FB98", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - WT SG No Doxo")

ggplot(alternew.ALSWC_D, aes(x=alternew.ALSWC_D)) + geom_density(fill="#87CEFA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - ALS WC Doxo")

ggplot(alternew.ALSSG_D, aes(x=alternew.ALSSG_D)) + geom_density(fill="#EEE8AA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - ALS SG Doxo")

ggplot(alternew.WTWC_D, aes(x=alternew.WTWC_D)) + geom_density(fill="#F08080", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.25, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - WT WC Doxo")

ggplot(alternew.WTSG_D, aes(x=alternew.WTSG_D)) + geom_density(fill="#B0E0E6", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - WT SG Doxo")

```



By seeing the shape of the graph, we can see where we should cut. We have a **binomial distribution** (low reads and noise vs actual biological meaningful reads). The ideal place to cut them is right before the wedge.

### ***After Filtration***

I didn't do it this time, but I've maintained the code to filter by sample type. This time, I considered that it was more correct to filter all samples together.

```{r}
filterednew <- subset(read.counts, alternew>3.5)
filterednew.ALSWC_ND <- subset(read.counts, alternew.ALSWC_ND>2.5)
filterednew.ALSSG_ND <- subset(read.counts, alternew.ALSSG_ND>2.5)
filterednew.WTWC_ND <- subset(read.counts, alternew.WTWC_ND>2.5)
filterednew.WTSG_ND <- subset(read.counts, alternew.WTSG_ND>2.5)
filterednew.ALSWC_D <- subset(read.counts, alternew.ALSWC_D>2.5)
filterednew.ALSSG_D <- subset(read.counts, alternew.ALSSG_D>2.5)
filterednew.WTWC_D <- subset(read.counts, alternew.WTWC_D>2.5)
filterednew.WTSG_D <- subset(read.counts, alternew.WTSG_D>2.5)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))
impgenes.ALSWC_ND <- c(row.names(filterednew.ALSWC_ND))
impgenes.ALSSG_ND <- c(row.names(filterednew.ALSSG_ND))
impgenes.WTWC_ND <- c(row.names(filterednew.WTWC_ND))
impgenes.WTSG_ND <- c(row.names(filterednew.WTSG_ND))
impgenes.ALSWC_D <- c(row.names(filterednew.ALSWC_D))
impgenes.ALSSG_D <- c(row.names(filterednew.ALSSG_D))
impgenes.WTWC_D <- c(row.names(filterednew.WTWC_D))
impgenes.WTSG_D <- c(row.names(filterednew.WTSG_D))

#Now we need to select only the rows which have those genes.
finaltable <- read.counts[row.names(read.counts)%in%impgenes,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.ALSWC_ND,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.ALSSG_ND,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WTWC_ND,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WTSG_ND,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.ALSWC_D,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.ALSSG_D,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WTWC_D,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WTSG_D,]
```


## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#Let us prepare metadata to add to our samples
#I start by normalizing them all together
#This one has all samples
D0 <- DGEList(finaltable)


#Then we calculate the normalization factors for these new generated data frames
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)


#Lets create a dataframe with this information
#For potential Batch Effects
CellType <- c("ALS","WT","ALS","WT","WT","ALS","ALS","WT","WT","ALS","ALS","WT","WT","ALS","ALS","WT")
MethilType <- c("CT","CT","KO","KO","CT","KO","CT","KO","CT","KO","CT","KO","CT","KO","CT","KO")
SampleType <- c(rep("WC",3),rep("SG",4),rep("WC",4),rep("SG",4),"WC")



metadata <- data.frame(Cell=CellType, M6A=MethilType, Sample=SampleType)


#Now we do the matrixes
#the one with everything
mm <- model.matrix(~0+SampleColouring)
#the one without pbs


#and now we normalize. This step also turns our counts into log2-counts per million.
#The object with all simple
y <- voom(D0, mm, plot = T)
```


## Normalisation

For normalisation I convert the read counts into log2CPMs. Its a good way to visualise how the reads are. 

```{r}
dfforboxy <- finaltable
orderforplots <- SampleColouring
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
# Convert 'Types' to factor with custom order
dfforboxy$Types <- factor(dfforboxy$Types, levels = unique(SampleColouring))

prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c(custom_palette), labels=c(unique(orderforplots))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

#Using the Standard Matrix

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
# Convert 'Types' to factor with custom order
dfforboxy$Types <- factor(dfforboxy$Types, levels = unique(SampleColouring))
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c(custom_palette), labels=c(unique(orderforplots))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```

## PCA

Before we move on, lets do a **Principal Component analysis**. It is a good exploratory tool to find out how our samples relate to each other and if there are any obvious outliers. We would expect our replicates to be nearer to one another than to other samples, and more different samples should be further away. 

```{r, fig.height=7, fig.width=9}
# Define color palette
custom_palette <- qualitative_hcl(4, palette = "Pastel")
custom_palette <- sample(custom_palette, 4)

#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- SampleType
dfforpca$M6A <- MethilType
dfforpca$Cell <- CellType

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-3)], scale = FALSE)

p_Sample <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Sample") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_M6A <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "M6A", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "M6A", values = custom_palette) +
  ggtitle("PCA - Coloured by M6A") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Cell <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Cell", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Cell", values = custom_palette) +
  ggtitle("PCA - Coloured by Cell") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Sample
p_M6A
p_Cell

#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])


```

As we can observe on the plot, PC1 is driven by sample type, and PC2 is driven by m6A status.

## Linear Models

```{r}

SampleType <- factor(SampleType)
SampleType <- relevel(SampleType,"WC")

CellType <- factor(CellType)
CellType <- relevel(CellType,"WT")

MethilType <- factor(MethilType)
MethilType <- relevel(MethilType,"KO")


mm <- model.matrix(~0+Sample+Cell+M6A, metadata)
fitted <- lmFit(y, mm)
head(coef(fitted))



contrastas <- makeContrasts(SampleSG   - SampleWC    , levels = coef(fitted))
# 
tmp <- contrasts.fit(fitted, contrastas)
# 
# #Correction for multiple testing
tmp <- eBayes(tmp)
# 
top.table <- topTable(tmp, sort.by = "logFC", n=Inf)

DEGenes <- top.table



ndegs=20
```

# SG vs Non-SG

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}

write.table(x = DEGenes,
            file = "~/Projects/StressGranules/SGPAPER/SGvsWC/SGm6A.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2.5 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -4 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```

## Volcano Plot with Transcript Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "pearson")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)

```


# With m6A SG vs WC

Now, I compare samples separately. First I will compare samples with m6A. Since the authors link m6A to SG enrichment and ALS, I focused on the ALS samples. 

(Additionally, it appears the effect of m6A is very very light, but it appears to be slightly less so on ALS samples)

```{r}

mm <- model.matrix(~0+SampleColouring)
fitted <- lmFit(y, mm)
head(coef(fitted))

contrastas <- makeContrasts( ((SampleColouringALSGCT))    - ((SampleColouringALWCCT))     , levels = coef(fitted))
# 
tmp <- contrasts.fit(fitted, contrastas)
# 
# #Correction for multiple testing
tmp <- eBayes(tmp)
# 
top.table <- topTable(tmp, sort.by = "logFC", n=Inf)

DEGenes <- top.table



ndegs=20
```


## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2 & DEGenes$B>5,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -3.5 & DEGenes$B>5,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC (WT)")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#A10702",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#03256C",max.overlaps = 20)+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702")+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C")

wt_t <- data.frame(Gene=row.names(DEGenes), wt_t=DEGenes$t, wt_logFC=DEGenes$logFC)
```

## Volcano Plot with Transcript Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs No-SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG-Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) + ggtitle("With m6A")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)



```

There is a very strong lenght bias. m6A WT Stress granules are enriched in longer rnas.

# Without m6A SG vs WC


Now I will compare samples without m6A. Since the authors link m6A to SG enrichment and ALS, I focused on the ALS samples. 

(Additionally, it appears the effect of m6A is very very light, but it appears to be slightly less so on ALS samples)

```{r}

mm <- model.matrix(~0+SampleColouring)
fitted <- lmFit(y, mm)
head(coef(fitted))

contrastas <- makeContrasts( ((SampleColouringALSGEX))    - ((SampleColouringALWCEX))     , levels = coef(fitted))
# 
tmp <- contrasts.fit(fitted, contrastas)
# 
# #Correction for multiple testing
tmp <- eBayes(tmp)
# 
top.table <- topTable(tmp, sort.by = "logFC", n=Inf)

DEGenes <- top.table



ndegs=20
```


## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2 & DEGenes$B>5,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -3.5 & DEGenes$B>5,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC (ko)")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#A10702",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#03256C",max.overlaps = 20)+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702")+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C")

nt_t <- data.frame(Gene=row.names(DEGenes), nt_t=DEGenes$t, nt_logFC=DEGenes$logFC)
```

## Volcano Plot with Transcript Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs No-SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG-Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) + ggtitle("Without m6A")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)




```

There is a very strong lenght bias. m6A KO Stress granules are enriched in longer rnas.

# Comparisons

```{r}
m6A <- fread("~/Projects/StressGranules/Simulator/MolecularWeights_m6A.csv")[,c(1,3,4)]


wtvsnt <- merge(wt_t, nt_t, by="Gene")
colnames(m6A) <- c("Gene", "NM6A", "Ratio")
wtvsnt <- merge(wtvsnt, m6A, all.y=F, all.x=T)
wtvsnt[is.na(wtvsnt)] <- 0

smooth.SG <- lm(wt_t~(nt_t), wtvsnt)
liner.SG <- rstandard(smooth.SG)
wtvsnt <-cbind(wtvsnt, liner.SG)
colnames(wtvsnt)[8] <- "Residuals"
ogres <- wtvsnt[,c(1,8)]
#row.names(wtvsnt) <- wtvsnt$Gene
linemarkers <- as.data.frame(wtvsnt[abs(wtvsnt$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(wtvsnt$wt_t, (wtvsnt$nt_t), method = "spearman")



ggplot(wtvsnt, aes(y=wt_t, x=(nt_t))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)  +   theme_bw()+xlab("M6AKO SG Enrichment")+
  ylab("WT SG Enrichment")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(text=element_text(size=15,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=5)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")


ggplot(wtvsnt, aes(y=NM6A, x=(liner.SG))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ 
 ylab("Number of m6A") + xlab("Residuals") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  + stat_cor(method = "spearman", cor.coef.name = "rho", size=5)

wtvsnt$bins <- ifelse(wtvsnt$NM6A <= 4, wtvsnt$NM6A,
                      ifelse(wtvsnt$NM6A>=5 & wtvsnt$NM6A<=9, "[5,9]","[10,["))


wtvsnt$bins <- factor(wtvsnt$bins, levels=c(0:5, "[5,9]","[10,[" ))
ggplot(wtvsnt, aes(x=bins, y=(Residuals)))  + geom_jitter(alpha=0.2) + geom_boxplot(fill="#43F3D5", alpha=0.4) +theme_bw()+ 
 xlab("Number of m6A") + ylab("Residuals") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) 

```
Enrichment is very similar between WT and m6A KO stress granules. There does seem to be a very slight enrichment for transcripts with many m6A modifications.

# Spin-Sim

```{r}


goalsum_WT1 <- sum(read.counts[,7])
goalsum_KO1 <- sum(read.counts[,6])
goalsum_WT2 <- sum(read.counts[,15])
goalsum_KO2 <- sum(read.counts[,14])

colnames(genelength) <- c("Length", "Gene")


#For the new samples. Comp samples are "complex" and weight was calculate using the ratio of m6A. Simp samples simply assume that all transcripts of the same type have the same number of m6A modifications
simwt1s <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/m6A/ALSWTSG1_Simp.txt", sep = ",")[,c(1,3)]
colnames(simwt1s) <- c("Gene", "SGWT1S")
simwt2s <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/m6A/ALSWTSG2_Simp.txt", sep = ",")[,c(1,3)]
colnames(simwt2s) <- c("Gene", "SGWT2S")
simwt1c <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/m6A/ALSWTSG1_Comp.txt", sep = ",")[,c(1,3)]
colnames(simwt1c) <- c("Gene", "SGWT1C")
simwt2c <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/m6A/ALSWTSG2_Comp.txt", sep = ",")[,c(1,3)]
colnames(simwt2c) <- c("Gene", "SGWT2C")

simko1 <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/m6A/ALSKOSG1_Simp.txt", sep = ",")[,c(1,3)]
colnames(simko1) <- c("Gene", "SGKO1")
simko2 <- read.table("~/Projects/StressGranules/Simulator/OriginalSamples/m6A/ALSKOSG2_Simp.txt", sep = ",")[,c(1,3)]
colnames(simko2) <- c("Gene", "SGKO2")

simsamples <- merge(simwt1s, simwt2s, by="Gene")
#simsamples <- merge(simwt2, simwt3, by="Gene")
simsamples <- merge(simsamples, simwt1c, by="Gene")
simsamples <- merge(simsamples, simwt2c, by="Gene")
simsamples <- merge(simsamples, simko1, by="Gene")
simsamples <- merge(simsamples, simko2, by="Gene")
simsamples <- merge(simsamples, genelength, by="Gene")

simsamples <- simsamples %>%
  group_by(Gene) %>%
  slice(which.max(Length)) %>%
  ungroup()

simsamples <- as.data.frame(simsamples)

colnames(simsamples)[8] <- "Length"

simsamples$WTCounts_1 <- simsamples$SGWT1S*simsamples$Length
simsamples$WTCounts_1 <- simsamples$WTCounts_1/sum(simsamples$WTCounts_1)*goalsum_WT1
simsamples$WTCounts_2 <- simsamples$SGWT2S*simsamples$Length
simsamples$WTCounts_2 <- simsamples$WTCounts_2/sum(simsamples$WTCounts_2)*goalsum_WT2
simsamples$WTCounts_3 <- simsamples$SGWT1C*simsamples$Length
simsamples$WTCounts_3 <- simsamples$WTCounts_3/sum(simsamples$WTCounts_3)*goalsum_WT1
simsamples$WTCounts_4 <- simsamples$SGWT2C*simsamples$Length
simsamples$WTCounts_4 <- simsamples$WTCounts_4/sum(simsamples$WTCounts_4)*goalsum_WT2

simsamples$NTCounts_1 <- simsamples$SGKO1*simsamples$Length
simsamples$NTCounts_1 <- simsamples$NTCounts_1/sum(simsamples$NTCounts_1)*goalsum_KO1
simsamples$NTCounts_2 <- simsamples$SGKO2*simsamples$Length
simsamples$NTCounts_2 <- simsamples$NTCounts_2/sum(simsamples$NTCounts_2)*goalsum_KO2


simsamples <- simsamples[,c(1,9:14)] 

OGsamples <- read.counts[,c(1,11,3,10)]
OGsamples$Gene <- row.names(OGsamples)

simandog <- merge(simsamples, OGsamples, by="Gene")
row.names(simandog) <- simandog$Gene
simandog$Gene <- NULL
```

# Filtration

```{r}
alternew <- log10(rowSums(simandog))
alternew <- as.data.frame(alternew)
filterednew <- subset(simandog, alternew>3.5)
impgenes <- c(row.names(falternew))
finaltable <- simandog[row.names(simandog)%in%impgenes,]
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

```

# Matrices

```{r}
sampy <- c(rep("SGWTS",2), rep("SGWTC",2), rep("KOSG",2), rep("WTWC",2),rep("KOWC",2))

#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)

#Then we calculate the normalization factors.
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)


sampy <- sampy

#Now we do our matrix itself
mm <- model.matrix(~0+sampy)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```



## PCA

```{r}
#this is done to be able to do traditional PCA

sampy <- sampy
dfforpca <- y$E
dfforpca <- rbind(dfforpca, sampy)
dfforpca <- t(dfforpca)
numcol <- c(1:(ncol(dfforpca)-1))
dfforpca <- as.data.frame(dfforpca) #for some reason I can only change colum type if its a dataframe

dfforpca[ ,numcol] <- apply(dfforpca[ , numcol], 2,            # to turn our columns into numeric
                            function(x) as.numeric(as.character(x)))

#sapply(dfforpca, class) #to check class

# Or specify the factor levels in the order you want
dfforpca$sampy <- factor(dfforpca$sampy, levels = c("SGWTS", "SGWTC", "KOSG", "WTWC", "KOWC"))

pca_res <- prcomp(dfforpca[,numcol], scale=FALSE) #prepare the pca

#to make everything pretty
p <- autoplot(pca_res, data = dfforpca, x=1, y=2, colour = "sampy", size=8, alpha=0.7)
p <- p + theme_bw() + scale_colour_manual((name = "Samples"), values = c("black", "#769fca", "green", "purple", "orange"), labels=c("SGWTS", "KOSG", "WTWC", "KOWC", "SGWTC"))
p <- p + ggtitle("PCA Plot") + theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())
p


#in case you want to see them yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])
```


## Linear Models

Now we must perform linear model fitting. 

```{r Linear Models}
fitted <- lmFit(y, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr <- makeContrasts(sampySGWTS      - sampyWTWC     , levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)

top.table <- topTable(tmp, sort.by = "logFC", n = Inf)

DEGenes <- top.table

ndegs <- 20
```

# WT Samples

## Volcano Plots

```{r , results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$B>0,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -3 & DEGenes$B>5,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Sim SG vs WC (m6A WT)")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#769fca",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#a2d1e6",max.overlaps = 20)+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#769fca")+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#a2d1e6")

simwt <- data.frame(Gene=row.names(DEGenes), simwt_t=DEGenes$t, simwt_logFC=DEGenes$logFC)
#simnt <- data.frame(Gene=row.names(DEGenes), simnt_t=DEGenes$t, simnt_logFC=DEGenes$logFC)
```

# KO Samples

```{r}
fitted <- lmFit(y, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr <- makeContrasts(sampyKOSG       - sampyKOWC      , levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)

top.table <- topTable(tmp, sort.by = "logFC", n = Inf)

DEGenes <- top.table

ndegs <- 20
```


## Volcano Plots

```{r , results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$B>0,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -3 & DEGenes$B>5,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Sim SG vs WC (m6A KO)")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#769fca",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#a2d1e6",max.overlaps = 20)+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#769fca")+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#a2d1e6")

simnt <- data.frame(Gene=row.names(DEGenes), simnt_t=DEGenes$t, simnt_logFC=DEGenes$logFC)
```

# Comparisons

```{r}
wtvsnt <- merge(simwt, simnt, by="Gene")


smooth.SG <- lm(simwt_t~(simnt_t), wtvsnt)
liner.SG <- rstandard(smooth.SG)
wtvsnt <-cbind(wtvsnt, liner.SG)
colnames(wtvsnt)[6] <- "Residuals"
row.names(wtvsnt) <- wtvsnt$Gene
linemarkers <- as.data.frame(wtvsnt[abs(wtvsnt$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(wtvsnt$simwt_t, (wtvsnt$simnt_t), method = "spearman")



ggplot(wtvsnt, aes(y=simwt_t, x=(simnt_t))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("WT Enrichment") + xlab("KO Enrichemnt") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)


colnames(m6A) <- c("Gene", "NM6A", "Ratio")
wtvsnt <- merge(wtvsnt, m6A, all.y=F, all.x=T)
wtvsnt[is.na(wtvsnt)] <- 0

wtvsnt$bins <- ifelse(wtvsnt$NM6A <= 4, wtvsnt$NM6A,
                      ifelse(wtvsnt$NM6A>=5 & wtvsnt$NM6A<=9, "[5,9]","[10,["))


wtvsnt$bins <- factor(wtvsnt$bins, levels=c(0:5, "[6,8]","[9,[" ))
ggplot(wtvsnt, aes(x=bins, y=(Residuals)))  + geom_jitter(alpha=0.2) + geom_boxplot(fill="#43F3D5", alpha=0.4) +theme_bw()+ 
 xlab("Number of m6a") + ylab("Residuals") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(wtvsnt, aes(y=NM6A, x=(Residuals))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ 
 xlab("Number of m6a") + ylab("Residuals") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  + stat_cor(method = "spearman", cor.coef.name = "rho", size=5)

simres <- wtvsnt[,c(1,6)]



resvres <- merge(ogres, simres, by="Gene")


smooth.SG <- lm(Residuals.y~(Residuals.x), resvres)
liner.SG <- rstandard(smooth.SG)
resvres <-cbind(resvres, liner.SG)
colnames(resvres)[4] <- "Residuals"
#row.names(resvres) <- resvres$Gene
linemarkers <- as.data.frame(resvres[abs(resvres$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(resvres$Residuals.y, (resvres$Residuals.x), method = "spearman")


resvres <- merge(resvres, m6A, by="Gene", all.y=F, all.x=T)
resvres[is.na(resvres)] <- 0

resvres$bins <- ifelse(resvres$NM6A <= 8, resvres$NM6A,
                      ifelse(resvres$NM6A>=9 & resvres$NM6A<=12, "[6,8]","[9,["))


resvres$bins <- factor(resvres$bins, levels=c(0:12, "[6,8]","[9,[" ))

#resvres$bins <- resvres$NM6A
#resvres$bins <- factor(resvres$bins, levels=c(0:12))

upmarkers <- as.data.frame(resvres[resvres$NM6A>=9,] )

trial <- subset(resvres, NM6A>=5)

resvres$proxy <- resvres$NM6A*resvres$Ratio


ggplot(resvres, aes(x=(Residuals.x),y=Residuals.y, colour=as.factor(bins))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21)  +   theme_bw()+xlab("Original Residuals")+
  ylab("Simulated Residuals")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) + stat_cor(method = "spearman", cor.coef.name = "rho", size=5)+
  theme(text=element_text(size=15,  family="Arial MS")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_point(data=upmarkers, aes(x=Residuals.x, y=  Residuals.y), color="#43F3D5",fill="#43F3D5",alpha=1, size=5, shape=21) 

```
Honestly, the effect of m6A is so light, that its hard to make any observations. In my personal opinion, its because it a very light modification (in terms of weight), and so we have trouble seeing differences in DC isolated samples.

# "Real" SG

After the spin-sim pipeline, we can then compared the original SGs to our simulated ones, which allows us to determine genes which are more or less enriched than what was expected by centrifugation alone.


```{r}
read.counts$Gene <- row.names(read.counts)

goalsum_WTSGEX1 <- sum(read.counts[,4])
goalsum_WTSGEX2 <- sum(read.counts[,12])
goalsum_WTSGCT1 <- sum(read.counts[,5])
goalsum_WTSGCT2 <- sum(read.counts[,13])
goalsum_ALSGEX1 <- sum(read.counts[,6])
goalsum_ALSGEX2 <- sum(read.counts[,14])
goalsum_ALSGCT1 <- sum(read.counts[,7])
goalsum_ALSGCT2 <- sum(read.counts[,15])

colnames(genelength) <- c("Length","Gene")

#For the simulated samples
SGALCT1 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/m6A/ALCT1.txt", sep = ",")
SGALCT1 <- na.omit(SGALCT1)
colnames(SGALCT1) <- c("Gene", "ALCT1")
SGALCT2 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/m6A/ALCT2.txt", sep = ",")
SGALCT2 <- na.omit(SGALCT2)
colnames(SGALCT2) <- c("Gene", "ALCT2")
SGALEX1 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/m6A/ALEX1.txt", sep = ",")
SGALEX1 <- na.omit(SGALEX1)
colnames(SGALEX1) <- c("Gene", "ALEX1")
SGALEX2 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/m6A/ALEX2.txt", sep = ",")
SGALEX2 <- na.omit(SGALEX2)
colnames(SGALEX2) <- c("Gene", "ALEX2")
SGWTEX1 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/m6A/WTEX1.txt", sep = ",")
SGWTEX1 <- na.omit(SGWTEX1)
colnames(SGWTEX1) <- c("Gene", "WTEX1")
SGWTEX2 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/m6A/WTEX2.txt", sep = ",")
SGWTEX2 <- na.omit(SGWTEX2)
colnames(SGWTEX2) <- c("Gene", "WTEX2")

SGWTWC1 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/m6A/WTWC1.txt", sep = ",")
SGWTWC1 <- na.omit(SGWTWC1)
colnames(SGWTWC1) <- c("Gene", "WTWC1")
SGWTWC2 <- read.table("~/Projects/StressGranules/Simulator/SimulatedSamples/m6A/WTWC2.txt", sep = ",")
SGWTWC2 <- na.omit(SGWTWC2)
colnames(SGWTWC2) <- c("Gene", "WTWC2")


#merging simulated samples up
simsamples <- merge(SGALCT1, SGALCT2, by="Gene")
simsamples <- merge(simsamples, SGALEX1, by="Gene")
simsamples <- merge(simsamples, SGALEX2, by="Gene")
simsamples <- merge(simsamples, SGWTEX1, by="Gene")
simsamples <- merge(simsamples, SGWTEX2, by="Gene")
simsamples <- merge(simsamples, SGWTWC1, by="Gene")
simsamples <- merge(simsamples, SGWTWC2, by="Gene")
simsamples <- merge(simsamples, genelength, by="Gene")

#De-Tpming old samples
simsamples$Counts_ALCT1 <- simsamples$ALCT1*simsamples$Length
simsamples$Counts_ALCT1 <- simsamples$Counts_ALCT1/sum(simsamples$Counts_ALCT1)*goalsum_ALSGCT1
simsamples$Counts_ALCT2 <- simsamples$ALCT2*simsamples$Length
simsamples$Counts_ALCT2 <- simsamples$Counts_ALCT2/sum(simsamples$Counts_ALCT2)*goalsum_ALSGCT2
simsamples$Counts_ALEX1 <- simsamples$ALEX1*simsamples$Length
simsamples$Counts_ALEX1 <- simsamples$Counts_ALEX1/sum(simsamples$Counts_ALEX1)*goalsum_ALSGEX1
simsamples$Counts_ALEX2 <- simsamples$ALEX2*simsamples$Length
simsamples$Counts_ALEX2 <- simsamples$Counts_ALEX2/sum(simsamples$Counts_ALEX2)*goalsum_ALSGEX2
simsamples$Counts_WTCT1 <- simsamples$WTWC1*simsamples$Length
simsamples$Counts_WTCT1 <- simsamples$Counts_WTCT1/sum(simsamples$Counts_WTCT1)*goalsum_WTSGCT1
simsamples$Counts_WTCT2 <- simsamples$WTWC2*simsamples$Length
simsamples$Counts_WTCT2 <- simsamples$Counts_WTCT2/sum(simsamples$Counts_WTCT2)*goalsum_WTSGCT2
simsamples$Counts_WTEX1 <- simsamples$WTEX1*simsamples$Length
simsamples$Counts_WTEX1 <- simsamples$Counts_WTEX1/sum(simsamples$Counts_WTEX1)*goalsum_WTSGEX1
simsamples$Counts_WTEX2 <- simsamples$WTEX2*simsamples$Length
simsamples$Counts_WTEX2 <- simsamples$Counts_WTEX2/sum(simsamples$Counts_WTEX2)*goalsum_WTSGEX2


simsamples <- simsamples[,c(1,11:18)] 


OGsamples <- read.counts

SG_City <- merge(OGsamples, simsamples, by="Gene")
SG_City <- SG_City %>% distinct(Gene, .keep_all = TRUE)
row.names(SG_City) <- SG_City$Gene
SG_City$Gene <- NULL
```


## Filtration

### ***Before Filtration***

Let us begin with the filtration. Before moving on, it is important to note that according to the authors, these RNA granules are not pure and contain several contaminants. 

```{r, warning=FALSE}
alternew <- log10(rowSums(SG_City))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")
```

### ***After Filtration***

```{r}
filterednew <- subset(SG_City, alternew>3.75)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))

#Now we need to select only the rows which have those genes.
finaltable <- SG_City[row.names(SG_City)%in%impgenes,]
```

## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)

#Then we calculate the normalization factors.
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)

forall <- c("ALWCCT","WTWCCT","ALWCEX","WTSGEX","WTSGCT","ALSGEX","ALSGCT","WTWCEX","WTWCCT","ALWCEX","ALWCCT","WTSGEX","WTSGCT","ALSGEX","ALSGCT","WTWCEX", rep("SimALSGCT",2), rep("SimALSGEX",2), rep("SimWTSGCT",2), rep("SimWTSGEX",2))

CellType <- c("ALS","WT","ALS","WT","WT","ALS","ALS","WT","WT","ALS","ALS","WT","WT","ALS","ALS","WT", rep("ALS",4), rep("WT",4))
StressType <- c("CT","CT","KO","KO","CT","KO","CT","KO","CT","KO","CT","KO","CT","KO","CT","KO", rep("CT",2), rep("EX",2),rep("CT",2), rep("EX",2))
SampleType <- c(rep("WC",3),rep("SG",4),rep("WC",4),rep("SG",4),"WC", rep("SimSG",8))



#Now we do our matrix itself
mm <- model.matrix(~0+SampleType+CellType+StressType)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```


## Normalisation

For normalisation I convert the read counts into log2CPMs. Its a good way to visualise how the reads are. 

```{r}

dfforboxy <- finaltable
custom_palette <- qualitative_hcl(12, palette = "Pastel")
custom_palette <- sample(custom_palette,12)


orderforplots <- forall
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = (custom_palette), labels=unique(forall), breaks =unique(forall)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalization")

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = unique(custom_palette), labels=unique(forall), breaks =unique(forall)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalization")

ggarrange(prenorm, postnorm, nrow=1)

```


## PCA

```{r}
#this is done to be able to do traditional PCA
dfforpca <- y$E
dfforpca <- rbind(dfforpca, forall)
row.names(dfforpca)[nrow(dfforpca)] <- "forall"
dfforpca <- t(dfforpca)
numcol <- c(1:(ncol(dfforpca)-1))
dfforpca <- as.data.frame(dfforpca) #for some reason I can only change colum type if its a dataframe

dfforpca[ ,numcol] <- apply(dfforpca[ , numcol], 2,            # to turn our columns into numeric
                            function(x) as.numeric(as.character(x)))

#sapply(dfforpca, class) #to check class

# Or specify the factor levels in the order you want
dfforpca$forall <- factor(dfforpca$forall, levels = unique(forall))

pca_res <- prcomp(dfforpca[,numcol], scale=FALSE) #prepare the pca

#to make everything pretty
p <- autoplot(pca_res, data = dfforpca, x=1, y=2, colour = "forall", size=8, alpha=0.7)
p <- p + theme_bw() + scale_colour_manual((name = "Samples"), values = custom_palette, labels=unique(forall))
p <- p + ggtitle("PCA Plot") + theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())
p


#in case you want to see them yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])
```

```{r}

fitted <- lmFit(y, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr <- makeContrasts(SampleTypeSG - SampleTypeSimSG, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)

top.table <- topTable(tmp, sort.by = "logFC", n = Inf)

DEGenes <- top.table

ndegs <- 20
```


## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}

write.table(x = DEGenes,
            file = "~/Projects/StressGranules/SGPAPER/RealSGs/Realm6A.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)


upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -3 & DEGenes$adj.P.Val<0.05,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs Sim SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))


```


## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs simSG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs simSG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

Corrected SGs show a minor lenght bias

## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score (NES)",
       title= "Hallmark Pathways",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")
```


# Session Info

```{r}
sessionInfo()
```