---
title: "Analyses of GSE223295"
subtitle: "Profiling stress-triggered RNA condensation with photocatalytic proximity labeling."
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

In this document, I present all analyses pertaining to the dataset **GSE223295**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE223295)),  
originally published by **Z. Ren, 2023**.

## Methodology


```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)
library(readxl)


set.seed(1906)
```


## First Steps

```{r }
#This section here is to load up the genelenght file I will be using throughout the analyses.
genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)

genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")

genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()

genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")



#Get the table of read counts by indicating the path to the file. These files were obtained from the GEO dataset itself.

sg.read.counts_U2OS <- read_excel("~/Projects/StressGranules/SG_Transcriptome/CAP_Seq/GSE223295_U2OS_allcounts.xlsx")[,-1]
sg.read.counts_HEK <- read_excel("~/Projects/StressGranules/SG_Transcriptome/CAP_Seq/GSE223295_HEK293T_allcounts.xlsx")[,-1]

raw.read.counts <- merge(sg.read.counts_U2OS, sg.read.counts_HEK, by="geneId")


biomart <- fread("~/biomart.txt")[,-2]


colnames(raw.read.counts)[1] <- "ensembl_gene_id"
colnames(biomart) <- c("ensembl_gene_id", "hgnc_symbol")

eltable <- merge(raw.read.counts,biomart,by = "ensembl_gene_id", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="") #this removes reads that dont have a corresponding hgnc symbol.

eltable$ensembl_gene_id <- NULL

read.counts <- eltable %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

read.counts <- as.data.frame(read.counts)

row.names(read.counts) <- read.counts$hgnc_symbol #changes rownames into hgnc symbols
read.counts$hgnc_symbol <- NULL #We no longer need this column, as we have the rownames.

Met_frame <- fread("~/Projects/StressGranules/SG_Transcriptome/CAP_Seq/SraRunTable-3.csv")
```


## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, warning=FALSE, message=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(24, palette = "Pastel")
custom_palette <- sample(custom_palette, 24)

# Use SampleColouring from metadata
SampleColouring <- c(rep("U2G3AsLabEn",2), rep("U2G3AsLabIn",2),rep("U2G3AsNoLabEn",2),rep("U2CtAsLabEn",2),rep("HekG3AsLabEn",2), rep("HekG3AsLabIn",2),rep("HekG3AsNoLabEn",2),rep("HekCtAsLabEn",4),rep("HekG3UnLabEn",2), rep("HekG3UnLabIn",2),rep("HekG3UnNoLabEn",2),rep("HekCtUnLabEn",4),rep("HekG3SoLabEn",2), rep("HekG3SoLabIn",2),rep("HekG3SoNoLabEn",2),rep("HekCtSoLabEn",4),rep("HekG3T1LabEn",3), rep("HekG3T1LabIn",3),rep("HekG3T1NoLabEn",3),rep("HekCtT1LabEn",4),rep("HekG3T3LabEn",2), rep("HekG3T3LabIn",2),rep("HekG3T3NoLabEn",2),rep("HekCtT3LabEn",4))

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(4, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )
```

```{r}
decode_samples <- function(codes) {
  bait   <- ifelse(grepl("G3", codes), "G3BP1", "Untargeted")
  stress <- dplyr::case_when(
    grepl("As", codes) ~ "Stressed",
    grepl("Un", codes) ~ "Unstressed",
    grepl("So", codes) ~ "Stressed",
    grepl("T1", codes) ~ "Stressed",
    grepl("T3", codes) ~ "Unstressed"
  )
  probe <- ifelse(grepl("NoLab", codes), "NoProbe", "Label")
  frac  <- ifelse(grepl("En$", codes), "Enriched", "Input")

  role <- dplyr::case_when(
    frac == "Input"                      ~ "Whole Cell",
    probe == "NoProbe"                   ~ "Background Control",
    bait == "Untargeted"                 ~ "Untargeted Control",
    bait == "G3BP1" ~ "Stress Granule"  
  )

  paste0(role, "_", stress)
}

sample_codes <- rep(
  c("U2G3AsLabEn","U2G3AsLabIn","U2G3AsNoLabEn","U2CtAsLabEn",
    "HekG3AsLabEn","HekG3AsLabIn","HekG3AsNoLabEn","HekCtAsLabEn",
    "HekG3UnLabEn","HekG3UnLabIn","HekG3UnNoLabEn","HekCtUnLabEn",
    "HekG3SoLabEn","HekG3SoLabIn","HekG3SoNoLabEn","HekCtSoLabEn",
    "HekG3T1LabEn","HekG3T1LabIn","HekG3T1NoLabEn","HekCtT1LabEn",
    "HekG3T3LabEn","HekG3T3LabIn","HekG3T3NoLabEn","HekCtT3LabEn"),
  times = c(2,2,2,2, 2,2,2,4, 2,2,2,4, 2,2,2,4, 3,3,3,4, 2,2,2,4)
)

SampleColouring_readable <- decode_samples(sample_codes)
```


As we can see in these plots, similar samples have similar complexity. As in, they achieve roughly the same amount of genes covered using 50% of their reads. There is one sample with a lower read depth though.

# Differential Expression Analysis

## Filtration

### ***Before Filtration***

Let us begin with the filtration.

```{r, warning=FALSE}

alternew.U2G3AsLabEn <- log10(rowSums(read.counts[,1:2]))
alternew.U2G3AsLabEn <- as.data.frame(alternew.U2G3AsLabEn)

alternew.U2G3AsLabIn <- log10(rowSums(read.counts[,3:4]))
alternew.U2G3AsLabIn <- as.data.frame(alternew.U2G3AsLabIn)

alternew.U2G3AsNoLabEn <- log10(rowSums(read.counts[,5:6]))
alternew.U2G3AsNoLabEn <- as.data.frame(alternew.U2G3AsNoLabEn)

alternew.U2CtAsLabEn <- log10(rowSums(read.counts[,7:8]))
alternew.U2CtAsLabEn <- as.data.frame(alternew.U2CtAsLabEn)

alternew.HekG3AsLabEn <- log10(rowSums(read.counts[,9:10]))
alternew.HekG3AsLabEn <- as.data.frame(alternew.HekG3AsLabEn)

alternew.HekG3AsLabIn <- log10(rowSums(read.counts[,11:12]))
alternew.HekG3AsLabIn <- as.data.frame(alternew.HekG3AsLabIn)

alternew.HekG3AsNoLabEn <- log10(rowSums(read.counts[,13:14]))
alternew.HekG3AsNoLabEn <- as.data.frame(alternew.HekG3AsNoLabEn)

alternew.HekCtAsLabEn <- log10(rowSums(read.counts[,15:18]))
alternew.HekCtAsLabEn <- as.data.frame(alternew.HekCtAsLabEn)

alternew.HekG3UnLabEn <- log10(rowSums(read.counts[,19:20]))
alternew.HekG3UnLabEn <- as.data.frame(alternew.HekG3UnLabEn)

alternew.HekG3UnLabIn <- log10(rowSums(read.counts[,21:22]))
alternew.HekG3UnLabIn <- as.data.frame(alternew.HekG3UnLabIn)

alternew.HekG3UnNoLabEn <- log10(rowSums(read.counts[,23:24]))
alternew.HekG3UnNoLabEn <- as.data.frame(alternew.HekG3UnNoLabEn)

alternew.HekCtUnLabEn <- log10(rowSums(read.counts[,25:28]))
alternew.HekCtUnLabEn <- as.data.frame(alternew.HekCtUnLabEn)

alternew.HekG3SoLabEn <- log10(rowSums(read.counts[,29:30]))
alternew.HekG3SoLabEn <- as.data.frame(alternew.HekG3SoLabEn)

alternew.HekG3SoLabIn <- log10(rowSums(read.counts[,31:32]))
alternew.HekG3SoLabIn <- as.data.frame(alternew.HekG3SoLabIn)

alternew.HekG3SoNoLabEn <- log10(rowSums(read.counts[,33:34]))
alternew.HekG3SoNoLabEn <- as.data.frame(alternew.HekG3SoNoLabEn)

alternew.HekCtSoLabEn <- log10(rowSums(read.counts[,35:38]))
alternew.HekCtSoLabEn <- as.data.frame(alternew.HekCtSoLabEn)

alternew.HekG3T1LabEn <- log10(rowSums(read.counts[,39:41]))
alternew.HekG3T1LabEn <- as.data.frame(alternew.HekG3T1LabEn)

alternew.HekG3T1LabIn <- log10(rowSums(read.counts[,42:44]))
alternew.HekG3T1LabIn <- as.data.frame(alternew.HekG3T1LabIn)

alternew.HekG3T1NoLabEn <- log10(rowSums(read.counts[,45:47]))
alternew.HekG3T1NoLabEn <- as.data.frame(alternew.HekG3T1NoLabEn)

alternew.HekCtT1LabEn <- log10(rowSums(read.counts[,48:51]))
alternew.HekCtT1LabEn <- as.data.frame(alternew.HekCtT1LabEn)

alternew.HekG3T3LabEn <- log10(rowSums(read.counts[,52:53]))
alternew.HekG3T3LabEn <- as.data.frame(alternew.HekG3T3LabEn)

alternew.HekG3T3LabIn <- log10(rowSums(read.counts[,54:55]))
alternew.HekG3T3LabIn <- as.data.frame(alternew.HekG3T3LabIn)

alternew.HekG3T3NoLabEn <- log10(rowSums(read.counts[,56:57]))
alternew.HekG3T3NoLabEn <- as.data.frame(alternew.HekG3T3NoLabEn)

alternew.HekCtT3LabEn <- log10(rowSums(read.counts[,58:61]))
alternew.HekCtT3LabEn <- as.data.frame(alternew.HekCtT3LabEn)

alternew <- log10(rowSums(read.counts))
alternew <- as.data.frame(alternew)


ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3.25, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

# U2OS Cells
ggplot(alternew.U2G3AsLabEn, aes(x=alternew.U2G3AsLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - U2OS G3BP1 Arsenite Label Enriched")

ggplot(alternew.U2G3AsLabIn, aes(x=alternew.U2G3AsLabIn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - U2OS G3BP1 Arsenite Label Input")

ggplot(alternew.U2G3AsNoLabEn, aes(x=alternew.U2G3AsNoLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - U2OS G3BP1 Arsenite NoLabel Enriched")

ggplot(alternew.U2CtAsLabEn, aes(x=alternew.U2CtAsLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - U2OS Control Arsenite Label Enriched")

# HEK ARSENITE STRESS

ggplot(alternew.HekG3AsLabEn, aes(x=alternew.HekG3AsLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 Arsenite Label Enriched")

ggplot(alternew.HekG3AsLabIn, aes(x=alternew.HekG3AsLabIn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 Arsenite Label Input")

ggplot(alternew.HekG3AsNoLabEn, aes(x=alternew.HekG3AsNoLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 Arsenite NoLabel Enriched")

ggplot(alternew.HekCtAsLabEn, aes(x=alternew.HekCtAsLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 Control Arsenite Label Enriched")


# HEK NO STRESS

ggplot(alternew.HekG3UnLabEn, aes(x=alternew.HekG3UnLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 Unstressed Label Enriched")

ggplot(alternew.HekG3UnLabIn, aes(x=alternew.HekG3UnLabIn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 Unstressed Label Input")

ggplot(alternew.HekG3UnNoLabEn, aes(x=alternew.HekG3UnNoLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 Unstressed NoLabel Enriched")

ggplot(alternew.HekCtUnLabEn, aes(x=alternew.HekCtUnLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 Control Unstressed Label Enriched")


# HEK SORBITOL

ggplot(alternew.HekG3SoLabEn, aes(x=alternew.HekG3SoLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 Sorbitol Label Enriched")

ggplot(alternew.HekG3SoLabIn, aes(x=alternew.HekG3SoLabIn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 Sorbitol Label Input")

ggplot(alternew.HekG3SoNoLabEn, aes(x=alternew.HekG3SoNoLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 Sorbitol NoLabel Enriched")

ggplot(alternew.HekCtSoLabEn, aes(x=alternew.HekCtSoLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 Control Sorbitol Label Enriched")


# HEK 1H RECOVERY

ggplot(alternew.HekG3T1LabEn, aes(x=alternew.HekG3T1LabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 1H Recovery Label Enriched")

ggplot(alternew.HekG3T1LabIn, aes(x=alternew.HekG3T1LabIn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 1H Recovery Label Input")

ggplot(alternew.HekG3T1NoLabEn, aes(x=alternew.HekG3T1NoLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 1H Recovery NoLabel Enriched")

ggplot(alternew.HekCtT1LabEn, aes(x=alternew.HekCtT1LabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 Control 1H Recovery Label Enriched")


# HEK 3H RECOVERY

ggplot(alternew.HekG3T3LabEn, aes(x=alternew.HekG3T3LabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 3H Recovery Label Enriched")

ggplot(alternew.HekG3T3LabIn, aes(x=alternew.HekG3T3LabIn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 3H Recovery Label Input")

ggplot(alternew.HekG3T3NoLabEn, aes(x=alternew.HekG3T3NoLabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 G3BP1 3H Recovery NoLabel Enriched")

ggplot(alternew.HekCtT3LabEn, aes(x=alternew.HekCtT3LabEn)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Hek293 Control 3H Recovery Label Enriched")
```

By seeing the shape of the graph, we can see where we should cut. We have a **binomial distribution** (low reads and noise vs actual biological meaningful reads). The ideal place to cut them is right before the wedge.

### ***After Filtration***

This time, I considered that it was more correct to filter all samples together.


```{r}
filterednew <- subset(read.counts, alternew>3.25)
MitoGenes <- fread("~/MitoGenes.txt", header = FALSE)[[1]]

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- row.names(falternew)


#Now we need to select only the rows which have those genes.
finaltable <- read.counts[row.names(read.counts)%in%impgenes,]
```


## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)


#Then we calculate the normalization factors for these new generated data frames
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)

colnames(sg.read.counts_U2OS)

#Lets create a dataframe with all of the medatata information
CellType <- c(rep("U2",8), rep("Hek",53))
TargetType <- c(rep("G3",6), rep("Ct",2),rep("G3",6), rep("Ct",4),rep("G3",6), rep("Ct",4),rep("G3",6), rep("Ct",4),rep("G3",9), rep("Ct",4),rep("G3",6), rep("Ct",4))
StressType <- c(rep("Arsenite",18),rep("Unstressed",10),rep("Sorbitol",10),rep("H1",13),rep("H3",10))
LabType <- c(rep("Yes",4),rep("No",2),rep("Yes",2), #U2
             rep("Yes",4),rep("No",2),rep("Yes",4), #HekAs
             rep("Yes",4),rep("No",2),rep("Yes",4), #HekUn
             rep("Yes",4),rep("No",2),rep("Yes",4), #HekSo
             rep("Yes",6),rep("No",3),rep("Yes",4), 
             rep("Yes",4),rep("No",2),rep("Yes",4))
SampleType <- c(rep("Enriched",2),rep("Input",2),rep("Enriched",4), #U2
             rep("Enriched",2),rep("Input",2),rep("Enriched",6), #HekAs
             rep("Enriched",2),rep("Input",2),rep("Enriched",6), #HekUn
             rep("Enriched",2),rep("Input",2),rep("Enriched",6), #HekSo
             rep("Enriched",3),rep("Input",3),rep("Enriched",7), 
             rep("Enriched",2),rep("Input",2),rep("Enriched",6))
Biotype <- c(rep("SG",2),rep("WC",4),rep("Vec",2), #U2
             rep("SG",2),rep("WC",4),rep("Vec",4), #HekAs
             rep("SG",2),rep("WC",4),rep("Vec",4), #HekUn
             rep("SG",2),rep("WC",4),rep("Vec",4), #HekSo
             rep("SG",3),rep("WC",6),rep("Vec",4), 
             rep("SG",2),rep("WC",4),rep("Vec",4))


metadata <- data.frame(Cell=CellType, Stress=StressType, Target=TargetType, Label=LabType, Sample=SampleType)


#Now we do the matrixes

mm <- model.matrix(~0+CellType+StressType+TargetType+LabType+SampleType)


#and now we normalize. This step also turns our counts into log2-counts per million.
#The object with all simple
y <- voom(D0, mm, plot = T)
```


## Normalisation

For normalisation I convert the read counts into log2CPMs. Its a good way to visualize how the reads are. 

```{r fig.height=6, fig.width=12}
custom_palette <- qualitative_hcl(24, palette = "Dark3")

dfforboxy <- finaltable
orderforplots <- SampleColouring
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
# Convert 'Types' to factor with custom order
dfforboxy$Types <- factor(dfforboxy$Types, levels = unique(SampleColouring))

prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = custom_palette, labels=c(unique(orderforplots))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

#Using the Standard Matrix

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
# Convert 'Types' to factor with custom order
dfforboxy$Types <- factor(dfforboxy$Types, levels = unique(SampleColouring))
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = custom_palette, labels=c(unique(orderforplots))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```
In this case, from the boxplots, we can see that there no samples which can be considered outliers.

## PCA

Before we move on, lets do a **Principal Component analysis**. It is a good exploratory tool to find out how our samples relate to each other and if there are any obvious outliers. We would expect our replicates to be nearer to one another than to other samples, and more different samples should be further away. 

```{r, fig.height=7, fig.width=9}
# Define color palette
custom_palette <- qualitative_hcl(24, palette = "Pastel")
custom_palette <- sample(custom_palette, 24)


#This prepares my data for the PCA
dfforpca <- y$E[,-c(1:8)]
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- SampleType[-c(1:8)]
dfforpca$Stress <- StressType[-c(1:8)]
dfforpca$Cell <- CellType[-c(1:8)]
dfforpca$Target <- TargetType[-c(1:8)]
dfforpca$Label <- LabType[-c(1:8)]
dfforpca$All <- SampleColouring_readable[-c(1:8)]

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-6)], scale = FALSE)

p_Sample <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Sample") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Stress <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Stress", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Stress", values = custom_palette) +
  ggtitle("PCA - Coloured by Stress") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Cell <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Cell", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Cell", values = custom_palette) +
  ggtitle("PCA - Coloured by Cell") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Target <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Target", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Target", values = custom_palette) +
  ggtitle("PCA - Coloured by Target") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Label <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Label", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Label", values = custom_palette) +
  ggtitle("PCA - Coloured by Label") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_All <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "All", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "All", values = custom_palette) +
  ggtitle("PCA - Coloured by All") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Sample
p_Stress
p_Cell
p_Target
p_Label
p_All

#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])


```
As one can observe from the PCAs, our PC1 is mostly explained by sample type, and PC2 by if a sample is WC or not.
```{r fig.height=20, fig.width=20}
#this is done to be able to do traditional PCA
library(pheatmap)
pheatmap(cor(y$E, method = "spearman"),
         labels_row = SampleColouring,
         labels_col = colnames(y$E))
```



```{r}

get_variance_explained_manual <- function(mat, metadata, variables) {
  # mat: genes x samples (rows = genes, cols = samples)
  # metadata: dataframe of sample annotations
  # variables: character vector of column names in metadata
  
  message("Calculating variance explained per gene...")
  
  # Ensure sample ordering matches
  stopifnot(ncol(mat) == nrow(metadata))
  
  results <- lapply(variables, function(var) {
    message("Processing ", var, "...")
    sapply(1:nrow(mat), function(i) {
      y <- as.numeric(mat[i, ])
      df <- data.frame(y = y, x = metadata[[var]])
      # handle categorical or numeric
      model <- lm(y ~ x, data = df)
      summary(model)$r.squared
    })
  })
  
  names(results) <- variables
  varMat <- as.data.frame(results)
  rownames(varMat) <- rownames(mat)
  return(varMat)
}
plot_variance_explained_density <- function(varMat, variables, colours, scale = "log") {
  df_plot <- reshape2::melt(varMat, variable.name = "Variable", value.name = "VarianceExplained")
  
  # Convert to percentage variance explained
  df_plot$VarianceExplained <- df_plot$VarianceExplained * 100
  
  # Convert input colours to pastel and name them to ensure correct mapping
  pastel_colours <- colorspace::lighten(colours, amount = 0.5)
  names(pastel_colours) <- variables
  
  # Build x scale based on argument
  x_scale <- if (scale == "log") {
    scale_x_log10(
      name = "% variance explained",
      breaks = c(0.001, 0.01, 0.1, 1, 10, 100),
      labels = c("0.001", "0.01", "0.1", "1", "10", "100")
    )
  } else if (scale == "linear") {
    scale_x_continuous(
      name = "% variance explained",
      breaks = c(0, 20, 40, 60, 80, 100),
      labels = c("0", "20", "40", "60", "80", "100"),
      limits = c(0, 100)
    )
  } else {
    stop("Invalid scale argument. Use 'log' or 'linear'.")
  }
  
  ggplot(df_plot, aes(x = VarianceExplained, color = Variable)) +
    geom_density(size = 1.2, adjust = 1.2) +
    x_scale +
    ylab("Density") +
    theme_bw(base_size = 16) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 20),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank()
    )  +
    scale_color_manual(values = pastel_colours)
}
#Example Usage
   #expression is your normalised expression table, with genes as rows and samples as coloumns (most likely y$E if you used voom)
   #metadata is your metadata frame where each column is a variable, and each row a sample. Must be in the same order as expression
   #variables must be a vector of the variables (columns in metadata) that you want to test. So something like  c("Centrifugation", "Stress", "SG")
   #colours must be a vector of the colours you want to use. Must be the same number as the variables. So something like c("#229954", "#EC7063", "blue")
varMat <- get_variance_explained_manual(y$E, metadata, colnames(metadata))
# Log scale (default — no need to specify)
p <- plot_variance_explained_density(varMat, colnames(metadata), c("red", "blue", "green", "purple", "black"))
print(p)
# Linear scale
p <- plot_variance_explained_density(varMat, colnames(metadata), c("red", "blue", "green", "purple", "black"), scale = "linear")
print(p)

```

## Batch Effect Correction

Looking at the PCA I considered there was a clear batch effect owing to the cell type. I opted to remove this batch effect.

```{r}
batch_cols <- c("Cell")
metadata_cols <- c("Stress", "Target", "Label", "Sample")
mmbatch <- model.matrix(reformulate(paste0("~0 + ",paste0( batch_cols,collapse = " + "))), data=metadata)
mmkeep <- model.matrix(reformulate(paste0("~0 + ",paste0( metadata_cols,collapse =  "+" ))), data=metadata)
mm <- cbind(mmkeep,mmbatch)
D0 <- DGEList(2^y$E) #senescence_normcounts
D0 <- calcNormFactors(D0)
y <- voom(D0, mm, plot = F)
fit <- lmFit(y, mm)
beta <- fit$coefficients[,-(1:ncol(mmkeep)),drop=FALSE]
#beta <- fit$coefficients
beta[is.na(beta)] <- 0
corrcounts_v1 <- as.matrix(y$E) - beta %*% t(mmbatch)
offset <- apply(corrcounts_v1,1,min) - apply(y$E,1,min)
corrcounts_v1 <- as.data.frame(corrcounts_v1-offset)
corrcounts_v1 <- 2^corrcounts_v1
y <- voom(corrcounts_v1, mm)
```


## Batch Corrected PCA

```{r fig.height=8, fig.width=12}
# Define color palette
custom_palette <- qualitative_hcl(24, palette = "Pastel")
custom_palette <- sample(custom_palette, 24)


#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- SampleType
dfforpca$Stress <- StressType
dfforpca$Cell <- CellType
dfforpca$Target <- TargetType
dfforpca$Label <- LabType
dfforpca$All <- SampleColouring_readable

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-6)], scale = FALSE)

p_Sample <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Sample") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Stress <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Stress", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Stress", values = custom_palette) +
  ggtitle("PCA - Coloured by Stress") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Cell <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Cell", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Cell", values = custom_palette) +
  ggtitle("PCA - Coloured by Cell") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Target <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Target", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Target", values = custom_palette) +
  ggtitle("PCA - Coloured by Target") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Label <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Label", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Label", values = custom_palette) +
  ggtitle("PCA - Coloured by Label") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_All <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "All", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "All", values = custom_palette) +
  ggtitle("PCA - Coloured by All") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())


p_Sample
p_Stress
p_Cell
p_Target
p_Label
p_All

#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])

```
Now PC1 is driven by sample type instead.

# Simple Linear Models

Now we must perform linear model fitting.

```{r}
mm <- model.matrix(~0+SampleColouring)
y <- voom(finaltable, mm)
fitted <- lmFit(y$E, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
U2.AS <- makeContrasts(SampleColouringU2G3AsLabEn  - SampleColouringU2G3AsLabIn, levels = (colnames(coef(fitted))))
HEK.AS<- makeContrasts(SampleColouringHekG3AsLabEn  - SampleColouringHekG3AsLabIn , levels = (colnames(coef(fitted))))
HEK.SO<- makeContrasts(SampleColouringHekG3SoLabEn  - SampleColouringHekG3SoLabIn , levels = (colnames(coef(fitted))))


#next we estimate the contrast for each gene
tmp.U2.AS  <- contrasts.fit(fitted, U2.AS )
tmp.HEK.AS <- contrasts.fit(fitted, HEK.AS)
tmp.HEK.SO <- contrasts.fit(fitted, HEK.SO)

#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp.U2.AS <- eBayes(tmp.U2.AS)
tmp.HEK.AS <- eBayes(tmp.HEK.AS)
tmp.HEK.SO <- eBayes(tmp.HEK.SO)

top.table.U2.AS <- topTable(tmp.U2.AS, sort.by = "logFC", n=Inf)
top.table.HEK.AS <- topTable(tmp.HEK.AS, sort.by = "logFC", n=Inf)
top.table.HEK.SO<- topTable(tmp.HEK.SO, sort.by = "logFC", n=Inf)

DEGenes.U2.AS <- top.table.U2.AS
DEGenes.HEK.AS <- top.table.HEK.AS
DEGenes.HEK.SO <- top.table.HEK.SO





ndegs=20
```

# U2-OS Arsenite

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
write.table(x = DEGenes.U2.AS,
          file = "~/Projects/StressGranules/SGPAPER/SGvsWC/Ren_U2Ar.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes.U2.AS[DEGenes.U2.AS$logFC>2.5 & DEGenes.U2.AS$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.U2.AS[DEGenes.U2.AS$logFC< -2 & DEGenes.U2.AS$adj.P.Val<0.05,] )


ggplot(data=DEGenes.U2.AS,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#DDDDDD")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs Whole-Cell")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#de425b",max.overlaps = 200, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#769fca",max.overlaps = 200, fontface="italic")+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#de425b", size=3)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#769fca", size=3) + xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") + theme(text=element_text(family = "Arial", size = 20))

```

## Volcano Plot with Gene Lenght

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes.U2.AS$Gene <- row.names(DEGenes.U2.AS)
colnames(genelength) <- c("Length","Gene")
DEGenes.U2.AS <- merge(DEGenes.U2.AS, genelength, by="Gene")

ggplot(data=DEGenes.U2.AS,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.U2.AS)
liner.SG <- rstandard(smooth.SG)
DEGenes.U2.AS <-cbind(DEGenes.U2.AS, liner.SG)
colnames(DEGenes.U2.AS)[9] <- "Residuals"
row.names(DEGenes.U2.AS) <- DEGenes.U2.AS$Gene
linemarkers <- as.data.frame(DEGenes.U2.AS[abs(DEGenes.U2.AS$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.U2.AS$t, log10(DEGenes.U2.AS$Length), method = "pearson")



ggplot(DEGenes.U2.AS, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
We see a positive lenght bias.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}
# Define color palette
custom_palette <- qualitative_hcl(24, palette = "Pastel")
custom_palette <- sample(custom_palette, 24)

finagenes <- rbind(upmarkers, downmarkers)


dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:61))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=SampleColouring)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes.U2.AS <- DEGenes.U2.AS[order(DEGenes.U2.AS[,4]),]
gseatool <- DEGenes.U2.AS[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.U2.AS$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


## GSEA Length

Now, since SGs are enriched for longer transcripts, these pathways may simply be the ones which are tipically enriched in longer transcripts. Therefore, I will repeat the analysis, but using transcript lenght instead of t-statistic. 

```{r, results=FALSE}
#for all genes
DEGenes.U2.AS <- DEGenes.U2.AS[order(DEGenes.U2.AS[,8]),]
gseatool <- DEGenes.U2.AS[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.U2.AS$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)


#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg_pval", "sg", "size_pval", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")

```

Even though there is some lenght bias in terms of enrichment, it does not seem to be linked in terms of gene set enrichment.

## Most altered pathways

I opted to see altered pathways as a whole. That is, ignoring bidirectionally and simply looking at what pathways are more altered as a whole (i.e, more expressed genes on either side).

```{r, echo=FALSE, fig.height=20, fig.width=30, message=FALSE, warning=FALSE, results='hide'}
#for all genes
DEGenes.U2.AS <- DEGenes.U2.AS[order(DEGenes.U2.AS[,4]),]
gseatool <- DEGenes.U2.AS[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes.U2.AS)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol
forcorrel <- vectorranks
vectorranks <- abs(vectorranks)


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")



gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```


# HEK-293T Arsenite

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
write.table(x = DEGenes.HEK.AS,
          file = "~/Projects/StressGranules/SGPAPER/SGvsWC/Ren_HEKAr.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes.HEK.AS[DEGenes.HEK.AS$logFC>2 & DEGenes.HEK.AS$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.HEK.AS[DEGenes.HEK.AS$logFC< -2 & DEGenes.HEK.AS$adj.P.Val<0.05,] )


ggplot(data=DEGenes.HEK.AS,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#DDDDDD")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs Whole-Cell")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#de425b",max.overlaps = 200, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#769fca",max.overlaps = 200, fontface="italic")+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#de425b", size=3)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#769fca", size=3) + xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") + theme(text=element_text(family = "Arial", size = 20))

```

## Volcano Plot with Gene Lenght

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes.HEK.AS$Gene <- row.names(DEGenes.HEK.AS)
colnames(genelength) <- c("Length","Gene")
DEGenes.HEK.AS <- merge(DEGenes.HEK.AS, genelength, by="Gene")

ggplot(data=DEGenes.HEK.AS,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.HEK.AS)
liner.SG <- rstandard(smooth.SG)
DEGenes.HEK.AS <-cbind(DEGenes.HEK.AS, liner.SG)
colnames(DEGenes.HEK.AS)[9] <- "Residuals"
row.names(DEGenes.HEK.AS) <- DEGenes.HEK.AS$Gene
linemarkers <- as.data.frame(DEGenes.HEK.AS[abs(DEGenes.HEK.AS$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.HEK.AS$t, log10(DEGenes.HEK.AS$Length), method = "pearson")



ggplot(DEGenes.HEK.AS, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
We see a positive lenght bias.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}
# Define color palette
custom_palette <- qualitative_hcl(24, palette = "Pastel")
custom_palette <- sample(custom_palette, 24)

finagenes <- rbind(upmarkers, downmarkers)


dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:61))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=SampleColouring)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes.HEK.AS <- DEGenes.HEK.AS[order(DEGenes.HEK.AS[,4]),]
gseatool <- DEGenes.HEK.AS[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.HEK.AS$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


## GSEA Length

Now, since SGs are enriched for longer transcripts, these pathways may simply be the ones which are tipically enriched in longer transcripts. Therefore, I will repeat the analysis, but using transcript lenght instead of t-statistic. 

```{r, results=FALSE}
#for all genes
DEGenes.HEK.AS <- DEGenes.HEK.AS[order(DEGenes.HEK.AS[,8]),]
gseatool <- DEGenes.HEK.AS[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.HEK.AS$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)


#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg_pval", "sg", "size_pval", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")

```

Even though there is some lenght bias in terms of enrichment, it does not seem to be linked in terms of gene set enrichment.

## Most altered pathways

I opted to see altered pathways as a whole. That is, ignoring bidirectionally and simply looking at what pathways are more altered as a whole (i.e, more expressed genes on either side).

```{r, echo=FALSE, fig.height=20, fig.width=30, message=FALSE, warning=FALSE, results='hide'}
#for all genes
DEGenes.HEK.AS <- DEGenes.HEK.AS[order(DEGenes.HEK.AS[,4]),]
gseatool <- DEGenes.HEK.AS[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes.HEK.AS)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol
forcorrel <- vectorranks
vectorranks <- abs(vectorranks)


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")



gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```


# HEK293T Sorbitol

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
write.table(x = DEGenes.HEK.SO,
          file = "~/Projects/StressGranules/SGPAPER/SGvsWC/Ren_HEKSo.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes.HEK.SO[DEGenes.HEK.SO$logFC>2 & DEGenes.HEK.SO$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.HEK.SO[DEGenes.HEK.SO$logFC< -2 & DEGenes.HEK.SO$adj.P.Val<0.05,] )


ggplot(data=DEGenes.HEK.SO,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#DDDDDD")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs Whole-Cell")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#de425b",max.overlaps = 200, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#769fca",max.overlaps = 200, fontface="italic")+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#de425b", size=3)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#769fca", size=3) + xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") + theme(text=element_text(family = "Arial", size = 20))

```


## Volcano Plot with Gene Lenght

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes.HEK.SO$Gene <- row.names(DEGenes.HEK.SO)
colnames(genelength) <- c("Length","Gene")
DEGenes.HEK.SO <- merge(DEGenes.HEK.SO, genelength, by="Gene")

ggplot(data=DEGenes.HEK.SO,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.HEK.SO)
liner.SG <- rstandard(smooth.SG)
DEGenes.HEK.SO <-cbind(DEGenes.HEK.SO, liner.SG)
colnames(DEGenes.HEK.SO)[9] <- "Residuals"
row.names(DEGenes.HEK.SO) <- DEGenes.HEK.SO$Gene
linemarkers <- as.data.frame(DEGenes.HEK.SO[abs(DEGenes.HEK.SO$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.HEK.SO$t, log10(DEGenes.HEK.SO$Length), method = "pearson")



ggplot(DEGenes.HEK.SO, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
We see a positive lenght bias.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}
# Define color palette
custom_palette <- qualitative_hcl(24, palette = "Pastel")
custom_palette <- sample(custom_palette, 24)

finagenes <- rbind(upmarkers, downmarkers)


dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:61))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=SampleColouring)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes.HEK.SO <- DEGenes.HEK.SO[order(DEGenes.HEK.SO[,4]),]
gseatool <- DEGenes.HEK.SO[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.HEK.SO$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


## GSEA Length

Now, since SGs are enriched for longer transcripts, these pathways may simply be the ones which are tipically enriched in longer transcripts. Therefore, I will repeat the analysis, but using transcript lenght instead of t-statistic. 

```{r, results=FALSE}
#for all genes
DEGenes.HEK.SO <- DEGenes.HEK.SO[order(DEGenes.HEK.SO[,8]),]
gseatool <- DEGenes.HEK.SO[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.HEK.SO$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)


#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg_pval", "sg", "size_pval", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")

```

Even though there is some lenght bias in terms of enrichment, it does not seem to be linked in terms of gene set enrichment.

## Most altered pathways

I opted to see altered pathways as a whole. That is, ignoring bidirectionally and simply looking at what pathways are more altered as a whole (i.e, more expressed genes on either side).

```{r, echo=FALSE, fig.height=20, fig.width=30, message=FALSE, warning=FALSE, results='hide'}
#for all genes
DEGenes.HEK.SO <- DEGenes.HEK.SO[order(DEGenes.HEK.SO[,4]),]
gseatool <- DEGenes.HEK.SO[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes.HEK.SO)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol
forcorrel <- vectorranks
vectorranks <- abs(vectorranks)


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")



gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```


# Complex Linear Models

Now we must perform linear model fitting.

```{r}
#Lets create a dataframe with all of the medatata information
CellType <- c(rep("U2",8), rep("Hek",53))
TargetType <- c(rep("G3",6), rep("Ct",2),rep("G3",6), rep("Ct",4),rep("G3",6), rep("Ct",4),rep("G3",6), rep("Ct",4),rep("G3",9), rep("Ct",4),rep("G3",6), rep("Ct",4))
StressType <- c(rep("Arsenite",18),rep("Unstressed",10),rep("Sorbitol",10),rep("H1",13),rep("H3",10))
LabType <- c(rep("Yes",4),rep("No",2),rep("Yes",2), #U2
             rep("Yes",4),rep("No",2),rep("Yes",4), #HekAs
             rep("Yes",4),rep("No",2),rep("Yes",4), #HekUn
             rep("Yes",4),rep("No",2),rep("Yes",4), #HekSo
             rep("Yes",6),rep("No",3),rep("Yes",4), 
             rep("Yes",4),rep("No",2),rep("Yes",4))
SampleType <- c(rep("Enriched",2),rep("Input",2),rep("Enriched",4), #U2
             rep("Enriched",2),rep("Input",2),rep("Enriched",6), #HekAs
             rep("Enriched",2),rep("Input",2),rep("Enriched",6), #HekUn
             rep("Enriched",2),rep("Input",2),rep("Enriched",6), #HekSo
             rep("Enriched",3),rep("Input",3),rep("Enriched",7), 
             rep("Enriched",2),rep("Input",2),rep("Enriched",6))
Biotype <- c(rep("SG",2),rep("WC",4),rep("Vec",2), #U2
             rep("SG",2),rep("WC",4),rep("Vec",4), #HekAs
             rep("SG",2),rep("WC",4),rep("Vec",4), #HekUn
             rep("SG",2),rep("WC",4),rep("Vec",4), #HekSo
             rep("SG",3),rep("WC",6),rep("Vec",4), 
             rep("SG",2),rep("WC",4),rep("Vec",4))


metadata <- data.frame(Cell=CellType, Stress=StressType, Target=TargetType, Label=LabType, Sample=SampleType)

mm <- model.matrix(~0+Biotype+StressType)
y <- voom(corrcounts_v1, mm)
fitted <- lmFit(y$E, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
SG.vs.WC <- makeContrasts(BiotypeSG - BiotypeWC, levels = (colnames(coef(fitted))))
SG.vs.VEC <- makeContrasts(BiotypeSG - BiotypeVec, levels = (colnames(coef(fitted))))

#next we estimate the contrast for each gene
tmp.SG.vs.WC  <- contrasts.fit(fitted, SG.vs.WC )
tmp.SG.vs.VEC <- contrasts.fit(fitted, SG.vs.VEC)

#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp.SG.vs.WC <- eBayes(tmp.SG.vs.WC)
tmp.SG.vs.VEC <- eBayes(tmp.SG.vs.VEC)

top.table.SG.vs.WC <- topTable(tmp.SG.vs.WC, sort.by = "logFC", n=Inf)
top.table.SG.vs.VEC <- topTable(tmp.SG.vs.VEC, sort.by = "logFC", n=Inf)

DEGenes.SG.vs.WC <- top.table.SG.vs.WC
DEGenes.SG.vs.VEC <- top.table.SG.vs.VEC




ndegs=20
```


# SG vs WC

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
write.table(x = DEGenes.SG.vs.WC,
          file = "~/Projects/StressGranules/SGPAPER/SGvsWC/CapSeq.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes.SG.vs.WC[DEGenes.SG.vs.WC$logFC>2 & DEGenes.SG.vs.WC$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.SG.vs.WC[DEGenes.SG.vs.WC$logFC< -2 & DEGenes.SG.vs.WC$adj.P.Val<0.05,] )


ggplot(data=DEGenes.SG.vs.WC,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#DDDDDD")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs Whole-Cell")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#de425b",max.overlaps = 200, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#769fca",max.overlaps = 200, fontface="italic")+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#de425b", size=3)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#769fca", size=3) + xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") + theme(text=element_text(family = "Arial", size = 20))

```

## Volcano Plot with Gene Lenght

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes.SG.vs.WC$Gene <- row.names(DEGenes.SG.vs.WC)
colnames(genelength) <- c("Length","Gene")
DEGenes.SG.vs.WC <- merge(DEGenes.SG.vs.WC, genelength, by="Gene")

ggplot(data=DEGenes.SG.vs.WC,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.SG.vs.WC)
liner.SG <- rstandard(smooth.SG)
DEGenes.SG.vs.WC <-cbind(DEGenes.SG.vs.WC, liner.SG)
colnames(DEGenes.SG.vs.WC)[9] <- "Residuals"
row.names(DEGenes.SG.vs.WC) <- DEGenes.SG.vs.WC$Gene
linemarkers <- as.data.frame(DEGenes.SG.vs.WC[abs(DEGenes.SG.vs.WC$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.SG.vs.WC$t, log10(DEGenes.SG.vs.WC$Length), method = "pearson")



ggplot(DEGenes.SG.vs.WC, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
We see a positive lenght bias.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}
# Define color palette
custom_palette <- qualitative_hcl(24, palette = "Pastel")
custom_palette <- sample(custom_palette, 24)

finagenes <- rbind(upmarkers, downmarkers)


dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:61))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=SampleColouring)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes.SG.vs.WC <- DEGenes.SG.vs.WC[order(DEGenes.SG.vs.WC[,4]),]
gseatool <- DEGenes.SG.vs.WC[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.SG.vs.WC$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


## GSEA Length

Now, since SGs are enriched for longer transcripts, these pathways may simply be the ones which are tipically enriched in longer transcripts. Therefore, I will repeat the analysis, but using transcript lenght instead of t-statistic. 

```{r, results=FALSE}
#for all genes
DEGenes.SG.vs.WC <- DEGenes.SG.vs.WC[order(DEGenes.SG.vs.WC[,8]),]
gseatool <- DEGenes.SG.vs.WC[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.SG.vs.WC$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)


#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg_pval", "sg", "size_pval", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")

```

Even though there is some lenght bias in terms of enrichment, it does not seem to be linked in terms of gene set enrichment.

## Most altered pathways

I opted to see altered pathways as a whole. That is, ignoring bidirectionally and simply looking at what pathways are more altered as a whole (i.e, more expressed genes on either side).

```{r, echo=FALSE, fig.height=20, fig.width=30, message=FALSE, warning=FALSE, results='hide'}
#for all genes
DEGenes.SG.vs.WC <- DEGenes.SG.vs.WC[order(DEGenes.SG.vs.WC[,4]),]
gseatool <- DEGenes.SG.vs.WC[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes.SG.vs.WC)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol
forcorrel <- vectorranks
vectorranks <- abs(vectorranks)


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")



gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```

# SG vs VEC

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes.SG.vs.VEC[DEGenes.SG.vs.VEC$logFC>1 & DEGenes.SG.vs.VEC$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.SG.vs.VEC[DEGenes.SG.vs.VEC$logFC< -1 & DEGenes.SG.vs.VEC$adj.P.Val<0.05,] )


ggplot(data=DEGenes.SG.vs.VEC,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#DDDDDD")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs VEC")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#de425b",max.overlaps = 200, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#769fca",max.overlaps = 200, fontface="italic")+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#de425b", size=3)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#769fca", size=3) + xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") + theme(text=element_text(family = "Arial", size = 20))

```

## Volcano Plot with Gene Lenght

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes.SG.vs.VEC$Gene <- row.names(DEGenes.SG.vs.VEC)
colnames(genelength) <- c("Length","Gene")
DEGenes.SG.vs.VEC <- merge(DEGenes.SG.vs.VEC, genelength, by="Gene")

ggplot(data=DEGenes.SG.vs.VEC,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes.SG.vs.VEC)
liner.SG <- rstandard(smooth.SG)
DEGenes.SG.vs.VEC <-cbind(DEGenes.SG.vs.VEC, liner.SG)
colnames(DEGenes.SG.vs.VEC)[9] <- "Residuals"
row.names(DEGenes.SG.vs.VEC) <- DEGenes.SG.vs.VEC$Gene
linemarkers <- as.data.frame(DEGenes.SG.vs.VEC[abs(DEGenes.SG.vs.VEC$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes.SG.vs.VEC$t, log10(DEGenes.SG.vs.VEC$Length), method = "pearson")



ggplot(DEGenes.SG.vs.VEC, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
We see a positive lenght bias.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}
# Define color palette
custom_palette <- qualitative_hcl(24, palette = "Pastel")
custom_palette <- sample(custom_palette, 24)

finagenes <- rbind(upmarkers, downmarkers)


dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:61))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=SampleColouring)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(SampleColouring)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes.SG.vs.VEC <- DEGenes.SG.vs.VEC[order(DEGenes.SG.vs.VEC[,4]),]
gseatool <- DEGenes.SG.vs.VEC[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.SG.vs.VEC$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


## GSEA Length

Now, since SGs are enriched for longer transcripts, these pathways may simply be the ones which are tipically enriched in longer transcripts. Therefore, I will repeat the analysis, but using transcript lenght instead of t-statistic. 

```{r, results=FALSE}
#for all genes
DEGenes.SG.vs.VEC <- DEGenes.SG.vs.VEC[order(DEGenes.SG.vs.VEC[,8]),]
gseatool <- DEGenes.SG.vs.VEC[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes.SG.vs.VEC$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)


#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg_pval", "sg", "size_pval", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")

```

Even though there is some lenght bias in terms of enrichment, this time it does seem to be linked with gene set enrichment

## Most altered pathways

I opted to see altered pathways as a whole. That is, ignoring bidirectionally and simply looking at what pathways are more altered as a whole (i.e, more expressed genes on either side).

```{r, echo=FALSE, fig.height=20, fig.width=30, message=FALSE, warning=FALSE, results='hide'}
#for all genes
DEGenes.SG.vs.VEC <- DEGenes.SG.vs.VEC[order(DEGenes.SG.vs.VEC[,4]),]
gseatool <- DEGenes.SG.vs.VEC[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes.SG.vs.VEC)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol
forcorrel <- vectorranks
vectorranks <- abs(vectorranks)


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")



gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```

# Final Version

```{r}
# ── Build full metadata from sample-name structure ──────────────────────
CellType <- c(rep("U2",8), rep("Hek",53))
TargetType <- c(rep("G3",6), rep("Ct",2),rep("G3",6), rep("Ct",4),rep("G3",6), rep("Ct",4),rep("G3",6), rep("Ct",4),rep("G3",9), rep("Ct",4),rep("G3",6), rep("Ct",4))
StressType <- c(rep("Arsenite",18),rep("Unstressed",10),rep("Sorbitol",10),rep("H1",13),rep("H3",10))
LabType <- c(rep("Yes",4),rep("No",2),rep("Yes",2), #U2
             rep("Yes",4),rep("No",2),rep("Yes",4), #HekAs
             rep("Yes",4),rep("No",2),rep("Yes",4), #HekUn
             rep("Yes",4),rep("No",2),rep("Yes",4), #HekSo
             rep("Yes",6),rep("No",3),rep("Yes",4), 
             rep("Yes",4),rep("No",2),rep("Yes",4))
SampleType <- c(rep("Enriched",2),rep("Input",2),rep("Enriched",4), #U2
             rep("Enriched",2),rep("Input",2),rep("Enriched",6), #HekAs
             rep("Enriched",2),rep("Input",2),rep("Enriched",6), #HekUn
             rep("Enriched",2),rep("Input",2),rep("Enriched",6), #HekSo
             rep("Enriched",3),rep("Input",3),rep("Enriched",7), 
             rep("Enriched",2),rep("Input",2),rep("Enriched",6))

# Rename to the real biological roles instead of one collapsed Yes/No flag
metadata <- data.frame(
  CellType = CellType,
  Bait     = factor(ifelse(TargetType == "G3", "G3BP1", "Untargeted"),
                    levels = c("Untargeted", "G3BP1")),
  Stress   = StressType,
  Probe    = factor(ifelse(LabType == "Yes", "label", "noprobe"),
                    levels = c("noprobe", "label")),
  Fraction = SampleType   # "Enriched" / "Input"
)
stopifnot(nrow(metadata) == ncol(finaltable))

# ── Step 1: background-correct G3BP1/Enriched/label samples using ───────
#    matched G3BP1/Enriched/noprobe samples, per Stress level
#    (Untargeted has no noprobe arm in this design, so this only applies
#    to G3BP1 — exactly the samples it's meant to correct)
y <- voom(finaltable, model.matrix(~0 + CellType + Bait + Stress + Probe + Fraction, data = metadata))
logexpr <- y$E

correct_with_noprobe <- function(logexpr, metadata) {
  out <- logexpr
  for (st in unique(metadata$Stress)) {
    label_idx   <- which(metadata$Bait == "G3BP1" & metadata$Probe == "label" &
                         metadata$Fraction == "Enriched" & metadata$Stress == st)
    noprobe_idx <- which(metadata$Bait == "G3BP1" & metadata$Probe == "noprobe" &
                         metadata$Fraction == "Enriched" & metadata$Stress == st)
    if (length(label_idx) == 0 || length(noprobe_idx) == 0) next
    noprobe_mean <- rowMeans(logexpr[, noprobe_idx, drop = FALSE])
    out[, label_idx] <- logexpr[, label_idx, drop = FALSE] - noprobe_mean
  }
  out
}
logexpr_corrected <- correct_with_noprobe(logexpr, metadata)

# ── Step 2: main model on background-corrected Enriched/label samples ───
#    Drop H3 — IF/microscopy data shows granules are dissolved by 3h, so
#    pooling it into "Stressed" would dilute the interaction term
keep <- metadata$Probe == "label" & metadata$Fraction == "Enriched" & metadata$Stress != "H3"
meta_main <- metadata[keep, ]
expr_main <- logexpr_corrected[, keep]

meta_main$StressGroup <- factor(ifelse(meta_main$Stress == "Unstressed", "Basal", "Stressed"),
                                levels = c("Basal", "Stressed"))
meta_main$Bait     <- factor(meta_main$Bait, levels = c("Untargeted", "G3BP1"))
meta_main$CellType <- factor(meta_main$CellType)

mm <- model.matrix(~ CellType + Bait * StressGroup, data = meta_main)
fitted <- lmFit(expr_main, mm)
head(coef(fitted))

# Empirical Bayes smoothing of standard errors (shrinks extreme SEs)
tmp.SG.vs.VEC <- eBayes(fitted)

# The interaction term is the SG-specific signal: stress-induced change in
# G3BP1 proximity, with (a) baseline G3BP1 proximity and (b) the generic
# cytoplasmic stress response (shared with Untargeted) both subtracted out
top.table.SG.vs.VEC <- topTable(tmp.SG.vs.VEC,
                                coef = "BaitG3BP1:StressGroupStressed",
                                sort.by = "logFC", n = Inf)
DEGenes.SG.vs.VEC <- top.table.SG.vs.VEC
ndegs <- 20
```

```{r}
top_interact <- topTable(tmp.SG.vs.VEC, coef="BaitG3BP1:StressGroupStressed", sort.by="logFC", n=Inf)
top_bait     <- topTable(tmp.SG.vs.VEC, coef="BaitG3BP1", sort.by="logFC", n=Inf)
top_stress   <- topTable(tmp.SG.vs.VEC, coef="StressGroupStressed", sort.by="logFC", n=Inf)

interact_t <- data.frame(Gene=row.names(top_interact), interact_t=top_interact$t)
bait_t     <- data.frame(Gene=row.names(top_bait), bait_t=top_bait$t)
stress_t   <- data.frame(Gene=row.names(top_stress), stress_t=top_stress$t)

MitoGenes <- fread("~/MitoGenes.txt")[[1]]
NucGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]

## ---- Interaction vs Bait (G3BP1) component ----

comp1_df <- merge(interact_t, bait_t, by="Gene")
row.names(comp1_df) <- comp1_df$Gene
MitoData_c1 <- comp1_df[comp1_df$Gene %in% MitoGenes, ]
NucData_c1  <- comp1_df[comp1_df$Gene %in% NucGenes, ]

spearman_c1 <- cor(comp1_df$interact_t, comp1_df$bait_t, method = "spearman")

(p_c1 <- ggplot(comp1_df, aes(x=bait_t, y=interact_t)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData_c1, aes(x=bait_t, y=interact_t), color="#73937E", size=4) +
  geom_point(data=NucData_c1, aes(x=bait_t, y=interact_t), color="#CB643F", size=4) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color="grey40") +
  theme_bw() + ylab("Interaction (t-stat)") + xlab("Bait Component (t-stat)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Interaction vs Bait Component") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=15, family="Arial MS")) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size=5))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/ZRen_InteractionVsBait.png", width = 7, height=6, units = "in", res=300)
p_c1
dev.off()

## ---- Interaction vs Stress component ----

comp2_df <- merge(interact_t, stress_t, by="Gene")
row.names(comp2_df) <- comp2_df$Gene
MitoData_c2 <- comp2_df[comp2_df$Gene %in% MitoGenes, ]
NucData_c2  <- comp2_df[comp2_df$Gene %in% NucGenes, ]

spearman_c2 <- cor(comp2_df$interact_t, comp2_df$stress_t, method = "spearman")

(p_c2 <- ggplot(comp2_df, aes(x=stress_t, y=interact_t)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData_c2, aes(x=stress_t, y=interact_t), color="#73937E", size=4) +
  geom_point(data=NucData_c2, aes(x=stress_t, y=interact_t), color="#CB643F", size=4) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color="grey40") +
  theme_bw() + ylab("Interaction (t-stat)") + xlab("Stress Component (t-stat)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Interaction vs Stress Component") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=15, family="Arial MS")) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size=5))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/ZRen_InteractionVsStress.png", width = 7, height=6, units = "in", res=300)
p_c2
dev.off()
```


## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
write.table(x = DEGenes.SG.vs.VEC,
          file = "~/Projects/StressGranules/SGPAPER/RealSGs/CapSeq.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes.SG.vs.VEC[DEGenes.SG.vs.VEC$logFC>4 & DEGenes.SG.vs.VEC$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes.SG.vs.VEC[DEGenes.SG.vs.VEC$logFC< -4 & DEGenes.SG.vs.VEC$adj.P.Val<0.05,] )


ggplot(data=DEGenes.SG.vs.VEC,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#DDDDDD")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs VEC")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#de425b",max.overlaps = 200, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#769fca",max.overlaps = 200, fontface="italic")+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#de425b", size=3)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#769fca", size=3) + xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") + theme(text=element_text(family = "Arial", size = 20))

DEGenes <- DEGenes.SG.vs.VEC
MitoGenes <- fread("~/MitoGenes.txt")[[1]]
MitoData <- DEGenes[rownames(DEGenes) %in% MitoGenes, ]
NucGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]
NucData <- DEGenes[rownames(DEGenes) %in% NucGenes, ]
(m <- ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=B)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=NucData, aes(x=logFC, y=B), color="#CB643F", size=4) +
  geom_point(data=MitoData, aes(x=logFC, y=B), color="#73937E", size=4) +
  ylab("B-Statistic") +
  xlab(expression(Log[2]~"Fold-Change")) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),   # remove major gridlines
    panel.grid.minor = element_blank(),   # remove minor gridlines
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))


png("~/Projects/StressGranules/SGPAPER/Images/PNG/ZRenMito.png", width = 7, height=6, units = "in", res=300)
m
dev.off()

```

# MA Plot

```{r}
# MA plot: mean expression (x) vs logFC (y), same highlighting scheme as the volcano

ggplot(data=DEGenes,
                 aes(x=AveExpr,
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=AveExpr, y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=AveExpr, y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  theme_bw() + ylab(expression(Log[2]~"Fold-Change")) + xlab("Average Expression (Log2 CPM)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("T. Matheny Both - MA Plot") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,], mapping=aes(x=AveExpr, y=logFC, label=row.names(upmarkers)[1:ndegs]), size=4, point.padding=-0.5, color="#A10702", max.overlaps=20, fontface="italic") +
  geom_label_repel(downmarkers[1:ndegs,], mapping=aes(x=AveExpr, y=logFC, label=row.names(downmarkers)[1:ndegs]), size=4, point.padding=-0.5, color="#03256C", max.overlaps=20, fontface="italic") +
  geom_point(data=upmarkers, aes(x=AveExpr, y=logFC), color="#A10702", size=4) +
  geom_point(data=downmarkers, aes(x=AveExpr, y=logFC), color="#03256C", size=4) +
  theme(text=element_text(size=15, family="Arial MS"))

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=AveExpr,
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=AveExpr, y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=AveExpr, y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) +
  xlab("Average Expression (Log2 CPM)") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/RenMito_MA.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()
```

## MA Plot

```{r}
MitoGenes <- fread("~/MitoGenes.txt")[[1]]

genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)
genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")
genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()
genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")

genelength$Cromo <- ifelse(genelength$Gene %in% MitoGenes, "Mito", "Other")

# Apply the length-based weight formula
genelength$S <- ifelse(
  genelength$Cromo == "Mito",
  (genelength$Length)^(1/3) * 3.23e-11 + 1.17e-9,   # Formula for mitochondrial RNAs
  (genelength$Length)^(1/3) * 6.09e-11 + 3.88e-10    # Formula for other RNAs
)
DEGenes$Gene <- row.names(DEGenes)
DEGenes <- merge(DEGenes, genelength, by="Gene")
MitoGenes <- fread("~/MitoGenes.txt")[[1]]
MitoData <- DEGenes[DEGenes$Gene %in% MitoGenes, ]
NucGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]
NucData <- DEGenes[DEGenes$Gene %in% NucGenes, ]

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=log10(Length),
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=log10(Length), y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=log10(Length), y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) +
  xlab("RNA Length (log10)") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/ZRenMito_MA_length.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=(S),
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=(S), y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=(S), y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) +
  xlab("S") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/ZRenMito_MA_S.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()
```

## Volcano Plot with Gene Lenght

```{r, message=FALSE, warning=FALSE, results='hide'}
# 
# DEGenes.SG.vs.VEC$Gene <- row.names(DEGenes.SG.vs.VEC)
# colnames(genelength) <- c("Length","Gene")
# DEGenes.SG.vs.VEC <- merge(DEGenes.SG.vs.VEC, genelength, by="Gene")
# 
# ggplot(data=DEGenes.SG.vs.VEC,
#                  aes(x=logFC,
#                      y=  B)) +
#   geom_point(aes(color=log10(Length)))+
#   theme_bw() +
#   theme(axis.line = element_line(colour = "black"),
#         panel.grid.minor = element_blank(),
#         panel.border = element_blank(),
#         panel.background = element_blank()) +
#   ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
#   theme(plot.title = element_text(hjust = 0.5))  +
#   scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))
# 
# 
# smooth.SG <- lm(t~log10(Length), DEGenes.SG.vs.VEC)
# liner.SG <- rstandard(smooth.SG)
# DEGenes.SG.vs.VEC <-cbind(DEGenes.SG.vs.VEC, liner.SG)
# colnames(DEGenes.SG.vs.VEC)[9] <- "Residuals"
# row.names(DEGenes.SG.vs.VEC) <- DEGenes.SG.vs.VEC$Gene
# linemarkers <- as.data.frame(DEGenes.SG.vs.VEC[abs(DEGenes.SG.vs.VEC$Residuals)>4,])
# 
# # Calculate the Spearman correlation coefficient
# spearman_corr <- cor(DEGenes.SG.vs.VEC$t, log10(DEGenes.SG.vs.VEC$Length), method = "pearson")
# 
# 
# 
# ggplot(DEGenes.SG.vs.VEC, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
#  ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
#   theme(axis.line = element_line(colour = "black"),
#         panel.grid.minor = element_blank(),
#         panel.border = element_blank(),
#         panel.background = element_blank()) + 
#   theme(plot.title = element_text(hjust = 0.5))  +
#   geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(rstatix)

# expr_main / meta_main are exactly the samples used to fit tmp.SG.vs.VEC
expression.df <- expr_main
bait_grp   <- meta_main$Bait
stress_grp <- meta_main$StressGroup
Condition  <- interaction(bait_grp, stress_grp, sep=".")

custom_palette <- c(
  "Untargeted.Basal"    = "#a2d1e6",
  "Untargeted.Stressed" = "#769fca",
  "G3BP1.Basal"         = "#e88184",
  "G3BP1.Stressed"      = "#de425b"
)

# ── 1. Gene sets already loaded earlier in the script (MitoGenes, NucGenes) ──
mito_genes_present <- MitoGenes[MitoGenes %in% rownames(expression.df)]
nuc_genes_present  <- NucGenes[NucGenes %in% rownames(expression.df)]
cat(length(mito_genes_present), "Mito genes found in expression matrix\n")
cat(length(nuc_genes_present), "Nuc genes found in expression matrix\n")

# ── 2. Rank all genes per sample (1 = least expressed) ───────────────────────
rank_matrix <- apply(expression.df, 2, function(x) rank(x, ties.method = "average"))
rank_matrix <- as.data.frame(rank_matrix)

# ── 3. Compute a rank-sum score per sample, one gene set at a time ───────────
compute_score <- function(genes, rank_matrix, n_total) {
  colSums(rank_matrix[genes, , drop = FALSE]) / n_total
}

mito_score <- compute_score(mito_genes_present, rank_matrix, nrow(expression.df))
nuc_score  <- compute_score(nuc_genes_present, rank_matrix, nrow(expression.df))

# ── 4. Assemble plot dataframe, long format across both gene sets ────────────
score_df <- rbind(
  data.frame(Sample = colnames(expression.df), Score = as.numeric(mito_score),
             BaitGrp = bait_grp, StressGrp = stress_grp,
             Condition = Condition, GeneSet = "Mito"),
  data.frame(Sample = colnames(expression.df), Score = as.numeric(nuc_score),
             BaitGrp = bait_grp, StressGrp = stress_grp,
             Condition = Condition, GeneSet = "Nuc")
)

# ── 5. Statistics per gene set: two-way ANOVA (Bait*StressGroup), interaction term ──
stat_labels <- score_df %>%
  group_by(GeneSet) %>%
  group_modify(~{
    aov_res  <- aov(Score ~ BaitGrp * StressGrp, data = .x)
    aov_tidy <- summary(aov_res)

    # The interaction term is the third row of the ANOVA table
    p_raw  <- aov_tidy[[1]][["Pr(>F)"]][3]
    p_adj  <- p.adjust(p_raw, method = "BH")

    ss_interact <- aov_tidy[[1]][["Sum Sq"]][3]
    ss_total    <- sum(aov_tidy[[1]][["Sum Sq"]])
    eta_sq      <- ss_interact / ss_total
    cohens_f    <- sqrt(eta_sq / (1 - eta_sq))

    data.frame(
      label = paste0("Interaction Cohen's f = ", round(cohens_f, 3),
                      "\nadj. p = ", formatC(p_adj, format = "e", digits = 2))
    )
  })

# ── 6. Plot ───────────────────────────────────────────────────────────────────
vilsigplot <- ggplot(score_df, aes(x = Condition, y = Score, fill = Condition)) +
  geom_violin(colour = "black", alpha = 0.6, trim = FALSE) +
  geom_point(
    position = position_jitter(width = 0.1),
    alpha    = 0.7,
    size     = 1.5
  ) +
  stat_summary(fun = mean,   geom = "point", shape = 18,
               size = 3, colour = "black") +
  stat_summary(fun = median, geom = "point", shape = 17,
               size = 3, colour = "black") +
  geom_text(
    data  = stat_labels,
    aes(x = -Inf, y = Inf, label = label),
    hjust = -0.05, vjust = 1.2,
    size  = 4, fontface = "bold", lineheight = 1.1,
    inherit.aes = FALSE
  ) +
  facet_wrap(~GeneSet, scales = "free_y") +
  scale_fill_manual(values = custom_palette) +
  labs(
    x     = "Bait : Stress",
    y     = "Signature Score (Rank Sum)",
    title = "Mitochondrial vs Nuclear-Localized Gene Scores"
  ) +
  theme_bw() +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x     = element_text(size = 12, angle=20, hjust=1),
    strip.text      = element_text(size = 12, face = "bold"),
    legend.position = "none"
  )

png("~/Projects/StressGranules/SGPAPER/Images/PNG/ZRen_MitoSigs.png", width = 14, height=6, units = "in", res=300)
vilsigplot
dev.off()

vilsigplot
```


### GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
# #for all genes
# DEGenes.SG.vs.VEC <- DEGenes.SG.vs.VEC[order(DEGenes.SG.vs.VEC[,4]),]
# gseatool <- DEGenes.SG.vs.VEC[,4]
# gseatool <- as.data.frame(gseatool)
# colnames(gseatool)[1] <- "t" 
# gseatool[,2] <- DEGenes.SG.vs.VEC$Gene
# colnames(gseatool)[2] <- "Symbol" 
# gseatool <- gseatool[,c(2,1)]
# 
# vectorranks <- gseatool$t
# names(vectorranks) <- gseatool$Symbol
# 
# ### CELL COMPONENT
# 
# #to load our pathways
# hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "C5", subcollection = "GO:CC")
# 
# hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)
# 
# for (set in names(hallmarks.gsets)){
#   hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
# }
# 
# hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
#                        stats = vectorranks)
# 
# hallmarks.res.tidy <- hallmarks.res %>%
#   as_tibble() %>%
#   arrange(desc(NES))
# 
# hallmarks.res.tidy %>% 
#   dplyr::select(-leadingEdge, -ES) %>% 
#   arrange(padj)
# 
# hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
# hallmark.sg.t <- as.data.frame(hallmark.sg.t)
# 
# #this small part is generic
# thresh <- 0.05
# 
# #this one is no longer generic
# data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)
# 
# statvec <- c()
# for (i in 1:nrow(data_to_plot)){
#   if (data_to_plot[i,6] > 0){
#     status <- "Upregulated"
#   }
#  else {
#    status <- "Downregulated"
#  }   
# statvec <- c(statvec, status)
# }
# 
# data_to_plot$Status <- statvec
# 
# ##GOCC
# gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
#   geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
#   coord_flip() +
#   labs(x="Pathway", y="Normalised Enrichment Score (NES)",
#        title= "Cell Component Gene Sets",
#        caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
#   theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")
# 
# 
# gsea_plot_hall+
#   theme(text=element_text(size=25,  family="Arial MS"))
# 
# 
# ### HALLMARK
# 
# #to load our pathways
# hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")
# 
# hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)
# 
# for (set in names(hallmarks.gsets)){
#   hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
# }
# 
# hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
#                        stats = vectorranks)
# 
# hallmarks.res.tidy <- hallmarks.res %>%
#   as_tibble() %>%
#   arrange(desc(NES))
# 
# hallmarks.res.tidy %>% 
#   dplyr::select(-leadingEdge, -ES) %>% 
#   arrange(padj)
# 
# hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
# hallmark.sg.t <- as.data.frame(hallmark.sg.t)
# 
# #this small part is generic
# thresh <- 0.05
# 
# #this one is no longer generic
# data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)
# 
# statvec <- c()
# for (i in 1:nrow(data_to_plot)){
#   if (data_to_plot[i,6] > 0){
#     status <- "Upregulated"
#   }
#  else {
#    status <- "Downregulated"
#  }   
# statvec <- c(statvec, status)
# }
# 
# data_to_plot$Status <- statvec
# 
# ##-----HALMARK-----
# gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
#   geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
#   coord_flip() +
#   labs(x="Pathway", y="Normalised Enrichment Score (NES)",
#        title= "Hallmark Gene Sets",
#        caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
#   theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")
# 
# 
# gsea_plot_hall+
#   theme(text=element_text(size=25,  family="Arial MS"))
```
From the Cell Component Gene Sets, we see a lot of cytoskeletal mRNAs.

### For Mito

```{r}
# --- Rank genes by t-statistic as usual ---
DEGenes.SG.vs.VEC <- DEGenes.SG.vs.VEC[order(DEGenes.SG.vs.VEC[,4]),]
gseatool <- DEGenes.SG.vs.VEC[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes.SG.vs.VEC)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]
vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol

# --- Define MitoGenes as a custom gene set ---
MitoGenes <- fread("~/MitoGenes.txt")[[1]]
custom.gsets <- list(MitoGenes = toupper(MitoGenes))

# --- Run fgsea ---
mito.res <- fgsea(pathways = custom.gsets,
                  stats = vectorranks)

mito.plot.base <- plotEnrichment(custom.gsets[["MitoGenes"]], vectorranks)
curve.data <- mito.plot.base$data

# --- key fix: match the curve's internal descending order ---
ord <- order(vectorranks, decreasing = TRUE)
ranked.names <- names(vectorranks)[ord]
gene.positions <- which(ranked.names %in% custom.gsets[["MitoGenes"]])

rug.top    <- min(curve.data$y) - 0.03
rug.bottom <- rug.top - 0.04

y.min <- rug.bottom - 0.03
y.max <- max(curve.data$y) + 0.05

(mito.plot <- ggplot(curve.data, aes(x = x, y = y)) +
  geom_area(fill = "#73937E", alpha = 0.4) +
  geom_line(color = "#73937E", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_segment(data = data.frame(x = gene.positions),
               aes(x = x, xend = x, y = rug.top, yend = rug.bottom),
               color = "#73937E", alpha = 0.6, linewidth = 0.9,
               inherit.aes = FALSE) +
  annotate("text", x = 1500, y = -0.6,
           label = paste0("NES = ", round(mito.res$NES, 2),
                          "\nadj.p ≈ ", round(mito.res$padj, 4)),
           size = 5, family = "Arial MS", fontface = "bold") +
  coord_cartesian(ylim = c(y.min, y.max), clip = "off") +
  theme_bw() +
  theme(text = element_text(size = 25, family = "Arial MS"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/MitoGSEA_ZRen.png", width=7, height=6, units="in", res=300)
mito.plot
dev.off()
```

# 2026



```{r}
decode_samples <- function(codes) {
  bait   <- ifelse(grepl("G3", codes), "G3BP1", "Untargeted")
  stress <- dplyr::case_when(
    grepl("As", codes) ~ "Stressed",
    grepl("Un", codes) ~ "Unstressed",
    grepl("So", codes) ~ "Stressed",
    grepl("T1", codes) ~ "Stressed",
    grepl("T3", codes) ~ "Unstressed"
  )
  probe <- ifelse(grepl("NoLab", codes), "NoProbe", "Label")
  frac  <- ifelse(grepl("En$", codes), "Enriched", "Input")

  role <- dplyr::case_when(
    frac == "Input"                      ~ "Whole Cell",
    probe == "NoProbe"                   ~ "Background Control",
    bait == "Untargeted"                 ~ "Untargeted Control",
    bait == "G3BP1" ~ "Stress Granule"  
  )

  paste0(role, "_", stress)
}

sample_codes <- rep(
  c("U2G3AsLabEn","U2G3AsLabIn","U2G3AsNoLabEn","U2CtAsLabEn",
    "HekG3AsLabEn","HekG3AsLabIn","HekG3AsNoLabEn","HekCtAsLabEn",
    "HekG3UnLabEn","HekG3UnLabIn","HekG3UnNoLabEn","HekCtUnLabEn",
    "HekG3SoLabEn","HekG3SoLabIn","HekG3SoNoLabEn","HekCtSoLabEn",
    "HekG3T1LabEn","HekG3T1LabIn","HekG3T1NoLabEn","HekCtT1LabEn",
    "HekG3T3LabEn","HekG3T3LabIn","HekG3T3NoLabEn","HekCtT3LabEn"),
  times = c(2,2,2,2, 2,2,2,4, 2,2,2,4, 2,2,2,4, 3,3,3,4, 2,2,2,4)
)

SampleColouring_readable <- decode_samples(sample_codes)

## =====================================================================
## CORE SG-SPECIFIC SIGNAL — HEK293T ONLY, POOLED Arsenite+Sorbitol(+H1)
## v3: Probe (label/noprobe) fit as a covariate instead of manually
##     subtracted — fixes the asymmetric-background bug from v2, where
##     background was only ever removed from the G3BP1 arm (Untargeted
##     has no noprobe control), producing an artificial, near-uniform
##     negative shift on every Bait contrast.
## =====================================================================
## Requires finaltable, SampleColouring, genelength, MitoGenes, NucGenes
## already in the environment (built earlier in the Rmd).
## =====================================================================

library(limma)
library(edgeR)
library(tidyverse)
library(ggrepel)
library(ggpubr)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)
library(rstatix)

set.seed(1906)
out_dir <- "~/Projects/StressGranules/SGPAPER/Images/PNG/"

## ---- 1. Metadata ------------------------------------------------------
CellType   <- c(rep("U2",8), rep("Hek",53))
TargetType <- c(rep("G3",6), rep("Ct",2), rep("G3",6), rep("Ct",4),
                 rep("G3",6), rep("Ct",4), rep("G3",6), rep("Ct",4),
                 rep("G3",9), rep("Ct",4), rep("G3",6), rep("Ct",4))
StressType <- c(rep("Arsenite",18), rep("Unstressed",10), rep("Sorbitol",10),
                 rep("H1",13), rep("H3",10))
LabType <- c(rep("Yes",4),rep("No",2),rep("Yes",2),
             rep("Yes",4),rep("No",2),rep("Yes",4),
             rep("Yes",4),rep("No",2),rep("Yes",4),
             rep("Yes",4),rep("No",2),rep("Yes",4),
             rep("Yes",6),rep("No",3),rep("Yes",4),
             rep("Yes",4),rep("No",2),rep("Yes",4))
SampleType <- c(rep("Enriched",2),rep("Input",2),rep("Enriched",4),
                 rep("Enriched",2),rep("Input",2),rep("Enriched",6),
                 rep("Enriched",2),rep("Input",2),rep("Enriched",6),
                 rep("Enriched",2),rep("Input",2),rep("Enriched",6),
                 rep("Enriched",3),rep("Input",3),rep("Enriched",7),
                 rep("Enriched",2),rep("Input",2),rep("Enriched",6))

metadata <- data.frame(
  CellType = CellType,
  Bait     = factor(ifelse(TargetType == "G3", "G3BP1", "Untargeted"),
                     levels = c("Untargeted", "G3BP1")),
  Stress   = StressType,
  Probe    = factor(ifelse(LabType == "Yes", "label", "noprobe"),
                     levels = c("noprobe", "label")),
  Fraction = SampleType
)
stopifnot(nrow(metadata) == ncol(finaltable))

## ---- 2. Voom — NO manual background subtraction this time --------------
## Probe goes into the model as a covariate instead (see step 4). Keeping
## raw y$E means the same additive Probe term is shared by every sample,
## so it can't create the arm-asymmetric shift the old correction did.
y <- voom(finaltable, model.matrix(~0 + CellType + Bait + Stress + Probe + Fraction,
                                    data = metadata))
logexpr <- y$E

## ---- 3a. Subset for FITTING: label + noprobe, Enriched, HEK, no H3 -----
## (Untargeted still has no noprobe arm — that's fine, Probe is estimated
## off the G3BP1 label/noprobe pairs and applied as a single shared
## coefficient, so it doesn't bias the Bait comparison one way or another.)
keep_fit <- metadata$CellType == "Hek" &
            metadata$Fraction == "Enriched" &
            metadata$Stress != "H3"

meta_fit <- metadata[keep_fit, ]
expr_fit <- logexpr[, keep_fit]

meta_fit$StressGroup <- factor(
  ifelse(meta_fit$Stress == "Unstressed", "Basal", "Stressed"),
  levels = c("Basal", "Stressed")
)
meta_fit$Stress3 <- factor(
  dplyr::case_when(
    meta_fit$Stress == "Unstressed"          ~ "Unstressed",
    meta_fit$Stress %in% c("Arsenite", "H1") ~ "ArsActive",
    meta_fit$Stress == "Sorbitol"            ~ "Sorbitol"
  ),
  levels = c("Unstressed", "ArsActive", "Sorbitol")
)
meta_fit$Bait  <- factor(meta_fit$Bait,  levels = c("Untargeted", "G3BP1"))
meta_fit$Probe <- factor(meta_fit$Probe, levels = c("noprobe", "label"))

table(meta_fit$Bait, meta_fit$Probe)   # sanity check: noprobe only within G3BP1

## ---- 3b. Subset for DESCRIPTIVE plots: label only, same as before ------
## (violin scores, expression QC bars — we don't want noprobe "background
## floor" samples showing up as if they were biological expression values)
keep_desc <- keep_fit & metadata$Probe == "label"
meta_main <- metadata[keep_desc, ]
expr_main <- logexpr[, keep_desc]
meta_main$StressGroup <- factor(ifelse(meta_main$Stress == "Unstressed", "Basal", "Stressed"),
                                 levels = c("Basal", "Stressed"))
meta_main$Bait <- factor(meta_main$Bait, levels = c("Untargeted", "G3BP1"))

## ---- 4a. POOLED model — Bait * StressGroup + Probe, CENTERED coding ----
## Coding note: for a saturated design (intercept + both main effects +
## their interaction, i.e. `~Bait*StressGroup`), the interaction
## coefficient is mathematically identical whether the main effects are
## 0/1 dummy-coded or mean-centered — centering only reallocates variance
## between the intercept and the main-effect terms, it can't change the
## interaction. So this does NOT move DEGenes.SG.vs.VEC / the volcano /
## MA plots / GSEA at all. What it DOES change: BaitG3BP1 and
## StressGroupStressed now represent the *average* effect across the
## other factor's levels rather than the effect at its reference level —
## which is exactly what feeds the Interaction-vs-Bait/Stress component
## plots in section 5.
Bait_c   <- ifelse(meta_fit$Bait == "G3BP1", 1, 0)
Bait_c   <- Bait_c - mean(Bait_c)
Stress_c <- ifelse(meta_fit$StressGroup == "Stressed", 1, 0)
Stress_c <- Stress_c - mean(Stress_c)
Probe_c  <- ifelse(meta_fit$Probe == "label", 1, 0)
Probe_c  <- Probe_c - mean(Probe_c)

mm_pooled <- cbind(
  "(Intercept)"                   = 1,
  "BaitG3BP1"                     = Bait_c,
  "StressGroupStressed"           = Stress_c,
  "Probelabel"                    = Probe_c,
  "BaitG3BP1:StressGroupStressed" = Bait_c * Stress_c
)

fitted <- lmFit(expr_fit, mm_pooled)
head(coef(fitted))

tmp.SG.vs.VEC <- eBayes(fitted)

top.table.SG.vs.VEC <- topTable(tmp.SG.vs.VEC, coef = "BaitG3BP1:StressGroupStressed",
                                 sort.by = "logFC", n = Inf)
DEGenes.SG.vs.VEC <- top.table.SG.vs.VEC
ndegs <- 20

## Quick check the fix worked: Bait component should no longer be
## almost-uniformly negative.
top_bait_check <- topTable(tmp.SG.vs.VEC, coef = "BaitG3BP1", sort.by = "none", n = Inf)
summary(top_bait_check$t)
hist(top_bait_check$t, breaks = 60, main = "BaitG3BP1 t-stat distribution (should be ~centered, not all-negative)")

## ---- 4b. CONCORDANCE model — Arsenite vs Sorbitol components ------------
## Same centered-coding logic as above; the two interaction coefficients
## are unaffected since Bait*Stress3 is also saturated (all 6 Bait x
## Stress3 cells are populated in this design).
Ars_c <- ifelse(meta_fit$Stress3 == "ArsActive", 1, 0)
Ars_c <- Ars_c - mean(Ars_c)
Sor_c <- ifelse(meta_fit$Stress3 == "Sorbitol", 1, 0)
Sor_c <- Sor_c - mean(Sor_c)

mm_concord <- cbind(
  "(Intercept)"                = 1,
  "BaitG3BP1"                  = Bait_c,
  "Stress3ArsActive"           = Ars_c,
  "Stress3Sorbitol"            = Sor_c,
  "Probelabel"                 = Probe_c,
  "BaitG3BP1:Stress3ArsActive" = Bait_c * Ars_c,
  "BaitG3BP1:Stress3Sorbitol"  = Bait_c * Sor_c
)
fit_concord <- eBayes(lmFit(expr_fit, mm_concord))
top_ars <- topTable(fit_concord, coef = "BaitG3BP1:Stress3ArsActive", sort.by = "none", n = Inf)
top_sor <- topTable(fit_concord, coef = "BaitG3BP1:Stress3Sorbitol",  sort.by = "none", n = Inf)

concord_df <- data.frame(
  Gene        = rownames(DEGenes.SG.vs.VEC),
  pooled_t    = DEGenes.SG.vs.VEC$t,
  pooled_adjP = DEGenes.SG.vs.VEC$adj.P.Val,
  ars_t       = top_ars[rownames(DEGenes.SG.vs.VEC), "t"],
  ars_p       = top_ars[rownames(DEGenes.SG.vs.VEC), "P.Value"],
  sor_t       = top_sor[rownames(DEGenes.SG.vs.VEC), "t"],
  sor_p       = top_sor[rownames(DEGenes.SG.vs.VEC), "P.Value"]
)
concord_df$Core <- with(concord_df,
  pooled_adjP < 0.05 & sign(ars_t) == sign(sor_t) & sign(ars_t) == sign(pooled_t) &
  ars_p < 0.05 & sor_p < 0.05
)
cat(sum(concord_df$Core), "genes pass the pooled+concordance 'core' filter\n")

ggplot(concord_df, aes(x = ars_t, y = sor_t, color = Core)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("FALSE" = "#CCCCCC", "TRUE" = "#de425b")) +
  theme_bw() + xlab("Arsenite-active component (t-stat)") + ylab("Sorbitol component (t-stat)") +
  ggtitle("Concordance check: pooled 'core' hits vs per-stressor components") +
  theme(plot.title = element_text(hjust = 0.5), axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(), panel.border = element_blank(),
        panel.background = element_blank()) +
  stat_cor(method = "spearman", cor.coef.name = "rho")

## =========================================================================
## ---- 5. Interaction vs Bait / Interaction vs Stress components ---------
## =========================================================================
top_interact <- topTable(tmp.SG.vs.VEC, coef = "BaitG3BP1:StressGroupStressed", sort.by = "logFC", n = Inf)
top_bait     <- topTable(tmp.SG.vs.VEC, coef = "BaitG3BP1",                     sort.by = "logFC", n = Inf)
top_stress   <- topTable(tmp.SG.vs.VEC, coef = "StressGroupStressed",          sort.by = "logFC", n = Inf)

interact_t <- data.frame(Gene = row.names(top_interact), interact_t = top_interact$t)
bait_t     <- data.frame(Gene = row.names(top_bait),     bait_t     = top_bait$t)
stress_t   <- data.frame(Gene = row.names(top_stress),   stress_t   = top_stress$t)

MitoGenes <- fread("~/MitoGenes.txt")[[1]]
NucGenes  <- fread("~/rnalocate_mito_filt.csv")[[1]]

## ---- Interaction vs Bait (G3BP1) component ----
comp1_df <- merge(interact_t, bait_t, by = "Gene")
row.names(comp1_df) <- comp1_df$Gene
MitoData_c1 <- comp1_df[comp1_df$Gene %in% MitoGenes, ]
NucData_c1  <- comp1_df[comp1_df$Gene %in% NucGenes, ]

spearman_c1 <- cor(comp1_df$interact_t, comp1_df$bait_t, method = "spearman")

(p_c1 <- ggplot(comp1_df, aes(x = bait_t, y = interact_t)) +
  geom_point(color = "#CCCCCC") +
  geom_point(data = MitoData_c1, aes(x = bait_t, y = interact_t), color = "#73937E", size = 4) +
  geom_point(data = NucData_c1, aes(x = bait_t, y = interact_t), color = "#CB643F", size = 4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  theme_bw() + ylab("Interaction (t-stat)") + xlab("Bait Component (t-stat)") +
  theme(axis.line = element_line(colour = "black"), panel.grid.minor = element_blank(),
        panel.border = element_blank(), panel.background = element_blank()) +
  ggtitle("Interaction vs Bait Component — Core Model") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text = element_text(size = 15, family = "Arial MS")) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 5))

png(paste0(out_dir, "CoreSG_InteractionVsBait.png"), width = 7, height = 6, units = "in", res = 300)
p_c1
dev.off()

## ---- Interaction vs Stress component ----
comp2_df <- merge(interact_t, stress_t, by = "Gene")
row.names(comp2_df) <- comp2_df$Gene
MitoData_c2 <- comp2_df[comp2_df$Gene %in% MitoGenes, ]
NucData_c2  <- comp2_df[comp2_df$Gene %in% NucGenes, ]

spearman_c2 <- cor(comp2_df$interact_t, comp2_df$stress_t, method = "spearman")

(p_c2 <- ggplot(comp2_df, aes(x = stress_t, y = interact_t)) +
  geom_point(color = "#CCCCCC") +
  geom_point(data = MitoData_c2, aes(x = stress_t, y = interact_t), color = "#73937E", size = 4) +
  geom_point(data = NucData_c2, aes(x = stress_t, y = interact_t), color = "#CB643F", size = 4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  theme_bw() + ylab("Interaction (t-stat)") + xlab("Stress Component (t-stat)") +
  theme(axis.line = element_line(colour = "black"), panel.grid.minor = element_blank(),
        panel.border = element_blank(), panel.background = element_blank()) +
  ggtitle("Interaction vs Stress Component — Core Model") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text = element_text(size = 15, family = "Arial MS")) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 5))

png(paste0(out_dir, "CoreSG_InteractionVsStress.png"), width = 7, height = 6, units = "in", res = 300)
p_c2
dev.off()

## =========================================================================
## ---- 6. Volcano plot (exploratory + publication Mito/Nuc panel) --------
## =========================================================================
write.table(x = DEGenes.SG.vs.VEC,
            file = "~/Projects/StressGranules/SGPAPER/CoreSG_HEK_ArsSorb.txt",
            sep = "\t", col.names = TRUE, row.names = TRUE, quote = FALSE)

upmarkers   <- as.data.frame(DEGenes.SG.vs.VEC[DEGenes.SG.vs.VEC$logFC > 4  & DEGenes.SG.vs.VEC$adj.P.Val < 0.05, ])
downmarkers <- as.data.frame(DEGenes.SG.vs.VEC[DEGenes.SG.vs.VEC$logFC < -4 & DEGenes.SG.vs.VEC$adj.P.Val < 0.05, ])

ggplot(data = DEGenes.SG.vs.VEC, aes(x = logFC, y = B)) +
  geom_point(color = "#DDDDDD") +
  theme_bw() +
  theme(axis.line = element_line(colour = "black"), panel.grid.minor = element_blank(),
        panel.border = element_blank(), panel.background = element_blank()) +
  ggtitle("Core SG Signal (HEK293T, Arsenite+Sorbitol pooled)") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs, ], mapping = aes(label = row.names(upmarkers)[1:ndegs]),
                    size = 4, point.padding = -0.5, color = "#de425b", max.overlaps = 200, fontface = "italic") +
  geom_label_repel(downmarkers[1:ndegs, ], mapping = aes(label = row.names(downmarkers)[1:ndegs]),
                    size = 4, point.padding = -0.5, color = "#769fca", max.overlaps = 200, fontface = "italic") +
  geom_point(data = upmarkers, aes(x = logFC, y = B), color = "#de425b", size = 3) +
  geom_point(data = downmarkers, aes(x = logFC, y = B), color = "#769fca", size = 3) +
  xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(text = element_text(family = "Arial", size = 20))

DEGenes  <- DEGenes.SG.vs.VEC
MitoData <- DEGenes[rownames(DEGenes) %in% MitoGenes, ]
NucData  <- DEGenes[rownames(DEGenes) %in% NucGenes, ]

(m <- ggplot(data = DEGenes, aes(x = logFC, y = B)) +
  geom_point(color = "#CCCCCC") +
  geom_point(data = NucData, aes(x = logFC, y = B), color = "#CB643F", size = 4) +
  geom_point(data = MitoData, aes(x = logFC, y = B), color = "#73937E", size = 4) +
  ylab("B-Statistic") + xlab(expression(Log[2] ~ "Fold-Change")) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text = element_text(size = 25, family = "Arial MS")))

png(paste0(out_dir, "CoreSGMito_Volcano.png"), width = 7, height = 6, units = "in", res = 300)
m
dev.off()

## =========================================================================
## ---- 7. Three MA plots (AveExpr / raw length / weighted length "S") ----
## =========================================================================

## -- MA #1: AveExpr vs logFC, exploratory + publication versions ----------
ggplot(data = DEGenes, aes(x = AveExpr, y = logFC)) +
  geom_point(color = "#CCCCCC") +
  geom_point(data = MitoData, aes(x = AveExpr, y = logFC), color = "#73937E", size = 4) +
  geom_point(data = NucData, aes(x = AveExpr, y = logFC), color = "#CB643F", size = 4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  theme_bw() + ylab(expression(Log[2] ~ "Fold-Change")) + xlab("Average Expression (Log2 CPM)") +
  theme(axis.line = element_line(colour = "black"), panel.grid.minor = element_blank(),
        panel.border = element_blank(), panel.background = element_blank()) +
  ggtitle("Core SG Signal - MA Plot") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs, ], mapping = aes(x = AveExpr, y = logFC, label = row.names(upmarkers)[1:ndegs]),
                    size = 4, point.padding = -0.5, color = "#A10702", max.overlaps = 20, fontface = "italic") +
  geom_label_repel(downmarkers[1:ndegs, ], mapping = aes(x = AveExpr, y = logFC, label = row.names(downmarkers)[1:ndegs]),
                    size = 4, point.padding = -0.5, color = "#03256C", max.overlaps = 20, fontface = "italic") +
  geom_point(data = upmarkers, aes(x = AveExpr, y = logFC), color = "#A10702", size = 4) +
  geom_point(data = downmarkers, aes(x = AveExpr, y = logFC), color = "#03256C", size = 4) +
  theme(text = element_text(size = 15, family = "Arial MS"))

(ma1 <- ggplot(data = DEGenes, aes(x = AveExpr, y = logFC)) +
  geom_point(color = "#CCCCCC") +
  geom_point(data = MitoData, aes(x = AveExpr, y = logFC), color = "#73937E", size = 4) +
  geom_point(data = NucData, aes(x = AveExpr, y = logFC), color = "#CB643F", size = 4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  ylab(expression(Log[2] ~ "Fold-Change")) + xlab("Average Expression (Log2 CPM)") +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 6,
           color = "#800020", label.y.npc = 0.98) + geom_density2d(color = "#800020", size = 1.2)+
  stat_cor(data = MitoData, aes(x = AveExpr, y = logFC), method = "spearman",
           cor.coef.name = "rho", size = 6, color = "#73937E", label.y.npc = 0.90) +
  stat_cor(data = NucData, aes(x = AveExpr, y = logFC), method = "spearman",
           cor.coef.name = "rho", size = 6, color = "#CB643F", label.y.npc = 0.82) +
  theme(axis.line = element_line(colour = "black"), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), panel.border = element_blank(),
        panel.background = element_blank(), axis.title = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text = element_text(size = 25, family = "Arial MS")))

png(paste0(out_dir, "CoreSGMito_MA_AveExpr.png"), width = 7, height = 6, units = "in", res = 300)
ma1
dev.off()

## -- MA #2 / #3 need gene length + the Mito-weighted "S" score ------------
genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header = T)
genelength <- subset(genelength, Ensembl.Canonical == 1)
genelength <- subset(genelength, Gene.name != "")
genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()
genelength <- genelength[, -c(2, 3)]
colnames(genelength) <- c("Length", "Gene")
genelength$Cromo <- ifelse(genelength$Gene %in% MitoGenes, "Mito", "Other")
genelength$S <- ifelse(
  genelength$Cromo == "Mito",
  (genelength$Length)^(1/3) * 3.23e-11 + 1.17e-9,
  (genelength$Length)^(1/3) * 6.09e-11 + 3.88e-10
)

DEGenes$Gene <- row.names(DEGenes)
DEGenes <- merge(DEGenes, genelength, by = "Gene")
MitoData <- DEGenes[DEGenes$Gene %in% MitoGenes, ]
NucData  <- DEGenes[DEGenes$Gene %in% NucGenes, ]

## -- MA #2: raw length -----------------------------------------------------
(ma2 <- ggplot(data = DEGenes, aes(x = log10(Length), y = logFC)) +
  geom_point(color = "#CCCCCC") +
  geom_point(data = MitoData, aes(x = log10(Length), y = logFC), color = "#73937E", size = 4) +
  geom_point(data = NucData, aes(x = log10(Length), y = logFC), color = "#CB643F", size = 4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  ylab(expression(Log[2] ~ "Fold-Change")) + xlab("RNA Length (log10)") + geom_density2d(color = "#800020", size = 1.2)+
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 6,
           color = "#800020", label.y.npc = 0.98) +
  stat_cor(data = MitoData, aes(x = log10(Length), y = logFC), method = "spearman",
           cor.coef.name = "rho", size = 6, color = "#73937E", label.y.npc = 0.90) +
  stat_cor(data = NucData, aes(x = log10(Length), y = logFC), method = "spearman",
           cor.coef.name = "rho", size = 6, color = "#CB643F", label.y.npc = 0.82) +
  theme(axis.line = element_line(colour = "black"), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), panel.border = element_blank(),
        panel.background = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text = element_text(size = 25, family = "Arial MS")))

png(paste0(out_dir, "CoreSGMito_MA_length.png"), width = 7, height = 6, units = "in", res = 300)
ma2
dev.off()

## -- MA #3: weighted-length "S" ---------------------------------------------
(ma3 <- ggplot(data = DEGenes, aes(x = S, y = logFC)) +
  geom_point(color = "#CCCCCC") +
  geom_point(data = MitoData, aes(x = S, y = logFC), color = "#73937E", size = 4) +
  geom_point(data = NucData, aes(x = S, y = logFC), color = "#CB643F", size = 4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  ylab(expression(Log[2] ~ "Fold-Change")) + xlab("S") +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 6,
           color = "#800020", label.y.npc = 0.98) + geom_density2d(color = "#800020", size = 1.2)+
  stat_cor(data = MitoData, aes(x = S, y = logFC), method = "spearman",
           cor.coef.name = "rho", size = 6, color = "#73937E", label.y.npc = 0.90) +
  stat_cor(data = NucData, aes(x = S, y = logFC), method = "spearman",
           cor.coef.name = "rho", size = 6, color = "#CB643F", label.y.npc = 0.82) +
  theme(axis.line = element_line(colour = "black"), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), panel.border = element_blank(),
        panel.background = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text = element_text(size = 25, family = "Arial MS")))

png(paste0(out_dir, "CoreSGMito_MA_S.png"), width = 7, height = 6, units = "in", res = 300)
ma3
dev.off()

## =========================================================================
## ---- 8. Volcano with gene length + standardized residuals --------------
## =========================================================================
DEGenes.SG.vs.VEC$Gene <- row.names(DEGenes.SG.vs.VEC)
colnames(genelength)[1:2] <- c("Length", "Gene")
DEGenes.SG.vs.VEC <- merge(DEGenes.SG.vs.VEC, genelength[, c("Length", "Gene")], by = "Gene")

ggplot(data = DEGenes.SG.vs.VEC, aes(x = logFC, y = B)) +
  geom_point(aes(color = log10(Length))) +
  theme_bw() +
  theme(axis.line = element_line(colour = "black"), panel.grid.minor = element_blank(),
        panel.border = element_blank(), panel.background = element_blank()) +
  ggtitle("Core SG Signal") + xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_colour_gradient(low = "black", high = "red") + theme(text = element_text(family = "Quicksand"))

smooth.SG <- lm(t ~ log10(Length), DEGenes.SG.vs.VEC)
liner.SG <- rstandard(smooth.SG)
DEGenes.SG.vs.VEC <- cbind(DEGenes.SG.vs.VEC, liner.SG)
colnames(DEGenes.SG.vs.VEC)[ncol(DEGenes.SG.vs.VEC)] <- "Residuals"
row.names(DEGenes.SG.vs.VEC) <- DEGenes.SG.vs.VEC$Gene
linemarkers <- as.data.frame(DEGenes.SG.vs.VEC[abs(DEGenes.SG.vs.VEC$Residuals) > 4, ])

spearman_corr <- cor(DEGenes.SG.vs.VEC$t, log10(DEGenes.SG.vs.VEC$Length), method = "pearson")

ggplot(DEGenes.SG.vs.VEC, aes(y = t, x = log10(Length))) +
  geom_point(color = "#CCCCCC", alpha = 1) + theme_bw() + geom_density_2d() +
  ylab("Core SG Signal (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"), panel.grid.minor = element_blank(),
        panel.border = element_blank(), panel.background = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope = smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed") +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 7)

## =========================================================================
## ---- 9. Violin plot — Mito vs Nuc rank-sum scores, exact ANOVA style ----
## =========================================================================
expression.df <- expr_main
bait_grp   <- meta_main$Bait
stress_grp <- meta_main$StressGroup
Condition  <- interaction(bait_grp, stress_grp, sep = ".")

custom_palette <- c(
  "Untargeted.Basal"    = "#a2d1e6",
  "Untargeted.Stressed" = "#769fca",
  "G3BP1.Basal"         = "#e88184",
  "G3BP1.Stressed"      = "#de425b"
)

mito_genes_present <- MitoGenes[MitoGenes %in% rownames(expression.df)]
nuc_genes_present  <- NucGenes[NucGenes %in% rownames(expression.df)]
cat(length(mito_genes_present), "Mito genes found in expression matrix\n")
cat(length(nuc_genes_present), "Nuc genes found in expression matrix\n")

rank_matrix <- apply(expression.df, 2, function(x) rank(x, ties.method = "average"))
rank_matrix <- as.data.frame(rank_matrix)

compute_score <- function(genes, rank_matrix, n_total) {
  colSums(rank_matrix[genes, , drop = FALSE]) / n_total
}
mito_score <- compute_score(mito_genes_present, rank_matrix, nrow(expression.df))
nuc_score  <- compute_score(nuc_genes_present, rank_matrix, nrow(expression.df))

score_df <- rbind(
  data.frame(Sample = colnames(expression.df), Score = as.numeric(mito_score),
             BaitGrp = bait_grp, StressGrp = stress_grp, Condition = Condition, GeneSet = "Mito"),
  data.frame(Sample = colnames(expression.df), Score = as.numeric(nuc_score),
             BaitGrp = bait_grp, StressGrp = stress_grp, Condition = Condition, GeneSet = "Nuc")
)

stat_labels <- score_df %>%
  group_by(GeneSet) %>%
  group_modify(~{
    aov_res  <- aov(Score ~ BaitGrp * StressGrp, data = .x)
    aov_tidy <- summary(aov_res)
    p_raw  <- aov_tidy[[1]][["Pr(>F)"]][3]
    p_adj  <- p.adjust(p_raw, method = "BH")
    ss_interact <- aov_tidy[[1]][["Sum Sq"]][3]
    ss_total    <- sum(aov_tidy[[1]][["Sum Sq"]])
    eta_sq      <- ss_interact / ss_total
    cohens_f    <- sqrt(eta_sq / (1 - eta_sq))
    data.frame(label = paste0("Interaction Cohen's f = ", round(cohens_f, 3),
                               "\nadj. p = ", formatC(p_adj, format = "e", digits = 2)))
  })

vilsigplot <- ggplot(score_df, aes(x = Condition, y = Score, fill = Condition)) +
  geom_violin(colour = "black", alpha = 0.6, trim = FALSE) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.7, size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, colour = "black") +
  stat_summary(fun = median, geom = "point", shape = 17, size = 3, colour = "black") +
  geom_text(data = stat_labels, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.05, vjust = 1.2, size = 4, fontface = "bold", lineheight = 1.1,
            inherit.aes = FALSE) +
  facet_wrap(~GeneSet, scales = "free_y") +
  scale_fill_manual(values = custom_palette) +
  labs(x = "Bait : Stress", y = "Signature Score (Rank Sum)",
       title = "Mitochondrial vs Nuclear-Localized Gene Scores — Core Model") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        axis.text.x = element_text(size = 12, angle = 20, hjust = 1),
        strip.text = element_text(size = 12, face = "bold"),
        legend.position = "none")

png(paste0(out_dir, "CoreSG_MitoSigs.png"), width = 14, height = 6, units = "in", res = 300)
vilsigplot
dev.off()
vilsigplot

## =========================================================================
## ---- 10. GSEA (Hallmark, GO:CC) + Mito enrichment curve -----------------
## =========================================================================
DEGenes.SG.vs.VEC <- DEGenes.SG.vs.VEC[order(DEGenes.SG.vs.VEC$t), ]
vectorranks <- setNames(DEGenes.SG.vs.VEC$t, DEGenes.SG.vs.VEC$Gene)

## -- GO:CC --
gocc.gs <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:CC")
gocc.gsets <- lapply(split(gocc.gs$gene_symbol, gocc.gs$gs_name), toupper)
gocc.res <- fgsea(pathways = gocc.gsets, stats = vectorranks)
gocc.tidy <- as_tibble(gocc.res) %>% arrange(desc(NES))
thresh <- 0.05
data_to_plot <- gocc.tidy %>% filter(padj < thresh) %>%
  mutate(Status = ifelse(NES > 0, "Upregulated", "Downregulated"))

ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = Status), alpha = 0.75) +
  scale_fill_manual(values = c("#53A2BE", "#BA3B46"), name = "Regulation Status",
                     labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x = "Pathway", y = "Normalised Enrichment Score (NES)",
       title = "Cell Component Gene Sets — Core Model", caption = paste0("padj < ", thresh)) +
  theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text = element_text(family = "Quicksand")) + theme(legend.position = "none") +
  theme(text = element_text(size = 25, family = "Arial MS"))

## -- Hallmark --
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")
hallmarks.gsets <- lapply(split(hallmarks.gs$gene_symbol, hallmarks.gs$gs_name), toupper)
hallmarks.res <- fgsea(pathways = hallmarks.gsets, stats = vectorranks)
hallmarks.res.tidy <- as_tibble(hallmarks.res) %>% arrange(desc(NES))
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh) %>%
  mutate(Status = ifelse(NES > 0, "Upregulated", "Downregulated"))

ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = Status), alpha = 0.75) +
  scale_fill_manual(values = c("#53A2BE", "#BA3B46"), name = "Regulation Status",
                     labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x = "Pathway", y = "Normalised Enrichment Score (NES)",
       title = "Hallmark Gene Sets — Core Model", caption = paste0("padj < ", thresh)) +
  theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text = element_text(family = "Quicksand")) + theme(legend.position = "none") +
  theme(text = element_text(size = 25, family = "Arial MS"))

## -- Mito enrichment curve --
custom.gsets <- list(MitoGenes = toupper(MitoGenes))
mito.res <- fgsea(pathways = custom.gsets, stats = vectorranks)
mito.plot.base <- plotEnrichment(custom.gsets[["MitoGenes"]], vectorranks)
curve.data <- mito.plot.base$data

ord <- order(vectorranks, decreasing = TRUE)
ranked.names <- names(vectorranks)[ord]
gene.positions <- which(ranked.names %in% custom.gsets[["MitoGenes"]])
rug.top    <- min(curve.data$y) - 0.03
rug.bottom <- rug.top - 0.04
y.min <- rug.bottom - 0.03
y.max <- max(curve.data$y) + 0.05

(mito.plot <- ggplot(curve.data, aes(x = x, y = y)) +
  geom_area(fill = "#73937E", alpha = 0.4) +
  geom_line(color = "#73937E", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_segment(data = data.frame(x = gene.positions),
               aes(x = x, xend = x, y = rug.top, yend = rug.bottom),
               color = "#73937E", alpha = 0.6, linewidth = 0.9, inherit.aes = FALSE) +
  annotate("text", x = 1500, y = -0.6,
           label = paste0("NES = ", round(mito.res$NES, 2), "\nadj.p \u2248 ", round(mito.res$padj, 4)),
           size = 5, family = "Arial MS", fontface = "bold") +
  coord_cartesian(ylim = c(y.min, y.max), clip = "off") +
  theme_bw() +
  theme(text = element_text(size = 25, family = "Arial MS"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_blank(), axis.text = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major = element_blank()))

png(paste0(out_dir, "CoreSG_MitoGSEA.png"), width = 7, height = 6, units = "in", res = 300)
mito.plot
dev.off()

## -- Nuc enrichment curve (same ranking, same style, orange) --------------
custom.gsets.nuc <- list(NucGenes = toupper(NucGenes))
nuc.res <- fgsea(pathways = custom.gsets.nuc, stats = vectorranks)
nuc.plot.base <- plotEnrichment(custom.gsets.nuc[["NucGenes"]], vectorranks)
curve.data.nuc <- nuc.plot.base$data

gene.positions.nuc <- which(ranked.names %in% custom.gsets.nuc[["NucGenes"]])
rug.top.nuc    <- min(curve.data.nuc$y) - 0.03
rug.bottom.nuc <- rug.top.nuc - 0.04
y.min.nuc <- rug.bottom.nuc - 0.03
y.max.nuc <- max(curve.data.nuc$y) + 0.05

(nuc.plot <- ggplot(curve.data.nuc, aes(x = x, y = y)) +
  geom_area(fill = "#CB643F", alpha = 0.4) +
  geom_line(color = "#CB643F", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_segment(data = data.frame(x = gene.positions.nuc),
               aes(x = x, xend = x, y = rug.top.nuc, yend = rug.bottom.nuc),
               color = "#CB643F", alpha = 0.6, linewidth = 0.9, inherit.aes = FALSE) +
  annotate("text", x = 2000, y = -0.2,
           label = paste0("NES = ", round(nuc.res$NES, 2), "\nadj.p = ", round(nuc.res$padj, 4)),
           size = 5, family = "Arial MS", fontface = "bold") +
  coord_cartesian(ylim = c(y.min.nuc, y.max.nuc), clip = "off") +
  theme_bw() +
  theme(text = element_text(size = 25, family = "Arial MS"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_blank(), axis.text = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major = element_blank()))

png(paste0(out_dir, "CoreSG_NucGSEA.png"), width = 7, height = 6, units = "in", res = 300)
nuc.plot
dev.off()
```

#Trial

```{r}
## ---- Metadata: HEK293T only, no batch correction ------------------------
CellType   <- c(rep("U2",8), rep("Hek",53))
TargetType <- c(rep("G3",6), rep("Ct",2), rep("G3",6), rep("Ct",4),
                 rep("G3",6), rep("Ct",4), rep("G3",6), rep("Ct",4),
                 rep("G3",9), rep("Ct",4), rep("G3",6), rep("Ct",4))
StressType <- c(rep("Arsenite",18), rep("Unstressed",10), rep("Sorbitol",10),
                 rep("H1",13), rep("H3",10))
LabType <- c(rep("Yes",4),rep("No",2),rep("Yes",2),
             rep("Yes",4),rep("No",2),rep("Yes",4),
             rep("Yes",4),rep("No",2),rep("Yes",4),
             rep("Yes",4),rep("No",2),rep("Yes",4),
             rep("Yes",6),rep("No",3),rep("Yes",4),
             rep("Yes",4),rep("No",2),rep("Yes",4))
SampleType <- c(rep("Enriched",2),rep("Input",2),rep("Enriched",4),
                 rep("Enriched",2),rep("Input",2),rep("Enriched",6),
                 rep("Enriched",2),rep("Input",2),rep("Enriched",6),
                 rep("Enriched",2),rep("Input",2),rep("Enriched",6),
                 rep("Enriched",3),rep("Input",3),rep("Enriched",7),
                 rep("Enriched",2),rep("Input",2),rep("Enriched",6))

## Same 61-slot code vector decode_samples() expects
sample_codes_full <- rep(
  c("U2G3AsLabEn","U2G3AsLabIn","U2G3AsNoLabEn","U2CtAsLabEn",
    "HekG3AsLabEn","HekG3AsLabIn","HekG3AsNoLabEn","HekCtAsLabEn",
    "HekG3UnLabEn","HekG3UnLabIn","HekG3UnNoLabEn","HekCtUnLabEn",
    "HekG3SoLabEn","HekG3SoLabIn","HekG3SoNoLabEn","HekCtSoLabEn",
    "HekG3T1LabEn","HekG3T1LabIn","HekG3T1NoLabEn","HekCtT1LabEn",
    "HekG3T3LabEn","HekG3T3LabIn","HekG3T3NoLabEn","HekCtT3LabEn"),
  times = c(2,2,2,2, 2,2,2,4, 2,2,2,4, 2,2,2,4, 3,3,3,4, 2,2,2,4)
)

metadata <- data.frame(
  CellType = CellType,
  Bait     = factor(ifelse(TargetType == "G3", "G3BP1", "Untargeted"), levels = c("Untargeted", "G3BP1")),
  Stress   = StressType,
  Probe    = factor(ifelse(LabType == "Yes", "label", "noprobe"), levels = c("noprobe", "label")),
  Fraction = SampleType
)
stopifnot(nrow(metadata) == ncol(finaltable))

metadata$StressGroup <- factor(
  ifelse(metadata$Stress %in% c("Arsenite", "Sorbitol", "H1"), "Stressed", "Unstressed"),
  levels = c("Unstressed", "Stressed")
)

## ---- Keep: HEK only, Enriched + Label only, NO batch correction --------
keep       <- metadata$CellType == "Hek" & metadata$Fraction == "Enriched" & metadata$Probe == "label"
meta_hek   <- droplevels(metadata[keep, ])
counts_hek <- finaltable[, keep]
codes_hek  <- sample_codes_full[keep]

## ---- voom, straight up, no correction step ------------------------------
D0 <- DGEList(counts_hek)
D0 <- calcNormFactors(D0)
mm_simple <- model.matrix(~ Bait * StressGroup, data = meta_hek)
y   <- voom(D0, mm_simple, plot = FALSE)
fit <- eBayes(lmFit(y, mm_simple))

DEGenes.SG.HekOnly <- topTable(fit, coef = "BaitG3BP1:StressGroupStressed",
                                sort.by = "logFC", n = Inf)

upmarkers   <- DEGenes.SG.HekOnly[DEGenes.SG.HekOnly$logFC > 2  & DEGenes.SG.HekOnly$adj.P.Val < 0.05, ]
downmarkers <- DEGenes.SG.HekOnly[DEGenes.SG.HekOnly$logFC < -2 & DEGenes.SG.HekOnly$adj.P.Val < 0.05, ]
cat(nrow(upmarkers), "up /", nrow(downmarkers), "down at logFC>2, adj.P<0.05\n")

ggplot(DEGenes.SG.HekOnly, aes(x = logFC, y = B)) +
  geom_point(color = "#DDDDDD") +
  theme_bw() +
  ggtitle("SG signal — HEK293T only, no batch correction, no balancing") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  geom_point(data = upmarkers, color = "#de425b", size = 2) +
  geom_point(data = downmarkers, color = "#769fca", size = 2) +
  xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic")

## =========================================================================
## ---- PCA, full readable labels, same style as your original PCA block --
## =========================================================================
readable_labels <- decode_samples(codes_hek)   # your function from before

custom_palette <- qualitative_hcl(length(unique(readable_labels)), palette = "Pastel")
custom_palette <- sample(custom_palette, length(unique(readable_labels)))

dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca)
dfforpca <- apply(dfforpca, 2, as.numeric)
dfforpca <- as.data.frame(dfforpca)

dfforpca$Bait   <- meta_hek$Bait
dfforpca$Stress <- meta_hek$StressGroup
dfforpca$All    <- readable_labels

pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca) - 3)], scale = FALSE)

p_All <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "All", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Full Label") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Bait <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Bait", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Bait", values = custom_palette) +
  ggtitle("PCA - Coloured by Bait") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Stress <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Stress", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Stress", values = custom_palette) +
  ggtitle("PCA - Coloured by Stress") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_All
p_Bait
p_Stress

## Top PC1/PC2 contributors, same as your original diagnostic block
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]
(pos1genes <- sort(contribs1, decreasing = TRUE)[1:10])
(neg1genes <- sort(contribs1, decreasing = FALSE)[1:10])
(pos2genes <- sort(contribs2, decreasing = TRUE)[1:10])
(neg2genes <- sort(contribs2, decreasing = FALSE)[1:10])

dfforpca$StressFull <- meta_hek$Stress   # original 5-level factor, not collapsed into Stressed/Unstressed

custom_palette_stress <- qualitative_hcl(length(unique(dfforpca$StressFull)), palette = "Pastel")
custom_palette_stress <- sample(custom_palette_stress, length(unique(dfforpca$StressFull)))

p_StressFull <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StressFull", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Stress", values = custom_palette_stress) +
  ggtitle("PCA - Coloured by Stress (undecomposed)") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_StressFull
```


# Session Info

```{r}
sessionInfo()
```