---
title: "Analyses of GSE245209"
subtitle: "Mapping RNA-protein interactions with subcellular resolution using colocalization CLIP."
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

In this document, I present all analyses pertaining to the datasets **GSE245209**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE245209)),  
originally published by **Y. Soon, 2023**.

## Methodology


```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)

set.seed(1906)
```

# CLIP-Seq Samples

## First Steps

```{r}
#This section here is to load up the genelenght file I will be using throughout the analyses.
genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)

genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")

genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()

genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")

#Get the table of read counts by indicating the path to the file. This file is what we obtained from the running featureCounts.

sg.read.counts.SE <- read.table("~/Projects/StressGranules/SG_Transcriptome/HuR-APEX/alignment_STAR/featureCounts_Results_SE.txt", header=TRUE)

# Afterwards, I discard the columns which don't have read counts. These columns are only informations about the gene.
sg.read.counts.SE <- sg.read.counts.SE[,-c(2:6)]

sg.read.counts <- sg.read.counts.SE

biomart <- fread("~/biomart.txt")[,-1]


colnames(sg.read.counts)[1] <- "ensembl_gene_id_version"
colnames(biomart) <- c("ensembl_gene_id_version", "hgnc_symbol")

eltable <- merge(sg.read.counts,biomart,by = "ensembl_gene_id_version", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="") #this removes reads that dont have a corresponding hgnc symbol.

eltable$ensembl_gene_id_version <- NULL

read.counts <- eltable %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

read.counts <- as.data.frame(read.counts)

row.names(read.counts) <- read.counts$hgnc_symbol #changes rownames into hgnc symbols
read.counts$hgnc_symbol <- NULL #We no longer need this column, as we have the rownames.
```

## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, warning=FALSE, message=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(16, palette = "Pastel")
custom_palette <- sample(custom_palette, 16)

# Use SampleColouring from metadata
SampleColouring <- c(rep("CU",3),rep("CS",3),rep("NU",3),rep("NS",3),rep("G3UI",2),rep("G3SI",3),rep("G3UE",2),rep("G3SE",3),"NLUI","NLSI",rep("NLUE",2),rep("NLSE",2),"NEUI","NESI", rep("NEUE",2), rep("NESE",2),"G3UI","G3SI",rep("G3UE",2),rep("G3SE",2),"NLUI","NLSI",rep("NLUE",2),rep("NLSE",2),"NEUI","NESI",rep("NEUE",2),rep("NESE",2),"G3UI","G3SI",rep("G3UE",2),rep("G3SE",2))

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(0, 7) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )
```

This time, we see massive differences, both in complexity and read depth. This is mostly likely because eCLIP samples end up being "simple peaks" of binding. Analysing them directly like I am, is probably not the best procedure. Nevertheless, since the problem is present on most samples, comparing these eclip samples directly should still be informative.

# Differential Expression Analysis

## Filtration

### ***Before Filtration***

Let us begin with the filtration. Personally, I find that its easier to do visually. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bi-modal distributions, with the first mode corresponding to lowly expressed genes. I remove these by cutting just before the second mode.

```{r, warning=FALSE}

#Cyto and Nuclear Fractions Mock
alternew.CytoUN <- log10(rowSums(read.counts[,c(1:3)]))
alternew.CytoUN <- as.data.frame(alternew.CytoUN)

alternew.CytoST <- log10(rowSums(read.counts[,c(4:6)]))
alternew.CytoST <- as.data.frame(alternew.CytoST)

alternew.NuclearUN <- log10(rowSums(read.counts[,c(7:9)]))
alternew.NuclearUN <- as.data.frame(alternew.NuclearUN)

alternew.NuclearST <- log10(rowSums(read.counts[,c(10:12)]))
alternew.NuclearST <- as.data.frame(alternew.NuclearST)

#G3BP1 Apex Samples
alternew.G3UN_Input <- log10(rowSums(read.counts[,c(13:14,35,53)]))
alternew.G3UN_Input <- as.data.frame(alternew.G3UN_Input)

alternew.G3ST_Input <- log10(rowSums(read.counts[,c(15:17,36,54)]))
alternew.G3ST_Input <- as.data.frame(alternew.G3ST_Input)

alternew.G3UN_Enrich <- log10(rowSums(read.counts[,c(18:19,37:38,55:56)]))
alternew.G3UN_Enrich <- as.data.frame(alternew.G3UN_Enrich)

alternew.G3ST_Enrich <- log10(rowSums(read.counts[,c(20:22, 39:40,57:58)]))
alternew.G3ST_Enrich <- as.data.frame(alternew.G3ST_Enrich)

#NLS Apex Samples
alternew.NLUN_Input <- log10(rowSums(read.counts[,c(23,41)]))
alternew.NLUN_Input <- as.data.frame(alternew.NLUN_Input)

alternew.NLST_Input <- log10(rowSums(read.counts[,c(24,42)]))
alternew.NLST_Input <- as.data.frame(alternew.NLST_Input)

alternew.NLUN_Enrich <- log10(rowSums(read.counts[,c(25:26,43:44)]))
alternew.NLUN_Enrich <- as.data.frame(alternew.NLUN_Enrich)

alternew.NLST_Enrich <- log10(rowSums(read.counts[,c(27:28,45:46)]))
alternew.NLST_Enrich <- as.data.frame(alternew.NLST_Enrich)

#NES Apex Samples
alternew.NEUN_Input <- log10(rowSums(read.counts[,c(29,47)]))
alternew.NEUN_Input <- as.data.frame(alternew.NEUN_Input)

alternew.NEST_Input <- log10(rowSums(read.counts[,c(30,48)]))
alternew.NEST_Input <- as.data.frame(alternew.NEST_Input)

alternew.NEUN_Enrich <- log10(rowSums(read.counts[,c(31:32,49:50)]))
alternew.NEUN_Enrich <- as.data.frame(alternew.NEUN_Enrich)

alternew.NEST_Enrich <- log10(rowSums(read.counts[,c(33:34,51:52)]))
alternew.NEST_Enrich <- as.data.frame(alternew.NEST_Enrich)

#All samples
alternew <- log10(rowSums(read.counts))
alternew <- as.data.frame(alternew)


ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 1.25, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

#Cytoplasmic and Nuclear Fractions
ggplot(alternew.CytoUN, aes(x=alternew.CytoUN)) + geom_density(fill="#FFB6C1", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Unstressed Cytoplasm")

ggplot(alternew.CytoST, aes(x=alternew.CytoST)) + geom_density(fill="#FFDAB9", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Stressed Cytoplasm")

ggplot(alternew.NuclearUN, aes(x=alternew.NuclearUN)) + geom_density(fill="#E6E6FA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Unstressed Nucleus")

ggplot(alternew.NuclearST, aes(x=alternew.NuclearST)) + geom_density(fill="#98FB98", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Stressed Nucleus")

#G3BP1 Apex
ggplot(alternew.G3UN_Input, aes(x=alternew.G3UN_Input)) + geom_density(fill="#87CEFA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - G3 Unstressed Input")

ggplot(alternew.G3ST_Input, aes(x=alternew.G3ST_Input)) + geom_density(fill="#EEE8AA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - G3 Stressed Input")

ggplot(alternew.G3UN_Enrich, aes(x=alternew.G3UN_Enrich)) + geom_density(fill="#87CEFA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - G3 Unstressed Enrich")

ggplot(alternew.G3ST_Enrich, aes(x=alternew.G3ST_Enrich)) + geom_density(fill="#EEE8AA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - G3 Stressed Enrich")

#NLS Apex
ggplot(alternew.NLUN_Input, aes(x=alternew.NLUN_Input)) + geom_density(fill="#87CEFA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - NLS Unstressed Input")

ggplot(alternew.NLST_Input, aes(x=alternew.NLST_Input)) + geom_density(fill="#EEE8AA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - NLS Stressed Input")

ggplot(alternew.NLUN_Enrich, aes(x=alternew.NLUN_Enrich)) + geom_density(fill="#87CEFA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - NLS Unstressed Enrich")

ggplot(alternew.NLST_Enrich, aes(x=alternew.NLST_Enrich)) + geom_density(fill="#EEE8AA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - NLS Stressed Enrich")

#NES Apex
ggplot(alternew.NEUN_Input, aes(x=alternew.NEUN_Input)) + geom_density(fill="#87CEFA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - NES Unstressed Input")

ggplot(alternew.NEST_Input, aes(x=alternew.NEST_Input)) + geom_density(fill="#EEE8AA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - NES Stressed Input")

ggplot(alternew.NEUN_Enrich, aes(x=alternew.NEUN_Enrich)) + geom_density(fill="#87CEFA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - NES Unstressed Enrich")

ggplot(alternew.NEST_Enrich, aes(x=alternew.NEST_Enrich)) + geom_density(fill="#EEE8AA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - NES Stressed Enrich")

```
This time, due to the huge differences in complexity, I could not filter as I usually do. Instead, I did a very simple, and very soft filtration. 

### ***After Filtration***

```{r}
filterednew <- subset(read.counts, alternew>0)


#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))


#Now we need to select only the rows which have those genes.
finaltable <- read.counts[row.names(read.counts)%in%impgenes,]
```


## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#Let us prepare metadata to add to our samples
#I start by normalizing them all together
#This one has all samples
D0 <- DGEList(finaltable)


#Then we calculate the normalization factors for these new generated data frames
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)


#Now we do the matrixes
#the one with everything
mm <- model.matrix(~0+SampleColouring)
#the one without pbs


#and now we normalize. This step also turns our counts into log2-counts per million.
#The object with all simple
y <- voom(D0, mm, plot = T)
```


## Normalisation

For normalisation I convert the read counts into log2CPMs. Its a good way to visualise how the reads are. 

```{r fig.height=8, fig.width=20}
dfforboxy <- finaltable
orderforplots <- SampleColouring
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
# Convert 'Types' to factor with custom order
dfforboxy$Types <- factor(dfforboxy$Types, levels = unique(SampleColouring))

custom_palette <- qualitative_hcl(16, palette = "Dark2")

prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = custom_palette, labels=c(unique(orderforplots))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

#Using the Standard Matrix

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
# Convert 'Types' to factor with custom order
dfforboxy$Types <- factor(dfforboxy$Types, levels = unique(SampleColouring))
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = custom_palette, labels=c(unique(orderforplots))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```


## PCA

```{r}
#this is done to be able to do traditional PCA

custom_palette <- qualitative_hcl(22, palette = "Pastel")
custom_palette <- sample(custom_palette,22, FALSE)
dfforpca <- y$E
dfforpca <- rbind(dfforpca, SampleColouring)
dfforpca <- t(dfforpca)
numcol <- c(1:(ncol(dfforpca)-1))
dfforpca <- as.data.frame(dfforpca) #for some reason I can only change colum type if its a dataframe

dfforpca[ ,numcol] <- apply(dfforpca[ , numcol], 2,            # to turn our columns into numeric
                            function(x) as.numeric(as.character(x)))

#sapply(dfforpca, class) #to check class

# Or specify the factor levels in the order you want
dfforpca$SampleColouring <- factor(dfforpca$SampleColouring, levels = c(unique(SampleColouring)))

pca_res <- prcomp(dfforpca[,numcol], scale=FALSE) #prepare the pca

#to make everything pretty
p <- autoplot(pca_res, data = dfforpca, x=1, y=2, colour = "SampleColouring", size=8, alpha=0.7)
p <- p + theme_bw() + scale_colour_manual(name = "Sample Type", values = custom_palette, labels=c(unique(SampleColouring)))
p <- p + ggtitle("PCA Plot") + theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())
p


#in case you want to see them yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])
```
# Simple Linear Models

Now we must perform linear model fitting.

```{r}
SampleColouring <- c(rep("CU",3),rep("CS",3),rep("NU",3),rep("NS",3),rep("G3UI",2),rep("G3SI",3),rep("G3UE",2),rep("G3SE",3),"NLUI","NLSI",rep("NLUE",2),rep("NLSE",2),"NEUI","NESI", rep("NEUE",2), rep("NESE",2),"G3UI","G3SI",rep("G3UE",2),rep("G3SE",2),"NLUI","NLSI",rep("NLUE",2),rep("NLSE",2),"NEUI","NESI",rep("NEUE",2),rep("NESE",2),"G3UI","G3SI",rep("G3UE",2),rep("G3SE",2))


mm <- model.matrix(~0+SampleColouring)
fitted <- lmFit(y, mm)
head(coef(fitted))
# 
G3BP1 <- makeContrasts(SampleColouringG3SE  - SampleColouringG3SI , levels =colnames(coef(fitted)))
# 
tmp_G3BP1 <- contrasts.fit(fitted, G3BP1)
# 
# #Correction for multiple testing
tmp_G3BP1 <- eBayes(tmp_G3BP1)
# 
top.table <- topTable(tmp_G3BP1,  sort.by = "logFC", n=Inf)
DEGenes <- top.table

ndegs <- 20
```


## SG vs Non-SG

```{r, results='hide', warning=FALSE, message=FALSE}
write.table(x = DEGenes,
            file = "~/Projects/StressGranules/SGPAPER/SGvsWC/HurAPEX.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)


upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$B>20,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -10 & DEGenes$B>30,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  -log10(adj.P.Val))) +
  geom_point(color="#CCCCCC")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#A10702",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#03256C",max.overlaps = 20)+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702")+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C")

```


```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs No-SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "pearson")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("Stress-Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)
  #geom_label_repel(linemarkers[1:ndegs,] , mapping=aes(label=rownames(linemarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  #geom_point(data=linemarkers, aes(x=t, y= log10(Length)), color="#399e5a")



```

# Complex Linear Models

Now we must perform linear model fitting. Once more, we will have 4, one for each matrix we previously designed.

```{r}
SampleColouring <- c(rep("CU",3),rep("CS",3),rep("NU",3),rep("NS",3),rep("G3UI",2),rep("G3SI",3),rep("G3UE",2),rep("G3SE",3),"NLUI","NLSI",rep("NLUE",2),rep("NLSE",2),"NEUI","NESI", rep("NEUE",2), rep("NESE",2),"G3UI","G3SI",rep("G3UE",2),rep("G3SE",2),"NLUI","NLSI",rep("NLUE",2),rep("NLSE",2),"NEUI","NESI",rep("NEUE",2),rep("NESE",2),"G3UI","G3SI",rep("G3UE",2),rep("G3SE",2))

Target <- c(rep("Cyto",3),rep("Cyto",3),rep("Nucleus",3),rep("Nucleus",3),rep("G3BP1",2),rep("G3BP1",3),rep("G3BP1",2),rep("G3BP1",3),"NL","NL",rep("NL",2),rep("NL",2),"NE","NE", rep("NE",2), rep("NE",2),"G3BP1","G3BP1",rep("G3BP1",2),rep("G3BP1",2),"NL","NL",rep("NL",2),rep("NL",2),"NE","NE",rep("NE",2),rep("NE",2),"G3BP1","G3BP1",rep("G3BP1",2),rep("G3BP1",2))

Type <- c(rep("Frac",3),rep("Frac",3),rep("Frac",3),rep("Frac",3),rep("Input",2),rep("Input",3),rep("Enri",2),rep("Enri",3),"Input","Input",rep("Enri",2),rep("Enri",2),"Input","Input", rep("Enri",2), rep("Enri",2),"Input","Input",rep("Enri",2),rep("Enri",2),"Input","Input",rep("Enri",2),rep("Enri",2),"Input","Input",rep("Enri",2),rep("Enri",2),"Input","Input",rep("Enri",2),rep("Enri",2))

Stress <- c(rep("Unstressed",3),rep("Stressed",3),rep("Unstressed",3),rep("Stressed",3),rep("Unstressed",2),rep("Stressed",3),rep("Unstressed",2),rep("Stressed",3),"Unstressed","Stressed",rep("Unstressed",2),rep("Stressed",2),"Unstressed","Stressed", rep("Unstressed",2), rep("Stressed",2),"Unstressed","Stressed",rep("Unstressed",2),rep("Stressed",2),"Unstressed","Stressed",rep("Unstressed",2),rep("Stressed",2),"Unstressed","Stressed",rep("Unstressed",2),rep("Stressed",2),"Unstressed","Stressed",rep("Unstressed",2),rep("Stressed",2))

SG <- c(rep("NO",3),rep("NO",3),rep("NO",3),rep("NO",3),rep("NO",2),rep("NO",3),rep("NO",2),rep("YES",3),"NO","NO",rep("NO",2),rep("NO",2),"NO","NO", rep("NO",2), rep("NO",2),"NO","NO",rep("NO",2),rep("YES",2),"NO","NO",rep("NO",2),rep("NO",2),"NO","NO",rep("NO",2),rep("NO",2),"NO","NO",rep("NO",2),rep("YES",2))


mm <- model.matrix(~0+Target+Type+Stress+SG)
fitted <- lmFit(y, mm)
head(coef(fitted))
# 
# #Correction for multiple testing
tmp <- eBayes(fitted)
# 
top.table <- topTable(tmp,9,  sort.by = "logFC", n=Inf)
DEGenes <- top.table

```


## SG vs Non-SG

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$B>20,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -10 & DEGenes$B>30,] )

ndegs <- 20

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  -log10(adj.P.Val))) +
  geom_point(color="#CCCCCC")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#A10702",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#03256C",max.overlaps = 20)+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702")+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C")

```
# Session Info

```{r}
sessionInfo()
```

