---
title: "Analyses of GSE99304"
subtitle: "The Stress Granule Transcriptome Reveals Principles of mRNA Accumulation in Stress Granules"
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---


# Introduction

In this document, I present all analyses pertaining to the dataset **GSE99304**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE99304)),  
originally published by **A. Khong, 2017**.

## Libraries

```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)

set.seed(1906)
```


## Data Loading

```{r}
#This section here is to load up the genelenght file I will be using throughout the analyses.
genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)

genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")

genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()

genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")



#Get the table of read counts by indicating the path to the file. This file is what we obtained from the running featureCounts.

sg.read.counts <- read.table("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2017/TrimmedSamples/alignment_STAR/featureCounts_results.txt", header=TRUE)

# Afterwards, I discard the columns which don't have read counts. These columns are only informations about the gene.
sg.read.counts <- sg.read.counts[,-c(2:6)]

#Loading up our biomart file, obtained from Ensembl, where we have the equivalence between Ensembl IDs and HGNC names
biomart <- fread("~/biomart.txt")[,-1]


colnames(sg.read.counts)[1] <- "ensembl_gene_id_version"
colnames(biomart) <- c("ensembl_gene_id_version", "hgnc_symbol")

eltable <- merge(sg.read.counts,biomart,by = "ensembl_gene_id_version", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="") #this removes reads that dont have a corresponding hgnc symbol.

eltable$ensembl_gene_id_version <- NULL

stress.read.counts <- eltable %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

stress.read.counts <- as.data.frame(stress.read.counts)

row.names(stress.read.counts) <- stress.read.counts$hgnc_symbol #changes rownames into hgnc symbols
stress.read.counts$hgnc_symbol <- NULL #We no longer need this column, as we have the rownames.

```

### Bulk TPM for Centrifugation

In this section I prepare the bulk samples that I then use to simulate SGs from.

```{r}
read.counts <- subset(stress.read.counts, !(SRR5605166Aligned.sortedByCoord.out.bam == 0 | SRR5605167Aligned.sortedByCoord.out.bam == 0 | SRR5605168Aligned.sortedByCoord.out.bam == 0))



mof <- fread("~/Projects/StressGranules/Simulator/MolecularWeights.csv")
colnames(mof)[2] <- "GeneID" 
mof <- mof[!duplicated(mof$GeneID), ]




#Lets grab the samples to use in our simulator
#sample number 1
#First the Unstressed Samp1
stress.tpm <- read.counts[,4]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/SG2017/17Bulk_1.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)

#sample number 2
stress.tpm <- read.counts[,5]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/SG2017/17Bulk_2.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)

#sample number 3
stress.tpm <- read.counts[,6]
stress.tpm <- as.data.frame(stress.tpm)
stress.tpm$GeneID <- row.names(read.counts)
colnames(genelength) <- c("Length", "GeneID")
stress.tpm <- merge(stress.tpm, genelength, by="GeneID")
stress.tpm$TPM <- (stress.tpm$stress.tpm)/stress.tpm$Length
stress.tpm$TPM <- (stress.tpm$TPM * 1e6)
stress.tpm$TPM <- stress.tpm$TPM / sum(stress.tpm$TPM)
stress.tpm <- stress.tpm[,-2]
stress.tpm <- merge(stress.tpm, mof, by="GeneID")
write.table(x = stress.tpm,
            file = "~/Projects/StressGranules/Simulator/OriginalSamples/SG2017/17Bulk_3.txt",
            sep="\t", col.names = TRUE, row.names = FALSE, quote=FALSE)
```


## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, warning=FALSE, message=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(2, palette = "Pastel")
custom_palette <- sample(custom_palette, 2)

# Use SampleColouring from metadata
SampleColouring <- c("SG", "SG", "SG", "WC", "WC", "WC")

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(6, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

compplot <- ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_colour_manual(values = c("#53A2BE","#BA3B46"), name = "Sample Type", labels = c("WC", "SG"),breaks = c("WC","SG")) +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") + 
  xlim(0, 5000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )+
  theme(text=element_text(size=20,  family="Arial MS")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),legend.position = c(0.7, 0.3))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Fig2A.png", width = 7, height=4, units = "in", res=300)
compplot
dev.off()
```

As we can see in these plots, even though there is a sample type with lower complexity (Bulk samples), in this case there are no individual sample outliers in terms of complexity or read depth. Excellent!

# Differential Expression Analysis

## Filtration

### ***Before Filtration***

Let us begin with the filtration. Personally, I find that its easier to do visually. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bi-modal distributions, with the first mode corresponding to lowly expressed genes. I remove these by cutting just before the second mode. This time, I opted to also observe every sample type by itself.

```{r, warning=FALSE}

alternew.sg <- log10(rowSums(stress.read.counts[,1:3]))
alternew.sg <- as.data.frame(alternew.sg)

alternew.bu <- log10(rowSums(stress.read.counts[,4:6]))
alternew.bu <- as.data.frame(alternew.bu)

alternew <- log10(rowSums(stress.read.counts))
alternew <- as.data.frame(alternew)


ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.bu, aes(x=alternew.bu)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Bulk Samples")

ggplot(alternew.sg, aes(x=alternew.sg)) + geom_density(fill="#a2d1e6", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Stress Granules Samples")
```

By seeing the shape of the graph, we can see where we should cut. We have a **binomial distribution** (low reads and noise vs actual biological meaningful reads). The ideal place to cut them is right before the wedge.

### ***After Filtration***

I didn't do it this time, but I've maintained the code to filter by sample type. This time, I considered that it was more correct to filter all samples together.


```{r}
filterednew <- subset(stress.read.counts, alternew>3.5)
filterednew.sg <- subset(stress.read.counts, alternew.sg>3)
filterednew.bulk <- subset(stress.read.counts, alternew.bu>3)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))
impgenes.sg <- c(row.names(filterednew.sg))
impgenes.bulk <- c(row.names(filterednew.bulk))

#Now we need to select only the rows which have those genes.
finaltable <- stress.read.counts[row.names(stress.read.counts)%in%impgenes,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.sg,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.bulk,]

write.csv(finaltable, "~/Projects/StressGranules/Simulator/OriginalSamples/SG2017/Filtcounts.txt", row.names = T)

```

## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable) 

#Then we calculate the normalisation factors for these new generated data frames
#Bear in mind this step calculates the normalisation factors, but does not normalise itself.
D0 <- calcNormFactors(D0)


#Lets create the metadata
Sample <- c(rep("SG",3), rep("WC",3))

mm <- model.matrix(~0+Sample)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```

## Alternative DEA

By plotting log2CPMs of SG vs Bulk. This will allows us to see which genes disobey the general trend. This is how I generated some images, but for that I filtered the sample types separately. Please see above.

```{r}
#Calculating the average expression of SG samples
sglogs <- y$E[,c(1:3)]

med <- c()
aver <- c()
for (i in 1:nrow(sglogs)){
  med <- mean(sglogs[i,])
  aver <- c(aver, med)
}

sglogs <- cbind(sglogs, aver)
sglogs <- sglogs[,4]

#Calculating the average expression of WC samples
bulklogs <- y$E[,c(4:6)]

med <- c()
aver <- c()
for (i in 1:nrow(bulklogs)){
  med <- mean(bulklogs[i,])
  aver <- c(aver, med)
}

bulklogs <- cbind(bulklogs, aver)
bulklogs <- bulklogs[,4]

#Merging both together
sglogs <- as.data.frame(sglogs)

bulklogs <- as.data.frame(bulklogs)

sglogs$Gene <- row.names(sglogs)
bulklogs$Gene <- row.names(bulklogs)

sglogs <- reshape2::melt(sglogs)
bulklogs <- reshape2::melt(bulklogs)

sgVbulklogs <- cbind(sglogs, bulklogs)
sgVbulklogs <- sgVbulklogs[,-c(2,4,5)]
colnames(sgVbulklogs) <- c("Gene","SG", "Bulk")

res_logs <- lm(Bulk ~ SG, sgVbulklogs)

cor.test(sgVbulklogs$SG, sgVbulklogs$Bulk, method = "spearman")


ggplot(sgVbulklogs, aes(y=SG, x=Bulk)) + geom_point(color="#CCCCCC", alpha=0.3) +   theme_bw()+
 ylab("SG Expression (log2 CPM)") + xlab("Whole-Cell Expression (log2 CPM)") + geom_density2d()+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("SG and Whole-Cell Expression Correlation")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_logs$coefficients[2], intercept = res_logs$coefficients[1], linetype = "dashed")

write.table(sgVbulklogs, "~/Projects/StressGranules/SGPAPER/SGvsWC/Expression.txt", row.names = F, col.names = T)
```


## Normalisation

Now, we are almost ready to move on to the differential gene expression analysis. I will first check to see if any issues arose during the normalisation process, by looking at samples before and after this process.

```{r}

dfforboxy <- finaltable

orderforplots <- c("SG", "SG", "SG", "Bulk", "Bulk", "Bulk")
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#769fca", "#de425b"), labels=c("SG", "Bulk"), breaks =c("SG", "Bulk")) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#769fca", "#de425b"), labels=c("SG", "Bulk"), breaks =c("SG", "Bulk")) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```


In this case, from the boxplots, we can see that there no samples which can be considered outliers.

## PCA

I like to start with a Principal Component Analysis (PCA), which helps to identify what drives the largest variation in the data.


```{r PCA}

#In here, I am just defining some colours to use
custom_palette <- c("#43F3D5", "#769fca")

#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- Sample

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-1)], scale = FALSE)

p_Sample <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Sample") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Sample

#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])

```


As one can observe from the PCAs, our PC1 is mostly explained by sample type (95.27%).

## Linear Models


Finally, we move on to differential gene expression! In this chunk, I define the linear model I will actually use for my analyses. 

```{r}
#We now fit the linear model to our expression
fitted <- lmFit(y, mm)

head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr <- makeContrasts(SampleSG - SampleWC, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)

#Correction for multiple testing. This is very important, as with so many tests, a ton of transcripts will otherwise achieve very low p-values, simply by random chance.
tmp <- eBayes(tmp)
```


## Differential Expression Analysis

Lets move on to the main analyses.

```{r}
#time to see which genes are differentially expressed!
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)

DEGenes <- top.table

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -4 & DEGenes$adj.P.Val<0.05,] )

ndegs <- 20
```

## Volcano Plots

Now, we simply see all of this on a volcano plot. Personally, I prefer my significance to be given by B-statistic, since adjusted p-values easily saturate at values closer to 0. As B-statistic is an "infinite" scale, this doesn't occur.

### Stress Granule vs Whole Cell RNA ("Normal" Volcano Plot)

```{r, message=FALSE, warning=FALSE, results='hide'}
write.table(x = DEGenes,
            file = "~/Projects/StressGranules/SGPAPER/SGvsWC/Akhong2017.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#DDDDDD")+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs Whole-Cell (DC)")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#de425b",max.overlaps = 200, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#769fca",max.overlaps = 200, fontface="italic")+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#de425b", size=3)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#769fca", size=3) + xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") + theme(text=element_text(family = "Arial", size = 20))

stress_t <- data.frame(Gene=row.names(DEGenes), s_t=DEGenes$t, s_logFC=DEGenes$logFC)
```

### Stress Granule vs Whole Cell RNA (Volcano Plot with Gene Lenght)

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "pearson")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

As seen by the original authors, there is a massive lenght bias in SG enrichment.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -4 & DEGenes$adj.P.Val<0.05,] )
finagenes <- rbind(upmarkers, downmarkers)
finagenes <- as.data.frame(finagenes[abs(finagenes$t)>19,] )

dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1, 2, 3, 5, 6, 7))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=c("SG", "SG", "SG", "Bulk", "Bulk", "Bulk"))
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = c("#769fca", "#de425b"), breaks = c("SG", "Bulk")) + scale_colour_manual(name = "Sample Type", values = c("#769fca", "#de425b"), breaks = c("SG", "Bulk")) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

write.table(hallmark.sg.t, "~/Projects/StressGranules/SGPAPER/SGvsWC/KhongSGHallmark.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```


## GSEA Length

Now, since SGs are enriched for longer transcripts, these pathways may simply be the ones which are tipically enriched in longer transcripts. Therefore, I will repeat the analysis, but using transcript lenght instead of t-statistic. 

```{r, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,8]),]
gseatool <- DEGenes[,8]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- log10(gseatool$t)
vectorranks <- scale(vectorranks)
names(vectorranks) <- gseatool$Symbol


#to load our pathways

##-----HALLMARK-----
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.size.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.size.t <- as.data.frame(hallmark.size.t)

write.table(hallmark.size.t, "~/Projects/StressGranules/SGPAPER/SGvsWC/LengthHallmark.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)


#HALLMARK
hallmark.scatter_ak <- merge(hallmark.sg.t, hallmark.size.t, by="pathway")
colnames(hallmark.scatter_ak) <- c("pathway", "sg_pval", "sg", "size_pval", "size")
hallmark.scatter_ak <- na.omit(hallmark.scatter_ak)
res_hallmark <- lm(size ~ sg, hallmark.scatter_ak)

vip.pathways_ak <- as.data.frame(hallmark.scatter_ak[(hallmark.scatter_ak$sg/hallmark.scatter_ak$size)<0,] )



ggplot(hallmark.scatter_ak, aes(x=sg, y=size)) + geom_point(color="black", alpha=01) +   theme_bw()+
 xlab("NES SG") + ylab("NES Size") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("hallmark SG vs Size")+
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope=res_hallmark$coefficients[2], intercept = res_hallmark$coefficients[1], linetype = "dashed") +
  geom_label_repel(vip.pathways_ak[1:ndegs,] , mapping=aes(label=vip.pathways_ak$pathway[1:ndegs]),size=3,point.padding=-0.5, color="#399e5a",max.overlaps = 200)+
  geom_point(data=vip.pathways_ak, aes(x=sg, y= size), color="#399e5a") + stat_cor(method = "spearman", cor.coef.name = "rho")

```

Barring some exceptions which I labeled on the plot, most gene set enrichment is driven by transcript size.

## Most altered pathways

I opted to see altered pathways as a whole. That is, ignoring bidirectionally and simply looking at what pathways are more altered as a whole (i.e, more expressed genes on either side).

```{r, echo=FALSE, fig.height=20, fig.width=30, message=FALSE, warning=FALSE, results='hide'}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol
forcorrel <- vectorranks
vectorranks <- abs(vectorranks)


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")



gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))

```

# After SpinSim v.2.

After my SpinSim pipeline, this is how I compared these simulated SGs with bulk and their original counterparts. For more details on this pipeline, please refer to its corresponding files in the Simulations/ folder in the GitHub repository.

```{r}
stress.read.counts$Gene <- row.names(stress.read.counts)

#To maintain equitable read depth, I base my final total count number per sample on the original samples
goalsum_SG1 <- sum(stress.read.counts$SRR5605163Aligned.sortedByCoord.out.bam)
goalsum_SG2 <- sum(stress.read.counts$SRR5605164Aligned.sortedByCoord.out.bam)
goalsum_SG3 <- sum(stress.read.counts$SRR5605165Aligned.sortedByCoord.out.bam)


colnames(genelength) <- c("Length","Gene")

#For the simulated samples. 
sim_sg1 <- read.table("~/Projects/StressGranules/Simulator/New/SG2017/NewSG_1.txt", sep = ",", header = T)
sim_sg1 <- na.omit(sim_sg1)
colnames(sim_sg1) <- c("Gene", "SimSG1")
sim_sg2 <- read.table("~/Projects/StressGranules/Simulator/New/SG2017/NewSG_2.txt", sep = ",", header = T)
sim_sg2 <- na.omit(sim_sg2)
colnames(sim_sg2) <- c("Gene", "SimSG2")
sim_sg3 <- read.table("~/Projects/StressGranules/Simulator/New/SG2017/NewSG_3.txt", sep = ",", header = T)
sim_sg3 <- na.omit(sim_sg3)
colnames(sim_sg3) <- c("Gene", "SimSG3")


#merging old samples up
simsamples <- merge(sim_sg1, sim_sg2, by="Gene")
simsamples <- merge(simsamples, sim_sg3, by="Gene")
simsamples <- merge(simsamples, genelength, by="Gene")

#De-Tpming old samples
simsamples$Counts_SimSG1 <- simsamples$SimSG1*simsamples$Length
simsamples$Counts_SimSG1 <- simsamples$Counts_SimSG1/sum(simsamples$Counts_SimSG1)*goalsum_SG1
simsamples$Counts_SimSG2 <- simsamples$SimSG2*simsamples$Length
simsamples$Counts_SimSG2 <- simsamples$Counts_SimSG2/sum(simsamples$Counts_SimSG2)*goalsum_SG2
simsamples$Counts_SimSG3 <- simsamples$SimSG3*simsamples$Length
simsamples$Counts_SimSG3 <- simsamples$Counts_SimSG3/sum(simsamples$Counts_SimSG3)*goalsum_SG3
simsamples <- simsamples[,c(1,6:8)] 

OGsamples <- stress.read.counts

SG_City <- merge(OGsamples, simsamples, by="Gene")
SG_City <- SG_City %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(Gene) %>%
  summarise(across(everything(), sum, na.rm = TRUE))
SG_City <- as.data.frame(SG_City)
row.names(SG_City) <- SG_City$Gene
SG_City$Gene <- NULL
```


## Filtration

### ***Before Filtration***

Let us begin with the filtration.

```{r, warning=FALSE}
alternew.bu <- log10(rowSums(SG_City[,4:6]))
alternew.bu <- as.data.frame(alternew.bu)

alternew.rsg <- log10(rowSums(SG_City[,1:3]))
alternew.rsg <- as.data.frame(alternew.rsg)

alternew.ssg <- log10(rowSums(SG_City[,7:9]))
alternew.ssg <- as.data.frame(alternew.ssg)

alternew <- log10(rowSums(SG_City))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.bu, aes(x=alternew.bu)) + geom_density(fill="#B0D9B1", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Bulk Sample")

ggplot(alternew.rsg, aes(x=alternew.rsg)) + geom_density(fill="#618264", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Real SG")

ggplot(alternew.ssg, aes(x=alternew.ssg)) + geom_density(fill="#EA4848", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Sim SG")

dffordensy <- SG_City
sampy <- c(rep("RSG",3), rep("BU",3), rep("SimSG",3))
names(sampy) <- colnames(dffordensy)
dffordensy <- reshape2::melt(t(dffordensy))
colnames(dffordensy) <- c("Sample", "Gene", "Expression")
```

### ***After Filtration***

This time, on the filtration, I filter by sample type. My SpinSim pipeline leads to a few outliers (some very low expressed genes), which would mess with the normalisation.

```{r}
MitoGenes <- fread("~/MitoGenes.txt", header = FALSE)[[1]]

filterednew <- subset(SG_City, alternew>3.5)
filterednew.bu <- subset(SG_City, alternew.bu>2.75)
filterednew.rsg <- subset(SG_City, alternew.rsg>2.75)
filterednew.ssg <- subset(SG_City, alternew.ssg>2.75)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew), MitoGenes)
impgenes.bu <- c(row.names(filterednew.bu))
impgenes.rsg <- c(row.names(filterednew.rsg))
impgenes.ssg <- c(row.names(filterednew.ssg))

#Now we need to select only the rows which have those genes.
finaltable <- SG_City[row.names(SG_City)%in%impgenes,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.bu,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.rsg,]
finaltable <- finaltable[row.names(finaltable)%in%impgenes.ssg,]

```


## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)

#Then we calculate the normalization factors.
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)


#Now we do our matrix itself
mm <- model.matrix(~0+sampy)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```


## Normalisation

For normalisation I convert the read counts into log2CPMs. Its a good way to visualize how the reads are. 

```{r}

dfforboxy <- finaltable
coolors <-c("#B0D9B1", "#B0D9B1", "#B0D9B1", "#618264", "#618264", "#618264", "#EA4848", "#EA4848", "#EA4848", "#3BABBA", "#3BABBA", "#3BABBA")


orderforplots <- sampy
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = unique(coolors), labels=unique(sampy), breaks =unique(sampy)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = unique(coolors), labels=unique(sampy), breaks =unique(sampy)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```



## PCA

```{r}
#this is done to be able to do traditional PCA
dfforpca <- y$E
dfforpca <- rbind(dfforpca, sampy)
row.names(dfforpca)[nrow(dfforpca)] <- "sampy"
dfforpca <- t(dfforpca)
numcol <- c(1:(ncol(dfforpca)-1))
dfforpca <- as.data.frame(dfforpca) #for some reason I can only change colum type if its a dataframe

dfforpca[ ,numcol] <- apply(dfforpca[ , numcol], 2,            # to turn our columns into numeric
                            function(x) as.numeric(as.character(x)))

#sapply(dfforpca, class) #to check class

# Or specify the factor levels in the order you want
dfforpca$forall <- factor(dfforpca$sampy, levels = unique(sampy))

pca_res <- prcomp(dfforpca[,numcol], scale=FALSE) #prepare the pca

#to make everything pretty
p <- autoplot(pca_res, data = dfforpca, x=1, y=2, colour = "sampy", size=8, alpha=0.7)
p <- p + theme_bw() + scale_colour_manual((name = "Samples"), values = unique(coolors), labels=unique(sampy))
p <- p + ggtitle("PCA Plot") + theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())
p


#in case you want to see them yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])
```


## Linear Models

Now we must perform linear model fitting. Although we can perform several comparisons, which I leave the contrasts off here, I will focus on the comparison of the Simulated SGs with the Original SGs, as this is what reveals what is more or less enriched than what would be expected with centrifugation alone.

```{r}
fitted <- lmFit(y, mm)

head(coef(fitted))

#_____________________Original Enrichment
#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr <- makeContrasts(sampyRSG  - sampyBU, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)

top.table <- topTable(tmp, sort.by = "logFC", n = Inf)

DEGenes <- top.table

originalenrichment <- data.frame(Gene=row.names(DEGenes), og_t=DEGenes$t, og_logFC=DEGenes$logFC)

#_____________________Simulated Enrichment
#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr <- makeContrasts(sampySimSG  - sampyBU, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)

top.table <- topTable(tmp, sort.by = "logFC", n = Inf)

DEGenes <- top.table

simulatedenrichment <- data.frame(Gene=row.names(DEGenes), sim_t=DEGenes$t, sim_logFC=DEGenes$logFC)

#_____________________Corrected Enrichment
#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr <- makeContrasts(sampyRSG  - sampySimSG, levels = colnames(coef(fitted)))

#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)

top.table <- topTable(tmp, sort.by = "logFC", n = Inf)

DEGenes <- top.table



ndegs <- 20
```


## Original SG vs Simulated SG

```{r, results='hide', warning=FALSE, message=FALSE}

write.table(x = DEGenes,
            file = "~/Projects/StressGranules/SGPAPER/RealSGs/RealAkhong.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)


upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -3 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Original vs Simulated")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))  +
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))


correctedsignal <- data.frame(Gene=row.names(DEGenes), cor_t=DEGenes$t, cor_logFC=DEGenes$logFC)
```

## Volcano Length

```{r, message=FALSE, warning=FALSE, results='hide'}


DEGenes$Gene <- row.names(DEGenes)
colnames(genelength)[2] <- "Gene"
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs Whole-Cell RNA - Sodium Arsenite")+ ylab("B stat")+xlab("Log 2 Fold Change") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "blue", high = "red")



smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene

linemarkers <- rbind(upmarkers, downmarkers)
linemarkers <- DEGenes[DEGenes$Gene %in% row.names(linemarkers),]

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")


firstsimcentri <- ggplot(DEGenes, aes(y=t, x=log10(Length))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)+ theme_bw()+
 xlab("RNA Length") + ylab("Corrected SG Enrichment") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)+ 
  theme(text=element_text(size=15,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=5)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

#tiff("~/Projects/StressGranules/SGPAPER/Figure3A.tiff", width = 7, height=4, units = "in", res=300)
firstsimcentri
#dev.off()

```

## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+ 
  theme(text=element_text(size=25,  family="Arial MS"))
```

# Comparison Between Original and Simulated Enrichment

Is the SpinSim pipeline, which only takes centrifugation into account, a good recapitulation of DC's SG enrichment? Here, I directly check this.

```{r fig.height=6, fig.width=7}
cor_sim_df <- merge(simulatedenrichment, originalenrichment, by="Gene")
cor_sim_df <- merge(cor_sim_df, genelength, by="Gene")


smooth.SG <- lm(sim_t ~ og_t, cor_sim_df)
liner.SG <- rstandard(smooth.SG)
cor_sim_df <-cbind(cor_sim_df, liner.SG)
colnames(cor_sim_df)[7] <- "Residuals"
row.names(cor_sim_df) <- cor_sim_df$Gene
linemarkers <- as.data.frame(cor_sim_df[abs(cor_sim_df$Residuals)>4,])

cor.test(cor_sim_df$sim_t, cor_sim_df$og_t, method="spearman")

SimCorrel <- ggplot(cor_sim_df, aes(x=og_t, y=sim_t)) + geom_point(aes(colour=log10(Length)),alpha=0.5, size=5) +   theme_bw()+
 xlab("Original SG Enrichment") + ylab("Simulated SG Enrichment") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Simulation Results")+
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+
  theme(text=element_text(size=15,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope = smooth.SG$coefficients[2], 
              intercept = smooth.SG$coefficients[1], 
              linetype = "dashed") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  scale_colour_gradient(low="#729EA1", high = "#DB5375")+
  labs(colour = "RNA Length")

SimCorrel

#tiff("../Projects/StressGranules/SGPAPER/SimCorrel.tiff", width = 8, height=6, units = "in", res=300)
#SimCorrel
#dev.off()
```

As we can see, there is a strong correlation, which implies that the hypothesis that the signal is mainly from centrifugation cannot be discarded.

# Session Info

```{r}
sessionInfo()
```