---
title: "Analyses of PRJNA1152621"
subtitle: "Subcellular level spatial transcriptomics with PHOTON"
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---


# Introduction

In this document, I present all analyses pertaining to the dataset **PRJNA1152621**  
([NCBI BioProject Link](https://www.ncbi.nlm.nih.gov/bioproject/?term=PRJNA1152621)),  
originally published by **S. Rajachandran, 2024**.

```{r , message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)
library(readxl)


set.seed(1906)
```


## Loading up the data

```{r}
#This section here is to load up the genelenght file I will be using throughout the analyses.
genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)

genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")

genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()

genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")


#Loading up the data, from the featureCounts table.
raw.read.counts <- read.table("~/Projects/StressGranules/SG_Transcriptome/PHOTON/alignment_STAR/featureCounts_results.txt", header = T)[,-c(2:6)]


biomart <- fread("~/biomart.txt")[,-1]


colnames(raw.read.counts)[1] <- "ensembl_gene_id_version"
colnames(biomart) <- c("ensembl_gene_id_version", "hgnc_symbol")

eltable <- merge(raw.read.counts,biomart,by = "ensembl_gene_id_version", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="") #this removes reads that dont have a corresponding hgnc symbol.

eltable$ensembl_gene_id_version <- NULL

read.counts <- eltable %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

read.counts <- as.data.frame(read.counts)

row.names(read.counts) <- read.counts$hgnc_symbol #changes rownames into hgnc symbols
read.counts$hgnc_symbol <- NULL #We no longer need this column, as we have the rownames.

```



## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, warning=FALSE, message=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(6, palette = "Pastel")
custom_palette <- sample(custom_palette, 6)

# Use SampleColouring from metadata
SampleColouring <- c(rep("SG",3), rep("Nucin",2), rep("Nuc",2), rep("SGkoin",3), rep("SGko",3), rep("SGin",3))

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(4, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )
```

As we can see in these plots, even though there is a sample type with less complexity, in this case there are no sample outliers in terms of complexity or read depth. Excellent!

# Differential Expression Analysis

## Filtration

### ***Before Filtration***

Let us begin with the filtration. Personally, I find that its easier to do visually. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bi-modal distributions, with the first mode corresponding to lowly expressed genes. I remove these by cutting just before the second mode.

```{r, warning=FALSE}
alternew.sg <- log10(rowSums(read.counts[,1:3]))
alternew.sg <- as.data.frame(alternew.sg)

alternew.nuc.inp <- log10(rowSums(read.counts[,4:5]))
alternew.nuc.inp <- as.data.frame(alternew.nuc.inp)

alternew.nuc <- log10(rowSums(read.counts[,6:7]))
alternew.nuc <- as.data.frame(alternew.nuc)

alternew.sg.m6ako.input <- log10(rowSums(read.counts[,8:10]))
alternew.sg.m6ako.input <- as.data.frame(alternew.sg.m6ako.input)

alternew.sg.m6ako <- log10(rowSums(read.counts[,11:13]))
alternew.sg.m6ako <- as.data.frame(alternew.sg.m6ako)

alternew.sg.input <- log10(rowSums(read.counts[,14:16]))
alternew.sg.input <- as.data.frame(alternew.sg.input)

alternew <- log10(rowSums(read.counts))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.sg, aes(x=alternew.sg)) + geom_density(fill="#a2d1e6", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - SG")

ggplot(alternew.nuc.inp, aes(x=alternew.nuc.inp)) + geom_density(fill="#769fca", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.25, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Nucleolus Input")

ggplot(alternew.nuc, aes(x=alternew.nuc)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - Nucleolus")

ggplot(alternew.sg.m6ako.input, aes(x=alternew.sg.m6ako.input)) + geom_density(fill="#de425b", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - SG m6A KO Input")

ggplot(alternew.sg.m6ako, aes(x=alternew.sg.m6ako)) + geom_density(fill="#de425b", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - SG m6A KO")


ggplot(alternew.sg.input, aes(x=alternew.sg.input)) + geom_density(fill="#a2d1e6", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - SG Input")
```


### ***After Filtration***

By seeing the shape of the graph, we can see where we should cut. We have a **binomial distribution** (low reads and noise vs actual biological meaningful reads). The ideal place to cut them is right before the wedge. We could filter on a sample type per sample type basis, but seeing as most samples had very similar read depths, I opted to filter them all together.

```{r}
#Here is my filtration. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bimodal distributions, with the first mode corresponding to lowly expressed or noisy genes. I remove these by cutting just before the second mode.

MitoGenes <- fread("~/MitoGenes.txt", header = FALSE)[[1]]
filterednew <- subset(read.counts, alternew>3.5)
#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- row.names(falternew)

#Now we need to select only the rows which have those genes.
finaltable <- read.counts[row.names(read.counts)%in%impgenes,]
```


## Matrices

Good! Now we can proceed by designing our matrices. 

```{r, fig.height=8, fig.width=4}
#Let us prepare metadata to add to our samples
Sample_names <- c(rep("SGm6AKOInput",3), rep("Nucleus",2), rep("NucleusInput",2), rep("SGm6AKO",3), rep("SG",3), rep("SGInput",3))
Sample_type <- c(rep("IN",3), rep("Out",2), rep("IN",2), rep("Out",3), rep("Out",3), rep("IN",3))
Sample_target <- c(rep("SG",3), rep("NUC",2), rep("NUC",2), rep("SG",3), rep("SG",3), rep("SG",3))
Sample_genome <- c(rep("KO",3), rep("WT",2), rep("WT",2), rep("KO",3), rep("WT",3), rep("WT",3))

#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)

#Then we calculate the normalization factors.
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)


#Now we do our matrix itself
mm <- model.matrix(~0+Sample_names)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)

#I did this simply to check if replicates were indeed similar between each other
heatmap(cor(y$E, method="spearman"), symm=T, labCol = colnames(y$E), labRow = Sample_names)

```

Replicates seem more similar to one another than to other sample types. Good!

## Normalisation

For normalisation I convert the read counts into log2CPMs. Its a good way to visualise how the reads are. 

```{r fig.height=4, fig.width=14}
#To visualise how the normalisation went.


sample_order <- c("Nucleus", "NucleusInput", "SG", "SGInput", "SGm6AKO", "SGm6AKOInput")

dfforboxy <- finaltable

orderforplots <- Sample_names
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
dfforboxy$Types <- factor(
  dfforboxy$Types,
  levels = sample_order
)
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab(expression(Log[2]~Read~Counts~"\n")) + xlab("Samples") + scale_fill_manual(
  name = "",
  values = c(
    "Nucleus" = "#FFB3BA",
    "Nucleus Input   " = "#FFDFBA",
    "SG" = "#FFFFBA",
    "SG Input   " = "#BAFFC9",
    "SG m6A KO" = "#BAE1FF",
    "SG m6A KO Input   " = "#CBAACB"
  ),
  breaks = sample_order
) + theme_classic() +
  theme(axis.line = element_line(colour = "black"),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + ggtitle("Pre-Normalisation")+
  theme(text=element_text(size=20,  family="Arial MS"))+theme(
  legend.title = element_text(size = 16),
  legend.text = element_text(size = 14)
)

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
dfforboxy$Types <- factor(
  dfforboxy$Types,
  levels = sample_order
)
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + ylab(expression(Log[2]~Read~Counts~"\n")) + xlab("Samples") + scale_fill_manual(
  name = "",
  values = c(
    "Nucleus" = "#FFB3BA",
    "Nucleus Input   " = "#FFDFBA",
    "SG" = "#FFFFBA",
    "SG Input   " = "#BAFFC9",
    "SG m6A KO" = "#BAE1FF",
    "SG m6A KO Input   " = "#CBAACB"
  ),
  breaks = sample_order
) + theme_classic() +
  theme(axis.line = element_line(colour = "black"),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())  + ggtitle("Post-Normalisation")+
  theme(text=element_text(size=20,  family="Arial MS"))+theme(
  legend.title = element_text(size = 16),
  legend.text = element_text(size = 14)
)

normplot <- ggarrange(
  prenorm,
  postnorm,
  nrow = 1,
  common.legend = TRUE,
  legend = "bottom"
)

normplot
png("~/Images/Normalisation_Example.png", width = 14, height=4, units = "in", res=300)
normplot
dev.off()
```

Normalisation process went very well!

## PCA

Before we move on, lets do a **Principal Component analysis**. It is a good exploratory tool to find out how our samples relate to each other and if there are any obvious outliers. We would expect our replicates to be nearer to one another than to other samples, and more different samples should be further away. 

```{r}
# Define color palette
custom_palette <- qualitative_hcl(6, palette = "Pastel")
custom_palette <- sample(custom_palette, 6)

#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Genome <- Sample_genome
dfforpca$Target <- Sample_target
dfforpca$Type <- Sample_type
dfforpca$Name <- Sample_names

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-4)], scale = FALSE)

p_Genome <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Genome", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Genome", values = custom_palette) +
  ggtitle("PCA - Coloured by Genome") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Target <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Target", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Target", values = custom_palette) +
  ggtitle("PCA - Coloured by Target") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Type <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Type", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Type", values = custom_palette) +
  ggtitle("PCA - Coloured by Type") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Name <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Name", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Name", values = custom_palette) +
  ggtitle("PCA - Coloured by Name") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_Genome
p_Target
p_Type
p_Name

#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])


```

From the PCA, we can see that biggest variable is the "Target", that is, what PHOTON aims to capture.

# Nucleolus

## Linear Models

Now we must perform linear model fitting.

```{r}

fitted <- lmFit(y, mm)
head(coef(fitted))


#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(Sample_namesNucleus  - Sample_namesNucleusInput, levels = mm)


#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```
## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -3 & DEGenes$adj.P.Val<0.05,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Nucleolus")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```
## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Nucleolus")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("Nucleolus") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
The nucleolus has a small negative lenght bias.

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}
finagenes <- rbind(upmarkers, downmarkers)


dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:16))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=Sample_names)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(Sample_names)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(Sample_names)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```

## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*. 

While I usually check Hallmark Gene Sets, for the Nucleolus, I considered GO:Cell Component to be more interesting.

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "C5", subcollection = "GO:CC")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Cell Component Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

As expected, we have an enrichment of gene sets related to ribosome. Considering the nucleolus known role in ribosome biosynthesis, this is a good sign!

# SG WT

## Linear Models

Now we must perform linear model fitting.

```{r}
fitted <- lmFit(y, mm)
head(coef(fitted))


#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(Sample_namesSG       - Sample_namesSGInput      , levels = mm)


#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```


## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -3 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("WT SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))


write.table(x = DEGenes,
           file = "~/Projects/StressGranules/SGPAPER/SGvsWC/Rajachandran_WT.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)
```
## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("WT SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("WT SG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

The SG has a "modest" lenght bias.

## Correlation with m6A

```{r}
m6A <- read_excel("~/Projects/StressGranules/SG_Transcriptome/m6A/41467_2022_31358_MOESM26_ESM.xlsx")[,c(4,16,17)]
colnames(m6A)[1] <- "Gene"

DEGenes <- merge(DEGenes, m6A, by="Gene")

ggplot(DEGenes, aes(y=t, x=(m6a_sites))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab("Number of m6A Sites") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) + stat_cor(method = "spearman", cor.coef.name = "rho", size=5)
```
There is indeed a correlation of SG enrichment with number of m6A sites. However, as there was one with overall transcript lenght, and transcript lenght is linked to number of m6A sites, this was to be expected.


## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}
finagenes <- rbind(upmarkers, downmarkers)


dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:16))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=Sample_names)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(Sample_names)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(Sample_names)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```

## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# SG m6A KO

## Linear Models

Now we must perform linear model fitting.

```{r}
fitted <- lmFit(y, mm)
head(coef(fitted))

#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(Sample_namesSGm6AKO      - Sample_namesSGm6AKOInput     , levels = mm)


#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)

#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>4 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -4 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("m6A KO SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

write.table(x = DEGenes,
           file = "~/Projects/StressGranules/SGPAPER/SGvsWC/Rajachandran_KO.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)
```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("m6A KO SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("m6A KO SG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
m6A KO Stress granules no longer exhibit a lenght bias.

## Correlation with m6A

```{r}
m6A <- read_excel("~/Projects/StressGranules/SG_Transcriptome/m6A/41467_2022_31358_MOESM26_ESM.xlsx")[,c(4,16,17)]
colnames(m6A)[1] <- "Gene"

DEGenes <- merge(DEGenes, m6A, by="Gene")

ggplot(DEGenes, aes(y=t, x=(m6a_sites))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) + stat_cor(method = "spearman", cor.coef.name = "rho", size=5)
```
And as expected, since the lenght bias dissapeared, we also lost the correlation with the number of m6A sites.


## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}
finagenes <- rbind(upmarkers, downmarkers)


dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:16))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=Sample_names)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(Sample_names)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(Sample_names)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```

## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# SG vs SG

## Linear Models

Now we must perform linear model fitting. Now, I will directly compare the SG with m6A with the SG without m6A.

```{r}
fitted <- lmFit(y, mm)
head(coef(fitted))


#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(Sample_namesSG      - Sample_namesSGm6AKO      , levels = mm)


#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```


## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>1 & DEGenes$adj.P.Val<0.05,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -1 & DEGenes$adj.P.Val<0.05,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```
The differences between them are very small. Can we really say m6A is important for SG enrichment when apparently they are "the same"?

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs SG")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs SG") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

## Correlation with m6A

```{r}
m6A <- read_excel("~/Projects/StressGranules/SG_Transcriptome/m6A/41467_2022_31358_MOESM26_ESM.xlsx")[,c(4,16,17)]
colnames(m6A)[1] <- "Gene"

DEGenes <- merge(DEGenes, m6A, by="Gene")

ggplot(DEGenes, aes(y=t, x=(m6a_sites))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG Enrichment (t-stat)") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) + stat_cor(method = "spearman", cor.coef.name = "rho", size=5)
```

There were barely any differences, and it doesn't seem like they are linked to number of m6A modifications at all.

## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# WC vs WC

## Linear Models

Now we must perform linear model fitting.

```{r}



fitted <- lmFit(y, mm)
head(coef(fitted))


#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(Sample_namesSGInput       - Sample_namesSGm6AKOInput, levels = mm)


#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -3 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("WC vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("WC vs WC")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("WC vs WC") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```

Cells without m6A are "much longer" than their WT counterparts.

## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# Stress vs Unstress (proper)

## Linear Models

Now we must perform linear model fitting. 

```{r Linear Models}

sampnamer <- c(rep("SGkoin",3), rep("Nuc",2), rep("Nucin",2), rep("SGko",3), rep("SG",3), rep("SGin",3))

Stress <- c(rep("Yes",3), rep("No",2), rep("No",2), rep("Yes",3), rep("Yes",3), rep("Yes",3))

Sample <- c(rep("WC",3), rep("Nuc",2), rep("WC",2), rep("SG",3), rep("SG",3), rep("WC",3))

m6A <- c(rep("KO",3), rep("WT",2), rep("WT",2), rep("KO",3), rep("WT",3), rep("WT",3))

mm <- model.matrix(~0+Stress+Sample+m6A)

fitted <- lmFit(y, mm)
head(coef(fitted))


#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(StressYes         - StressNo      , levels = mm)


#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -3 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Stress vs Control")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```

## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("WC vs WC")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("WC vs WC") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
As usual, stressed cells are smaller.

## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# M6A Effect

## Linear Models

Now we must perform linear model fitting.

```{r}

sampnamer <- c(rep("SGkoin",3), rep("Nuc",2), rep("Nucin",2), rep("SGko",3), rep("SG",3), rep("SGin",3))

Stress <- c(rep("Yes",3), rep("No",2), rep("No",2), rep("Yes",3), rep("Yes",3), rep("Yes",3))

Sample <- c(rep("WC",3), rep("Nuc",2), rep("WC",2), rep("SG",3), rep("SG",3), rep("WC",3))

m6A <- c(rep("KO",3), rep("WT",2), rep("WT",2), rep("KO",3), rep("WT",3), rep("WT",3))

mm <- model.matrix(~1+Stress+Sample*m6A)

mm[,2] <- ifelse(mm[,2] == 1, 0.5, -0.5)
mm[,3] <- ifelse(mm[,3] == 1, 0.5, -0.5)
mm[,4] <- ifelse(mm[,4] == 1, 0.5, -0.5)
mm[,5] <- ifelse(mm[,5] == 1, 0.5, -0.5)
mm[,6] <- mm[,5]*mm[,3]*-1


fitted <- lmFit(y, mm)
head(coef(fitted))


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(fitted)
top.table <- topTable(tmp,6, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}
upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>2 & DEGenes$B>5,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -2 & DEGenes$B>5,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("M6A")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))

```
## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

DEGenes$Gene <- row.names(DEGenes)
colnames(genelength) <- c("Length","Gene")
DEGenes <- merge(DEGenes, genelength, by="Gene")

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("m6A")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[9] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("m6A") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

# SG (proper)

## Linear Models

Now we must perform linear model fitting. Once more, we will have 4, one for each matrix we previously designed.

```{r}

sampnamer <- c(rep("SGkoin",3), rep("Nuc",2), rep("Nucin",2), rep("SGko",3), rep("SG",3), rep("SGin",3))

Stress <- c(rep("Yes",3), rep("No",2), rep("No",2), rep("Yes",3), rep("Yes",3), rep("Yes",3))

Sample <- c(rep("Input",3), rep("Nuc",2), rep("Input",2), rep("SG",3), rep("SG",3), rep("Input",3))

m6A <- c(rep("KO",3), rep("WT",2), rep("WT",2), rep("KO",3), rep("WT",3), rep("WT",3))

proc <- c(rep("NO",3), rep("YES",2), rep("NO",2), rep("YES",3), rep("YES",3), rep("NO",3))

metadata <- data.frame(Stress=Stress,Sample=Sample, Metil=m6A, PHOTON=proc)


batch_cols <- c("Stress", "Metil", "PHOTON")
metadata_cols <- c("Sample")
mmbatch <- model.matrix(reformulate(paste0("~0 + ",paste0( batch_cols,collapse = " + "))), data=metadata)
mmkeep <- model.matrix(reformulate(paste0("~0 + ",paste0( metadata_cols,collapse =  "+" ))), data=metadata)
mm <- cbind(mmkeep,mmbatch)
D0 <- DGEList(2^y$E) #senescence_normcounts
D0 <- calcNormFactors(D0)
y <- voom(D0, mm, plot = F)
fit <- lmFit(y, mm)
beta <- fit$coefficients[,-(1:ncol(mmkeep)),drop=FALSE]
#beta <- fit$coefficients
beta[is.na(beta)] <- 0
corrcounts_v1 <- as.matrix(y$E) - beta %*% t(mmbatch)
offset <- apply(corrcounts_v1,1,min) - apply(y$E,1,min)
corrcounts_v1 <- as.data.frame(corrcounts_v1-offset)
corrcounts_v1 <- 2^corrcounts_v1
y <- voom(corrcounts_v1, mm, plot = T)


mm <- model.matrix(~0+Sample)

fitted <- lmFit(y, mm)
head(coef(fitted))


#next we measure contrasts between groups (log-fold change is obtained from these fitted linear models)
contr<- makeContrasts(SampleSG            - SampleInput              , levels = mm)


#next we estimate the contrast for each gene
tmp <- contrasts.fit(fitted, contr)


#use empirical bayes smoothing of standard erros (shrinks extreme standard erros)
tmp <- eBayes(tmp)
top.table <- topTable(tmp, sort.by = "logFC", n = Inf)
DEGenes <- top.table
ndegs=20
```

## Volcano Plot

```{r, results='hide', warning=FALSE, message=FALSE}

write.table(x = DEGenes,
          file = "~/Projects/StressGranules/SGPAPER/RealSGs/Photon.txt", sep="\t", col.names = TRUE, row.names = TRUE, quote=FALSE)

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>3 & DEGenes$B>10,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -3 & DEGenes$B>10,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + 
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  theme(text=element_text(size=15,  family="Arial MS"))


MitoGenes <- fread("~/MitoGenes.txt")[[1]]
MitoData <- DEGenes[rownames(DEGenes) %in% MitoGenes, ]
NucGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]
NucData <- DEGenes[rownames(DEGenes) %in% NucGenes, ]
(m <- ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=B)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=NucData, aes(x=logFC, y=B), color="#CB643F", size=4) +
  geom_point(data=MitoData, aes(x=logFC, y=B), color="#73937E", size=4) +
  ylab("B-Statistic") +
  xlab(expression(Log[2]~"Fold-Change")) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),   # remove major gridlines
    panel.grid.minor = element_blank(),   # remove minor gridlines
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))


png("~/Projects/StressGranules/SGPAPER/Images/PNG/RajaMito.png", width = 7, height=6, units = "in", res=300)
m
dev.off()

```

# MA Plot

```{r}
# MA plot: mean expression (x) vs logFC (y), same highlighting scheme as the volcano

ggplot(data=DEGenes,
                 aes(x=AveExpr,
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=AveExpr, y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=AveExpr, y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  theme_bw() + ylab(expression(Log[2]~"Fold-Change")) + xlab("Average Expression (Log2 CPM)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("T. Matheny Both - MA Plot") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,], mapping=aes(x=AveExpr, y=logFC, label=row.names(upmarkers)[1:ndegs]), size=4, point.padding=-0.5, color="#A10702", max.overlaps=20, fontface="italic") +
  geom_label_repel(downmarkers[1:ndegs,], mapping=aes(x=AveExpr, y=logFC, label=row.names(downmarkers)[1:ndegs]), size=4, point.padding=-0.5, color="#03256C", max.overlaps=20, fontface="italic") +
  geom_point(data=upmarkers, aes(x=AveExpr, y=logFC), color="#A10702", size=4) +
  geom_point(data=downmarkers, aes(x=AveExpr, y=logFC), color="#03256C", size=4) +
  theme(text=element_text(size=15, family="Arial MS"))

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=AveExpr,
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=AveExpr, y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=AveExpr, y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) +
  xlab("Average Expression (Log2 CPM)") + geom_density2d(color = "#800020", size = 1.2) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+
  stat_cor(data = MitoData, aes(x=AveExpr, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#73937E", label.y.npc = 0.1) +
  stat_cor(data = NucData, aes(x=AveExpr, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#CB643F", label.y.npc = 0.2) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/RajaMito_MA.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()
```

## MA Plot

```{r}
MitoGenes <- fread("~/MitoGenes.txt")[[1]]

genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)
genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")
genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()
genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")

genelength$Cromo <- ifelse(genelength$Gene %in% MitoGenes, "Mito", "Other")

# Apply the length-based weight formula
genelength$S <- ifelse(
  genelength$Cromo == "Mito",
  (genelength$Length)^(1/3) * 3.23e-11 + 1.17e-9,   # Formula for mitochondrial RNAs
  (genelength$Length)^(1/3) * 6.09e-11 + 3.88e-10    # Formula for other RNAs
)
DEGenes$Gene <- row.names(DEGenes)
DEGenes <- merge(DEGenes, genelength, by="Gene")
MitoGenes <- fread("~/MitoGenes.txt")[[1]]
MitoData <- DEGenes[DEGenes$Gene %in% MitoGenes, ]
NucGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]
NucData <- DEGenes[DEGenes$Gene %in% NucGenes, ]

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=log10(Length),
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=log10(Length), y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=log10(Length), y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) + geom_density2d(color = "#800020", size = 1.2) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+
  stat_cor(data = MitoData, aes(x=log10(Length), y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#73937E", label.y.npc = 0.1) +
  stat_cor(data = NucData, aes(x=log10(Length), y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#CB643F", label.y.npc = 0.2) +
  scale_x_continuous(
    breaks = log10(c(10, 100, 1000, 10000)),  # tick positions (in log10 scale)
    labels = c("10", "100", "1000", "10000") # what you want to show
  )+
  xlab("RNA Length (log10)") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/RajaMito_MA_length.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()

# Publication-style version, stripped of axis titles, matching your "m" plot
(ma <- ggplot(data=DEGenes,
                 aes(x=(S),
                     y=logFC)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=MitoData, aes(x=(S), y=logFC), color="#73937E", size=4) +
  geom_point(data=NucData, aes(x=(S), y=logFC), color="#CB643F", size=4) +
  geom_hline(yintercept = 0, linetype="dashed", color="grey40") +
  ylab(expression(Log[2]~"Fold-Change")) + geom_density2d(color = "#800020", size = 1.2) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+
  stat_cor(data = MitoData, aes(x=S, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#73937E", label.y.npc = 0.1) +
  stat_cor(data = NucData, aes(x=S, y=logFC), method = "spearman",
           cor.coef.name = "rho", size=7, color = "#CB643F", label.y.npc = 0.2) +
  xlab("S") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25, family="Arial MS")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/RajaMito_MA_S.png", width = 7, height=6, units = "in", res=300)
ma
dev.off()
```


# Violins

```{r}
## Signature Violin (Mito/Nuc) - SG vs WC

library(data.table)
library(dplyr)
library(ggplot2)
library(rstatix)

# Restrict to just the SG and WC samples (exclude the Nuc fraction samples)
sgwc_idx <- which(Sample %in% c("SG", "Input"))

expression.df <- y$E[, sgwc_idx]
cond <- factor(Sample[sgwc_idx], levels = c("Input", "SG"))

custom_palette <- c(
  "Input" = "#399e5a",
  "SG" = "#769fca"
)

# ── 1. Gene sets already loaded earlier in the script (MitoGenes, NucGenes) ──
mito_genes_present <- MitoGenes[MitoGenes %in% rownames(expression.df)]
nuc_genes_present  <- NucGenes[NucGenes %in% rownames(expression.df)]
cat(length(mito_genes_present), "Mito genes found in expression matrix\n")
cat(length(nuc_genes_present), "Nuc genes found in expression matrix\n")

# ── 2. Rank all genes per sample (1 = least expressed) ───────────────────────
rank_matrix <- apply(expression.df, 2, function(x) rank(x, ties.method = "average"))
rank_matrix <- as.data.frame(rank_matrix)

# ── 3. Compute a rank-sum score per sample, one gene set at a time ───────────
compute_score <- function(genes, rank_matrix, n_total) {
  colSums(rank_matrix[genes, , drop = FALSE]) / n_total
}

mito_score <- compute_score(mito_genes_present, rank_matrix, nrow(expression.df))
nuc_score  <- compute_score(nuc_genes_present, rank_matrix, nrow(expression.df))

# ── 4. Assemble plot dataframe, long format across both gene sets ────────────
score_df <- rbind(
  data.frame(Sample = colnames(expression.df), Score = as.numeric(mito_score),
             Condition = cond, GeneSet = "Mito"),
  data.frame(Sample = colnames(expression.df), Score = as.numeric(nuc_score),
             Condition = cond, GeneSet = "Nuc")
)

# ── 5. Statistics per gene set: one-way ANOVA (equivalent to t-test, 2 groups) ──
stat_labels <- score_df %>%
  group_by(GeneSet) %>%
  group_modify(~{
    aov_res  <- aov(Score ~ Condition, data = .x)
    aov_tidy <- summary(aov_res)
    p_raw  <- aov_tidy[[1]][["Pr(>F)"]][1]
    p_adj  <- p.adjust(p_raw, method = "BH")
    ss_between <- aov_tidy[[1]][["Sum Sq"]][1]
    ss_total   <- sum(aov_tidy[[1]][["Sum Sq"]])
    eta_sq     <- ss_between / ss_total
    cohens_f   <- sqrt(eta_sq / (1 - eta_sq))
    data.frame(
      label = paste0("Cohen's f = ", round(cohens_f, 3),
                      "\nadj. p = ", formatC(p_adj, format = "e", digits = 2))
    )
  })

# ── 6. Plot ───────────────────────────────────────────────────────────────────
vilsigplot <- ggplot(score_df, aes(x = Condition, y = Score, fill = Condition)) +
  geom_violin(colour = "black", alpha = 0.6, trim = FALSE) +
  geom_point(
    position = position_jitter(width = 0.1),
    alpha    = 0.7,
    size     = 2
  ) +
  stat_summary(fun = mean,   geom = "point", shape = 18,
               size = 4, colour = "blue") +
  stat_summary(fun = median, geom = "point", shape = 17,
               size = 4, colour = "blue") +
  geom_text(
    data  = stat_labels,
    aes(x = -Inf, y = Inf, label = label),
    hjust = -0.05, vjust = 1.2,
    size  = 4, fontface = "bold", lineheight = 1.1,
    inherit.aes = FALSE
  ) +
  facet_wrap(~GeneSet, scales = "free_y") +
  scale_fill_manual(values = custom_palette) +
  labs(
    x     = "Fraction",
    y     = "Signature Score (Rank Sum)",
    title = "Mitochondrial vs Nuclear-Localised Gene Scores"
  ) +
  theme_bw() +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x     = element_text(size = 12),
    strip.text      = element_text(size = 12, face = "bold"),
    legend.position = "none",    
    text = element_text(size = 20, family = "Arial MS", face = "bold")
  )

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Raja_MitoSigs.png", width = 14, height=6, units = "in", res=300)
vilsigplot
dev.off()

vilsigplot
```


## Volcano Plot with Gene Length

```{r, message=FALSE, warning=FALSE, results='hide'}

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(aes(color=log10(Length)))+
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs WC")+ xlab(expression(Log[2] ~ Fold ~ Change)) + ylab("B-Statistic") +
  theme(plot.title = element_text(hjust = 0.5))  +
  scale_colour_gradient(low = "black", high = "red")+ theme(text=element_text(family = "Quicksand"))


smooth.SG <- lm(t~log10(Length), DEGenes)
liner.SG <- rstandard(smooth.SG)
DEGenes <-cbind(DEGenes, liner.SG)
colnames(DEGenes)[11] <- "Residuals"
row.names(DEGenes) <- DEGenes$Gene
linemarkers <- as.data.frame(DEGenes[abs(DEGenes$Residuals)>4,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(DEGenes$t, log10(DEGenes$Length), method = "spearman")



ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(color="#CCCCCC", alpha=1) +   theme_bw()+ geom_density_2d()+
 ylab("SG vs WC") + xlab(expression(RNA ~ Length ~ Log[10] ~ Base ~ Pairs)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho",size=7)
```
## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression. This helps to see if the expression differences are robust, and not just driven by one sample.

```{r, fig.height=12, fig.width=12}
finagenes <- rbind(upmarkers, downmarkers)


dfforexpression <- dfforpca
samplecodes <- data.frame(samples=row.names(dfforexpression), purity=c(1:16))
sampletype <- data.frame(samples=row.names(dfforexpression), sampletype=Sample_names)
dfforexpression <- t(dfforexpression)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]


winnyplot <- ggplot(genesforexp, aes(x=purity, y=expression, fill=sampletype, colour=sampletype)) + geom_bar(position="dodge", stat="identity", alpha=0.5) + facet_wrap(.~ Gene, ncol=5)
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = custom_palette, breaks = unique(Sample_names)) + scale_colour_manual(name = "Sample Type", values = custom_palette, breaks = unique(Sample_names)) 
winnyplot <- winnyplot + theme_bw() + xlab("Samples (SG to Bulk Rna)") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```

## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.001

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalised Enrichment Score (NES)",
       title= "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")


gsea_plot_hall+
  theme(text=element_text(size=25,  family="Arial MS"))
```

### For Mito

```{r}
# --- Rank genes by t-statistic as usual ---
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]
vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol

# --- Define MitoGenes as a custom gene set ---
MitoGenes <- fread("~/MitoGenes.txt")[[1]]
custom.gsets <- list(MitoGenes = toupper(MitoGenes))

# --- Run fgsea ---
mito.res <- fgsea(pathways = custom.gsets,
                  stats = vectorranks)

mito.plot.base <- plotEnrichment(custom.gsets[["MitoGenes"]], vectorranks)
curve.data <- mito.plot.base$data

# --- key fix: match the curve's internal descending order ---
ord <- order(vectorranks, decreasing = TRUE)
ranked.names <- names(vectorranks)[ord]
gene.positions <- which(ranked.names %in% custom.gsets[["MitoGenes"]])

rug.top    <- min(curve.data$y) - 0.03
rug.bottom <- rug.top - 0.04

y.min <- rug.bottom - 0.03
y.max <- max(curve.data$y) + 0.05

(mito.plot <- ggplot(curve.data, aes(x = x, y = y)) +
  geom_area(fill = "#73937E", alpha = 0.4) +
  geom_line(color = "#73937E", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_segment(data = data.frame(x = gene.positions),
               aes(x = x, xend = x, y = rug.top, yend = rug.bottom),
               color = "#73937E", alpha = 0.6, linewidth = 0.9,
               inherit.aes = FALSE) +
  annotate("text", x = 9000, y = 0.5,
           label = paste0("NES = ", round(mito.res$NES, 2),
                          "\nadj.p ≈ ", round(mito.res$padj, 4)),
           size = 5, family = "Arial MS", fontface = "bold") +
  coord_cartesian(ylim = c(y.min, y.max), clip = "off") +
  theme_bw() +
  theme(text = element_text(size = 25, family = "Arial MS"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/MitoGSEA_Raja.png", width=7, height=6, units="in", res=300)
mito.plot
dev.off()
```
### For Nuc

```{r}
# --- Rank genes by t-statistic as usual ---
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]
vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol

# --- Define MitoGenes as a custom gene set ---
MitoGenes <- fread("~/rnalocate_mito_filt.csv")[[1]]
custom.gsets <- list(MitoGenes = toupper(MitoGenes))

# --- Run fgsea ---
mito.res <- fgsea(pathways = custom.gsets,
                  stats = vectorranks)

mito.plot.base <- plotEnrichment(custom.gsets[["MitoGenes"]], vectorranks)
curve.data <- mito.plot.base$data

# --- key fix: match the curve's internal descending order ---
ord <- order(vectorranks, decreasing = TRUE)
ranked.names <- names(vectorranks)[ord]
gene.positions <- which(ranked.names %in% custom.gsets[["MitoGenes"]])

rug.top    <- min(curve.data$y) - 0.03
rug.bottom <- rug.top - 0.04

y.min <- rug.bottom - 0.03
y.max <- max(curve.data$y) + 0.05

(mito.plot <- ggplot(curve.data, aes(x = x, y = y)) +
  geom_area(fill = "#CB643F", alpha = 0.4) +
  geom_line(color = "#CB643F", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_segment(data = data.frame(x = gene.positions),
               aes(x = x, xend = x, y = rug.top, yend = rug.bottom),
               color = "#CB643F", alpha = 0.6, linewidth = 0.9,
               inherit.aes = FALSE) +
  annotate("text", x = 2500, y = -0.4,
           label = paste0("NES = ", round(mito.res$NES, 2),
                          "\nadj.p = ", round(mito.res$padj, 4)),
           size = 5, family = "Arial MS", fontface = "bold") +
  coord_cartesian(ylim = c(y.min, y.max), clip = "off") +
  theme_bw() +
  theme(text = element_text(size = 25, family = "Arial MS"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/NucGSEA_Raja.png", width=7, height=6, units="in", res=300)
mito.plot
dev.off()
```

# Session Info

```{r}
sessionInfo()
```