---
title: "Uncouvering the Stressome"
subtitle: "A Computational Aproach to Define a Stress Granule Signature and its Implication in Cancer"
author: 
  - name: "Alexandre Afonso^[Disease Transcriptomics Group, Instituto de Medicina Molecular, Faculdade de Medicina da Universidade de Lisboa]"
    email: alexandreafonso@medicina.ulisboa.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

In this document I will be presenting and compiling all the hardcore computational parts of my PhD Project. For simplicity's sake, the research and general scientific knowledge *a priori* is provided on another document. The same holds true for all the quality control analysis, which is also kept elsewhere. 

## Libraries

All quality control was performed using **FASTQC**. Alignment was performed using **STAR** (parameters on a saved .sh file) to the human genome with **Gencode V.41 annotation**. To counts the reads **featureCounts** was used, and **DEA** was completed using **limma-voom**.

```{r preparation, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(rstatix)


genelength <- read.csv("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/TranscriptLenght.txt", header=T)

genelength <- subset(genelength, Ensembl.Canonical==1)
genelength <- subset(genelength, Gene.name!="")

genelength <- genelength %>%
  group_by(Gene.name) %>%
  slice(which.max(Transcript.length..including.UTRs.and.CDS.)) %>%
  ungroup()

genelength <- genelength[,-c(2,3)]
colnames(genelength) <- c("Length", "Gene")
```


# Figures

## Figure 1

### A

```{r fig.height=6, fig.width=14}
stress_t <- fread("~/Projects/StressGranules/SGPAPER/SGvsWC/Akhong2017.txt")[,c(1,4)]
colnames(stress_t) <- c("Gene", "s_t")
stressoma_t <- fread("~/Projects/StressGranules/SGPAPER/SGvsWC/Somasekharan.txt")[,c(1,4)]
colnames(stressoma_t) <- c("Gene", "so_t")

colnames(genelength) <- c("Length", "Gene")
fig1 <- merge(stress_t, stressoma_t, by="Gene")
fig1 <- merge(fig1, genelength, by="Gene")




smooth.SG <- lm(s_t~log10(Length), fig1)

# Plot the graph
A1 <- ggplot(fig1, aes(y = s_t, x = log10(Length))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2) +
  theme_bw() +
  scale_x_continuous(
    breaks = log10(c(10, 100, 1000, 10000)),  # tick positions (in log10 scale)
    labels = c("10", "100", "1000", "10000") # what you want to show
  )+
  ylab("DGE SG-WC (t-stat)") +
theme(axis.title.y = element_text(margin = margin(r = 10))) + 
  xlab(expression(paste("RNA Length (log"[10]*" (nt))"))) + # Subscript "10"
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("Differential Centrifugation") + 
  theme(text=element_text(size=20,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7) +
  theme(plot.title = element_text(hjust = 0.5)) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

smooth.SG <- lm(so_t~log10(Length), fig1)

# Plot the graph
A2 <- ggplot(fig1, aes(y = so_t, x = log10(Length))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2) +
  theme_bw() +
 ylab("DGE SG-WC (t-stat)") +
theme(axis.title.y = element_text(margin = margin(r = 10))) + 
  xlab(expression(paste("RNA Length (log"[10]*" (nt))"))) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),   # Remove major gridlines
    panel.grid.minor = element_blank(),   # Remove minor gridlines
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  scale_x_continuous(
    breaks = log10(c(10, 100, 1000, 10000)),  # tick positions (in log10 scale)
    labels = c("10", "100", "1000", "10000") # what you want to show
  )+
  ggtitle("Proximity Labelling") + 
  theme(
    text = element_text(size = 20, family = "Arial MS"),
    plot.title = element_text(hjust = 0.5)
  ) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 7)

(A <- ggarrange(A1, A2, ncol = 2, labels = c("A","", ""), font.label = list(size=25), widths = c(0.45,0.45)))

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure1/FigureA.png", width = 14, height=6, units = "in", res=300)
A
dev.off()
```

### B

```{r fig.width=28, fig.height=28}
PIC <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/PIC.txt")[,c(1,4)]
colnames(PIC) <- c("Gene","PIC")
artr <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Artr.txt")[,c(1,4)]
colnames(artr) <- c("Gene","artr")
padron_hs <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/APadron_Heatshock.txt")[,c(1,4)]
colnames(padron_hs) <- c("Gene","pad_hs")
padron_hip <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/APadron_Hippuristanol.txt")[,c(1,4)]
colnames(padron_hip) <- c("Gene","pad_hip")
soma <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Somasekharan.txt")[,c(1,4)]
colnames(soma) <- c("Gene","soma")
shenx <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Shen_X.txt")[,c(1,4)]
colnames(shenx) <- c("Gene","shenx")
sheny <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Shen_Y.txt")[,c(1,4)]
colnames(sheny) <- c("Gene","sheny")
renu2ar <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Ren_U2Ar.txt")[,c(1,4)]
colnames(renu2ar) <- c("Gene","renu2ar")
renhekar <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Ren_HEKAr.txt")[,c(1,4)]
colnames(renhekar) <- c("Gene","renhekar")
renhekso <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Ren_HEKSo.txt")[,c(1,4)]
colnames(renhekso) <- c("Gene","renhekso")
rajawt <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Rajachandran_WT.txt")[,c(1,4)]
colnames(rajawt) <- c("Gene","rajawt")
rajako <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Rajachandran_KO.txt")[,c(1,4)]
colnames(rajako) <- c("Gene","rajako")
fancias <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/FANCI_AS.txt")[,c(1,4)]
colnames(fancias) <- c("Gene","fancias")
fancihs <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/FANCI_HS.txt")[,c(1,4)]
colnames(fancihs) <- c("Gene","fancihs")
fancios <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/FANCI_OS.txt")[,c(1,4)]
colnames(fancios) <- c("Gene","fancios")
fanciuv <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/FANCI_UV.txt")[,c(1,4)]
colnames(fanciuv) <- c("Gene","fanciuv")
alzheclip <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/AlzheClip.txt")[,c(1,4)]
colnames(alzheclip) <- c("Gene","alzheclip")
alzhdc<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/AlzhDC.txt")[,c(1,4)]
colnames(alzhdc) <- c("Gene","alzhdc")
akhong<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Akhong2017.txt")[,c(1,4)]
colnames(akhong) <- c("Gene","akhong")
tmat19<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/TMatheny2019.txt")[,c(1,4)]
colnames(tmat19) <- c("Gene","tmat19")
tmat_wtarpa<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/TMatheny2021_WTARPA.txt")[,c(1,4)]
colnames(tmat_wtarpa) <- c("Gene","tmat_wtarpa")
tmat_wtsopa<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/TMatheny2021_WTSOPA.txt")[,c(1,4)]
colnames(tmat_wtsopa) <- c("Gene","tmat_wtsopa")
tmat_kosopa<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/TMatheny2021_KOSOPA.txt")[,c(1,4)]
colnames(tmat_kosopa) <- c("Gene","tmat_kosopa")
tmat_gfsog3<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/TMatheny2021_GFSOG3.txt")[,c(1,4)]
colnames(tmat_gfsog3) <- c("Gene","tmat_gfsog3")
viadevaia<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/VIadevaia.txt")[,c(1,4)]
colnames(viadevaia) <- c("Gene","viadevaia")
curdy<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/TcellSG.txt")[,c(1,4)]
colnames(curdy) <- c("Gene","curdy")
kudrinwt<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Kudrin_WT.txt")[,c(1,4)]
colnames(kudrinwt) <- c("Gene","kudrinwt")
kudrinko<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Kudrin_KO.txt")[,c(1,4)]
colnames(kudrinko) <- c("Gene","kudrinko")
timowtct<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Timo_WTCT.txt")[,c(1,4)]
colnames(timowtct) <- c("Gene","timowtct")
timowtex<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Timo_WTEX.txt")[,c(1,4)]
colnames(timowtex) <- c("Gene","timowtex")
timoalsct<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Timo_ALSCT.txt")[,c(1,4)]
colnames(timoalsct) <- c("Gene","timoalsct")
timoalsex<- fread("../Projects/StressGranules/SGPAPER/SGvsWC/Timo_ALSEX.txt")[,c(1,4)]
colnames(timoalsex) <- c("Gene","timoalsex")
fig2 <- merge(PIC, artr, by="Gene", all=T)
fig2 <- merge(fig2, padron_hs, by="Gene")
fig2 <- merge(fig2, padron_hip, by="Gene")
fig2 <- merge(fig2, soma, by="Gene")
fig2 <- merge(fig2, shenx, by="Gene")
fig2 <- merge(fig2, sheny, by="Gene")
fig2 <- merge(fig2, renu2ar, by="Gene")
fig2 <- merge(fig2, renhekar, by="Gene")
fig2 <- merge(fig2, renhekso, by="Gene")
fig2 <- merge(fig2, rajawt, by="Gene")
fig2 <- merge(fig2, rajako, by="Gene")
fig2 <- merge(fig2, fancias, by="Gene")
fig2 <- merge(fig2, fancihs, by="Gene")
fig2 <- merge(fig2, fancios, by="Gene")
fig2 <- merge(fig2, fanciuv, by="Gene")
fig2 <- merge(fig2, alzheclip, by="Gene")
fig2 <- merge(fig2, alzhdc, by="Gene")
fig2 <- merge(fig2, akhong, by="Gene")
fig2 <- merge(fig2, tmat19, by="Gene")
fig2 <- merge(fig2, tmat_wtarpa, by="Gene")
fig2 <- merge(fig2, tmat_wtsopa, by="Gene")
fig2 <- merge(fig2, tmat_kosopa, by="Gene")
fig2 <- merge(fig2, tmat_gfsog3, by="Gene")
fig2 <- merge(fig2, viadevaia, by="Gene")
fig2 <- merge(fig2, curdy, by="Gene")
fig2 <- merge(fig2, kudrinwt, by="Gene")
fig2 <- merge(fig2, kudrinko, by="Gene")
fig2 <- merge(fig2, timowtct, by="Gene")
fig2 <- merge(fig2, timowtex, by="Gene")
fig2 <- merge(fig2, timoalsct, by="Gene")
fig2 <- merge(fig2, timoalsex, by="Gene")
fig2 <- merge(fig2, genelength, by="Gene")


#Heatmap Itself

fig2$Length <- log10(fig2$Length)
colnames(fig2) <- c("Gene", "Honda 2021", "Xiao 2024", "Padrón 2019 (HS)", "Padrón 2019 (Hip)", "Somasekharan 2020", "Shen 2022 (X)", "Shen 2022 (Y)", "Ren 2023 (U2 AR)", "Ren 2023 (Hek AR)", "Ren 2023 (Hek SO)", "Rajachandran 2024 (WT)", "Rajachandran 2024 (KO)", "Zhou 2024 (AS)", "Zhou 2024 (HS)", "Zhou 2024 (OS)", "Zhou 2024 (UV)", "Sato 2023 (eClip)", "Sato 2023 (DC)", "Khong 2017", "Matheny 2019", "Matheny 2021 (WT AR)", "Matheny 2021 (WT Sor)", "Matheny 2021 (KO Sor)", "Matheny 2021 (GFP Sor)", "Iadevaia 2022", "Curdy 2023", "Kudrin 2024 (WT)", "Kudrin 2024 (KO)", "Timoteo 2024 (WT)", "Timoteo 2024 (KD)", "Timoteo 2024 (FUSMUT)", "Timoteo 2024 (FUSMUT KD)", "RNA Length")

fig2_cor <- cor(fig2[,-1],method="spearman", use = "complete.obs")

fig2_cor_melted <- reshape2::melt(fig2_cor, varnames = c("Var1", "Var2"))

(pl1 <- fig2_cor_melted %>%
  arrange(Var1) %>%
    filter(row_number() <= which(Var1 == Var2)) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "#5e5e5e", size = 0.5) +
  #geom_text(aes(label = ifelse(is.na(value), "", sprintf("%.2f", value))), size = 7) +
  scale_fill_gradient2(low = "#729EA1", mid = "white", high = "#DB5375", limit = c(-1, 1)) +
  labs(fill = "Spearman's Rho") +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(angle = 0, hjust = 1, size = 20, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 20, family = "Arial MS", face = "bold"),
    legend.position = "top"
  )+
  ylab("") + xlab("") +
  scale_y_discrete(limits = sort(unique(fig2_cor_melted$Var2), decreasing = TRUE)))


#Method
df <- data.frame(
  x = (c("pic", "artr", "pl", "pl", "pl", "pl", "pl", "pl", "pl", "pl", "photon", "photon", "fanci", "fanci", "fanci", "fanci", "eclip", "dc", "dc", "dc","dc", "dc","dc", "dc", "dc", "dc", "dc", "dc","dc", "dc", "dc", "dc",  "length")),
  colour = c("Honda 2021", "Xiao 2024", "Padrón 2019", "Padrón 2019", "Somasekharan 2020", "Shen 2022", "Shen 2022", "Ren 2023", "Ren 2023", "Ren 2023", "Rajachandran 2024", "Rajachandran 2024", "Zhou 2024","Zhou 2024","Zhou 2024","Zhou 2024", "Sato 2023","Sato 2023", "Khong 2017", "Matheny 2019", "Matheny 2021","Matheny 2021","Matheny 2021","Matheny 2021", "Iadevaia 2022", "Curdy 2023", "Kudrin 2024", "Kudrin 2024", "Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024", "RNA Length"),
  xpos = (c("Honda", "Xiao", "Padron_1", "Padron_2", "Somasekharan", "Shen_1", "Shen_2", "Ren_1", "Ren_2", "Ren_3", "Raja_1", "Raja_2", "Zhou_1", "Zhou_2", "Zhou_3", "Zhou_4", "Sato_1", "Sato_2", "Khong", "Matheny19","Matheny21_1","Matheny21_2","Matheny21_3","Matheny21_4", "Iadevaia", "Curdy", "Kudrin_1", "Kudrin_2", "Timo_1", "Timo_2", "Timo_3", "Timo_4", "Length"))
)
df$x <- factor(df$x, levels = unique(df$x))
df$xpos <- factor(df$xpos, levels = (df$xpos))

colour_vector <- c(
  "pic" = '#FFD1DC',
  "artr" = '#FFB185',
  "pl" = '#FFDAC1',
  "photon" = '#FF9AA2',
  "fanci" = '#FFC3A0',
  "eclip" ='#FFB347',
  "dc" = '#FFE5B4',
  "length" = 'white'
)


lebel_vector <- c(
  "pic" = "PIC",
  "artr" = "ARTR-Seq",
  "pl" = "P. Labelling",
  "photon" = 'PHOTON',
  "fanci" = 'FANCI',
  "eclip" ='eCLIP-Seq',
  "dc" = 'D. Centrifugation',
  "length" = 'RNA Length'
)

(pl2 <- ggplot(df, aes(x = xpos, y = 1, fill = x)) +
  geom_tile() +
  theme_void() +
  labs(fill = "SG Purification \nMethod") +
  scale_fill_manual(
    values = colour_vector,labels = lebel_vector,
    breaks = names(lebel_vector)
  ) +
  scale_x_discrete(
    labels = c("Honda 2021", "Xiao 2024", "Padrón 2019", "Padrón 2019", "Somasekharan 2020", "Shen 2022", "Shen 2022", "Ren 2023", "Ren 2023", "Ren 2023", "Rajachandran 2024", "Rajachandran 2024", "Zhou 2024","Zhou 2024","Zhou 2024","Zhou 2024","Sato 2023","Sato 2023", "Khong 2017", "Matheny 2019", "Matheny 2021","Matheny 2021","Matheny 2021","Matheny 2021", "Iadevaia 2022", "Curdy 2023", "Kudrin 2024", "Kudrin 2024","Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024", "RNA Length")
  ) +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1),
    text = element_text(size = 20, family = "Arial MS", color = "#505050", face = "bold")
  ) +
  ylab("") + xlab(""))


#Stress  
df2 <-data.frame(x = (c("oxi", "oxi", "hs", "hip", "oxi","oxi", "oxi", "oxi", "oxi", "sor", "oxi", "oxi", "oxi", "hs", "sor", "uv", "oxi", "oxi", "oxi", "oxi", "oxi", "sor", "sor", "sor", "viral", "tcellactivation", "oxi", "oxi","oxi", "oxi","oxi", "oxi", "length")),
  colour = c("Honda 2021", "Xiao 2024", "Padrón 2019", "Padrón 2019", "Somasekharan 2020", "Shen 2022", "Shen 2022", "Ren 2023", "Ren 2023", "Ren 2023", "Rajachandran 2024", "Rajachandran 2024", "Zhou 2024","Zhou 2024","Zhou 2024","Zhou 2024", "Sato 2023","Sato 2023", "Khong 2017", "Matheny 2019", "Matheny 2021","Matheny 2021","Matheny 2021","Matheny 2021", "Iadevaia 2022", "Curdy 2023", "Kudrin 2024", "Kudrin 2024", "Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024", "RNA Length"),
  xpos = (c("Honda", "Xiao", "Padron_1", "Padron_2", "Somasekharan", "Shen_1", "Shen_2", "Ren_1", "Ren_2", "Ren_3", "Raja_1", "Raja_2", "Zhou_1", "Zhou_2", "Zhou_3", "Zhou_4", "Sato_1", "Sato_2", "Khong", "Matheny19","Matheny21_1","Matheny21_2","Matheny21_3","Matheny21_4", "Iadevaia", "Curdy", "Kudrin_1", "Kudrin_2", "Timo_1", "Timo_2", "Timo_3", "Timo_4", "Length"))
)

df2$x <- factor(df2$x, levels = unique(df2$x))
df2$xpos <- factor(df2$xpos, levels = (df2$xpos))

colour_vector <- c(
  "oxi" = '#AEC6CF',
  "hs" = '#BFD8B8',
  "hip" = '#A2D2FF',
  "sor" = '#B5EAD7',
  "uv" = '#C7CEEA',
  "viral" ='#C1F0F6',
  "tcellactivation" = '#B5D3FF',
  "length" = 'white'
)


lebel_vector <- c(
  "oxi" = 'Oxidative',
  "hs" = 'Heat shock',
  "hip" = 'Hippuristanol',
  "sor" = "Hyperosmotic",
  "uv" = 'UV Irradiation',
  "viral" ='Viral Infection',
  "tcellactivation" = 'T Cell Activation',
  "length" = 'RNA Length'
)

(pl3 <- ggplot(df2, aes(x = xpos, y = 1, fill = x)) +
  geom_tile() +
  theme_void() +
  labs(fill = "Stress Stimuli") +
  scale_fill_manual(
    values = colour_vector,labels = lebel_vector,
    breaks = names(lebel_vector)
  ) +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_blank(),
    text = element_text(size = 20, family = "Arial MS", color = "#505050", face = "bold")
  ) +
  ylab("") + xlab(""))

#Target 
df3 <-data.frame(x = (c("g3bp", "g3bp", "eif4a", "eif4a", "g3bp", "ddx3x", "ddx3y", "g3bp", "g3bp", "g3bp", "eifeta", "eifeta", "g3bp", "g3bp", "g3bp", "g3bp", "g3bp", "g3bp", "g3bp", "none", "pabpc1", "pabpc1", "pabpc1", "g3bp", "g3bp", "g3bp", "g3bp", "g3bp","g3bp", "g3bp", "g3bp", "g3bp", "length")), 
  colour = c("Honda 2021", "Xiao 2024", "Padrón 2019", "Padrón 2019", "Somasekharan 2020", "Shen 2022", "Shen 2022", "Ren 2023", "Ren 2023", "Ren 2023", "Rajachandran 2024", "Rajachandran 2024", "Zhou 2024","Zhou 2024","Zhou 2024","Zhou 2024", "Sato 2023","Sato 2023", "Khong 2017", "Matheny 2019", "Matheny 2021","Matheny 2021","Matheny 2021","Matheny 2021", "Iadevaia 2022", "Curdy 2023", "Kudrin 2024", "Kudrin 2024", "Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024", "RNA Length"),
  xpos = (c("Honda", "Xiao", "Padron_1", "Padron_2", "Somasekharan", "Shen_1", "Shen_2", "Ren_1", "Ren_2", "Ren_3", "Raja_1", "Raja_2", "Zhou_1", "Zhou_2", "Zhou_3", "Zhou_4", "Sato_1", "Sato_2", "Khong", "Matheny19","Matheny21_1","Matheny21_2","Matheny21_3","Matheny21_4", "Iadevaia", "Curdy", "Kudrin_1", "Kudrin_2", "Timo_1", "Timo_2", "Timo_3", "Timo_4", "Length"))
)
df3$x <- factor(df3$x, levels = unique(df3$x))
df3$xpos <- factor(df3$xpos, levels = (df3$xpos))

colour_vector <- c(
  "g3bp" = '#B388EB',
  "eif4a" = '#DA70D6',
  "ddx3x" = '#9C38FA',
  "ddx3y" = '#D291BC',
  "eifeta" = '#957DAD',
  "none" ='#D8C8E5',
  "pabpc1" = '#B39EB5',
  "length" = 'white'
)



lebel_vector <-c(
  "g3bp" = 'G3BP',
  "eif4a" = 'eIF4A',
  "ddx3x" = 'DDX3X',
  "ddx3y" = 'DDX3Y',
  "eifeta" = 'eIFeta',
  "none" ='None',
  "pabpc1" = 'PABPC1',
  "length" = 'RNA Length'
)



(pl4 <- ggplot(df3, aes(x = xpos, y = 1, fill = x)) +
  geom_tile() +
  theme_void() +
  labs(fill = "Isolation Target") +
  scale_fill_manual(
    values = colour_vector,labels = lebel_vector,
    breaks = names(lebel_vector)
  ) +
  scale_x_discrete(
    labels = NULL
  ) +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1),
    text = element_text(size = 20, family = "Arial MS", color = "#505050", face = "bold")
  ) +
  ylab("") + xlab(""))

#Cell
df4 <-data.frame(x = (c("hela", "hela", "hek", "hek", "pc3", "hela", "hela", "u2os", "hek", "hek", "hela", "hela", "hela", "hela", "hela", "hela", "shsy5y","shsy5y", "u2os", "u2os","u2os", "u2os","u2os", "u2os", "u2os", "tcells", "hela", "hela", "sknbe","sknbe","sknbe","sknbe", "length")), 
  colour = c("Honda 2021", "Xiao 2024", "Padrón 2019", "Padrón 2019", "Somasekharan 2020", "Shen 2022", "Shen 2022", "Ren 2023", "Ren 2023", "Ren 2023", "Rajachandran 2024", "Rajachandran 2024", "Zhou 2024","Zhou 2024","Zhou 2024","Zhou 2024", "Sato 2023","Sato 2023", "Khong 2017", "Matheny 2019", "Matheny 2021","Matheny 2021","Matheny 2021","Matheny 2021", "Iadevaia 2022", "Curdy 2023", "Kudrin 2024", "Kudrin 2024", "Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024", "RNA Length"),
  xpos = (c("Honda", "Xiao", "Padron_1", "Padron_2", "Somasekharan", "Shen_1", "Shen_2", "Ren_1", "Ren_2", "Ren_3", "Raja_1", "Raja_2", "Zhou_1", "Zhou_2", "Zhou_3", "Zhou_4", "Sato_1", "Sato_2", "Khong", "Matheny19","Matheny21_1","Matheny21_2","Matheny21_3","Matheny21_4", "Iadevaia", "Curdy", "Kudrin_1", "Kudrin_2", "Timo_1", "Timo_2", "Timo_3", "Timo_4", "Length"))
)
df4$x <- factor(df4$x, levels = unique(df4$x))
df4$xpos <- factor(df4$xpos, levels = (df4$xpos))

colour_vector <- c(
  "hela" = '#B0E0E6',
  "hek" = '#A0D6B4',
  "pc3" = '#C5E384',
  "u2os" = '#D0F0C0',
  "shsy5y" = '#D3F8E2',
  "tcells" ='#B2F7EF',
  "sknbe" = '#AFE9FF',
  "length" = 'white'
)


lebel_vector <-c(
  "hela" = 'HeLa',
  "hek" = 'HEK293T',
  "pc3" = 'PC-3',
  "u2os" = 'U2-OS',
  "shsy5y" = 'SH-SY-5Y',
  "tcells" ='Donor T Cells',
  "sknbe" = 'SK-N-BE',
  "length" = 'RNA Length'
)


(pl5 <- ggplot(df4, aes(x = xpos, y = 1, fill = x)) +
  geom_tile() +
  theme_void() +
  labs(fill = "Cell Type") +
  scale_fill_manual(
    values = colour_vector,labels = lebel_vector,
    breaks = names(lebel_vector)
  ) +
  scale_x_discrete(
    labels = NULL
  ) +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1),
    text = element_text(size = 20, family = "Arial MS", color = "#505050", face = "bold")
  ) +
  ylab("") + xlab(""))

#Transcriptome
df5 <-data.frame(x = (c("wt", "wt", "ap-eif4a", "ap-eif4a", "ap-g3bp", "ap-ddx3x", "ap-ddx3y", "g3bp-mini", "g3bp-mini", "g3bp-mini", "wt", "mettl3ko", "dhx9-egfp","dhx9-egfp","dhx9-egfp","dhx9-egfp", "wt","wt", "g3bp-gfp", "wt", "wt", "wt", "g3bpko", "g3bp-gfp", "g3bp-gfp", "wt", "wt", "nat10ko", "wt", "mettl3ko", "FUS-mut", "FUS-mutko", "length")), 
  colour = c("Honda 2021", "Xiao 2024", "Padrón 2019", "Padrón 2019", "Somasekharan 2020", "Shen 2022", "Shen 2022", "Ren 2023", "Ren 2023", "Ren 2023", "Rajachandran 2024", "Rajachandran 2024", "Zhou 2024","Zhou 2024","Zhou 2024","Zhou 2024", "Sato 2023","Sato 2023", "Khong 2017", "Matheny 2019", "Matheny 2021","Matheny 2021","Matheny 2021","Matheny 2021", "Iadevaia 2022", "Curdy 2023", "Kudrin 2024", "Kudrin 2024", "Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024","Di Timoteo 2024", "RNA Length"),
  xpos = (c("Honda", "Xiao", "Padron_1", "Padron_2", "Somasekharan", "Shen_1", "Shen_2", "Ren_1", "Ren_2", "Ren_3", "Raja_1", "Raja_2", "Zhou_1", "Zhou_2", "Zhou_3", "Zhou_4", "Sato_1", "Sato_2", "Khong", "Matheny19","Matheny21_1","Matheny21_2","Matheny21_3","Matheny21_4", "Iadevaia", "Curdy", "Kudrin_1", "Kudrin_2", "Timo_1", "Timo_2", "Timo_3", "Timo_4", "Length"))
)
df5$x <- factor(df5$x, levels = unique(df5$x))
df5$xpos <- factor(df5$xpos, levels = (df5$xpos))

colour_vector <- c(
  "wt" = '#F3B0C3',
  "ap-eif4a" = '#F4A261',
  "ap-g3bp" = '#B8B8FF',
  "ap-ddx3x" = '#FFF275',
  "ap-ddx3y" = '#A5B4FC',
  "g3bp-mini" ='#F6BD60',
  "mettl3ko" = '#A9DEF9',
  "dhx9-egfp" = '#B2F7EF',
  "g3bp-gfp" = '#A3C4BC',
  "g3bpko" = '#90E0EF',
  "nat10ko" = '#CBAACB',
  "FUS-mut" = '#F7A9A8',
  "FUS-mutko" = '#FBC4AB',
  "length" = 'white'
)


lebel_vector <-c(
  "wt" = 'Wild-Type',
  "ap-eif4a" = 'APEX2-eIF4A',
  "ap-g3bp" = 'APEX2-G3BP',
  "ap-ddx3x" = 'APEX2-DDX3X',
  "ap-ddx3y" = 'APEX2-DDX3Y',
  "g3bp-mini" ='G3BP-miniSOG',
  "mettl3ko" = 'METTL3 KO',
  "dhx9-egfp" = 'DHX9-eGFP + G3BP-mCherry',
  "g3bp-gfp" = 'G3BP-GFP',
  "g3bpko" = 'G3BP KO',
  "nat10ko" = 'NAT10 KO',
  "FUS-mut" = 'FUS-P525L',
  "FUS-mutko" = 'FUS-P525L + METTL3 KO',
  "length" = 'RNA Length'
)


(pl6 <- ggplot(df5, aes(x = xpos, y = 1, fill = x)) +
  geom_tile() +
  theme_void() +
  labs(fill = "Transcriptome") +
  scale_fill_manual(
    values = colour_vector,labels = lebel_vector,
    breaks = names(lebel_vector)
  ) +
  scale_x_discrete(
    labels = NULL
  ) +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1),
    text = element_text(size = 20, family = "Arial MS", color = "#505050", face = "bold")
  ) +
  ylab("") + xlab(""))




  
heatmap <- pl1 + theme(axis.text.x = element_blank(),
                      axis.title.y = element_blank(),
         axis.ticks.x = element_blank(), legend.position="top", axis.title.x = element_blank())

 
heatmap <-  heatmap + theme(strip.text.x = element_blank(),
      strip.background = element_rect(colour="white", fill="white"),
      legend.position='none', legend.direction="vertical"
      )
  
heatmap <- heatmap +
  theme(
    plot.margin = grid::unit(c(1, 1, 1, 6), "lines"),
    axis.text.y = element_text(angle = 0, hjust = 1, size = 20, face = "bold")
  )

pl2 <- pl2 +
  theme(
    plot.margin = grid::unit(c(0, 0, 6, 0), "lines")
  )

(left <-   ggarrange(heatmap, pl3, pl4, pl5, pl6, pl2, heights = c(5,.1,.1,.1,.1,.85), ncol=1, align="v", labels = c("B"),font.label = list(size=25)))

heatmap_leg <- get_legend(pl1+ theme(text=element_text(size=25,  family="Arial MS", color="#505050"), face="bold"), position = 'bottom')
heatmap_leg <- as_ggplot(heatmap_leg)
leg1 <- get_legend(pl2+ theme(text=element_text(size=25,  family="Arial MS", color="#505050"), face="bold"), position = 'right')
leg1 <- as_ggplot(leg1)
leg2 <- get_legend(pl3+ theme(text=element_text(size=25,  family="Arial MS", color="#505050"), face="bold"), position = 'right')
leg2 <- as_ggplot(leg2)
leg3 <- get_legend(pl4+ theme(text=element_text(size=25,  family="Arial MS", color="#505050"), face="bold"), position = 'right')
leg3 <- as_ggplot(leg3)
leg4 <- get_legend(pl5+ theme(text=element_text(size=25,  family="Arial MS", color="#505050"), face="bold"), position = 'right')
leg4 <- as_ggplot(leg4)
leg5 <- get_legend(pl6+ theme(text=element_text(size=25,  family="Arial MS", color="#505050"), face="bold"), position = 'right')
leg5 <- as_ggplot(leg5)

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure1/FigureB.png", width = 28, height=28, units = "in", res=300)
left
dev.off()

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure1/FigureB_heatleg.png", width = 28, height=28, units = "in", res=300)
heatmap_leg
dev.off()

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure1/FigureB_leg1.png", width = 28, height=28, units = "in", res=300)
leg1
dev.off()

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure1/FigureB_leg2.png", width = 28, height=28, units = "in", res=300)
leg2
dev.off()

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure1/FigureB_leg3.png", width = 28, height=28, units = "in", res=300)
leg3
dev.off()

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure1/FigureB_leg4.png", width = 28, height=28, units = "in", res=300)
leg4
dev.off()

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure1/FigureB_leg5.png", width = 28, height=28, units = "in", res=300)
leg5
dev.off()

```
### C

```{r fig.height=6, fig.width=10}

hallmark.sg.t_df <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/KhongSGHallmark.txt")[,-1]
hallmark.size.t <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/LengthHallmark.txt")[,-1]
hallmark.sg.t_pl <- fread("../Projects/StressGranules/SGPAPER/SGvsWC/SomaSGHallmark.txt")[,-1]


colnames(hallmark.sg.t_df) <- c("pathway", "pval_dc", "nes_dc")
colnames(hallmark.size.t) <- c("pathway", "pval_size", "nes_size")
colnames(hallmark.sg.t_pl) <- c("pathway", "pval_pl", "nes_pl")

hallmark_df <- merge(hallmark.size.t, hallmark.sg.t_df, by="pathway")
hallmark_df <- merge(hallmark_df, hallmark.sg.t_pl, by="pathway")

vip_dc <- as.data.frame(hallmark_df[hallmark_df$pval_size<0.05 & hallmark_df$pval_dc<0.05,] )
vip_pl <- as.data.frame(hallmark_df[hallmark_df$pval_size<0.05 & hallmark_df$pval_pl<0.05,] )

(hallplot <- ggplot(hallmark_df)+geom_point(aes(x=nes_size, y=nes_pl), colour="#506863", shape=17,size=7,alpha=.5)+ geom_point(aes(x=nes_size, y=nes_dc), colour="#007EA7",size=7,alpha=.5, shape=15)+  theme_classic()+
 ylab("SG NES") + xlab("RNA Length NES") + geom_abline(slope = 1, intercept = 0, linetype="dashed", colour="black")+geom_abline(slope = -1, intercept = 0, linetype="dashed", colour="black", alpha=.5)+geom_vline(xintercept = 0, linetype="dotted")+geom_hline(yintercept = 0, linetype="dotted")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=20,  family="Arial MS")) +xlim(c(-3.5,3.5))+ylim(c(-3.5,3.5)))
#007EA7 blue sig
#AD3434 red sig

df_aux_dc <- hallmark_df[,c("pval_size","nes_size","pval_dc","nes_dc")]
df_aux_dc$sig <- ifelse(df_aux_dc$pval_dc <= 0.05 & df_aux_dc$pval_size <= 0.05, "Significant","Non Significant")
df_aux_dc <- as.data.frame(df_aux_dc)
df_aux_dc <- df_aux_dc[,!colnames(df_aux_dc) %in% c("pval_size","pval_dc")]
df_aux_dc$method <- "D. Centrifugation"
colnames(df_aux_dc) <- c("NES_size","NES_method","significance","method")
  
df_aux_pl <- hallmark_df[,c("pval_size","nes_size","pval_pl","nes_pl")]
df_aux_pl$sig <- ifelse(df_aux_pl$pval_pl <= 0.05 & df_aux_pl$pval_size <= 0.05, "Significant","Non Significant")
df_aux_pl <- as.data.frame(df_aux_pl)
df_aux_pl <- df_aux_pl[,!colnames(df_aux_pl) %in% c("pval_size","pval_pl")]
df_aux_pl$method <- "P. Labelling"
colnames(df_aux_pl) <- c("NES_size","NES_method","significance","method")

df_aux <- rbind(df_aux_dc,df_aux_pl)

pmet_leg <- ggplot(df_aux)+
  geom_point(aes(x=NES_size, y=NES_method, shape=method, color=method),size=7,alpha=.7)+   
  theme_classic()+
 xlab("SG NES") + ylab("RNA Length NES") + geom_abline(slope = 1, intercept = 0, linetype="dashed", colour="black")+geom_abline(slope = -1, intercept = 0, linetype="dashed", colour="black", alpha=.5)+geom_vline(xintercept = 0, linetype="dotted")+geom_hline(yintercept = 0, linetype="dotted")+
  labs(shape = "Purification\nMethod ")+
  labs(colour = "Purification\nMethod ")+
scale_colour_manual(values=c('#75958F','#007EA7'), breaks = c("P. Labelling", "D. Centrifugation"))+ scale_shape_manual(values = c(17,15),breaks = c("P. Labelling", "D. Centrifugation"))+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=20,  family="Arial MS")) +xlim(c(-3.5,3.5))+ylim(c(-3.5,3.5))

pmet_leg2 <- ggplot(df_aux)+
  geom_point(aes(x=NES_size, y=NES_method, shape=method, color=method),size=7,alpha=.5)+   
  theme_classic()+
 xlab("SG NES") + ylab("RNA Length NES") + geom_abline(slope = 1, intercept = 0, linetype="dashed", colour="black")+geom_abline(slope = -1, intercept = 0, linetype="dashed", colour="black", alpha=.5)+geom_vline(xintercept = 0, linetype="dotted")+geom_hline(yintercept = 0, linetype="dotted")+
  labs(shape = "Statistical\nSignificance")+
  labs(colour = "Statistical\nSignificance")+
scale_colour_manual(values=c('#ADEBFF','#AEC1BD'), breaks = c("P. Labelling", "D. Centrifugation"), labels=c("Padj ≤ 0.05", "Padj > 0.05"))+ scale_shape_manual(values = c(17,15),breaks = c("P. Labelling", "D. Centrifugation"), labels=c("Padj ≤ 0.05", "Padj > 0.05"))+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=20,  family="Arial MS")) +xlim(c(-3.5,3.5))+ylim(c(-3.5,3.5))

leggyboy <- get_legend(pmet_leg)
leggyboy <- as_ggplot(leggyboy)



(hallploty <- ggarrange(hallplot,ggplot()+theme_void(), leggyboy, ggplot()+theme_void(), nrow = 1, widths = c(0.8,0.05,0.2,0.05),labels = c("C"),font.label = list(size=25)))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure1/Hallplot.png", width = 10, height=6, units = "in", res=300)
hallploty
dev.off()

tiff("Projects/StressGranules/SGPAPER/Leg1.tiff", width = 8, height=4, units = "in", res=300)
pmet_leg
dev.off()

tiff("Projects/StressGranules/SGPAPER/Leg2.tiff", width = 8, height=4, units = "in", res=300)
pmet_leg2
dev.off()





#Fisher Test Example
PL_Q1 <- nrow(hallmark_df[hallmark_df$nes_size>0 & hallmark_df$nes_pl>0,])
PL_Q2 <- nrow(hallmark_df[hallmark_df$nes_size>0 & hallmark_df$nes_pl<0,] )
PL_Q3 <- nrow(hallmark_df[hallmark_df$nes_size<0 & hallmark_df$nes_pl>0,] )
PL_Q4 <- nrow(hallmark_df[hallmark_df$nes_size<0 & hallmark_df$nes_pl<0,] )

PL_matrix <- matrix(c(PL_Q1, PL_Q2, PL_Q3, PL_Q4), nrow = 2, byrow = TRUE)
fisher.test(PL_matrix)

DC_Q1 <- nrow(hallmark_df[hallmark_df$nes_size>0 & hallmark_df$nes_dc>0,])
DC_Q2 <- nrow(hallmark_df[hallmark_df$nes_size>0 & hallmark_df$nes_dc<0,] )
DC_Q3 <- nrow(hallmark_df[hallmark_df$nes_size<0 & hallmark_df$nes_dc>0,] )
DC_Q4 <- nrow(hallmark_df[hallmark_df$nes_size<0 & hallmark_df$nes_dc<0,] )

DC_matrix <- matrix(c(DC_Q1, DC_Q2, DC_Q3, DC_Q4), nrow = 2, byrow = TRUE)
fisher.test(DC_matrix)

dimnames(PL_matrix) <- list(
  Size_Pos_vs_Neg = c("Size_Pos", "Size_Neg"),
  PL_Pos_vs_Neg = c("PL_Pos", "PL_Neg")
)

dimnames(DC_matrix) <- list(
  Size_Pos_vs_Neg = c("Size_Pos", "Size_Neg"),
  PL_Pos_vs_Neg = c("PL_Pos", "PL_Neg")
)
```
# Supplementary Figure 1

### B

```{r fig.height=6, fig.width=16}
sgVbulklogs <- fread("~/Projects/StressGranules/SGPAPER/SGvsWC/Expression.txt")

res_logs <- lm(Bulk ~ SG, sgVbulklogs)

DCwc <- ggplot(sgVbulklogs, aes(y=SG, x=Bulk)) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)  +   theme_bw() +
  xlab(expression(paste("WC Expression (log"[2]*" CPM)")))+
  ylab(expression(paste("SG Expression (log"[2]*" CPM)")))+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Differential Centrifugation")+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=20,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho",size=7)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 20, l = 0)))


sgVbulklogs <- fread("~/Projects/StressGranules/SGPAPER/SGvsWC/Expression2.txt")

res_logs <- lm(Bulk ~ SG, sgVbulklogs)

PLwc <- ggplot(sgVbulklogs, aes(y=SG, x=Bulk)) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)  +   theme_bw()+xlab(expression(paste("WC Expression (log"[2]*" CPM)")))+
  ylab(expression(paste("SG Expression (log"[2]*" CPM)")))+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Proximity Labelling")+
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(text=element_text(size=20,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+theme(axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 20, l = 0)))


(C <- ggarrange(DCwc, PLwc, labels = c("D",""),font.label = list(size=25)))

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure1/FigureD.png", width = 14, height=6, units = "in", res=300)
C
dev.off()
```


### A

```{r}
u20s <- fread("~/Projects/StressGranules/SGPAPER/U2OS_StressResponse.txt")[,c(1,4)]
pc3 <- fread("~/Projects/StressGranules/SGPAPER/PC3_StressResponse.txt")[,c(1,4)]

stressresponde <- merge(u20s, pc3, by="V1")
colnames(stressresponde) <- c("Gene", "U2OS", "PC3")

smooth.SG <- lm(PC3~(U2OS), stressresponde)
res <- residuals(smooth.SG)
stressresponde$Residuals <- res
stressresponde$mul <- (stressresponde$U2OS + stressresponde$PC3)/2

stresspoints <- stressresponde[stressresponde$U2OS>25 & stressresponde$PC3>25,]

# Plot the graph
(sup1 <- ggplot(stressresponde, aes(y = PC3, x = (U2OS))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2) +
  theme_bw() +
  ylab("DGE in PC-3 (t-stat) ") + 
  xlab("DGE in U2-OS (t-stat)") + # Subscript "10"
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())  + 
  theme(text=element_text(size=20,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope = smooth.SG$coefficients[2], 
              intercept = smooth.SG$coefficients[1], 
              linetype = "dashed") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  geom_point(data=stresspoints, aes(x=U2OS, y=  PC3), fill="#800020",colour="black", size=5,alpha=0.8,shape=21)+
  geom_label_repel(stresspoints, mapping=aes(label=stresspoints$Gene),size=4,point.padding=-0.5, color="#800020",max.overlaps = 20, fontface='italic')+theme(
  axis.title.y = element_text(
    margin = margin(r = 20)
  )
))

png("../Projects/StressGranules/SGPAPER/Images/PNG/SupFigure1A.png", width = 7, height=4, units = "in", res=300)
sup1
dev.off()
```

```{r fig.height=6, fig.width=10}

#for u20s genes
u20s <- u20s[order(u20s[,2]),]
gseatool <- u20s[,2]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- u20s$V1
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.u20s <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.u20s <- as.data.frame(hallmark.u20s)

#for pc3 genes
pc3 <- pc3[order(pc3[,2]),]
gseatool <- pc3[,2]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- pc3$V1
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.pc3 <- hallmarks.res.tidy[,c(1,3,6)]
hallmark.pc3 <- as.data.frame(hallmark.pc3)

colnames(hallmark.pc3) <- c("pathway", "pval_dc", "nes_pc3")
colnames(hallmark.u20s) <- c("pathway", "pval_pl", "nes_u20s")

hallmark_df <- merge(hallmark.pc3, hallmark.u20s, by="pathway")

(hallplot <- ggplot(hallmark_df)+geom_point(aes(x=nes_u20s, y=nes_pc3), colour="#506863", shape=17,size=7,alpha=.5)+  theme_classic()+
 ylab("PC-3 NES") + xlab("U2-OS NES") + geom_abline(slope = 1, intercept = 0, linetype="dashed", colour="black")+geom_abline(slope = -1, intercept = 0, linetype="dashed", colour="black", alpha=.5)+geom_vline(xintercept = 0, linetype="dotted")+geom_hline(yintercept = 0, linetype="dotted")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=20,  family="Arial MS")) +xlim(c(-3.5,3.5))+ylim(c(-3.5,3.5)))

#Fisher Test Example
PL_Q1 <- nrow(hallmark_df[hallmark_df$nes_pc3>0 & hallmark_df$nes_u20s>0,])
PL_Q2 <- nrow(hallmark_df[hallmark_df$nes_pc3>0 & hallmark_df$nes_u20s<0,] )
PL_Q3 <- nrow(hallmark_df[hallmark_df$nes_pc3<0 & hallmark_df$nes_u20s>0,] )
PL_Q4 <- nrow(hallmark_df[hallmark_df$nes_pc3<0 & hallmark_df$nes_u20s<0,] )

PL_matrix <- matrix(c(PL_Q1, PL_Q2, PL_Q3, PL_Q4), nrow = 2, byrow = TRUE)
fisher.test(PL_matrix)

dimnames(PL_matrix) <- list(
  Size_Pos_vs_Neg = c("Size_Pos", "Size_Neg"),
  PL_Pos_vs_Neg = c("PL_Pos", "PL_Neg")
)

png("../Projects/StressGranules/SGPAPER/Images/PNG/SupFigure1B.png", width = 7, height=4, units = "in", res=300)
hallplot
dev.off()
```
```{r fig.height=6, fig.width=14}
# Arrange your plots
sup <- ggarrange(
  sup1, hallplot,
  ncol = 2, nrow = 1
)

# Add a common title
sup <- annotate_figure(
  sup,
  top = text_grob("Oxidative Stress vs Control", face = "bold", size = 25)
)


png("../Projects/StressGranules/SGPAPER/Images/PNG/SupFigure1.png", width = 14, height=6, units = "in", res=300)
sup
dev.off()
```


## Figure 2

### B

```{r}
#from the khong 2017 code
trial$condition <- ifelse(trial$variable %in% c("1", "2", "3"), "SG", "WC")

(compyplot <- ggplot(data = trial, aes(x=genes, y=value.total, group=SampleColouring, color=SampleColouring)) + geom_line() +
  theme_bw()  + geom_hline(yintercept = 50, linetype="dashed") + xlim(0, 20000) + xlab("Number of Genes Covered") + ylab("Number of Reads (%)")+
  scale_colour_manual(
    name = "Sample Type", 
    values = c("WC" = "#53A2BE", "SG" = "#BA3B46")
  ) +xlim(0,5000)+
  theme(text=element_text(size=20,  family="Arial MS"))+ggtitle("Sample Complexity")+
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),legend.position = c(0.5, 0.3)))
```
### C

```{r fig.height=6, fig.width=14}
#Left
#From Simulator Analysis

(ogdist <- LiBTab %>%
    pivot_longer(cols = -insert_size) %>%
    ggplot(aes(x = insert_size, y = value, group = name, color = name)) + 
      geom_line(size=1.5) + theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Empirical Fragment Distribution") + scale_colour_manual(values = c("#53A2BE","#BA3B46"), name = "Sample Type", labels = c("WC", "SG")) + xlab("Fragment Length (nt)") +ylab("Relative Distribution")+
  theme(text=element_text(size=20,  family="Arial MS"))+ylim(0,0.007)+
  theme(plot.title = element_text(hjust = 0.5))+ 
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),legend.position = c(0.7, 0.7))+theme(
  axis.title.y = element_text(
    margin = margin(r = 20),  # << adjust this value as needed
  )
))

#right
WG_WC <- subset(WG, Gene=="WC")
WG_WC_count <- as.data.frame(table(WG_WC))
WG_WC_count$Freq <- WG_WC_count$Freq / sum(WG_WC_count$Freq)
SG_WC <- subset(WG, Gene=="SG")
SG_WC_count <- as.data.frame(table(SG_WC))
SG_WC_count$Freq <- SG_WC_count$Freq / sum(SG_WC_count$Freq)

WG <- rbind(WG_WC_count, SG_WC_count)
(simdist <- ggplot(WG, aes(x=as.numeric(fragments), y=Freq, group=Gene, colour=Gene))+ geom_line(size=1.5)+
       theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Simulated Fragment Distribution") + scale_colour_manual(values = c("#53A2BE","#BA3B46"), name = "Sample Type", labels = c("WC", "SG"),breaks = c("WC","SG")) + xlab("Fragment Length (nt)") +ylab("Relative Distribution")+ xlim(0,1100)+
  theme(text=element_text(size=20,  family="Arial MS"))+
  theme(plot.title = element_text(hjust = 0.5))+ ylim(0,0.007)+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),legend.position = c(0.7, 0.7))+theme(
  axis.title.y = element_text(
    margin = margin(r = 20),  # << adjust this value as needed
  )
))

(fig2c <- ggarrange(ogdist, simdist, nrow = 1))

png("../Projects/StressGranules/SGPAPER/SupFigure2.png", width = 7, height=4, units = "in", res=300)
simdist
dev.off()
```
### D

```{r fig.height=6, fig.width=14}
DEGenes <- fread("~/Sim_DEGenes.txt")

(simsamplersg <- ggplot(DEGenes, aes(y=t, x=log10(Length))) + geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') + theme_bw()+ 
  scale_x_continuous(
    breaks = log10(c(10, 100, 1000, 10000)),  # tick positions (in log10 scale)
    labels = c("10", "100", "1000", "10000") # what you want to show
  )+ coord_cartesian(ylim=c(-600,700))+
 ylab("DGE Empirical-Simulated (t-stat)") + xlab(expression(RNA ~ Length ~ (log[10] ~ (nt))))+
      geom_density2d(color="#800020", size=1.2)  +   theme_bw()+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + stat_cor(method = "spearman", cor.coef.name = "rho",size=5, label.y=600)+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=20,  family="Arial MS")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+theme(
  axis.title.y = element_text(
    margin = margin(r = 20),  # << adjust this value as needed
  )
))

(fig2b <- ggarrange(compplot, simsamplersg, ncol = 2, font.label = list(size=25)))
```

```{r fig.height=12, fig.width=16}
(Figure2 <- ggarrange( fig2b,ggplot()+theme_void(), fig2c, ncol = 1, font.label = list(size=25),heights = c(0.30,0.01,0.30,0.01,0.30)))

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure2.png", width = 16, height=12, units = "in", res=300)
Figure2
dev.off()
```

## Figure 3

### A and B

```{r  fig.height=6, fig.width=14}
SimDegenes <- fread("~/Projects/StressGranules/SGPAPER/SGvsWC/OldSpinSim.txt")
colnames(SimDegenes)[1] <- "Gene"
SimDegenes <- merge(SimDegenes, genelength, by="Gene")

(firstsimcentri <- ggplot(SimDegenes, aes(y=t, x=log10(Length))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)+ theme_bw()+
 xlab(expression(paste("RNA Length (log"[10]*" (nt))"))) + ylab("DGE SimSG-WC (t-stat)") +
  scale_x_continuous(
    breaks = log10(c(10, 100, 1000, 10000)),  # tick positions (in log10 scale)
    labels = c("10", "100", "1000", "10000") # what you want to show
  ) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+ 
  theme(text=element_text(size=25,  family="Arial MS")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()))

(AB3 <- ggarrange(ggplot()+theme_void(), firstsimcentri, ncol = 2))

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure3B.png", width = 14, height=6, units = "in", res=300)
AB3
dev.off()

```
### C

```{r}
simhallmark <- fread("~/Projects/StressGranules/SGPAPER/SGvsWC/KhongSGHallmark_old.txt")[,-1]
hallmark.size.t <- fread("~/Projects/StressGranules/SGPAPER/SGvsWC/LengthHallmark.txt")[,-1]

colnames(simhallmark) <- c("pathway", "nes_sim")
colnames(hallmark.size.t) <- c("pathway", "pval_size", "nes_size")

hallmark_df <- merge(hallmark.size.t, simhallmark, by="pathway")




hallplot <- ggplot(hallmark_df)+geom_point(aes(x=nes_size, y=nes_sim), colour="#506863", shape=17,size=7,alpha=.5)+ theme_classic()+
 ylab("SIM SG NES") + xlab("RNA Length NES") + geom_abline(slope = 1, intercept = 0, linetype="dashed", colour="black")+geom_abline(slope = -1, intercept = 0, linetype="dashed", colour="black", alpha=.5)+geom_vline(xintercept = 0, linetype="dotted")+geom_hline(yintercept = 0, linetype="dotted")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25,  family="Arial MS")) +xlim(c(-3.5,3.5))+ylim(c(-3.5,3.5))


df_aux <- hallmark_df[,c("pval_size","nes_size","nes_sim")]
df_aux$sig <- ifelse(df_aux$pval_sim <= 0.05 & df_aux$pval_size <= 0.05, "Significant","Non Significant")
df_aux <- as.data.frame(df_aux)
df_aux <- df_aux[,!colnames(df_aux) %in% c("pval_size")]
df_aux$method <- "D. Centrifugation"
colnames(df_aux) <- c("NES_size","NES_method","significance","method")
  


pmet_leg2 <- ggplot(df_aux)+
  geom_point(aes(x=NES_size, y=NES_method, color=significance),size=7,alpha=.5,shape=16)+   
  theme_classic()+
 xlab("SG NES") + ylab("RNA Length NES") + geom_abline(slope = 1, intercept = 0, linetype="dashed", colour="black")+geom_abline(slope = -1, intercept = 0, linetype="dashed", colour="black", alpha=.5)+geom_vline(xintercept = 0, linetype="dotted")+geom_hline(yintercept = 0, linetype="dotted")+
  labs(shape = "Statistical\nSignificance")+
  labs(colour = "Statistical\nSignificance")+
scale_colour_manual(values=c('#A300A3','#FFB7FF'), breaks = c("Significant", "Non Significant"), labels=c("Padj ≤ 0.05", "Padj > 0.05"))+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=11,  family="Arial MS")) +xlim(c(-3.5,3.5))+ylim(c(-3.5,3.5))


(hallploty <- ggarrange(hallplot, leggyboy, nrow = 1, widths = c(0.8,0.2)))


#Fisher Test Example
DC_Q1 <- nrow(hallmark_df[hallmark_df$nes_size>0 & hallmark_df$nes_sim>0,])
DC_Q2 <- nrow(hallmark_df[hallmark_df$nes_size>0 & hallmark_df$nes_sim<0,] )
DC_Q3 <- nrow(hallmark_df[hallmark_df$nes_size<0 & hallmark_df$nes_sim>0,] )
DC_Q4 <- nrow(hallmark_df[hallmark_df$nes_size<0 & hallmark_df$nes_sim<0,] )

DC_FT <- data.frame(C1=c(DC_Q1,DC_Q3), C2=c(DC_Q2, DC_Q4))
fisher.test(DC_FT)

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure3C.png", width = 7, height=6, units = "in", res=300)
hallplot
dev.off()

```
### D

```{r fig.height=6, fig.width=7}
realts <- fread("~/Projects/StressGranules/SGPAPER/SGvsWC/Akhong2017.txt")[,c(1,4)]
colnames(realts)[1] <- "Gene"
SimDegenes <- fread("~/Projects/StressGranules/SGPAPER/SGvsWC/OldSpinSim.txt")[,c(1,4)]
colnames(SimDegenes)[1] <- "Gene"
tscompy <- merge(realts, SimDegenes, by="Gene")
tscompy <- merge(tscompy, genelength, by="Gene")

smooth.SG <- lm(t.y~(t.x), tscompy)


tscorrelplot <- ggplot(tscompy, aes(y = t.y, x = t.x)) + 
  geom_point(aes(color = Length), alpha = 0.5, size = 5) +   # map raw Length
  geom_density2d(color = "#800020", size = 1.2) + 
  theme_bw() +
  xlab("DGE SG-WC (t-stat)") + 
  ylab("DGE SimSG-WC (t-stat)") +  
  labs(color = expression(paste("RNA\nLength"))) + 
  scale_colour_gradient(
    low = "#729EA1", high = "#A82445",
    trans = "log10",                                # log scale
    breaks = c(10, 100, 1000, 10000, 100000),       # tick positions
    labels = c("10", "100", "1000", "10000", "100000")  # pretty labels
  ) +
  geom_abline(
    slope = smooth.SG$coefficients[2],
    intercept = smooth.SG$coefficients[1],
    linetype = "dashed"
  ) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 7) +
  theme(
    text = element_text(size = 25, family = "Arial MS"),
    plot.title = element_text(hjust = 0.5),
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    legend.text = element_text(size = 15),     # smaller font for legend labels
    legend.title = element_text(size = 17)     # smaller font for legend title
  )

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure3D.png", width = 7, height=6, units = "in", res=300)
tscorrelplot
dev.off()


```

```{r fig.height=6, fig.width=14}
(CD3 <- ggarrange(hallploty, tscorrelplot, ncol = 2, labels = c("C","D"), font.label = list(size=25)))
```

### E

```{r fig.height=4, fig.width=7}
RealSig <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/RealAkhong_old.txt")
colnames(RealSig)[1] <- "Gene"
RealSig <- merge(RealSig, genelength, by="Gene")

smooth.SG <- lm(t~log10(Length), RealSig)

(realsig <- ggplot(RealSig, aes(y=t, x=log10(Length))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)+ theme_bw()+
 ylab("DGE SG-SimSG (t-stat)") + xlab(expression(paste("RNA Length (log"[10]*"(nt))"))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ 
  theme(text=element_text(size=25,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  scale_x_continuous(
    breaks = log10(c(10, 100, 1000, 10000)),  # tick positions (in log10 scale)
    labels = c("10", "100", "1000", "10000") # what you want to show
  ))
```

### F

```{r fig.height=6, fig.width=7}
BU <- fread("~/Projects/StressGranules/SGPAPER/Stressresponse_u2os.txt")[,c(1,4)]
colnames(BU) <- c("Gene", "BU")

RNP <- fread("~/Projects/StressGranules/SGPAPER/Stressresponse_u2os_rnp.txt")[,c(1,4)]
colnames(RNP) <- c("Gene", "RNP")

Rnptables <- merge(BU, RNP, by="Gene")

smooth.SG <- lm(RNP~(BU), Rnptables)

(rnpy <- ggplot(Rnptables, aes(y=RNP, x=(BU))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)+ theme_bw()+
 ylab("DGE Stressed RNP (t-stat)") + xlab("DGE Stressed WC (t-stat)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ 
  theme(text=element_text(size=25,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()))
```

```{r fig.height=6, fig.width=14}
(EF3 <- ggarrange(realsig, rnpy, ncol = 2))

png("../Projects/StressGranules/SGPAPER/Images/PNG/Figure3EF.png", width = 14, height=6, units = "in", res=300)
EF3
dev.off()
```


## Figure 4

### A

```{r fig.height=6, fig.width=14}
#loading relevant data
actables <- fread("~/Projects/StressGranules/SGPAPER/AC4COGDATA.txt")
smooth.ac <- lm(nt_t~(wt_t), actables)

(accorrel <- ggplot(actables, aes(y=nt_t, x=(wt_t))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)  +   theme_bw()+ylab("DGE KO SG-WC (t-stat)")+
  xlab("DGE WT SG-WC (t-stat)")+ ggtitle("ac4C")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(text=element_text(size=25,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_abline(slope=smooth.ac$coefficients[2], intercept = smooth.ac$coefficients[1], linetype = "dashed"))

m6tables <- fread("~/Projects/StressGranules/SGPAPER/M6ADATAORIG.txt")
smooth.m6 <- lm(nt_t~(wt_t), m6tables)

(m6correl <- ggplot(m6tables, aes(y=nt_t, x=(wt_t))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)  +   theme_bw()+ylab("DGE KO SG-WC (t-stat)")+ ggtitle("m6A")+
  xlab("DGE WT SG-WC (t-stat)")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(text=element_text(size=25,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_abline(slope=smooth.m6$coefficients[2], intercept = smooth.m6$coefficients[1], linetype = "dashed"))

(A4 <- ggarrange(accorrel, m6correl, ncol = 2))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure4A.png", width = 14, height=6, units = "in", res=300)
A4
dev.off()
```

### B

```{r fig.height=6, fig.width=14}
#AC
actables$bins <- factor(actables$bins, levels=c(0:8, "[9,12]","[13,[" ))

trial <- subset(actables, NAC >= 5)

spearman_corr <- cor(trial$NAC, (trial$Residuals), method = "spearman")

(acbins <- ggplot(actables, aes(x=bins, y=(Residuals)))  + geom_jitter(alpha=0.1) + geom_violin(fill="#43F3D5", alpha=0.4) +theme_bw()+ 
 xlab("Number of ac4C Modifications") + ylab("Residuals") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) +annotate("text", x = 4.75, y = 4.5, label = expression("For NAC" >= 5 * "," ~ rho == 0.33 * "," ~ italic(p) < 8.6e-9), size = 7, color = "black") 
  +
  theme(
    text = element_text(size = 25, family = "Arial MS"),       # base size
    axis.text.x = element_text(size = 15),  # smaller x-axis label
    axis.text.y = element_text(size = 15),  # smaller y-axis label
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ))

#m6
m6tables$bins <- ifelse(m6tables$NM6A <= 4, m6tables$NM6A,
                      ifelse(m6tables$NM6A>=5 & m6tables$NM6A<=9, "[5,9]","[10,["))

m6tables$bins <- factor(m6tables$bins, levels=c(0:4, "[5,9]","[10,[" ))

trial <- subset(m6tables, NM6A >= 4)

spearman_corr <- cor.test(trial$NM6A, (trial$Residuals), method = "spearman")

(m6bins <- ggplot(m6tables, aes(x=bins, y=(Residuals)))  + geom_jitter(alpha=0.1) + geom_violin(fill="#43F3D5", alpha=0.4) +theme_bw() +
 xlab("Number of m6A Modifications") + ylab("Residuals") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))+annotate("text", x = 3, y = 2.8, label = expression("For NM" >= 4 * "," ~ rho == 0.16 * "," ~ italic(p) < 6.3e-5), size = 7, color = "black") +
  theme(
    text = element_text(size = 25, family = "Arial MS"),       # base size
    axis.text.x = element_text(size = 15),  # smaller x-axis label
    axis.text.y = element_text(size = 15),  # smaller y-axis label
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ))

(B4 <- ggarrange(acbins, m6bins, ncol = 2))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure4B.png", width = 14, height=6, units = "in", res=300)
B4
dev.off()

```
### C

```{r fig.height=7, fig.width=14}
#Starting with my boy AC
ac4cres <- fread("~/Projects/StressGranules/SGPAPER/ACSimRes.txt")

upmarkers <- ac4cres[
  !(ac4cres$bins %in% c(0, 1, 2, 3, 4,5)), 
  ]


spearman_corr <- cor.test(upmarkers$Residuals.y, (upmarkers$Residuals.x), method = "spearman")

(acres <- ggplot(ac4cres, aes(y=Residuals.y, x=(Residuals.x))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black')  +   theme_bw()+xlab("Original Residuals")+
  ylab("Simulated Residuals")+ylim(-12,12)+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(text=element_text(size=25,  family="Arial MS")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_point(data=upmarkers, aes(x=Residuals.x, y=  Residuals.y), color="#43F3D5",fill="#43F3D5",alpha=0.5, size=5, shape=21)+geom_density2d(color="#800020", size=1.2)+annotate("text", x = -1, y = 10, label = expression("For NAC" >= 6 * "," ~ rho == 0.41 * "," ~ italic(p) < 3.5e-9), size = 7, color = "black"))

#moving to m6a
m6ares <- fread("~/Projects/StressGranules/SGPAPER/m6SimRes.txt")


upmarkers <- m6ares[
  !(m6ares$bins %in% c(0, 1, 2, 3, 4)), 
  ]


spearman_corr <- cor.test(upmarkers$Residuals.y, (upmarkers$Residuals.x), method = "spearman")

(m6res <- ggplot(m6ares, aes(y=Residuals.y, x=(Residuals.x))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black')  +   theme_bw()+xlab("Original Residuals")+
  ylab("Simulated Residuals")+ylim(-13,13)+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(text=element_text(size=25,  family="Arial MS")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_point(data=upmarkers, aes(x=Residuals.x, y=  Residuals.y), color="#43F3D5",fill="#43F3D5",alpha=0.5, size=5, shape=21)+geom_density2d(color="#800020", size=1.2)+annotate("text", x = -2.5, y = 10.5, label = expression("For NM" >= 5 * "," ~ rho == 0.23 * "," ~ italic(p) < 3.3e-5), size = 7, color = "black"))

(C4 <- ggarrange(acres, m6res, ncol = 2))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure4C.png", width = 14, height=7, units = "in", res=300)
C4
dev.off()
```


## Figure 5

### A

```{r}
tstable <- fread("~/Projects/StressGranules/SGPAPER/PBoby_noBC.txt")

smooth.SG <- lm(sg_t~(pb_t), tstable)



# Calculate the Spearman correlation coefficient
spearman_corr <- cor(tstable$sg_t, (tstable$pb_t), method = "spearman")



(sgpcorrel <- ggplot(tstable, aes(y = sg_t, x = pb_t)) + 
  geom_point(aes(color = Length), alpha = 0.5, size = 5) +   # <-- raw Length
  theme_bw() +
  ylab("DGE SG-WC (t-stat)") + 
  xlab("DGE PB-WC (t-stat)") +  
  geom_density2d(color = "#800020", size = 1.2) +
  geom_abline(
    slope = smooth.SG$coefficients[2], 
    intercept = smooth.SG$coefficients[1], 
    linetype = "dashed"
  ) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 7) +
  scale_colour_gradient(
    low = "#729EA1", high = "#A82445",
    trans = "log10",
    breaks = c(10, 100, 1000, 10000, 100000),
    labels = c("10", "100", "1000", "10000", "100000")
  ) +
  labs(colour = "RNA\nLength") +
  theme(
    text = element_text(size = 25, family = "Arial MS"),
    plot.title = element_text(hjust = 0.5),
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure5A.png", width = 7, height=6, units = "in", res=300)
sgpcorrel
dev.off()

```
## C

```{r}

smooth.SG <- lm(pb_t~(rnp_t), tstable)


# Calculate the Spearman correlation coefficient
spearman_corr <- cor(tstable$pb_t, (tstable$rnp_t), method = "spearman")




(rnpcorrel <- ggplot(tstable, aes(y = pb_t, x = rnp_t)) + 
  geom_point(aes(color = Length), alpha = 0.5, size = 5) +   # <-- raw Length
  theme_bw() +
  ylab("DGE PB-WC (t-stat)") + 
  xlab("DGE RNP-WC (t-stat)") +  
  geom_density2d(color = "#800020", size = 1.2) +
  geom_abline(
    slope = smooth.SG$coefficients[2], 
    intercept = smooth.SG$coefficients[1], 
    linetype = "dashed"
  ) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 7) +
  scale_colour_gradient(
    low = "#729EA1", high = "#A82445",
    trans = "log10",
    breaks = c(10, 100, 1000, 10000, 100000),
    labels = c("10", "100", "1000", "10000", "100000")
  ) +
  labs(colour = "RNA\nLength") +
  theme(
    text = element_text(size = 25, family = "Arial MS"),
    plot.title = element_text(hjust = 0.5),
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure5B.png", width = 7, height=6, units = "in", res=300)
rnpcorrel
dev.off()
```

### B

```{r}

smooth.SG <- lm(sg_t~(rnp_t), tstable)


# Calculate the Spearman correlation coefficient
spearman_corr <- cor(tstable$sg_t, (tstable$rnp_t), method = "spearman")




(rnpsgcorrel <- ggplot(tstable, aes(y = sg_t, x = rnp_t)) + 
  geom_point(aes(color = Length), alpha = 0.5, size = 5) +   # <-- raw Length
  theme_bw() +
  ylab("DGE SG-WC (t-stat)") + 
  xlab("DGE RNP-WC (t-stat)") +  
  geom_density2d(color = "#800020", size = 1.2) +
  geom_abline(
    slope = smooth.SG$coefficients[2], 
    intercept = smooth.SG$coefficients[1], 
    linetype = "dashed"
  ) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 7) +
  scale_colour_gradient(
    low = "#729EA1", high = "#A82445",
    trans = "log10",
    breaks = c(10, 100, 1000, 10000, 100000),
    labels = c("10", "100", "1000", "10000", "100000")
  ) +
  labs(colour = "RNA\nLength") +
  theme(
    text = element_text(size = 25, family = "Arial MS"),
    plot.title = element_text(hjust = 0.5),
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure5.png", width = 7, height=6, units = "in", res=300)
rnpsgcorrel
dev.off()
```

# E and F

```{r}
tstable <- fread("~/Projects/StressGranules/SGPAPER/FAPSvsDC.txt")

smooth.SG <- lm(dc_t~(fc_t), tstable)



# Calculate the Spearman correlation coefficient
spearman_corr <- cor(tstable$dc_t, (tstable$fc_t), method = "spearman")



(sgpcorrel <- ggplot(tstable, aes(y = dc_t, x = fc_t)) + 
  geom_point(aes(color = Length), alpha = 0.5, size = 5) +   # <-- raw Length
  theme_bw() +
  ylab("DGE DC-SG-WC (t-stat)") + 
  xlab("DGE FAPS-SG-WC (t-stat)") +  
  geom_density2d(color = "#800020", size = 1.2) +
  geom_abline(
    slope = smooth.SG$coefficients[2], 
    intercept = smooth.SG$coefficients[1], 
    linetype = "dashed"
  ) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 7) +
  scale_colour_gradient(
    low = "#729EA1", high = "#A82445",
    trans = "log10",
    breaks = c(10, 100, 1000, 10000, 100000),
    labels = c("10", "100", "1000", "10000", "100000")
  ) +
  labs(colour = "RNA\nLength") +
  theme(
    text = element_text(size = 25, family = "Arial MS"),
    plot.title = element_text(hjust = 0.5),
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure5E.png", width = 7, height=6, units = "in", res=300)
sgpcorrel
dev.off()

tstable <- fread("~/Projects/StressGranules/SGPAPER/FAPSPB_batchCor.txt")

smooth.SG <- lm(sg_t~(pb_t), tstable)



# Calculate the Spearman correlation coefficient
spearman_corr <- cor(tstable$sg_t, (tstable$pb_t), method = "spearman")



(sgpcorrel <- ggplot(tstable, aes(y = sg_t, x = pb_t)) + 
  geom_point(aes(color = Length), alpha = 0.5, size = 5) +   # <-- raw Length
  theme_bw() +
  ylab("DGE FAPS-SG-WC (t-stat)") + 
  xlab("DGE FAPS-PB-WC (t-stat)") +  
  geom_density2d(color = "#800020", size = 1.2) +
  geom_abline(
    slope = smooth.SG$coefficients[2], 
    intercept = smooth.SG$coefficients[1], 
    linetype = "dashed"
  ) +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 7) +
  scale_colour_gradient(
    low = "#729EA1", high = "#A82445",
    trans = "log10",
    breaks = c(10, 100, 1000, 10000, 100000),
    labels = c("10", "100", "1000", "10000", "100000")
  ) +
  labs(colour = "RNA\nLength") +
  theme(
    text = element_text(size = 25, family = "Arial MS"),
    plot.title = element_text(hjust = 0.5),
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  ))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure5F.png", width = 7, height=6, units = "in", res=300)
sgpcorrel
dev.off()


```

### D

```{r fig.height=6, fig.width=7}
pbodydc <- fread("~/Projects/StressGranules/SGPAPER/Pboddata.txt")[,c(1,4)]
colnames(pbodydc) <- c("Gene", "DCPB")
newmeth <- fread("~/Projects/StressGranules/SGPAPER/BrunoPbo.txt")[,c(1,4)]
colnames(newmeth) <- c("Gene", "KodalPB")
oldmeth <- fread("~/Projects/StressGranules/SGPAPER/HubsPbo.txt")[,c(1,2)]
colnames(oldmeth) <- c("Gene", "HubsPB")

pbodystuff <- merge(pbodydc, newmeth, by="Gene")
pbodystuff <- merge(pbodystuff, oldmeth, by="Gene")
pbodystuff <- merge(pbodystuff, genelength, by="Gene")

SupFig4A <- ggplot(pbodystuff, aes(y = KodalPB, x = DCPB)) + 
  geom_point(aes(color = Length), alpha = 0.5, size = 5) +  # raw Length
  geom_density2d(color = "#800020", size = 1.2) +
  ylab("DGE PB-WC FAPS (t-stat)") + 
  xlab("DGE PB-WC DC (t-stat)") +
  stat_cor(method = "spearman", cor.coef.name = "rho", size = 7) +
  scale_colour_gradient(
    low = "#729EA1", high = "#A82445",
    trans = "log10",
    breaks = c(10, 100, 1000, 10000, 100000),      # example breaks
    labels = c("10", "100", "1000", "10000", "100000")  # human-readable
  ) +
  labs(colour = "RNA\nLength") +
  theme_bw() +
  theme(
    text = element_text(size = 25, family = "Arial MS"),
    plot.title = element_text(hjust = 0.5),
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure5D.png", width = 7, height=6, units = "in", res=300)
SupFig4A
dev.off()

```




## Figure 6

### A

```{r}
yeast_s <- fread("~/Projects/StressGranules/SGPAPER/YeastSs.txt")

#lm doesnt work if I dont do this first. Lets cubic root lenght and weight
yeast_s$Length <- ((yeast_s$Length)^(1/3))
yeast_s$MW <- ((yeast_s$MW)^(1/3))

yeast_s$average <- yeast_s$Scof_St - yeast_s$Length

yeast_s_mito <- as.data.frame(yeast_s[yeast_s$Cromo=="Mito",] )
# Calculate the Spearman correlation coefficient
spearman_corr <- cor(yeast_s$Scof_St, (yeast_s$Length), method = "spearman")

smooth.SG <- lm(Scof_St~((Length)), yeast_s)
smooth.SG.mito <- lm(Scof_St~((Length)), yeast_s_mito)

upmarkers <- as.data.frame(yeast_s[yeast_s$Cromo=="Mito",] )
smooth.upmarkers <- lm(Scof_St ~ Scof_Unst, upmarkers)

(A6 <- ggplot(yeast_s, aes(y=Scof_St, x=(Length))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)+  theme_bw()+xlab(expression(~sqrt(RNA~Length~log10(nt), 3)))+
  ylab("S")+ scale_colour_gradient(low="black", high="red")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(text=element_text(size=25,  family="Arial MS")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+
  geom_point(data=upmarkers, aes(y=Scof_St, x=  Length), color="#43F3D5",fill="#43F3D5",alpha=0.7, size=5, shape=21)+ stat_cor(method = "spearman", cor.coef.name = "rho", size=7))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure6A.png", width = 7, height=6, units = "in", res=300)
A6
dev.off()
```

### B


```{r}
#From AKhong code
human_s <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/RealAkhong.txt")
smooth.SG <- lm(old_t ~ ss_t, human_s)

(B6 <- ggplot(human_s, aes(y=old_t, x=(ss_t))) + 
  geom_point(aes(colour=log10(Length)),alpha=0.5, size=5) +   
  geom_density2d(color="#800020", size=1.2)+  theme_bw()+xlab("Original SG Enrichment (t-stat)") + ylab("Simulated SG Enrichment (t-stat)")
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(text=element_text(size=15,  family="Arial MS")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=5)+
  scale_colour_gradient(low="#729EA1", high = "#A82445")+
  labs(colour = "RNA\nLength"))
```

```{r fig.height=6, fig.width=14}
(Figure6AB <- ggarrange(A6,B6, ncol = 2,labels = c("A","B"), font.label = list(size=25)))
```


### C

```{r fig.height=6, fig.width=7}
RealDeg <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/RealAkhong.txt")[,c(1,4)]
colnames(RealDeg) <- c("Gene", "t")

RealDeg <- merge(RealDeg, genelength, by="Gene")

smooth.SG <- lm(t~log10(Length), RealDeg)

(C6 <- ggplot(RealDeg, aes(y=t, x=log10(Length))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)+ theme_bw()+
 ylab("DGE SimSG-WC (t-stat)") + xlab(expression(paste("RNA Length (log"[10]*" (nt))"))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  + ylim(-60,50)+
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+
  theme(text=element_text(size=25,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  scale_x_continuous(
    breaks = log10(c(10, 100, 1000, 10000)),  # tick positions (in log10 scale)
    labels = c("10", "100", "1000", "10000") # what you want to show
  ))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure6C.png", width = 7, height=6, units = "in", res=300)
C6
dev.off()
```
### D

```{r fig.height=6, fig.width=8}
hallmarks.res.tidy <- fread("~/Projects/StressGranules/SGPAPER/SGvsWC/CorrectedHallmarkSG17.txt")[,-1]


#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
(D6 <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) + 
  coord_flip() +
  labs(x="", y="Normalized Enrichment Score (NES)",
       title= "Hallmark Pathways",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none"))+ 
  theme(text=element_text(size=15,  family="Arial MS"))

```

```{r fig.height=6, fig.width=14}
(Figure6CD <- ggarrange(C6,D6, ncol = 2,labels = c("C","D"), font.label = list(size=25)))
```

### E

```{r}
somas <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/RealSoma.txt")[,c(1,4)]
colnames(somas) <- c("Gene","t")

somas <- merge(somas, genelength, by="Gene")

smooth.SG <- lm(t~log10(Length), somas)

(A8 <- ggplot(somas, aes(y=t, x=log10(Length))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)+ theme_bw()+
 ylab("DGE Controlled PL-SG (t-stat)") + xlab(expression(paste("RNA Length (Log"[10]*" (nt))"))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ 
  theme(text=element_text(size=25,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  scale_x_continuous(
    breaks = log10(c(10, 100, 1000, 10000)),  # tick positions (in log10 scale)
    labels = c("10", "100", "1000", "10000") # what you want to show
  ))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/FigureA8.png", width = 7, height=6, units = "in", res=300)
A8
dev.off()
```


### F

```{r}
hallmarks.res.tidy <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/RealSomaHallmark.txt")


#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
(B8 <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) + 
  coord_flip() +
  labs(x="", y="Normalized Enrichment Score (NES)",
       title= "Hallmark Pathways",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none"))+ 
  theme(text=element_text(size=15,  family="Arial MS"))
```

## Sup 4

#### B

```{r}
RealDeg <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/RealAkhong.txt")
colnames(RealDeg)[1] <- "Gene"

ndegs <- 20
upmarkers <- as.data.frame(RealDeg[(RealDeg$Gene) %in% c("NORAD", "GAPDH"),] )

(sup6 <- ggplot(data = RealDeg,
                aes(x = logFC, y = B)) +
  geom_point(color = "#CCCCCC") +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid = element_blank(),      # 🚀 remove ALL gridlines
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  ggtitle("DGE SimSG-SG") +
  ylab("B-Statistic") +
  xlab(expression(Log[2]~"Fold-Change")) +  
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(
    data = upmarkers[1:ndegs,], 
    mapping = aes(label = (upmarkers$Gene)[1:ndegs]), 
    size = 4,
    point.padding = 0.5,
    color = "#A10702",
    max.overlaps = 20,
    segment.color = "#A10702",
    fontface = "italic"
  ) +
  geom_point(data = upmarkers, aes(x = logFC, y = B), color = "#A10702", size = 4) +
  theme(text = element_text(size = 25, family = "Arial MS")) +
  geom_vline(xintercept = -0.75, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0.75, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/SupFigure6.png", width = 7, height=6, units = "in", res=300)
sup6
dev.off()
```





#### A

```{r}
#From Khong Code

smooth.SG <- lm(old_t ~ ss_t, simmyyeast)

(Sup5 <- SimCorrel <- ggplot(simmyyeast, aes(x=ss_t, y=old_t)) + geom_point(aes(colour=log10(Length)),alpha=0.5, size=5) +   theme_bw()+
 xlab("Original SG Enrichment (t-stat)") + ylab("Simulated SG Enrichment (t-stat)") +   
  geom_density2d(color="#800020", size=1.2)+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Yeast Simulation")+
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+
  theme(text=element_text(size=15,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_abline(slope = smooth.SG$coefficients[2], 
              intercept = smooth.SG$coefficients[1], 
              linetype = "dashed") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  scale_colour_gradient(low="#729EA1", high = "#A82445")+
  labs(colour = "RNA\nLength"))



tiff("~/Projects/StressGranules/SGPAPER/SupFigure5.tiff", width = 7, height=6, units = "in", res=300)
Sup5
dev.off()
```

#### C

```{r}
mat2019 <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/RealMatheny2019.txt")
colnames(mat2019)[1] <- "Gene"


upmarkers <- as.data.frame(mat2019[(mat2019$Gene) %in% c("NORAD", "GAPDH", "ACTB"),] )


(sup7 <- ggplot(data = mat2019,
                aes(x = logFC, y = B)) +
  geom_point(color = "#CCCCCC") +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid = element_blank(),      # 🚀 remove ALL gridlines
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  ggtitle("DGE Controlled RNP") +
  ylab("B-Statistic") +
  xlab(expression(Log[2]~"Fold-Change")) +  
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_label_repel(
    data = upmarkers[1:ndegs,], 
    mapping = aes(label = (upmarkers$Gene)[1:ndegs]), 
    size = 4,
    point.padding = 0.5,
    color = "#A10702",
    max.overlaps = 20,
    segment.color = "#A10702",
    fontface = "italic"
  ) +
  geom_point(data = upmarkers, aes(x = logFC, y = B), color = "#A10702", size = 4) +
  theme(text = element_text(size = 25, family = "Arial MS")) +
  geom_vline(xintercept = -1, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 1, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5))


png("~/Projects/StressGranules/SGPAPER/Images/PNG/SupFigure7.png", width = 7, height=6, units = "in", res=300)
sup7
dev.off()
```

#### D

```{r}
mat2019 <- merge(mat2019, genelength)

smooth.SG <- lm(t~log10(Length), mat2019)

(A7 <- ggplot(mat2019, aes(y=t, x=log10(Length))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2)+ theme_bw()+
 ylab("DGE Controlled RNP") + xlab(expression(paste("RNA Length (Log"[10]*" (nt))"))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5))  +
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ 
  theme(text=element_text(size=25,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7)+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  scale_x_continuous(
    breaks = log10(c(10, 100, 1000, 10000)),  # tick positions (in log10 scale)
    labels = c("10", "100", "1000", "10000") # what you want to show
  ))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/FigureA7.png", width = 7, height=6, units = "in", res=300)
A7
dev.off()
```
#### E

```{r fig.height=6, fig.width=8}
hallmarks.res.tidy <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/RealMatheny2019_Hallmarks.txt")


#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>%
  filter(padj < thresh) %>%
  mutate(pathway = gsub("^HALLMARK_", "", pathway),
         pathway = gsub("_", " ", pathway))

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = Status), alpha = 0.75) + 
  scale_fill_manual(values = c("#53A2BE", "#BA3B46"), 
                    name = "Regulation Status", 
                    labels = c("Downregulated", "Upregulated"), 
                    breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x = "Gene Set", 
       y = "Normalized Enrichment Score (NES)",
       title = "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) +
  theme_bw() +
  theme(legend.position = "none",
        text = element_text(size = 25, family = "Arial MS", face = "bold"),
        plot.title = element_text(hjust = 1, face = "bold"),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "bold"))

```

#### F


```{r}
hallmarks.res.tidy <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/WCHallmarks.txt")


#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
(fi9 <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) + 
  coord_flip() +
  labs(x="", y="Normalized Enrichment Score (NES)",
       title= "Hallmark Pathways",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none"))+ 
  theme(text=element_text(size=15,  family="Arial MS"))

tiff("~/Projects/StressGranules/SGPAPER/Figure9.tiff", width = 8, height=6, units = "in", res=300)
fi9
dev.off()
```
## Figure 7

### A

```{r fig.width=10, fig.height=10}
#I get the data I used for the previous figure but add the values from the other studies
#Opening the Different Studies
tcellpb <- fread("../Projects/StressGranules/SGPAPER/RealSGs/TcellPB.txt")[,c(1,4)]
colnames(tcellpb) <- c("Gene","tcellpb")

rcapseq <- fread("../Projects/StressGranules/SGPAPER/RealSGs/CapSeq.txt")[,c(1,4)]
colnames(rcapseq) <- c("Gene","rcapseq")

rm6a <- fread("../Projects/StressGranules/SGPAPER/RealSGs/Realm6A.txt")[,c(1,4)]
colnames(rm6a) <- c("Gene","rm6a")

wcdsrna <- fread("../Projects/StressGranules/SGPAPER/RealSGs/WCdsRNA.txt")[,c(1,4)]
colnames(wcdsrna) <- c("Gene","wcds")

rsoma <- fread("../Projects/StressGranules/SGPAPER/RealSGs/RealSoma.txt")[,c(1,4)]
colnames(rsoma) <- c("Gene","rsoma")

rac <- fread("../Projects/StressGranules/SGPAPER/RealSGs/Realac4c.txt")[,c(1,4)]
colnames(rac) <- c("Gene","rac")

mat21 <- fread("../Projects/StressGranules/SGPAPER/RealSGs/RealtMat21.txt")[,c(1,4)]
colnames(mat21) <- c("Gene","mat21")

wcog <- fread("../Projects/StressGranules/SGPAPER/RealSGs/RealWC.txt")[,c(1,4)]
colnames(wcog) <- c("Gene","wcog")

mat19 <- fread("../Projects/StressGranules/SGPAPER/RealSGs/Tmathenypb.txt")[,c(1,4)]
colnames(mat19) <- c("Gene","mat19")

ak17 <- fread("../Projects/StressGranules/SGPAPER/RealSGs/RealAkhong.txt")[,c(1,4)]
colnames(ak17) <- c("Gene","ak17")

iade <- fread("../Projects/StressGranules/SGPAPER/RealSGs/RealvIadevaia.txt")[,c(1,4)]
colnames(iade) <- c("Gene","iade")

raj <- fread("../Projects/StressGranules/SGPAPER/RealSGs/Photon.txt")[,c(1,4)]
colnames(raj) <- c("Gene","raj")

fanci <- fread("../Projects/StressGranules/SGPAPER/RealSGs/FANCI.txt")[,c(1,4)]
colnames(fanci) <- c("Gene","fanci")

fig10 <- merge(wcog, wcdsrna, by="Gene", all=T)
fig10 <- merge(fig10, rsoma, by="Gene", all=T)
fig10 <- merge(fig10, rcapseq, by="Gene", all=T)
fig10 <- merge(fig10, mat19, by="Gene", all=T)
fig10 <- merge(fig10, tcellpb, by="Gene", all=T)
fig10 <- merge(fig10, iade, by="Gene", all=T)
fig10 <- merge(fig10, ak17, by="Gene", all=T)
fig10 <- merge(fig10, mat21, by="Gene", all=T)
fig10 <- merge(fig10, rm6a, by="Gene", all=T)
fig10 <- merge(fig10, rac, by="Gene", all=T)
fig10 <- merge(fig10, raj, by="Gene", all=T)
fig10 <- merge(fig10, fanci, by="Gene", all=T)
fig10 <- merge(fig10, genelength, by="Gene")

fig10$Length <- log10(fig10$Length)

colnames(fig10) <- c("Gene", "T. Matheny Both", "M. Paget 2023", "SN. Somasekharan 2020", "Z. Ren 2023" , "T. Matheny 2019", "N. Curdy 2023", "V. Iadevaia 2022","A. Khong 2017", "T. Matheny 2021",  "G. Di Timoteo 2024","P. Kudrin 2024","S. Rajachandran 2024", "Y. Zhou 2024",  "RNA Length")


fig10_cor <- cor(fig10[,-1],method="spearman", use = "complete.obs")

fig10_cor <- reshape2::melt(fig10_cor)

(pl1 <- fig10_cor %>%
  arrange(Var1) %>%
    filter(row_number() <= which(Var1 == Var2)) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "#5e5e5e", size = 0.5) +
  #geom_text(aes(label = ifelse(is.na(value), "", sprintf("%.2f", value))), size = 7) +
  scale_fill_gradient2(low = "#729EA1", mid = "white", high = "#DB5375", limit = c(-1, 1)) +
  labs(fill = "Spearman's Rho") +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(angle = 0, hjust = 1, size = 20, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 20, family = "Arial MS", face = "bold"),
    legend.position = "top"
  )+
  ylab("") + xlab("") +
  scale_y_discrete(limits = sort(unique(fig10_cor$Var2), decreasing = TRUE)))


#Method
df <- data.frame(
  x = (c("wc", "wc", "pl", "pl" , "pb", "pb", "condc","cdc", "cdc",  "cdc","cdc","photon", "fanci",  "length")),
  colour = c("T. Matheny Both", "M. Paget 2023", "SN. Somasekharan 2020", "Z. Ren 2023" , "T. Matheny 2019", "N. Curdy 2023", "V. Iadevaia 2022","A. Khong 2017", "T. Matheny 2021",  "G. Di Timoteo 2024","P. Kudrin 2024","S. Rajachandran 2024", "Y. Zhou 2024",  "RNA Length"),
  xpos = (c("T. Matheny Both", "M. Paget 2023", "SN. Somasekharan 2020", "Z. Ren 2023" , "T. Matheny 2019", "N. Curdy 2023", "V. Iadevaia 2022","A. Khong 2017", "T. Matheny 2021",  "G. Di Timoteo 2024","P. Kudrin 2024","S. Rajachandran 2024", "Y. Zhou 2024",  "RNA Length"))
)

df$x <- factor(df$x, levels = unique(df$x))
df$xpos <- factor(df$xpos, levels = (df$xpos))

colour_vector <- c(
  "wc" = '#B5EAD7',
  "pl" = '#FFDAC1',
  "pb" = '#AFCBFF',
  "condc" = '#C7CEEA',
  "cdc" = '#FFE5B4',
  "photon" ='#FF9AA2',
  "fanci" = '#FFC3A0',
  "length" = 'white'
)

lebel_vector <- c(
  "wc" = 'Whole-Cell',
  "pl" = 'P. Labelling',
  "pb" = 'vs P-Body',
  "condc" = 'Controlled DC',
  "cdc" = 'Corrected DC',
  "photon" ='PHOTON',
  "fanci" = 'FANCI',
  "length" = 'RNA Length'
)

(pl2 <- ggplot(df, aes(x = xpos, y = 1, fill = x)) +
  geom_tile() +
  theme_void() +
  labs(fill = "SG Purification \nMethod") +
  scale_fill_manual(
    values = colour_vector,labels = lebel_vector,
    breaks = names(lebel_vector)
  ) +
  scale_x_discrete(
    labels = c("T. Matheny Both", "M. Paget 2023", "SN. Somasekharan 2020", "Z. Ren 2023" , "T. Matheny 2019", "N. Curdy 2023", "V. Iadevaia 2022","A. Khong 2017", "T. Matheny 2021",  "G. Di Timoteo 2024","P. Kudrin 2024","S. Rajachandran 2024", "Y. Zhou 2024",  "RNA Length")
  ) +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1),
    text = element_text(size = 20, family = "Arial MS", color = "#505050", face = "bold")
  ) +
  ylab("") + xlab(""))



heatmap <- pl1 + theme(axis.text.x = element_blank(),
                      axis.title.y = element_blank(),
         axis.ticks.x = element_blank(), legend.position="top", axis.title.x = element_blank())

 
heatmap <-  heatmap + theme(strip.text.x = element_blank(),
      strip.background = element_rect(colour="white", fill="white"),
      legend.position='none', legend.direction="vertical"
      )
  

heatmap <- heatmap +
  theme(
    plot.margin = grid::unit(c(0, 0, 0, 6), "lines"),
    axis.text.y = element_text(angle = 0, hjust = 1, size = 20, face = "bold")
  )

pl2 <- pl2 +
  theme(
    plot.margin = grid::unit(c(0, 0, 6, 0), "lines")
  )



(left <-   ggarrange(heatmap,pl2, heights = c(.575,.425), ncol=1, align="v"))

# png("../Projects/StressGranules/SGPAPER/Images/PNG/7A.png", width = 10, height=10, units = "in", res=300)
# left
# dev.off()

heatmap_leg <- get_legend(pl1+ theme(text=element_text(size=25,  family="Arial MS", color="#505050")), position = 'bottom')
heatmap_leg <- as_ggplot(heatmap_leg)
leg1 <- get_legend(pl2+ theme(text=element_text(size=25,  family="Arial MS", color="#505050")), position = 'right')
leg1 <- as_ggplot(leg1)

(right <- ggarrange(heatmap_leg,leg1,heights = c(.05,.2), nrow = 2))

(A <- ggarrange(left, right, ncol = 2, widths = c(.8,.3)))

# png("../Projects/StressGranules/SGPAPER/Images/PNG/7A_leg.png", width = 10, height=10, units = "in", res=300)
# right
# dev.off()
  
```

### C

```{r}
DEGenes <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/MergedWCSig.txt")
DEGenes <- as.data.frame(DEGenes)
colnames(DEGenes)[1] <- "Gene"
row.names(DEGenes) <- DEGenes$Gene

upmarkers <- DEGenes[DEGenes$logFC > 1 & DEGenes$B > 15, ]
downmarkers <- DEGenes[DEGenes$logFC < -2 & DEGenes$B > 15, ]
mitomarkers <- DEGenes[DEGenes$Gene %in% MitoGenes & DEGenes$logFC < -1, ]

ndegs <- 50  # number of top genes to label, adjust as needed

(mergedwc <- ggplot(DEGenes, aes(x=logFC, y=B)) +
  geom_point(color="#CCCCCC") +
  geom_point(data=upmarkers, aes(x=logFC, y=B), color="#BA3B46", size=4) +
  geom_point(data=downmarkers, aes(x=logFC, y=B), color="#53A2BE", size=4) +
  geom_point(data=mitomarkers, aes(x=logFC, y=B), color="#73937E", size=4) +
  geom_label_repel(data = upmarkers[1:ndegs,], aes(label=Gene), size=4, color="#BA3B46", fontface="italic") +
  geom_label_repel(data = downmarkers[1:ndegs,], aes(label=Gene), size=4, color="#53A2BE", fontface="italic") +
  geom_label_repel(data = mitomarkers[1:ndegs,], aes(label=Gene), size=4, color="#73937E", max.overlaps = 20, fontface="italic") +
  theme_bw() +
  labs(x=expression(Log[2]~"Fold-Change"), y="B-Statistic", title="WC-Transcriptome", subtitle = "With SG") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),  # remove major gridlines
    panel.grid.minor = element_blank(),  # remove minor gridlines
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    text = element_text(size=25, family="Arial MS")
  )
)

png("~/Projects/StressGranules/SGPAPER/Images/PNG/FigureB8.png", width = 7, height=6, units = "in", res=300)
mergedwc
dev.off()

```
### D

```{r fig.width=16, fig.height=6}
#fig10 <- fig10[,-c(15)]
#-------------------------------
# 1. Gene sets
#-------------------------------
for.chisq <- list()

MitoGenes <- fread("~/MitoGenes.txt")[[1]]
for.chisq[["Mito_MITOCHONDRIAL_ENCODED"]] <- MitoGenes

MitoNuclear <- fread("~/rnalocate_mito.csv")
MitoNuclear <- subset(MitoNuclear, Species == "Homo sapiens")
MitoNuclear <- subset(MitoNuclear, `RNA Category` %in% c("lncRNA","mRNA","rRNA","tRNA"))
MitoNuclear <- MitoNuclear[!duplicated(MitoNuclear$`RNA Symbol`), ]
MitoNuclear <- subset(MitoNuclear, Score > 0.35)
for.chisq[["Mito_NUCLEAR_ENCODED"]] <- MitoNuclear$`RNA Symbol`

# Loop over datasets
all_fgsea <- list()
for(dataset in colnames(fig10)[-1]) {  # skip GeneSymbol column
  ranks <- setNames(fig10[[dataset]], fig10$Gene)
  ranks <- sort(ranks, decreasing = TRUE)

  fgseaRes <- fgsea(
    pathways = for.chisq,
    stats = ranks,
    minSize = 10,
    maxSize = 500
  )

  fgseaRes$Dataset <- dataset
  all_fgsea[[dataset]] <- fgseaRes
}

# Combine results
fgsea_combined <- bind_rows(all_fgsea)

# Prepare long-format data for ggplot
fgsea_plot <- fgsea_combined %>%
  select(pathway, Dataset, NES, padj) %>%
  mutate(
    pathway = recode(pathway,
                     "Mito_MITOCHONDRIAL_ENCODED" = "Mito (mito-encoded)",
                     "Mito_NUCLEAR_ENCODED" = "Mito (nuclear-encoded)"),
    sig = ifelse(padj < 0.05, "*", "")  # mark significance
  )

# Define the desired dataset order
dataset_order <- c(
  "T. Matheny Both", "M. Paget 2023", "SN. Somasekharan 2020", "Z. Ren 2023",
  "T. Matheny 2019", "N. Curdy 2023", "V. Iadevaia 2022", "A. Khong 2017",
  "T. Matheny 2021", "G. Di Timoteo 2024", "P. Kudrin 2024", "S. Rajachandran 2024",
  "Y. Zhou 2024"
)

# Add a label column with NES rounded to 2 decimals
fgsea_plot <- fgsea_plot %>%
  mutate(NES_label = sprintf("%.2f", NES))

# Plot heatmap with NES values
(mito <- ggplot(fgsea_plot, aes(x = Dataset, y = pathway, fill = NES)) +
  geom_tile(color = "black")  +  # NES on tiles
  scale_fill_gradient2(low = "#729EA1", mid = "white", high = "#DB5375", midpoint = 0) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 25),
    axis.text.y = element_text(size = 25),
    axis.title = element_blank(),
    legend.title = element_text(size = 25),
    legend.text = element_text(size = 25)
  ) +
  labs(fill = "NES"))

png("../Projects/StressGranules/SGPAPER/Images/PNG/mito.png", width = 16, height=6, units = "in", res=300)
mito
dev.off()
```


### B

```{r}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,4]),]
gseatool <- DEGenes[,4]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- (DEGenes$Gene)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}



hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmarks.res.tidy$pathway <- sub("^[^_]*_", "", hallmarks.res.tidy$pathway)


#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

(gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = Status), alpha = 0.75) + 
  scale_fill_manual(values = c("#53A2BE", "#BA3B46"), 
                    name = "Regulation Status", 
                    labels = c("Downregulated", "Upregulated"), 
                    breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x = "Gene Set", 
       y = "Normalized Enrichment Score",
       title = "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) +
  theme_bw() +
  theme(legend.position = "none",
        text = element_text(size = 25, family = "Arial MS", face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "bold")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/MergedWC_genesets.png", width = 14, height=10, units = "in", res=300)
gsea_plot_hall
dev.off()
```


### E

```{r}


#for all genes
SG_Transcriptome <- SG_Transcriptome[order(SG_Transcriptome[,6]),]
gseatool <- SG_Transcriptome[,6]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- (SG_Transcriptome$Gene)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}



hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmarks.res.tidy$pathway <- sub("^[^_]*_", "", hallmarks.res.tidy$pathway)


#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

(gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = Status), alpha = 0.75) + 
  scale_fill_manual(values = c("#53A2BE", "#BA3B46"), 
                    name = "Regulation Status", 
                    labels = c("Downregulated", "Upregulated"), 
                    breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x = "Gene Set", 
       y = "Normalized Enrichment Score",
       title = "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) +
  theme_bw() +
  theme(legend.position = "none",
        text = element_text(size = 25, family = "Arial MS", face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "bold")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/sgenriched.png", width = 14, height=10, units = "in", res=300)
gsea_plot_hall
dev.off()
```

## Supplemenatary Figure 5

#### Function
```{r}
hallmarker <- function(table, dataset){
  df_ordered <- table[order(table[,..dataset]),]
  gseatool <- df_ordered[,..dataset]
  gseatool <- as.data.frame(gseatool)
  colnames(gseatool)[1] <- "t" 
  gseatool[,2] <- df_ordered$Gene
  colnames(gseatool)[2] <- "Symbol" 
  gseatool <- gseatool[,c(2,1)]
  
  vectorranks <- gseatool$t
  names(vectorranks) <- gseatool$Symbol
  vectorranks <- na.omit(vectorranks)
  
  
  #to load our pathways
  hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")
  
  hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)
  
  for (set in names(hallmarks.gsets)){
    hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
  }
  
  hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                         stats = vectorranks)
  
  hallmark.sg.t <- hallmarks.res[,c(1,6)]
  hallmark.sg.t <- as.data.frame(hallmark.sg.t)
  colnames(hallmark.sg.t)[2] <- colnames(table)[dataset]
  return(hallmark.sg.t)
}
```

```{r fig.width=25, fig.height=25}
pathwayer <- fig10

# Initialize an empty list to store the resulting data frames
results_list <- list()

# Loop through and store each dataframe in the list
for (n in 2:15) {
  pathn <- hallmarker(table=pathwayer, dataset=n)
  results_list[[n-1]] <- pathn  # Store each data frame in the list
}

# Merge all data frames in the list by the "pathway" column
merged_data <- Reduce(function(x, y) merge(x, y, by = "pathway", sort = FALSE), results_list)


# Assuming your 'merged_data' is in wide format and the first column is 'pathway'
merged_data$pathway <- sub("^[^_]*_", "", merged_data$pathway)
merged_data$pathway <- gsub("_", " ", merged_data$pathway)

rownames(merged_data) <- merged_data$pathway
mat <- as.matrix(merged_data[, -1])  # Convert to matrix without 'pathway' column

#what pathways are more important?
# Using sum of absolute NES
pathway_scores_sums <- rowSums(abs(merged_data[,-c(1,15)]), na.rm = TRUE)
pathway_scores_mean <- rowMeans(abs(merged_data[,-c(1,15)]), na.rm = TRUE)

# Sort descending
pathway_scores_sums <- sort(pathway_scores_sums, decreasing = TRUE)
pathway_scores_mean <- sort(pathway_scores_mean, decreasing = TRUE)

# Top 10 pathways
top_pathways_sum <- names(pathway_scores_sums)[1:10]
top_scores_mean <- names(pathway_scores_mean)[1:10]

common_top_pathways <- intersect(top_pathways_sum, top_scores_mean)
common_top_pathways


# Plot heatmap with clustering and dendrograms
library(grid)

supheat <- pheatmap::pheatmap(mat, color = colorRampPalette(c("#729EA1", "white", "#DB5375"))(50),
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         display_numbers = FALSE,
         cluster_rows = T,
         legend = T,
         fontsize = 25,
         cellwidth = 50,
         cellheight=25, border_color = "black") 

png("../Projects/StressGranules/SGPAPER/supheat.png", width = 25, height=25, units = "in", res=300)
supheat
dev.off()


#Adding extra label for method

#Method
df <- data.frame(
  x = (c("wc", "wc", "pl", "pl" , "pb", "pb", "condc","cdc", "cdc",  "cdc","cdc","photon", "fanci",  "length")),
  colour = c("T. Matheny Both", "M. Paget 2023", "SN. Somasekharan 2020", "Z. Ren 2023" , "T. Matheny 2019", "N. Curdy 2023", "V. Iadevaia 2022","A. Khong 2017", "T. Matheny 2021",  "G. Di Timoteo 2024","P. Kudrin 2024","S. Rajachandran 2024", "Y. Zhou 2024",  "RNA Length"),
  xpos = (c("T. Matheny Both", "M. Paget 2023", "SN. Somasekharan 2020", "Z. Ren 2023" , "T. Matheny 2019", "N. Curdy 2023", "V. Iadevaia 2022","A. Khong 2017", "T. Matheny 2021",  "G. Di Timoteo 2024","P. Kudrin 2024","S. Rajachandran 2024", "Y. Zhou 2024",  "RNA Length"))
)

df$x <- factor(df$x, levels = unique(df$x))
desired_order <- c("V. Iadevaia 2022", "Y. Zhou 2024", "T. Matheny 2021", "A. Khong 2017",
                   "G. Di Timoteo 2024", "S. Rajachandran 2024", "RNA Length", "Z. Ren 2023",
                   "T. Matheny Both", "N. Curdy 2023","SN. Somasekharan 2020","M. Paget 2023",
                   "T. Matheny 2019", "P. Kudrin 2024")

df$xpos <- factor(df$xpos, levels = desired_order)


colour_vector <- c(
  "wc" = '#B5EAD7',
  "pl" = '#FFDAC1',
  "pb" = '#AFCBFF',
  "condc" = '#C7CEEA',
  "cdc" = '#FFE5B4',
  "photon" ='#FF9AA2',
  "fanci" = '#FFC3A0',
  "length" = 'white'
)

lebel_vector <- c(
  "wc" = 'Whole-Cell',
  "pl" = 'P. Labelling',
  "pb" = 'vs P-Body',
  "condc" = 'Controlled DC',
  "cdc" = 'Corrected DC',
  "photon" ='PHOTON',
  "fanci" = 'FANCI',
  "length" = 'RNA Length'
)

(pl2 <- ggplot(df, aes(x = xpos, y = 1, fill = x)) +
  geom_tile() +
  theme_void() +
  labs(fill = "SG Purification \nMethod") +
  scale_fill_manual(
    values = colour_vector,labels = lebel_vector,
    breaks = names(lebel_vector)
  ) +
  scale_x_discrete(
    labels = c("V. Iadevaia 2022", "Y. Zhou 2024", "T. Matheny 2021", "A. Khong 2017" , "G. Di Timoteo 2024", "S. Rajachandran 2024", "RNA Length","Z. Ren 2023", "T. Matheny Both",  "N. Curdy 2023","SN. Somasekharan 2020","M. Paget 2023", "T. Matheny 2019",  "P. Kudrin 2024")
  ) +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1),
    text = element_text(size = 25, family = "Arial MS", color = "#505050", face = "bold")
  ) +
  ylab("") + xlab("")+
  theme(
    plot.margin = grid::unit(c(6, 6, 6, 6), "lines")
  )
)

png("../Projects/StressGranules/SGPAPER/supheat_leg.png", width = 25, height=25, units = "in", res=300)
pl2
dev.off()

leg1 <- get_legend(pl2+ theme(text=element_text(size=25,  family="Arial MS", color="#505050")), position = 'right')
leg1 <- as_ggplot(leg1)

png("../Projects/StressGranules/SGPAPER/supheat_leg_label.png", width = 25, height=25, units = "in", res=300)
leg1
dev.off()

```


## Figure 8

#### A

```{r}
#MitoGenes <- c(MitoGenes, "MT-TF")


SG_Transcriptome <- fread("~/Projects/StressGranules/SGPAPER/SGEnrichedTranscriptsv2.txt")[,-c(1)]
SG_Transcriptome <- na.omit(SG_Transcriptome)
SG_Transcriptome$Tot <- rowMeans(SG_Transcriptome[,-1], na.rm = T)
SG_nonMito <- SG_Transcriptome[!(SG_Transcriptome$Gene %in% MitoGenes), ]
WC_Transcriptome <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/MergedWCSig.txt")[,c(1,4)]
colnames(WC_Transcriptome) <- c("Gene", "WC")

Both_Transcriptomes <- merge(WC_Transcriptome, SG_Transcriptome, by="Gene")

smooth.SG <- lm(Tot~WC, Both_Transcriptomes)
Both_Transcriptomes$Inverser <- Both_Transcriptomes$WC * Both_Transcriptomes$Tot
row.names(Both_Transcriptomes) <- Both_Transcriptomes$Gene
linemarkers <- Both_Transcriptomes[abs(Both_Transcriptomes$Inverser) > 30, ]

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(Both_Transcriptomes$Tot, (Both_Transcriptomes$WC), method = "spearman")



(Fig8 <- ggplot(Both_Transcriptomes, aes(y = Tot, x = (WC))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2) +
  theme_bw() +
  ylab("DGE SG-Enriched (Mean of t-stats)") + 
  xlab("DGE WC with vs Without SG (t-stat)") + # Subscript "10"
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())  + 
  theme(text=element_text(size=25,  family="Arial MS"))  +
  theme(plot.title = element_text(hjust = 0.5))  + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed") +
  geom_point(data=linemarkers, aes(x = WC, y = Tot ),alpha = 0.5, size=5, shape=21, fill="#de425b")+ geom_label_repel(
    data = linemarkers, 
    mapping = aes(label = linemarkers$Gene), 
    size = 4,
    point.padding = 0.5,  # Positive point padding
    color = "#A10702",
    max.overlaps = 30,
    segment.color = "#A10702"  # Ensures the segments are visible
  )+ stat_cor(method = "spearman", cor.coef.name = "rho", size=7, label.y = -9))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure8.png", width = 8, height=7, units = "in", res=300)
Fig8
dev.off()
```







#### B

```{r fig.height=6, fig.width=14}
wcdsrnamer <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/MergedWCSig.txt")[,c(1,4)]
colnames(wcdsrnamer) <- c("Gene", "merg")
simfns <- fread("~/Projects/StressGranules/SGPAPER/siMFNboth.txt")[,c(1,4)]
colnames(simfns) <- c("Gene", "simfns")

dfforheat <- merge(wcdsrnamer, simfns, by="Gene")

smooth.SG <- lm(simfns~(merg), dfforheat)

dfforheat$dist <- (dfforheat$simfns) * (dfforheat$merg)

linemarkers <- as.data.frame(dfforheat[(dfforheat$dist)>20,])
 

(D <- ggplot(dfforheat, aes(y = simfns, x = (merg))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2) +
  theme_bw() +
  ylab("DGE MAVS KO vs WT (t-stat)") + 
  xlab("DGE WC with vs Without SG (t-stat)") + # Subscript "10"
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())  + 
  theme(text=element_text(size=25,  family="Arial MS")) + stat_cor(method = "spearman", cor.coef.name = "rho", size=7) +
  theme(plot.title = element_text(hjust = 0.5))  + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed") +
  geom_point(data=linemarkers, aes(x = merg, y = simfns ),alpha = 0.5, size=5, shape=21, fill="#de425b")+ geom_label_repel(
    data = linemarkers, 
    mapping = aes(label = linemarkers$Gene), 
    size = 4,
    point.padding = 0.5,  # Positive point padding
    color = "#A10702",
    max.overlaps = 30,
    segment.color = "#A10702"  # Ensures the segments are visible
  ) )

png("~/Projects/StressGranules/SGPAPER/Images/PNG/senesscatter.png", width = 8, height=7, units = "in", res=300)
D
dev.off()
```
#### C

```{r}

wcdsrnamer <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/MergedWCSig.txt")[,c(1,4)]
colnames(wcdsrnamer) <- c("Gene", "merg")
simfns <- fread("~/Projects/StressGranules/SGPAPER/siMFNboth.txt")[,c(1,4)]
colnames(simfns) <- c("Gene", "simfns")

dfforheat <- merge(wcdsrnamer, simfns, by="Gene")


#for all genes
DEGenes <- dfforheat[order(dfforheat[,3]),]
gseatool <- DEGenes[,3]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- DEGenes$Gene
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection  = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

#MergeSGPath <- hallmarks.res.tidy[,c(1,6)]
SenesPath <- hallmarks.res.tidy[,c(1,6)]

Paths <- merge(MergeSGPath, SenesPath, by="pathway")
colnames(Paths) <- c("Pathway", "MergedWC", "Senescence")

vip <- as.data.frame(Paths[Paths$MergedWC < -1.7 & Paths$Senescence < -1.7,] )

(E <- ggplot(Paths)+geom_point(aes(x=MergedWC, y=Senescence), colour="#506863", shape=17,size=7,alpha=.5)+  theme_classic()+
  ylab("MFN1/2 KO vs WT NES") + 
  xlab("WC with vs Without SG NES")  + geom_abline(slope = 1, intercept = 0, linetype="dashed", colour="black")+geom_abline(slope = -1, intercept = 0, linetype="dashed", colour="black", alpha=.5)+geom_vline(xintercept = 0, linetype="dotted")+geom_hline(yintercept = 0, linetype="dotted")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text=element_text(size=25,  family="Arial MS")) +xlim(c(-3.5,3.5))+ylim(c(-3.5,3.5))+
  geom_label_repel(
    data = vip[1:ndegs,], 
    mapping = aes(x=MergedWC, y=  Senescence, label = (vip$Pathway)[1:ndegs]), 
    size = 4,
    point.padding = 0.5,  # Positive point padding
    color = "#506863",
    max.overlaps = 50,
    segment.color = "#506863", fontface="bold"  # Ensures the segments are visible
  ))



#Fisher Test Example
PL_Q1 <- nrow(Paths[Paths$MergedWC>0 & Paths$Senescence>0,])
PL_Q2 <- nrow(Paths[Paths$MergedWC>0 & Paths$Senescence<0,] )
PL_Q3 <- nrow(Paths[Paths$MergedWC<0 & Paths$Senescence>0,] )
PL_Q4 <- nrow(Paths[Paths$MergedWC<0 & Paths$Senescence<0,] )

PL_matrix <- matrix(c(PL_Q1, PL_Q2, PL_Q3, PL_Q4), nrow = 2, byrow = TRUE)
fisher.test(PL_matrix)




senes <- ggarrange(D, E, ncol = 2, labels = c("A","B"), font.label = list(size=25))


png("~/Projects/StressGranules/SGPAPER/Images/PNG/Senescence.png", width = 8, height=7, units = "in", res=300)
E
dev.off()
```


# Supplementary Figure 6

## Pre Processing

```{r}
TIA1 <- fread("~/Projects/StressGranules/AS.WC_Transcriptome/shRNAExp/TIA1/TrimmedSamples/vast_out/cRPKM_AND_COUNTS-hg38-8.tab")

# Keep only even-numbered columns from 'counts'
counts <- as.data.frame(TIA1)

# Keep only even-numbered columns
counts_even <- counts[, seq(2, ncol(counts), by = 2)]

reads <- counts_even %>%
  group_by(NAME) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

read.count <- reads

read.count <- as.data.frame(read.count)
rownames(read.count) <- read.count$NAME
read.count <- read.count[,-1]


stress.read.counts <- read.count

alternew <- log10(rowSums(stress.read.counts))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

filterednew <- subset(stress.read.counts, alternew>3)
#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))

#Now we need to select only the rows which have those genes.
finaltable <- stress.read.counts[row.names(stress.read.counts)%in%impgenes,]


#Let us prepare metadata to add to our samples
#I start by normalizing them all together
#This one has all samples
D0 <- DGEList(finaltable)


#Then we calculate the normalization factors for these new generated data frames
#Bear in mind this step calculates the normalization factors, but does not normalize itself.
D0 <- calcNormFactors(D0)


#Lets create a dataframe with this information
Cell <- c(rep("HEPG2",4), rep("K562",4))

Stress <- c(rep("CTRL",2), rep("shRNA",2),rep("CTRL",2), rep("shRNA",2))

metadata <- data.frame(Stress=Stress, Cell=Cell)


mm <- model.matrix(~0+Cell+Stress,metadata)
#the one without pbs


#and now we normalize. This step also turns our counts into log2-counts per million.
#The object with all simple
y <- voom(D0, mm, plot = T)


batch_cols <- c("Cell")
metadata_cols <- c("Stress")
mmbatch <- model.matrix(reformulate(paste0("~0 + ",paste0( batch_cols,collapse = " + "))), data=metadata)
mmkeep <- model.matrix(reformulate(paste0("~0 + ",paste0( metadata_cols,collapse =  "+" ))), data=metadata)
mm <- cbind(mmkeep,mmbatch)
D0 <- DGEList(2^y$E) #senescence_normcounts
D0 <- calcNormFactors(D0)
y <- voom(D0, mm, plot = F)
fit <- lmFit(y, mm)
beta <- fit$coefficients[,-(1:ncol(mmkeep)),drop=FALSE]
#beta <- fit$coefficients
beta[is.na(beta)] <- 0
corrcounts_v1 <- as.matrix(y$E) - beta %*% t(mmbatch)
offset <- apply(corrcounts_v1,1,min) - apply(y$E,1,min)
corrcounts_v1 <- as.data.frame(corrcounts_v1-offset)
corrcounts_v1 <- 2^corrcounts_v1


y <- voom(corrcounts_v1, mm, plot = T)


mm <- model.matrix(~0+Stress)
#the one without pbs

y <- voom(corrcounts_v1, mm, plot = T)
#and now we normalize. This step also turns our counts into log2-counts per million.
#The object with all simple
#y <- voom(D0[,-c(20:22)], mm, plot = T)

fitted <- lmFit(y$E, mm)

head(coef(fitted))

contr <- makeContrasts(StressshRNA - StressCTRL , levels = colnames(coef(fitted)))

tmp_contr <- contrasts.fit(fitted, contr)

#tmp <- eBayes(fitted)
tmp <- eBayes(tmp_contr)
top.table <- topTable(tmp, sort.by = "logFC", n=Inf)
DEGenes <- top.table
ndegs=20

#write.table(DEGenes, "~/Projects/StressGranules/SGPAPER/Protein Stuff/TIA1KO_expression.txt", row.names = T, col.names = T)
```

### C

```{r fig.height=8, fig.width=8}
SG <- fread("~/Projects/StressGranules/SGPAPER/RealSGs/WCdsRNA.txt")[,c(1,4)]
colnames(SG) <- c("Gene", "t")
TIA1KO <- fread("~/Projects/StressGranules/SGPAPER/Protein Stuff/TIA1KO_expression.txt")[,c(1,4)]
colnames(TIA1KO) <- c("Gene", "TIAKO")


TIA1Effect <- merge(SG, TIA1KO, by="Gene")

smooth.SG <- lm(t~TIAKO, TIA1Effect)
#Both_Transcriptomes$Inverser <- Both_Transcriptomes$WC * Both_Transcriptomes$Tot
#row.names(Both_Transcriptomes) <- Both_Transcriptomes$Gene
#linemarkers <- as.data.frame(Both_Transcriptomes[(Both_Transcriptomes$Inverser)< -100,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(TIA1Effect$t, (TIA1Effect$TIAKO), method = "spearman")



(SupFig10 <- ggplot(TIA1Effect, aes(y = t, x = (TIAKO))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2) +
  theme_bw() +
  ylab("DGE WC with SGs - WC without SGs (t-stat)") + 
  xlab("DGE TIA1 KD - TIA1 WT (t-stat)") + # Subscript "10"
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())  + 
  theme(text=element_text(size=24,  family="Arial MS"))  +
  theme(plot.title = element_text(hjust = 0.5))  + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=7, label.y.npc = "top"))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/SupFigure10.png", width = 8, height=8, units = "in", res=300)
SupFig10
dev.off()

```

### D

```{r fig.height=8, fig.width=8}
SG <- fread("../Projects/StressGranules/SGPAPER/Protein Stuff/AllwSG.txt", header = T)[,c(1,3)]
colnames(SG) <- c("EVENT", "dPSI_SG")
TIA1KO <- fread("../Projects/StressGranules/AS.WC_Transcriptome/shRNAExp/TIA1/Vulcan_all.txt", header = T)[,c(1,3)]
colnames(TIA1KO) <- c("EVENT", "dPSI_KO")


TIA1Effect <- merge(SG, TIA1KO, by="EVENT")

smooth.SG <- lm(dPSI_SG~dPSI_KO, TIA1Effect)
#Both_Transcriptomes$Inverser <- Both_Transcriptomes$WC * Both_Transcriptomes$Tot
#row.names(Both_Transcriptomes) <- Both_Transcriptomes$Gene
#linemarkers <- as.data.frame(Both_Transcriptomes[(Both_Transcriptomes$Inverser)< -100,])

# Calculate the Spearman correlation coefficient
spearman_corr <- cor(TIA1Effect$dPSI_SG, (TIA1Effect$dPSI_KO), method = "spearman")



(SupFig11 <- ggplot(TIA1Effect, aes(y = dPSI_SG, x = (dPSI_KO))) + 
  geom_point(fill = "#DDDDDD", alpha = 0.3, size=5, shape=21, colour='black') +   
  geom_density2d(color="#800020", size=1.2) +
  theme_bw() +
  ylab("DSA TIA1 WT- TIA1 KD (dPSI)") + 
  xlab("DSA WC with vs Without SG (dPSI)") + # Subscript "10"
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())  + 
  theme(text=element_text(size=24,  family="Arial MS"))  +
  theme(plot.title = element_text(hjust = 0.5))  + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  geom_abline(slope=smooth.SG$coefficients[2], intercept = smooth.SG$coefficients[1], linetype = "dashed")+ stat_cor(method = "spearman", cor.coef.name = "rho", size=7, label.y.npc = "top"))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/SupFigure11.png", width = 8, height=8, units = "in", res=300)
SupFig11
dev.off()

```


## Figure 11

### A

```{r fig.height=6, fig.width=24}
wSG <- fread("~/Projects/StressGranules/SGPAPER/Protein Stuff/dsRNAExp.txt", data.table = F)
row.names(wSG) <- wSG$V1
wSG$V1 <- NULL
ndegs <- 20

upmarkers <- as.data.frame(wSG[row.names(wSG) %in% c("DDX3X", "KHSRP", "PCBP1", "QKI", "RBM15", "TIA1", "YBX3"),] )

withSG <- ggplot(data = wSG, aes(x = logFC, y = B)) + 
  xlim(-2,2) +
  geom_point(color = "#CCCCCC") + 
  geom_vline(xintercept = -0.5, linetype = "dashed") + geom_hline(yintercept = 0, linetype="dashed")+
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  theme_bw() +
  ylab("B-Statistic") +
  xlab(expression(Log[2]~"Fold-Change")) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),   # remove major gridlines
    panel.grid.minor = element_blank(),   # remove minor gridlines
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  ggtitle("Stressed Cells vs Unstressed Cells", "SG Competent") +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  geom_label_repel(
    data = upmarkers[1:ndegs,], 
    mapping = aes(label = row.names(upmarkers)[1:ndegs]), 
    size = 4,
    point.padding = 0.5,
    color = "#A10702",
    max.overlaps = 20, fontface = "italic",
    segment.color = "#A10702"
  ) +
  geom_point(data = upmarkers, aes(x = logFC, y = B), color = "#A10702", size = 4) +
  theme(text = element_text(size = 25, family = "Arial MS"))


woSG <- fread("~/Projects/StressGranules/SGPAPER/Protein Stuff/dsRNAExp_wo.txt", data.table = F)
row.names(woSG) <- woSG$V1
woSG$V1 <- NULL


upmarkers <- as.data.frame(woSG[row.names(woSG) %in% c("DDX3X", "KHSRP", "PCBP1", "QKI", "RBM15", "TIA1", "YBX3"),] )

withoutsg <- ggplot(data = woSG, aes(x = logFC, y = B)) + 
  xlim(-2,2) +
  geom_point(color = "#CCCCCC") + 
  geom_vline(xintercept = -0.5, linetype = "dashed") +geom_hline(yintercept = 0, linetype="dashed")+
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  theme_bw() +
  ylab("B-Statistic") +
  xlab(expression(Log[2]~"Fold-Change")) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),   # remove major gridlines
    panel.grid.minor = element_blank(),   # remove minor gridlines
    panel.border = element_blank(),
    panel.background = element_blank()
  ) +
  ggtitle("Stressed Cells vs Unstressed Cells", "SG Incompetent") +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  geom_label_repel(
    data = upmarkers[1:ndegs,], 
    mapping = aes(label = row.names(upmarkers)[1:ndegs]), 
    size = 4,
    point.padding = 0.5,
    color = "#A10702",
    max.overlaps = 20, fontface = "italic",
    segment.color = "#A10702"
  ) +
  geom_point(data = upmarkers, aes(x = logFC, y = B), color = "#A10702", size = 4) +
  theme(text = element_text(size = 25, family = "Arial MS"))



(FigureN1 <- ggarrange(withSG,withoutsg, ncol = 2))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Figure9_row1.png", width = 24, height=6, units = "in", res=300)
FigureN1
dev.off()


```

# SIGNATURE IMAGES

A great deal of this code is on the TCGA_SG_Signature files.

```{r}
#This part was performed on the "Signature*" RMD files.
Discovery <- fread("~/Projects/StressGranules/SGPAPER/DiscoveryDataset.txt")
Discovery <- as.data.frame(Discovery)
colnames(Discovery)[1] <- "Gene"
row.names(Discovery) <- Discovery$Gene
Validation <- fread("~/Projects/StressGranules/SGPAPER/ValidationDataset.txt")
Validation <- as.data.frame(Validation)
colnames(Validation)[1] <- "Gene"
row.names(Validation) <- Validation$Gene
SG1 <- fread("~/SG_SIGNATURE_LASSOISH.txt")
SG2 <- fread("~/SG_SIGNATURE_ELASTICNET.txt")


# FOR CONTROL PURPOSES
TNF <- hallmarks.gsets$HALLMARK_TNFA_SIGNALING_VIA_NFKB
write.table(TNF, "~/Projects/TNFSignallingGeneSet.txt", row.names = F, col.names = F)
EMT <- hallmarks.gsets$HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION
write.table(EMT, "~/Projects/EMTSignallingGeneSet.txt", row.names = F, col.names = F)


```

# Figure 14

## A

```{r fig.height=6, fig.width=16}
Discovery <- fread("~/Projects/StressGranules/SGPAPER/DiscoveryDataset_Scores.txt")[,-1]

library(ggplot2)
library(effsize)

# Subset scores by SG presence
with_sg  <- Discovery$score[Discovery$SGPres == "With SG"]
without_sg <- Discovery$score[Discovery$SGPres == "Without SG"]

# Cohen's d
d_val <- cohen.d(with_sg, without_sg)$estimate

# P-value (Welch t-test)
p_val <- t.test(with_sg, without_sg)$p.value

# Format nicely
subtitle_text <- paste0("Cohen's d = ", round(d_val, 2),"\np = ", signif(p_val, 3))

# Plot with subtitle
disc <- ggplot(Discovery, aes(x = SGPres, y = score)) +
  geom_violin(trim = FALSE, alpha = 0.6, color = "black") +
  geom_jitter(aes(color = StressType),
              width = 0.2, size = 5, alpha = 0.6) +
  theme_bw() +
  ylab("Signature Genes' Ranking\n") +
  xlab("Stress Granule Presence") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    text = element_text(size = 25, family = "Arial MS"),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, face="italic")
  ) +
  scale_color_manual(values = c("Arsenite" = "#7390BF", 
                                "Sorbitol" = "#40C997", 
                                "Unstressed" = "#FF8D85")) +
  labs(color = "Stress Type",
       title = "Discovery Dataset",
       subtitle = subtitle_text)

Validation <- fread("~/Projects/StressGranules/SGPAPER/ValidationDataset_Scores.txt")[,-1]


# Subset scores by SG presence
with_sg  <- Validation$score[Validation$SGPres == "With SG"]
without_sg <- Validation$score[Validation$SGPres == "Without SG"]

# Cohen's d
d_val <- cohen.d(with_sg, without_sg)$estimate

# P-value (Welch t-test)
p_val <- t.test(with_sg, without_sg)$p.value

# Format nicely
subtitle_text <- paste0("Cohen's d = ", round(d_val, 2),"\np = ", signif(p_val, 3))

# Plot with subtitle
val <- ggplot(Validation, aes(x = SGPres, y = score)) +
  geom_violin(trim = FALSE, alpha = 0.6, color = "black") +
  geom_jitter(aes(color = StressType),
              width = 0.2, size = 5, alpha = 0.6) +
  theme_bw() +
  ylab("Signature Genes' Ranking\n") +
  xlab("Stress Granule Presence") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    text = element_text(size = 25, family = "Arial MS"),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, face="italic")
  ) +
  scale_color_manual(values = c("dsRNA" = "#C2B90A", 
                                "Unstressed" = "#FF8D85")) +
  labs(color = "Stress Type",
       title = "Validation Dataset",
       subtitle = subtitle_text)


fig12a <- ggarrange(disc, val, nrow = 1)

png("~/Projects/StressGranules/SGPAPER/Images/PNG/fig12a.png", width = 16, height=6, units = "in", res=300)
fig12a
dev.off()
```

# Supplementary Figure 10

```{r fig.height=30, fig.width=16}
plot_list <- readRDS("~/Projects/StressGranules/SGPAPER/Signature_TCGA_Scores.RDS")
library(ggplot2)
library(ggpubr)
library(effsize)

plots_list <- list()  # to store all plots

for (tissue in names(plot_list)) {
  df <- plot_list[[tissue]][["SG1"]]
  
  # Ensure factor levels
  df$tumour <- factor(df$tumour,
                      levels = c("Tumor", "Normal"),
                      labels = c("Tumor", "Normal"))
  
  # Subset scores by SG presence
  with_sg  <- df$score[df$tumour == "Tumor"]
  without_sg <- df$score[df$tumour == "Normal"]
  
  # Cohen's d
  d_val <- cohen.d(with_sg, without_sg)$estimate
  
  # P-value (Welch t-test)
  p_val <- t.test(with_sg, without_sg)$p.value
  
  # Format subtitle
  subtitle_text <- paste0("Cohen's d = ", round(d_val, 2), "\n p = ", signif(p_val, 3))
  
  # Create plot
  p <- ggplot(df, aes(x = tumour, y = score)) +
    geom_jitter(aes(color = tumour), width = 0.2, size = 5, alpha = 0.6) +
    theme_bw()+
    geom_violin(trim = FALSE, alpha = 0.2, color = "black")  +
    ylab("Signature Genes' Ranking\n") +
    xlab("Sample Type") +
    theme(
      axis.line = element_line(colour = "black"),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      text = element_text(size = 25, family = "Arial MS"),
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, face = "italic"),
      legend.position = "none"
    ) +
    scale_color_manual(values = c("Tumor" = "#779ECB", 
                                  "Normal" = "#FFA071")) +
    labs(title = tissue, subtitle = subtitle_text)
  
  # Store in list
  plots_list[[tissue]] <- p
}

TCGASCORES <- ggarrange(plotlist = plots_list, ncol = 3,
          nrow = ceiling(length(plots_list)/3))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/sup10.png", width = 20, height=30, units = "in", res=300)
TCGASCORES
dev.off()

```
# Supplementary Figure 11

```{r fig.height=60, fig.width=20}
all_survival_plots <- readRDS("~/Projects/StressGranules/SGPAPER/Simulations/SurvivalResults_TNFEMT.RDS")

# ✅ Combine all survival plots into a panel
if (length(all_survival_plots) > 0) {
  final_survival_plot <- ggarrange(
    plotlist = all_survival_plots,
    ncol = 3, nrow = ceiling(length(all_survival_plots) / 3)
  )
  print(final_survival_plot)
} else {
  print("No valid survival plots generated.")
}

# png("~/Projects/StressGranules/SGPAPER/Images/PNG/sup11.png", width = 20, height=30, units = "in", res=300)
# final_survival_plot
# dev.off()
```

# Potential Prognosis Analsysis

```{r}
TCGAobj_tissue <- readRDS("~/Projects/TCGA/TCGAobj_tissue.RDS")

for (tissue in names(TCGAobj_tissue)) {
  
  # Get metadata vector
  meta_vec <- TCGAobj_tissue[[tissue]]$metadata
  
  # Identify indices where metadata is 01 or 06
  keep_idx <- which(meta_vec %in% c("01", "06"))
  
  # Filter metadata
  TCGAobj_tissue[[tissue]]$metadata <- meta_vec[keep_idx]
  
  # Filter logCPM (samples are columns)
  TCGAobj_tissue[[tissue]]$logCPM <- TCGAobj_tissue[[tissue]]$logCPM[, keep_idx, drop = FALSE]
  
  # --- Handle SGScore orientation ---
  SGScore <- TCGAobj_tissue[[tissue]]$SGScore
  if (ncol(SGScore) == length(meta_vec)) {
    SGScore <- SGScore[, keep_idx, drop = FALSE]
    scores <- as.numeric(SGScore[1, ])  # assume score is 1 row, samples are columns
  } else if (nrow(SGScore) == length(meta_vec)) {
    SGScore <- SGScore[keep_idx, , drop = FALSE]
    scores <- as.numeric(SGScore[, 1])  # assume score is 1 column, samples are rows
  } else {
    warning("SGScore dimensions don’t match metadata for tissue: ", tissue)
    next
  }
  
  # --- Compute top/bottom 25% ---
  q25 <- quantile(scores, 0.25, na.rm = TRUE)
  q75 <- quantile(scores, 0.75, na.rm = TRUE)
  
  # Keep only samples in bottom 25% or top 25%
  keep_SG <- which(scores <= q25 | scores >= q75)
  
  # Update SGScore to only those samples
  if (ncol(SGScore) >= length(scores)) {
    SGScore <- SGScore[, keep_SG, drop = FALSE]
  } else {
    SGScore <- SGScore[keep_SG, , drop = FALSE]
  }
  
  # Update metadata & logCPM accordingly
  TCGAobj_tissue[[tissue]]$metadata <- TCGAobj_tissue[[tissue]]$metadata[keep_SG]
  TCGAobj_tissue[[tissue]]$logCPM   <- TCGAobj_tissue[[tissue]]$logCPM[, keep_SG, drop = FALSE]
  TCGAobj_tissue[[tissue]]$SGScore  <- SGScore
  
  # Save SGStatus (Low / High labels)
  SGStatus <- ifelse(scores[keep_SG] <= q25, "Low", "High")
  TCGAobj_tissue[[tissue]]$SGStatus <- SGStatus
}


#First Test, checking what the 4 cancers have in common
keep_tissues <- c("ACC", "MESO", "CHOL", "STAD")
TCGAobj_tissue <- TCGAobj_tissue[names(TCGAobj_tissue) %in% keep_tissues]

for (tissue in names(TCGAobj_tissue)) {
  TCGAobj_tissue[[tissue]]$OGTissue <- rep(tissue, ncol(TCGAobj_tissue[[tissue]]$logCPM))
}

# Step 1. Extract all logCPM matrices and merge by rownames
logCPM_list <- lapply(TCGAobj_tissue, function(x) x$logCPM)

# Make sure all have same genes in the same order
all_genes <- Reduce(intersect, lapply(logCPM_list, rownames))
logCPM_list <- lapply(logCPM_list, function(df) df[all_genes, , drop = FALSE])

# Merge by column-binding
merged_logCPM <- do.call(cbind, logCPM_list)

# Step 2. Extract metadata and OGTissue for each tissue
metadata_vec <- unlist(lapply(TCGAobj_tissue, function(x) x$metadata))
ogtissue_vec <- unlist(lapply(TCGAobj_tissue, function(x) x$OGTissue))
SGScore_vec <- unlist(lapply(TCGAobj_tissue, function(x) x$SGStatus))
# Step 3. Add them as rows to the matrix
# Make sure lengths match number of columns in merged matrix
stopifnot(length(metadata_vec) == ncol(merged_logCPM))
stopifnot(length(ogtissue_vec) == ncol(merged_logCPM))

merged_logCPM_with_meta <- rbind(
  merged_logCPM,
  metadata = metadata_vec,
  OGTissue = ogtissue_vec,
  SGScore = SGScore_vec
)


SampleType <- merged_logCPM_with_meta["metadata",]
OriginTissue <- merged_logCPM_with_meta["OGTissue",]
SGSCore <- merged_logCPM_with_meta["SGScore",]

merged_logCPM_with_meta <- merged_logCPM_with_meta[-((nrow(merged_logCPM_with_meta)-2):nrow(merged_logCPM_with_meta)), ]

# Save rownames
rn <- rownames(merged_logCPM_with_meta)

# Convert all values to numeric
merged_logCPM_with_meta <- as.data.frame(lapply(merged_logCPM_with_meta, as.numeric))

# Restore rownames
rownames(merged_logCPM_with_meta) <- rn


```

## Batch Correction for Tissue Origin

```{r}
metadata <- data.frame(Tissue=t(OriginTissue), SG=t(SGSCore))
colnames(metadata) <- c("Tissue", "SGScore")
batch_cols <- c("Tissue")
metadata_cols <- c("SGScore")
mmbatch <- model.matrix(reformulate(paste0("~0 + ",paste0( batch_cols,collapse = " + "))), data=metadata)
mmkeep <- model.matrix(reformulate(paste0("~0 + ",paste0( metadata_cols,collapse =  "+" ))), data=metadata)
mm <- cbind(mmkeep,mmbatch)
D0 <- DGEList(2^merged_logCPM_with_meta) 
D0 <- calcNormFactors(D0)
y <- voom(D0, mm, plot = F)
fit <- lmFit(y, mm)
beta <- fit$coefficients[,-(1:ncol(mmkeep)),drop=FALSE]
#beta <- fit$coefficients
beta[is.na(beta)] <- 0
corrcounts_v1 <- as.matrix(y$E) - beta %*% t(mmbatch)
offset <- apply(corrcounts_v1,1,min) - apply(y$E,1,min)
corrcounts_v1 <- as.data.frame(corrcounts_v1-offset)
corrcounts_v1 <- 2^corrcounts_v1

mm <- model.matrix(~0+SGScore, metadata)

y <- voom(corrcounts_v1, mm, plot = T)
```

## Linear Models

Finally, we move on to differential gene expression! In this chunk, I define the linear model I will actually use for my analyses. 

```{r}
#Now I simply create the linear model that isolates the SG signal
fitted <- lmFit(y$E, mm)

head(coef(fitted))

contr <- makeContrasts(SGScoreHigh  - SGScoreLow  , levels = colnames(coef(fitted)))

tmp_contr <- contrasts.fit(fitted, contr)

tmp <- eBayes(tmp_contr)
top.table <- topTable(tmp, sort.by = "logFC", n=Inf)
DEGenes <- top.table
ndegs=20
```


## Figure 14

### E

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>1 & DEGenes$B>25,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -1.5 & DEGenes$B>30,] )


(VC <- ggplot(data=DEGenes,
             aes(x=logFC, y=B)) +
  geom_point(color="#CCCCCC") +
  ylab("B-Statistic") +
  xlab(expression(Log[2]~"Fold-Change")) +
  ggtitle("Tumor vs Normal") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),   # remove major gridlines
    panel.grid.minor = element_blank(),   # remove minor gridlines
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    text = element_text(size=25, family="Arial MS")
  ) +
  geom_point(data=upmarkers, aes(x=logFC, y=B),
             color="#A10702", size=4) +
  geom_point(data=downmarkers, aes(x=logFC, y=B),
             color="#03256C", size=4) +
  geom_label_repel(upmarkers[1:ndegs,],
                   mapping=aes(label=row.names(upmarkers)[1:ndegs]),
                   size=4, point.padding=-0.5, color="#A10702",
                   max.overlaps = 20, fontface="italic") +
  geom_label_repel(downmarkers[1:ndegs,],
                   mapping=aes(label=row.names(downmarkers)[1:ndegs]),
                   size=4, point.padding=-0.5, color="#03256C",
                   max.overlaps = 20, fontface="italic") )



png("~/Projects/StressGranules/SGPAPER/Images/PNG/TvsN_SigCancer.png", width = 8, height=7, units = "in", res=300)
VC
dev.off()

```

### F

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=10, fig.width=14, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,3]),]
gseatool <- DEGenes[,3]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>%
  filter(padj < thresh) %>%
  mutate(pathway = gsub("^HALLMARK_", "", pathway),
         pathway = gsub("_", " ", pathway))

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
(gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = Status), alpha = 0.75) + 
  scale_fill_manual(values = c("#53A2BE", "#BA3B46"), 
                    name = "Regulation Status", 
                    labels = c("Downregulated", "Upregulated"), 
                    breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x = "Gene Set", 
       y = "Normalized Enrichment Score",
       title = "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) +
  theme_bw() +
  theme(legend.position = "none",
        text = element_text(size = 25, family = "Arial MS", face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "bold")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/TvsN_SigCancer_GeneSets.png", width = 14, height=10, units = "in", res=300)
gsea_plot_hall
dev.off()
```

# What do these cancers have in common? (Alternate comparison)

```{r}
TCGAobj_tissue <- readRDS("~/Projects/TCGA/TCGAobj_tissue.RDS")

# Filter TCGAobj_tissue by metadata (keep only samples with 11 or 06)
for (tissue in names(TCGAobj_tissue)) {
  
  # Get metadata vector
  meta_vec <- TCGAobj_tissue[[tissue]]$metadata
  
  # Identify indices where metadata is 11 or 06
  keep_idx <- which(meta_vec %in% c("01", "06"))
  
  # Filter metadata
  TCGAobj_tissue[[tissue]]$metadata <- meta_vec[keep_idx]
  
  # Filter logCPM using the same indices (columns = samples)
  TCGAobj_tissue[[tissue]]$logCPM <- TCGAobj_tissue[[tissue]]$logCPM[, keep_idx, drop = FALSE]
}


for (tissue in names(TCGAobj_tissue)) {
  TCGAobj_tissue[[tissue]]$OGTissue <- rep(tissue, ncol(TCGAobj_tissue[[tissue]]$logCPM))
}

# Step 1. Extract all logCPM matrices and merge by rownames
logCPM_list <- lapply(TCGAobj_tissue, function(x) x$logCPM)

# Make sure all have same genes in the same order
all_genes <- Reduce(intersect, lapply(logCPM_list, rownames))
logCPM_list <- lapply(logCPM_list, function(df) df[all_genes, , drop = FALSE])

# Merge by column-binding
merged_logCPM <- do.call(cbind, logCPM_list)

# Step 2. Extract metadata and OGTissue for each tissue
metadata_vec <- unlist(lapply(TCGAobj_tissue, function(x) x$metadata))
ogtissue_vec <- unlist(lapply(TCGAobj_tissue, function(x) x$OGTissue))

# Step 3. Add them as rows to the matrix
# Make sure lengths match number of columns in merged matrix
stopifnot(length(metadata_vec) == ncol(merged_logCPM))
stopifnot(length(ogtissue_vec) == ncol(merged_logCPM))

merged_logCPM_with_meta <- rbind(
  merged_logCPM,
  metadata = metadata_vec,
  OGTissue = ogtissue_vec
)


SampleType <- merged_logCPM_with_meta["metadata",]
OriginTissue <- merged_logCPM_with_meta["OGTissue",]

merged_logCPM_with_meta <- merged_logCPM_with_meta[-((nrow(merged_logCPM_with_meta)-1):nrow(merged_logCPM_with_meta)), ]

# Save rownames
rn <- rownames(merged_logCPM_with_meta)

# Convert all values to numeric
merged_logCPM_with_meta <- as.data.frame(lapply(merged_logCPM_with_meta, as.numeric))

# Restore rownames
rownames(merged_logCPM_with_meta) <- rn


```

## Batch Correction for Tissue Origin

```{r}
metadata <- data.frame(Sample=t(SampleType), Tissue=t(OriginTissue))
colnames(metadata) <- c("Sample", "Tissue")
batch_cols <- c("Sample")
metadata_cols <- c("Tissue")
mmbatch <- model.matrix(reformulate(paste0("~0 + ",paste0( batch_cols,collapse = " + "))), data=metadata)
mmkeep <- model.matrix(reformulate(paste0("~0 + ",paste0( metadata_cols,collapse =  "+" ))), data=metadata)
mm <- cbind(mmkeep,mmbatch)
D0 <- DGEList(2^merged_logCPM_with_meta) 
D0 <- calcNormFactors(D0)
y <- voom(D0, mm, plot = F)
fit <- lmFit(y, mm)
beta <- fit$coefficients[,-(1:ncol(mmkeep)),drop=FALSE]
#beta <- fit$coefficients
beta[is.na(beta)] <- 0
corrcounts_v1 <- as.matrix(y$E) - beta %*% t(mmbatch)
offset <- apply(corrcounts_v1,1,min) - apply(y$E,1,min)
corrcounts_v1 <- as.data.frame(corrcounts_v1-offset)
corrcounts_v1 <- 2^corrcounts_v1

mm <- model.matrix(~0+Tissue, metadata)

y <- voom(corrcounts_v1, mm, plot = T)
```
## Linear Models

Finally, we move on to differential gene expression! In this chunk, I define the linear model I will actually use for my analyses. 

```{r}
#Now I simply create the linear model that isolates the SG signal
fitted <- lmFit(y$E, mm)

colnames(coef(fitted))
# Define groups
tissues_of_interest <- c("TissueACC", "TissueMESO", "TissueCHOL", "TissueSTAD")
all_tissues <- colnames(coef(fitted))
other_tissues <- setdiff(all_tissues, tissues_of_interest)

# Build numerator (average of 4 tissues of interest)
num <- paste(tissues_of_interest, collapse = " + ")
num <- paste0("(", num, ")/", length(tissues_of_interest))

# Build denominator (average of all other tissues)
den <- paste(other_tissues, collapse = " + ")
den <- paste0("(", den, ")/", length(other_tissues))

# Final contrast
contrast_formula <- paste(num, "-", den)

# Make contrast matrix
contr <- makeContrasts(contrasts = contrast_formula, levels = all_tissues)

tmp_contr <- contrasts.fit(fitted, contr)

tmp <- eBayes(tmp_contr)
top.table <- topTable(tmp, sort.by = "logFC", n=Inf)
DEGenes <- top.table
ndegs=20
```
## Figure 14

###C

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>1 & DEGenes$B>200,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -1 & DEGenes$B>150,] )


(TT <- ggplot(data=DEGenes,
             aes(x=logFC, y=B)) +
  geom_point(color="#CCCCCC") +
  ylab("B-Statistic") +
  xlab(expression(Log[2]~"Fold-Change")) +
  ggtitle("Tumor vs Tumor") +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),   # remove major gridlines
    panel.grid.minor = element_blank(),   # remove minor gridlines
    panel.border = element_blank(),
    panel.background = element_blank(),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    text = element_text(size=25, family="Arial MS")
  ) +
  geom_label_repel(upmarkers[1:ndegs,],
                   mapping=aes(label=row.names(upmarkers)[1:ndegs]),
                   size=4, point.padding=-0.5, color="#A10702",
                   max.overlaps = 20, fontface="italic") +
  geom_label_repel(downmarkers[1:ndegs,],
                   mapping=aes(label=row.names(downmarkers)[1:ndegs]),
                   size=4, point.padding=-0.5, color="#03256C",
                   max.overlaps = 20, fontface="italic") +
  geom_point(data=upmarkers, aes(x=logFC, y=B),
             color="#A10702", size=4) +
  geom_point(data=downmarkers, aes(x=logFC, y=B),
             color="#03256C", size=4))



png("~/Projects/StressGranules/SGPAPER/Images/PNG/TvsT_SigCancer.png", width = 8, height=7, units = "in", res=300)
TT
dev.off()

```
### D

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=10, fig.width=14, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,3]),]
gseatool <- DEGenes[,3]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>%
  filter(padj < thresh) %>%
  mutate(pathway = gsub("^HALLMARK_", "", pathway),
         pathway = gsub("_", " ", pathway))

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
(gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = Status), alpha = 0.75) + 
  scale_fill_manual(values = c("#53A2BE", "#BA3B46"), 
                    name = "Regulation Status", 
                    labels = c("Downregulated", "Upregulated"), 
                    breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x = "Gene Set", 
       y = "Normalized Enrichment Score",
       title = "Hallmark Gene Sets",
       caption = paste0("p-value < ", thresh)) +
  theme_bw() +
  theme(legend.position = "none",
        text = element_text(size = 25, family = "Arial MS", face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title = element_text(face = "bold"),
        axis.text = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "bold")))

png("~/Projects/StressGranules/SGPAPER/Images/PNG/TvsT_SigCancer_GeneSets.png", width = 14, height=10, units = "in", res=300)
gsea_plot_hall
dev.off()
```

# Figure 10

## A

```{r fig.height=4, fig.width=6}
# Your data
FC_pre  <- c(40.46, 32.93, 63.01)
FC_post <- c(23.89, 19.08, 37.49)
PC_pre  <- c(63.71, 74.33, 70.30)
PC_post <- c(48.94, 44.81, 44.88)

# 1) Compare PRE between FC and PC
t_pre <- t.test(FC_pre, PC_pre, var.equal = FALSE)
wil_pre <- wilcox.test(FC_pre, PC_pre)

# 2) Compare POST between FC and PC
t_post <- t.test(FC_post, PC_post, var.equal = FALSE)
wil_post <- wilcox.test(FC_post, PC_post)

# Print results
cat("PRE comparison (FC vs PC):\n")
print(t_pre)
print(wil_pre)

cat("\nPOST comparison (FC vs PC):\n")
print(t_post)
print(wil_post)

# ---- Plotting with ggplot ----
library(ggplot2)
library(dplyr)
library(ggpubr)

df_compare <- data.frame(
  Line = rep(c("Control","Sorbitol"), each = 6),
  Time = rep(rep(c("Mitochondria","SG"), each = 3), 2),
  Value = c(FC_pre, FC_post, PC_pre, PC_post)
)


Imaging1 <- ggplot(df_compare, aes(x = Line, y = Value, color = Line)) +
  geom_jitter(width = 0.1, size = 3, alpha = 0.5) +
  facet_wrap(~Time) +
  stat_compare_means(
    aes(group = Line),
    method = "t.test",
    label = "p.format",
    size = 8,       # big text (≈ 25pt)
    vjust = 1    # move p-value text closer to points
  ) +
  theme_bw() +
  labs(y = "Colocalization (%)", x = "Stress Status") +
  scale_color_manual(values = c("Control" = "#1F6F52", "Sorbitol" = "#E00F00")) +
  theme(
    strip.text = element_text(face = "bold", size = 20),  # facet titles bold
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 20),
    legend.position = "none"
  )

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Imaging1.png", width = 6, height=4, units = "in", res=300)
Imaging1
dev.off()
```

## B

```{r fig.height=4, fig.width=8}
# Data as proportions
PC_post <- c(48.94, 44.81, 44.88) / 100
FC_post <- c(23.89, 19.08, 37.49) / 100

# Function to estimate Beta params from data
beta_params <- function(x) {
  m <- mean(x)
  v <- var(x)
  alpha <- m * ((m * (1 - m) / v) - 1)
  beta <- (1 - m) * ((m * (1 - m) / v) - 1)
  return(list(alpha = alpha, beta = beta, mean = m, var = v))
}

params_PC <- beta_params(PC_post)
params_FC <- beta_params(FC_post)

params_PC
params_FC

# ---- Plotting the fitted distributions ----
library(ggplot2)

x <- seq(0, 1, length.out = 1000)
df <- data.frame(
  x = rep(x, 2),
  density = c(dbeta(x, params_PC$alpha, params_PC$beta),
              dbeta(x, params_FC$alpha, params_FC$beta)),
  StressStatus = rep(c("Sorbitol", "Control"), each = length(x))
)

Imaging2 <- ggplot(df, aes(x, density, color = StressStatus, fill = StressStatus)) +
  geom_line(size = 1.2, alpha = 0.5) +
  geom_area(alpha = 0.2, position = "identity") +
  scale_color_manual(values = c("Control" = "#1F6F52", "Sorbitol" = "#E00F00")) +
  scale_fill_manual(values = c("Control" = "#1F6F52", "Sorbitol" = "#E00F00")) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 20, family = "Arial MS"),
    axis.line = element_line(color = "black"),
    panel.border = element_blank()
  ) +
  labs(x = "Colocalization", y = "Density", color = "Stress Status", fill = "Stress Status")

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Imaging2.png", width = 8, height=4, units = "in", res=300)
Imaging2
dev.off()
```
```{r}
library(ggplot2)
library(effsize)

set.seed(1906)

# Simulate 1000 draws
sim_PC <- rbeta(1000, params_PC$alpha, params_PC$beta)
sim_FC <- rbeta(1000, params_FC$alpha, params_FC$beta)

# Compute Cohen's d
d_val <- cohen.d(sim_PC, sim_FC)$estimate

# Welch t-test p-value
p_val <- t.test(sim_PC, sim_FC)$p.value

# Format subtitle with scientific notation for p
subtitle_text <- bquote(italic("Cohen's d") == .(round(d_val, 2)) ~ "," ~ italic("p") == .(formatC(p_val, format = "e", digits = 2)))

# Combine into df for ggplot
df_sim <- data.frame(
  Proportion = c(sim_PC, sim_FC),
  Group = rep(c("Stressed", "Unstressed"), each = 1000)
)
# If p is extremely small, show "≈ 0"
p_label <- if (p_val < 1e-10) {
  "≈ 0"
} else {
  formatC(p_val, format = "e", digits = 2)
}

subtitle_text <- bquote(italic("Cohen's d") == .(round(d_val, 2)) ~ "," ~ italic("p") == .(p_label))
# Plot
ggplot(df_sim, aes(x = Group, y = Proportion, fill = Group)) +
  geom_violin(trim = FALSE, alpha = 0.6, color = "black") +
  geom_jitter(width = 0.1, alpha = 0.3, size = 1) +
  scale_fill_manual(values = c("Stressed" = "#7390BF", "Unstressed" = "#FF8D85")) +
  theme_bw() +
  theme(
    text = element_text(size = 20, family = "Arial MS"),
    axis.line = element_line(colour = "black"),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  ylab("dsRNA colocalised within SGs") +
  xlab("Cell State") +
  ggtitle(subtitle_text)


```


## C

```{r fig.height=6, fig.width=8}
library(ggplot2)
library(effsize)

set.seed(1906)

# --- Observed data
PC_post <- c(48.94, 44.81, 44.88)/100
FC_post <- c(23.89, 19.08, 37.49)/100

# --- Simulated draws
sim_PC <- rbeta(1000, params_PC$alpha, params_PC$beta)
sim_FC <- rbeta(1000, params_FC$alpha, params_FC$beta)

# --- Statistics
p_diff <- abs(mean(sim_PC) - mean(sim_FC))

BC <- sum(sqrt(density(sim_PC, from=0, to=1, n=512)$y *
               density(sim_FC, from=0, to=1, n=512)$y)) / 512

AUC <- mean(outer(sim_PC, sim_FC, ">"))

d_val <- cohen.d(sim_PC, sim_FC)$estimate

# Caption text
caption_text <- paste0(
  "p-diff = ", round(p_diff, 3),
  " | BC = ", round(BC, 3),
  " | AUC = ", round(AUC, 3),
  " | Cohen's d = ", round(d_val, 3)
)

# --- Dataframes
df_sim <- data.frame(
  Proportion = c(sim_PC, sim_FC),
  Group = rep(c("Sorbitol", "Control"), each = 1000)
)

df_obs <- data.frame(
  Proportion = c(PC_post, FC_post),
  Group = rep(c("Sorbitol", "Control"), each = 3)
)

# --- Plot
Imaging3 <- ggplot(df_sim, aes(x = Group, y = Proportion, fill = Group)) +
  geom_violin(trim = FALSE, alpha = 0.4, color = "black") +
  geom_point(
    data = df_obs, aes(x = Group, y = Proportion),
    width = 0.15, shape = 21, fill = "white", size = 4, stroke = 1.2
  ) +
  scale_fill_manual(values = c("Control" = "#1F6F52", "Sorbitol" = "#E00F00")) +
  theme_bw() +
  theme(
    text = element_text(size = 20, family = "Arial MS"),
    axis.line = element_line(colour = "black"),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  ylab("dsRNA colocalised within SGs") +
  xlab("Stress State") +
  labs(caption = caption_text)

png("~/Projects/StressGranules/SGPAPER/Images/PNG/Imaging3.png", width = 8, height=6, units = "in", res=300)
Imaging3
dev.off()
```

