---
title: "Analyses of GSE173953"
subtitle: "Stress granules are shock absorbers that prevent excessive innate immune responses to dsRNA"
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

In this document, I present all analyses pertaining to the dataset **GSE173953**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE173953)),  
originally published by **M. Paget, 2023**.

## Methodology

```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)
library(readxl)

set.seed(1906)
```

```{r}
raw.read.counts <- read.table("~/Projects/StressGranules/AS.WC_Transcriptome/GSE173953/TrimmedSamples/alignment_STAR/featureCounts_Results.txt", header = T)


# Afterwards, I discard the columns which don't have read counts. These columns are only informations about the gene.
raw.read.counts <- raw.read.counts[,-c(2:6)]

#Loading up our biomart file, obtained from Ensembl, where we have the equivalence between Ensembl IDs and HGNC names
biomart <- fread("~/biomart.txt")[,-1]


colnames(raw.read.counts)[1] <- "ensembl_gene_id"
colnames(biomart) <- c("ensembl_gene_id", "hgnc_symbol")

eltable <- merge(raw.read.counts,biomart,by = "ensembl_gene_id", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="") #this removes reads that dont have a corresponding hgnc symbol.

eltable$ensembl_gene_id <- NULL # We no longer need this column

read.counts <- eltable %>% # If two gene ids resulted in the same gene name, this will sum up their counts
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

read.counts <- as.data.frame(read.counts)

row.names(read.counts) <- read.counts$hgnc_symbol #changes rownames into hgnc symbols
read.counts$hgnc_symbol <- NULL #We no longer need this column, as we have the rownames.
```

## Extra Quality Control

```{r}

# Define color palette
custom_palette <- qualitative_hcl(11, palette = "Pastel")
custom_palette <- sample(custom_palette, 11)

# Use SampleColouring from metadata
SampleColouring <- c(rep("WT.U2.CT",2), rep("WT.U2.DS6",2), rep("KO.U2.CT",2),rep("KO.U2.DS4",2), rep("WT.A5.CT",2), rep("WT.A5.DS4",2), rep("KO.A5.CT",2), rep("KO.A5.DS4",2), rep("KO.U2.DS6",3), rep("KOKO.U2.DS6",3), rep("KO.U2.DS24",3))

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(6, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )
```

As we can see in these plots, even though there is a sample type with lower complexity, in this case there are no individual sample outliers in terms of complexity or read depth. Excellent!

## Filtration

Let us begin with the filtration. Personally, I find that its easier to do visually. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bi-modal distributions, with the first mode corresponding to lowly expressed genes. I remove these by cutting just before the second mode. This time, I opted to also observe every sample type by itself.

```{r}

#I found that the MAVS and G3BP1&2 KOs were not important for my own analyses so I don-t use them.
read.counts <- read.counts[,-c(20:22)]


alternew.WT.U2.CT <- log10(rowSums(read.counts[,1:2]))
alternew.WT.U2.CT <- as.data.frame(alternew.WT.U2.CT)

alternew.WT.U2.DS6 <- log10(rowSums(read.counts[,3:4]))
alternew.WT.U2.DS6 <- as.data.frame(alternew.WT.U2.DS6)

alternew.KO.U2.CT <- log10(rowSums(read.counts[,5:6]))
alternew.KO.U2.CT <- as.data.frame(alternew.KO.U2.CT)

alternew.KO.U2.DS4 <- log10(rowSums(read.counts[,7:8]))
alternew.KO.U2.DS4 <- as.data.frame(alternew.KO.U2.DS4)

alternew.WT.A5.CT <- log10(rowSums(read.counts[,9:10]))
alternew.WT.A5.CT <- as.data.frame(alternew.WT.A5.CT)

alternew.WT.A5.DS4 <- log10(rowSums(read.counts[,11:12]))
alternew.WT.A5.DS4 <- as.data.frame(alternew.WT.A5.DS4)

alternew.KO.A5.CT <- log10(rowSums(read.counts[,13:14]))
alternew.KO.A5.CT <- as.data.frame(alternew.KO.A5.CT)

alternew.KO.A5.DS4 <- log10(rowSums(read.counts[,15:16]))
alternew.KO.A5.DS4 <- as.data.frame(alternew.KO.A5.DS4)

alternew.KO.U2.DS6 <- log10(rowSums(read.counts[,17:19]))
alternew.KO.U2.DS6 <- as.data.frame(alternew.KO.U2.DS6)

alternew.KO.U2.DS24 <- log10(rowSums(read.counts[,20:22]))
alternew.KO.U2.DS24 <- as.data.frame(alternew.KO.U2.DS24)


alternew <- log10(rowSums(read.counts))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 3.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.WT.U2.CT, aes(x=alternew.WT.U2.CT)) + geom_density(fill="#a2d1e6", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("WT U2OS Control")

ggplot(alternew.WT.U2.DS6, aes(x=alternew.WT.U2.DS6)) + geom_density(fill="#769fca", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("WT U2OS dsRNA 6h")

ggplot(alternew.KO.U2.CT, aes(x=alternew.KO.U2.CT)) + geom_density(fill="#e88184", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("KO-G3BP U2OS Control")

ggplot(alternew.KO.U2.DS4, aes(x=alternew.KO.U2.DS4)) + geom_density(fill="#de425b", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("KO-G3BP U2OS dsRNA 4h")

ggplot(alternew.WT.A5.CT, aes(x=alternew.WT.A5.CT)) + geom_density(fill="#4e9c81", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("WT A549 Control")

ggplot(alternew.WT.A5.DS4, aes(x=alternew.WT.A5.DS4)) + geom_density(fill="#207567", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("WT A549 dsRNA 4h")

ggplot(alternew.KO.A5.CT, aes(x=alternew.KO.A5.CT)) + geom_density(fill="#878787", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("KO-G3BP A549 Control")

ggplot(alternew.KO.A5.DS4, aes(x=alternew.KO.A5.DS4)) + geom_density(fill="#555555", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("KO-G3BP A549 dsRNA 4h")

ggplot(alternew.KO.U2.DS6, aes(x=alternew.KO.U2.DS6)) + geom_density(fill="yellow", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("KO-G3BP U2OS dsRNA 6h")

ggplot(alternew.KO.U2.DS24, aes(x=alternew.KO.U2.DS24)) + geom_density(fill="orange", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("KO-G3BP U2OS dsRNA 24h")

```
I didn't do it this time, but I've maintained the code to filter by sample type. This time, I considered that it was more correct to filter all samples together.

```{r}
filterednew <- subset(read.counts, alternew>4)
filterednew.WT.U2.CT <- subset(read.counts, alternew.WT.U2.CT>2.75)
filterednew.WT.U2.DS6 <- subset(read.counts, alternew.WT.U2.DS6>2.75)
filterednew.KO.U2.CT <- subset(read.counts, alternew.KO.U2.CT>2.75)
filterednew.KO.U2.DS4 <- subset(read.counts, alternew.KO.U2.DS4>2.5)
filterednew.WT.A5.CT <- subset(read.counts, alternew.WT.A5.CT>2.75)
filterednew.WT.A5.DS4 <- subset(read.counts, alternew.WT.A5.DS4>2.75)
filterednew.KO.A5.CT <- subset(read.counts, alternew.KO.A5.CT>2.5)
filterednew.KO.A5.DS4 <- subset(read.counts, alternew.KO.A5.DS4>2.75)
filterednew.KO.U2.DS6 <- subset(read.counts, alternew.KO.U2.DS6>2.75)
filterednew.KO.U2.DS24 <- subset(read.counts, alternew.KO.U2.DS24>2.75)

#after cutting, its a good idea to visualise again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))
impgenes.WT.U2.CT <- c(row.names(filterednew.WT.U2.CT))
impgenes.WT.U2.DS6 <- c(row.names(filterednew.WT.U2.DS6))
impgenes.KO.U2.CT <- c(row.names(filterednew.KO.U2.CT))
impgenes.KO.U2.DS4 <- c(row.names(filterednew.KO.U2.DS4))
impgenes.WT.A5.CT <- c(row.names(filterednew.WT.A5.CT))
impgenes.WT.A5.DS4 <- c(row.names(filterednew.WT.A5.DS4))
impgenes.KO.A5.CT <- c(row.names(filterednew.KO.A5.CT))
impgenes.KO.A5.DS4 <- c(row.names(filterednew.KO.A5.DS4))
impgenes.KO.U2.DS6 <- c(row.names(filterednew.KO.U2.DS6))
impgenes.KO.U2.DS24 <- c(row.names(filterednew.KO.U2.DS24))

finaltable <- read.counts[row.names(read.counts)%in%impgenes,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WT.U2.CT,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WT.U2.DS6,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.KO.U2.CT,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.KO.U2.DS4,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WT.A5.CT,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.WT.A5.DS4,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.KO.A5.CT,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.KO.A5.DS4,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.KO.U2.DS6,]
# finaltable <- finaltable[row.names(finaltable)%in%impgenes.KO.U2.DS24,]
```



## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable) 

#Then we calculate the normalisation factors for these new generated data frames
#Bear in mind this step calculates the normalisation factors, but does not normalise itself.
D0 <- calcNormFactors(D0)


#Lets create a dataframe with all metadata
#Just in case I end up needing it, this is the full name of every sample
samplenaming <- c(rep("WT.U2.CT",2), rep("WT.U2.DS6",2), rep("KO.U2.CT",2),rep("KO.U2.DS4",2), rep("WT.A5.CT",2), rep("WT.A5.DS4",2), rep("KO.A5.CT",2), rep("KO.A5.DS4",2), rep("KO.U2.DS6",3), rep("KOKO.U2.DS6",3), rep("KO.U2.DS24",3))[-c(20:22)]

#In this version of stress, I use the time of stress as a variable
StressType <- c(rep("0",2), rep("6",2), rep("0",2),rep("4",2), rep("0",2), rep("4",2), rep("0",2), rep("4",2), rep("6",3), rep("6",3), rep("24",3))

#In this version, I only consider if its stressed or not
StressType_simple <- c(rep("NO",2), rep("YES",2), rep("NO",2),rep("YES",2), rep("NO",2), rep("YES",2), rep("NO",2), rep("YES",11))

CellType <- c(rep("U2OS",8), rep("A549",8), rep("U2OS",9))

KOStat <- c(rep("WT",4), rep("KO",4), rep("WT",4), rep("KO",7), rep("KOKO",3), rep("KO",3))

SGPres <- c(rep("NO",2), rep("YES",2), rep("NO",6), rep("YES",2), rep("NO",13))

StudyNum <- c(rep("Study1",8), rep("Study2",8), rep("Study3",9)) #I separate by study

metadata <- data.frame(Stress=StressType, Cell=CellType, KO=KOStat, SG=SGPres, Study=StudyNum) #Once again, I remove the KOKO samples

metadata <- metadata[-c(20:22),]

mm <- model.matrix(~0+Cell+Stress+KO+SG+Study,metadata)


#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```

The Mean-Variance Trend graph from voom also allows to determine if the filtration was appropriate. In this case, it was.


## Normalisation

Now, we are almost ready to move on to the differential gene expression analysis. I will first check to see if any issues arose during the normalisation process, by looking at samples before and after this process.

```{r}
dfforboxy <- finaltable
orderforplots <- samplenaming
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
# Convert 'Types' to factor with custom order
dfforboxy$Types <- factor(dfforboxy$Types, levels = unique(samplenaming))

prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type",values = c("#FFB6C1", "#FFB6C1", "#FFDAB9", "#FFDAB9", "#E6E6FA", "#E6E6FA", "#98FB98", "#98FB98", "#87CEEB", "#87CEEB", "#FFFF99", "#FFFF99", "#F08080", "#F08080", "#98FB98", "#98FB98", "#D8BFD8", "#D8BFD8", "#D8BFD8", "#FFA07A", "#FFA07A", "#FFA07A","#6495ED","#6495ED","#6495ED"), labels=c(unique(orderforplots))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

#Using the Standard Matrix

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
# Convert 'Types' to factor with custom order
dfforboxy$Types <- factor(dfforboxy$Types, levels = unique(samplenaming))
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c("#FFB6C1", "#FFB6C1", "#FFDAB9", "#FFDAB9", "#E6E6FA", "#E6E6FA", "#98FB98", "#98FB98", "#87CEEB", "#87CEEB", "#FFFF99", "#FFFF99", "#F08080", "#F08080", "#98FB98", "#98FB98", "#D8BFD8", "#D8BFD8", "#D8BFD8", "#FFA07A", "#FFA07A", "#FFA07A","#6495ED","#6495ED","#6495ED"), labels=c(unique(orderforplots))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```

In this case, from the boxplots, we can see that there no samples which can be considered outliers.

## PCA

I like to start with a Principal Component Analysis (PCA), which helps to identify what drives the largest variation in the data.


```{r}
#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$StressType <- StressType_simple[-c(20:22)]
dfforpca$CellType <- metadata$Cell
dfforpca$SGPres <- metadata$SG
dfforpca$KOStat <- metadata$KO
dfforpca$StudyNum <- metadata$Study


# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-5)], scale = FALSE)
p_stress<- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StressType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StressType", values = custom_palette) +
  ggtitle("PCA - Coloured by StressType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_sg <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "SGPres", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "SGPres", values = custom_palette) +
  ggtitle("PCA - Coloured by SGPres") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_cell <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "CellType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "CellType", values = custom_palette) +
  ggtitle("PCA - Coloured by CellType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_ko <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "KOStat", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "KOStat", values = custom_palette) +
  ggtitle("PCA - Coloured by KOStat") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_study <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StudyNum", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StudyNum", values = custom_palette) +
  ggtitle("PCA - Coloured by StudyNum") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_stress
p_sg
p_cell
p_ko
p_study


#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])

```

As one can observe from the PCAs, our PC1 is mostly explained by cell type.

Now, in my opinion, cell type, as well as the other variables, are not really relevant in this case. Therefore, I will treat them as "batch-effect", and will remove them. This way, we can compare samples from different cells, as they were the same.


## Batch Effect Correction


```{r}
batch_cols <- c("Cell", "Stress" ,"KO", "Study")
metadata_cols <- c("SG")
mmbatch <- model.matrix(reformulate(paste0("~0 + ",paste0( batch_cols,collapse = " + "))), data=metadata)
mmkeep <- model.matrix(reformulate(paste0("~0 + ",paste0( metadata_cols,collapse =  "+" ))), data=metadata)
mm <- cbind(mmkeep,mmbatch)
D0 <- DGEList(2^y$E) 
D0 <- calcNormFactors(D0)
y <- voom(D0, mm, plot = F)
fit <- lmFit(y, mm)
beta <- fit$coefficients[,-(1:ncol(mmkeep)),drop=FALSE]
beta[is.na(beta)] <- 0
corrcounts_v1 <- as.matrix(y$E) - beta %*% t(mmbatch)
offset <- apply(corrcounts_v1,1,min) - apply(y$E,1,min)
corrcounts_v1 <- as.data.frame(corrcounts_v1-offset)
corrcounts_v1 <- 2^corrcounts_v1

mm <- model.matrix(~0+SG, metadata)
y <- voom(corrcounts_v1, mm, plot = T)
```

## Batch Corrected PCA

```{r}
#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$StressType <- StressType_simple[-c(20:22)]
dfforpca$CellType <- metadata$Cell
dfforpca$SGPres <- metadata$SG
dfforpca$KOStat <- metadata$KO
dfforpca$StudyNum <- metadata$Study


# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-5)], scale = FALSE)
p_stress<- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StressType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StressType", values = custom_palette) +
  ggtitle("PCA - Coloured by StressType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_sg <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "SGPres", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "SGPres", values = custom_palette) +
  ggtitle("PCA - Coloured by SGPres") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_cell <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "CellType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "CellType", values = custom_palette) +
  ggtitle("PCA - Coloured by CellType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_ko <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "KOStat", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "KOStat", values = custom_palette) +
  ggtitle("PCA - Coloured by KOStat") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_study <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StudyNum", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StudyNum", values = custom_palette) +
  ggtitle("PCA - Coloured by StudyNum") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_stress
p_sg
p_cell
p_ko
p_study


#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])

#These are my corrected normalised counts, which I then use as this serves as my validation dataset for the signature
write.table(t(dfforpca), "~/Projects/StressGranules/SGPAPER/ValidationDataset.txt")
```

## Linear Modelling

Finally, we move on to differential gene expression! In this chunk, I define the linear model I will actually use for my analyses. 

```{r}
#Now I simply create the linear model that isolates the SG signal
fitted <- lmFit(y$E, mm)

head(coef(fitted))

contr <- makeContrasts(SGYES - SGNO , levels = colnames(coef(fitted)))

tmp_contr <- contrasts.fit(fitted, contr)

tmp <- eBayes(tmp_contr)
top.table <- topTable(tmp, sort.by = "logFC", n=Inf)
DEGenes <- top.table
ndegs=20
```


## SG vs Non-SG

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>1.5 & DEGenes$B>10,] )

downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -1.5 & DEGenes$B>10,] )



ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs No-SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20)+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  theme(text=element_text(size=15,  family="Arial MS"))

write.table(DEGenes, "~/Projects/StressGranules/SGPAPER/RealSGs/WCdsRNA.txt", row.names=T, col.names = T)
```
From the volcano, we can see downregulation of EGR transcripts, which are important in inflammation and stress signalling. 

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression.

```{r, fig.height=12, fig.width=12}
finagenes <- rbind(upmarkers, downmarkers)

dfforexpression <- y$E

samplecodes <- data.frame(samples=colnames(dfforexpression)[c(1:25)], purity=c(1:25))
sampletype <- data.frame(samples=colnames(dfforexpression), sampletype=samplenaming)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- as.data.frame(genesforexp)

genesforexp$Genes <- row.names(genesforexp)

genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]

codesforsgpres <- data.frame(sampletype=samplenaming, sgpresence=SGPres[-c(22:24)])
genesforexp <- merge(genesforexp, codesforsgpres, by="sampletype")

winnyplot <- ggplot(genesforexp, aes(x=sgpresence, y=expression)) + geom_violin(colour="black", alpha=0.5) + geom_point(aes(colour=sampletype)) + facet_wrap(.~ Gene, ncol=5, scales = "free")
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = unique(c("#FFB6C1", "#FFB6C1", "#FFDAB9", "#FFDAB9", "#E6E6FA", "#E6E6FA", "#98FB98", "#98FB98", "#87CEEB", "#87CEEB", "#FFFF99", "#FFFF99", "#F08080", "#F08080", "#98FB98", "#98FB98", "#D8BFD8", "#D8BFD8", "#D8BFD8", "#FFA07A", "#FFA07A", "#FFA07A","#6495ED","#6495ED","#6495ED", "#CCCCCC")), labels=c(unique(sampletype)))  + scale_colour_manual(name = "Sample Type", values = unique(c("#FFB6C1", "#FFB6C1", "#FFDAB9", "#FFDAB9", "#E6E6FA", "#E6E6FA", "#98FB98", "#98FB98", "#87CEEB", "#87CEEB", "#FFFF99", "#FFFF99", "#F08080", "#F08080", "#98FB98", "#98FB98", "#D8BFD8", "#D8BFD8", "#D8BFD8", "#FFA07A", "#FFA07A", "#FFA07A","#6495ED","#6495ED","#6495ED", "#CCCCCC")), labels=c(unique(sampletype))) 
winnyplot <- winnyplot + theme_bw() + xlab("Stress State") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,3]),]
gseatool <- DEGenes[,3]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score (NES)",
       title= "Hallmark Pathways",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall
```

This shows the importance of GSEA. Even though we have a downregulation of EGR, we still have an increase in inflammation pathways, even though we have downregulation of NFKB-signalling.

# Session Info

```{r}
sessionInfo()
```