---
title: "Analyses of GSE138988; GSE119977 and GSE173953"
subtitle: "Analyses comparing SG signal on WC transcriptome"
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

In this document, I analyse all 3 WC transcriptome regarding SGs, pertaining to the datasets **GSE138988**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE138988)),  **GSE119977**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE119977))
originally published by **T. Matheny, 2019 and 2021**, and **GSE173953**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE173953))
originally published by **M. Paget 2023**

## Methodology

```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)
library(readxl)

set.seed(1906)
```

```{r}
#From dsRNA dataset
ds.read.counts <- read.table("~/Projects/StressGranules/AS.WC_Transcriptome/GSE173953/TrimmedSamples/alignment_STAR/featureCounts_Results.txt", header = T)[,c(1,7:25,29:31)]

#From  Sorbitol testing
sorbit.read.counts <- read.table("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/featureCounts_results.txt", header = T)[,c(1,10:12,15,16,20:28,32:34)]

#From KO experiments
PE.read.counts <- read.table("~/Projects/StressGranules/AS.WC_Transcriptome/GSE138988/TrimmedSamples/alignment_STAR/featureCounts_results_2.txt", header = T)[,c(1,7:12)]
SE.read.counts <- read.table("~/Projects/StressGranules/AS.WC_Transcriptome/GSE138988/TrimmedSamples/alignment_STAR/featureCounts_results.txt", header = T)[,c(1,7:12)]

#Now we merge everything up
raw.read.counts <- merge(PE.read.counts, SE.read.counts, by="Geneid")
raw.read.counts <- merge(raw.read.counts, sorbit.read.counts, by="Geneid")
raw.read.counts <- merge(raw.read.counts, ds.read.counts, by="Geneid")

biomart <- fread("~/biomart.txt")[,-1]


colnames(raw.read.counts)[1] <- "ensembl_gene_id_version"
colnames(biomart) <- c("ensembl_gene_id_version", "hgnc_symbol")

eltable <- merge(raw.read.counts,biomart,by = "ensembl_gene_id_version", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="")#this removes reads that dont have a corresponding hgnc symbol.
eltable$ensembl_gene_id_version <- NULL


#I sum counts from rows with the same HGNC symbol.
read.counts <- eltable %>%
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))


read.counts <- as.data.frame(read.counts)

row.names(read.counts) <- read.counts$hgnc_symbol #changes rownames into hgnc symbols
read.counts$hgnc_symbol <- NULL
```


## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, warning=FALSE, message=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(20, palette = "Pastel")
custom_palette <- sample(custom_palette, 20)

# Use SampleColouring from metadata
SampleColouring <- c("PE.WTCT", "PE.WTCT", "PE.WTCT", "PE.WTAR", "PE.WTAR", "PE.WTAR", "SE.WTAR", "SE.WTAR", "SE.WTAR", "SE.KOAR", "SE.KOAR", "SE.KOAR", "WTAR", "WTAR", "WTAR", "WTSO", "WTSO", "WTSO", "KOSO", "KOSO", "GFUN", "GFUN", "GFUN", "KOUN", "KOUN", "KOUN", "GFSO", "GFSO", "GFSO", rep("WT.U2.CT",2), rep("WT.U2.DS6",2), rep("KO.U2.CT",2),rep("KO.U2.DS4",2), rep("WT.A5.CT",2), rep("WT.A5.DS4",2), rep("KO.A5.CT",2), rep("KO.A5.DS4",2), rep("KO.U2.DS6",3), rep("KO.U2.DS24",3))

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(6, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

```


As we can see in these plots, in this case there are no sample outliers in terms of complexity or read depth. Excellent!

# Differential Expression Analysis

## Filtration

### ***Before Filtration***

Let us begin with the filtration. Personally, I find that its easier to do visually. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bi-modal distributions, with the first mode corresponding to lowly expressed genes. I remove these by cutting just before the second mode.

```{r filtration, warning=FALSE}
alternew <- log10(rowSums(read.counts))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 4, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

```
By seeing the shape of the graph, we can see where we should cut. We have a **binomial distribution** (low reads and noise vs actual biological meaningful reads). The ideal place to cut them is right before the wedge.

### ***After Filtration***

```{r filtration 2}
#Here is my filtration. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bimodal distributions, with the first mode corresponding to lowly expressed or noisy genes. I remove these by cutting just before the second mode.
filterednew <- subset(read.counts, alternew>4)

MitoGenes <- fread("~/MitoGenes.txt", header = FALSE)[[1]]
#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(), MitoGenes)

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))

#Now we need to select only the rows which have those genes.
finaltable <- read.counts[row.names(read.counts)%in%impgenes,]
```


## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#Preparing the metadata I will use
#This is the name of the samples
Sample_names <- c("PE.WTCT", "PE.WTCT", "PE.WTCT", "PE.WTAR", "PE.WTAR", "PE.WTAR", "SE.WTAR", "SE.WTAR", "SE.WTAR", "SE.KOAR", "SE.KOAR", "SE.KOAR", "WTAR", "WTAR", "WTAR", "WTSO", "WTSO", "WTSO", "KOSO", "KOSO", "GFUN", "GFUN", "GFUN", "KOUN", "KOUN", "KOUN", "GFSO", "GFSO", "GFSO", rep("WT.U2.CT",2), rep("WT.U2.DS6",2), rep("KO.U2.CT",2),rep("KO.U2.DS4",2), rep("WT.A5.CT",2), rep("WT.A5.DS4",2), rep("KO.A5.CT",2), rep("KO.A5.DS4",2), rep("KO.U2.DS6",3), rep("KO.U2.DS24",3))

Cells <- c("U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", "U2-OS", rep("U2-OS",2), rep("U2-OS",2), rep("U2-OS",2),rep("U2-OS",2), rep("A524",2), rep("A524",2), rep("A524",2), rep("A524",2), rep("U2-OS",3), rep("U2-OS",3))

#"Complex" stress
#StressStat <- c("CT", "CT", "CT", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "SO", "SO", "SO", "SO", "SO", "CT", "CT", "CT", "CT", "CT", "CT", "SO", "SO", "SO", rep("CT",2), rep("DS6",2), rep("CT",2),rep("DS4",2), rep("CT",2), rep("DS4",2), rep("CT",2), rep("DS4",2), rep("DS6",3), rep("DS24",3))

#Simplified Stress
StressStat <- c("CT", "CT", "CT", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "AR", "SO", "SO", "SO", "SO", "SO", "CT", "CT", "CT", "CT", "CT", "CT", "SO", "SO", "SO", rep("CT",2), rep("DS",2), rep("CT",2),rep("DS",2), rep("CT",2), rep("DS",2), rep("CT",2), rep("DS",2), rep("DS",3), rep("DS",3))

G3Status <- c("WT", "WT", "WT", "WT", "WT", "WT", "WT", "WT", "WT", "KO", "KO", "KO", "WT", "WT", "WT", "WT", "WT", "WT", "KO", "KO", "GF", "GF", "GF", "KO", "KO", "KO", "GF", "GF", "GF", rep("WT",2), rep("WT",2), rep("KO",2),rep("KO",2), rep("WT",2), rep("WT",2), rep("KO",2), rep("KO",2), rep("KO",3), rep("KO",3))

StudyNum <-c(rep("Study1", 6), rep("Study2",6), rep("Study3", 17), rep("Study4",8),rep("Study5",8),rep("Study6",6))


SGStatus <- c("NO", "NO", "NO", "YES", "YES", "YES", "YES", "YES", "YES", "NO", "NO", "NO", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "NO", "NO", "NO", "NO", "NO", "NO", "YES", "YES", "YES", rep("NO",2), rep("YES",2), rep("NO",2),rep("NO",2), rep("NO",2), rep("YES",2), rep("NO",2), rep("NO",2), rep("NO",3), rep("NO",3))

metadata <- data.frame(Cell=Cells, Stress=StressStat, G3=G3Status, Study=StudyNum, SG=SGStatus)



#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)


#Then we calculate the normalisation factors for these new generated data frames
#Bear in mind this step calculates the normalisation factors, but does not normalise itself.
D0 <- calcNormFactors(D0)

mm <- model.matrix(~0+Sample_names)

#and now we normalize. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```

## Normalisation

For normalisation I convert the read counts into log2CPMs. Its a good way to visualise how the reads are. 

```{r normalization}
#To visualise how the normalisation went.

custom_palette <- qualitative_hcl(20, palette = "Dynamic")
custom_palette <- sample(custom_palette,20)

dfforboxy <- finaltable

orderforplots <- Sample_names
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = custom_palette, labels=unique(Sample_names), breaks =unique(Sample_names)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = custom_palette, labels=unique(Sample_names), breaks =unique(Sample_names)) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```


## PCA

```{r PCA, fig.height=8, fig.width=8}
#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- Sample_names
dfforpca$StressType <- metadata$Stress
dfforpca$CellType <- metadata$Cell
dfforpca$SGPres <- metadata$SG
dfforpca$StudyNum <- metadata$Study
dfforpca$G3 <- metadata$G3

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-6)], scale = FALSE)
p_sample<- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Sample") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_stress<- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StressType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StressType", values = custom_palette) +
  ggtitle("PCA - Coloured by StressType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_sg <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "SGPres", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "SGPres", values = custom_palette) +
  ggtitle("PCA - Coloured by SGPres") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_cell <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "CellType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "CellType", values = custom_palette) +
  ggtitle("PCA - Coloured by CellType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_study <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StudyNum", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StudyNum", values = custom_palette) +
  ggtitle("PCA - Coloured by StudyNum") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_g3 <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "G3", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "G3", values = custom_palette) +
  ggtitle("PCA - Coloured by G3") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_sample
p_stress
p_sg
p_cell
p_study
p_g3


#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])

```
As one can observe from the PCAs, our PC1 is mostly explained by cell type.

Now, in my opinion, cell type, as well as the other variables, are not really relevant in this case. Therefore, I will treat them as "batch-effect", and will remove them. This way, we can compare samples from different cells, as they were the same.


## Batch Effect Correction

```{r}
batch_cols <- c("Cell", "Study", "G3", "Stress")
metadata_cols <- c("SG")
mmbatch <- model.matrix(reformulate(paste0("~0 + ",paste0( batch_cols,collapse = " + "))), data=metadata)
mmkeep <- model.matrix(reformulate(paste0("~0 + ",paste0( metadata_cols,collapse =  "+" ))), data=metadata)
mm <- cbind(mmkeep,mmbatch)
D0 <- DGEList(2^y$E) #senescence_normcounts
D0 <- calcNormFactors(D0)
y <- voom(D0, mm, plot = F)
fit <- lmFit(y, mm)
beta <- fit$coefficients[,-(1:ncol(mmkeep)),drop=FALSE]
#beta <- fit$coefficients
beta[is.na(beta)] <- 0
corrcounts_v1 <- as.matrix(y$E) - beta %*% t(mmbatch)
offset <- apply(corrcounts_v1,1,min) - apply(y$E,1,min)
corrcounts_v1 <- as.data.frame(corrcounts_v1-offset)
corrcounts_v1 <- 2^corrcounts_v1


mm <- model.matrix(~0+SG, metadata)

y <- voom(corrcounts_v1, mm, plot = T)
```


```{r fig.height=8, fig.width=8}
#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- Sample_names
dfforpca$StressType <- metadata$Stress
dfforpca$CellType <- metadata$Cell
dfforpca$SGPres <- metadata$SG
dfforpca$StudyNum <- metadata$Study
dfforpca$G3 <- metadata$G3

# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-6)], scale = FALSE)
p_sample<- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Sample") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_stress<- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StressType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StressType", values = custom_palette) +
  ggtitle("PCA - Coloured by StressType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_sg <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "SGPres", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "SGPres", values = custom_palette) +
  ggtitle("PCA - Coloured by SGPres") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_cell <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "CellType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "CellType", values = custom_palette) +
  ggtitle("PCA - Coloured by CellType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_study <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StudyNum", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StudyNum", values = custom_palette) +
  ggtitle("PCA - Coloured by StudyNum") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_g3 <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "G3", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "G3", values = custom_palette) +
  ggtitle("PCA - Coloured by G3") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_sample
p_stress
p_sg
p_cell
p_study
p_g3


#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])

```


## Linear Modelling

Finally, we move on to differential gene expression! In this chunk, I define the linear model I will actually use for my analyses. 

```{r}


fitted <- lmFit(y$E, mm)

head(coef(fitted))

contr <- makeContrasts(SGYES - SGNO , levels = colnames(coef(fitted)))

tmp_contr <- contrasts.fit(fitted, contr)

tmp <- eBayes(tmp_contr)
top.table <- topTable(tmp, sort.by = "logFC", n=Inf)
DEGenes <- top.table
ndegs=40
```

## SG vs Non-SG

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>1 & DEGenes$adj.P.Val<0.05,] )

downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -1.5 & DEGenes$adj.P.Val<0.05,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#53A2BE",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#BA3B46",size=4)+
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#53A2BE",max.overlaps = 20)+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=3,point.padding=-0.5, color="#BA3B46",max.overlaps = 20)+
  theme(text=element_text(size=15,  family="Arial MS"))

write.table(DEGenes, "~/Projects/StressGranules/SGPAPER/RealSGs/MergedWCSig.txt", row.names = T, col.names = T)
```
From the volcano, we can see downregulation of FOS and IFIT transcripts, which are important in inflammation and stress signalling. 


## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered pathways** between enriched and depleted RNAs from the stress granules. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,3]),]
gseatool <- DEGenes[,3]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection =  "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.05

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score (NES)",
       title= "Hallmark Pathways",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall
```

We have a downregulation  in several inflammation pathways.

# Session Info

```{r}
sessionInfo()
```
