---
title: "markeR Based Signature Usage"
subtitle: "Creation of SGScore and Simulated Signatures"
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# Introduction

In this document, I use the markeR package (in development) to calculate the SGScore and Simulate Random Signatures as control. For more information, please refer to the original package developed by **Martins-Silva, R.** ([GitHub Link](https://github.com/DiseaseTranscriptomicsLab/markeR))

 
```{r}
library(markeR)
library(data.table)
library(msigdbr)
library(ggplot2)
library(ggpubr)
library(ggExtra)
```

 
# Preparation of TCGA data

First, I load the TCGA data. Please refer to the original resource in case of need, at the **GDC Portal** ([TCGA Link](https://portal.gdc.cancer.gov/)).

```{r}
#This object contains both counts and metadata. For preprocessing, please refer to markdown file "SIGNATURE_TCGA.Rmd". 
TCGAobj_tissue <- readRDS("~/Projects/TCGA/TCGAobj_tissue.RDS")

#Not all TCGA tissues have enough number of tumour and normal samples to warrant a direct comparsion. Herein, I select only those that do (at least 5 samples of each)
tissuestocheck <- c("BLCA", "BRCA", "CHOL", "COAD", "ESCA", "HNSC", "READ", "STAD", "UCEC", "KICH", "KIRC", "KIRP", "LIHC", "LUAD", "LUSC", "PRAD", "THCA")

TCGAobj_tissue_TvsN <- TCGAobj_tissue[names(TCGAobj_tissue) %in% tissuestocheck]


#Preprocessing of the TCGA data. For simplicity, I keep only either tumour or normal samples, discarding only sample types. 
for (tissue in names(TCGAobj_tissue_TvsN)) {

  # Retrieve metadata and logCPM for the tissue
  metadata <- as.character(TCGAobj_tissue_TvsN[[tissue]]$metadata)
  logCPM <- TCGAobj_tissue_TvsN[[tissue]]$logCPM
  
  # Replace "01" with "Tumor" and "11" with "Normal"
  metadata[metadata == "01"] <- "Tumor"
  metadata[metadata == "11"] <- "Normal"

  # Find the two most common tumour types (now updated to "Tumor" and "Normal")
  top_events <- names(sort(table(metadata), decreasing = TRUE))[1:2]

  # Filter metadata and logCPM to include only the two most common events
  keep_indices <- which(metadata %in% top_events)
  TCGAobj_tissue_TvsN[[tissue]]$metadata <- metadata[keep_indices]
  TCGAobj_tissue_TvsN[[tissue]]$logCPM <- logCPM[, keep_indices, drop = FALSE]
}



```

# Signatures

```{r}
# I load both of my signatures
Signature1 <- fread("~/SG_SIGNATURE_LASSOISH.txt")
Signature2 <- fread("~/SG_SIGNATURE_ELASTICNET.txt")

# Since my signature is bidirectional, I convert Pos or Neg signal to either 1 or -1
Signature1[, Signal := ifelse(Signal == "Pos", 1, -1)]
Signature2[, Signal := ifelse(Signal == "Pos", 1, -1)]


#In case I want to compare, I load some gene sets which may be interesting to compare my signature to
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "H")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}



InflammatoryResponse <- data.frame(Gene=hallmarks.gsets$HALLMARK_INFLAMMATORY_RESPONSE, Signal=1)
InterferonAlpha <- data.frame(Gene=hallmarks.gsets$HALLMARK_INTERFERON_ALPHA_RESPONSE, Signal=1)
InterferonGamma <- data.frame(Gene=hallmarks.gsets$HALLMARK_INTERFERON_GAMMA_RESPONSE, Signal=1)
TNFNFKB <- data.frame(Gene=hallmarks.gsets$HALLMARK_TNFA_SIGNALING_VIA_NFKB, Signal=1)
IL2 <- data.frame(Gene=hallmarks.gsets$HALLMARK_IL2_STAT5_SIGNALING, Signal=1)
IL6 <- data.frame(Gene=hallmarks.gsets$HALLMARK_IL6_JAK_STAT3_SIGNALING, Signal=1)
OXPHOS <- data.frame(Gene=hallmarks.gsets$HALLMARK_OXIDATIVE_PHOSPHORYLATION, Signal=1)

```
 


## Running
 
 
Here, I calculate the SGScore for both signatures, for all tumour types with enough sample types.

```{r fig.height=14, fig.width=16, message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)  # Install with install.packages("patchwork")

plot_list <- list()  # Create an empty list to store plots
plot_index <- 1  # Track the plot index

for (tissue in names(TCGAobj_tissue_TvsN)) {
  
  counts <- 2^TCGAobj_tissue_TvsN[[tissue]]$logCPM
  metadata <- data.frame(sample=colnames(counts), tumour = TCGAobj_tissue_TvsN[[tissue]]$metadata)
  
  # Check if there is more than one unique tumour type
  if(length(unique(metadata$tumour)) == 1) {
    message(paste("Skipping tissue", tissue, "because there is only one sample type"))
    next
  }

  # Create frequency table for top 2 most frequent instances
  freq_table <- sort(table(TCGAobj_tissue_TvsN[[tissue]]$metadata), decreasing = TRUE)
  top_instances <- names(freq_table)[1:2]
  cond_cohend <- list(A = top_instances[1], B = top_instances[2])

  # Wrap in tryCatch to handle errors
  tryCatch({
    plot_logmedian <- PlotScores(data = counts, metadata = metadata, 
                                 gene_sets = list(SG1=Signature1, SG2=Signature2),
                                 ColorVariable = "tumour",
                                 ColorValues = c("#779ECB", "#FFA071"),
                                 Variable="tumour",  
                                 method ="ranking",  
                                 ConnectGroups=TRUE, 
                                 ncol = NULL, 
                                 nrow = NULL, 
                                 widthTitle=20, 
                                 legend_nrow = 1, 
                                 pointSize=4,
                                 cond_cohend=cond_cohend,
                                 title=paste0(tissue),
                                 labsize=7, 
                                 titlesize = 10, 
                                 pvalcalc = TRUE)  
    
    plot_list[[plot_index]] <- plot_logmedian  # Store the plot
    plot_index <- plot_index + 1  # Increment the index
    
  }, error = function(e) {
    message(paste("Error in tissue", tissue, ":", e$message))
  })
}

# Combine all plots into a grid layout using patchwork
if (length(plot_list) > 0) {
  final_plot <- wrap_plots(plot_list, ncol = 5)  # Adjust ncol as needed
  print(final_plot)
}

tiff("~/Projects/StressGranules/SGPAPER/Images/SignatureCancer.tiff", width = 16, height=14, units = "in", res=300)
final_plot
dev.off()

```

# Function for Simulation

Since most gene signatures exhibit prognostic potential in cancer, I opted to compare both my signatures with random simulated signatures. If my signatures exhibit better prognostic potential than most random ones, they may therefore still hold potential. 

```{r}
# Helper function to compute Cohen's d
cohen_d <- function(x, y) {
  n1 <- length(x)
  n2 <- length(y)
  if(n1 < 2 || n2 < 2) return(NA)
  m1 <- mean(x)
  m2 <- mean(y)
  s1 <- sd(x)
  s2 <- sd(y)
  pooled_sd <- sqrt(((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / (n1 + n2 - 2))
  if (pooled_sd == 0) return(NA)
  d <- (m1 - m2) / pooled_sd
  d <- abs(d) # dor this application we don't need the direction of the classification, just if it works
  return(d)
}

compute_cohen_d <- function(dfScore, variable, quantitative_var="score") { 
  
  # Get unique group values
  unique_groups <- unique(dfScore[[variable]])
  
  # Ensure there are at least two groups
  if (length(unique_groups) < 2) {
    stop("The grouping column must have at least two unique values.")
  }
  
  
  
  # Store results
  results <- data.frame(Group1 = character(), Group2 = character(), 
                        CohenD = numeric(), PValue = numeric(), stringsAsFactors = FALSE)
  
  # Compute Cohen's d and p-value for all unique pairs
  combs <- combn(unique_groups, 2, simplify = FALSE)
  for (pair in combs) {
    x <- dfScore[dfScore[[variable]] == pair[1], quantitative_var, drop = TRUE]
    y <- dfScore[dfScore[[variable]] == pair[2], quantitative_var, drop = TRUE]
    
    d <- cohen_d(x, y)
    p_val <- t.test(x, y, var.equal = TRUE)$p.value
    
    results <- rbind(results, data.frame(Group1 = pair[1], Group2 = pair[2], 
                                         CohenD = d, PValue = p_val, stringsAsFactors = FALSE))
  }
  
  return(results)
}


# Function to flatten the list of lists into a data frame
flatten_results <- function(nested_list) {
  # Create an empty list to store results
  result_list <- list()
  
  # Iterate over methods (first level)
  for (method in names(nested_list)) {
    # Iterate over gene signatures (second level)
    for (signature in names(nested_list[[method]])) {
      # Extract the data frame (third level)
      df <- nested_list[[method]][[signature]]
      
      # Add columns for method and signature
      df$method <- method
      df$signature <- signature
      
      # Store the modified data frame in the list
      result_list[[paste(method, signature, sep = "_")]] <- df
    }
  }
  
  # Combine all data frames into one
  final_df <- do.call(rbind, result_list)
  
  # Reset row names
  rownames(final_df) <- NULL
  
  return(final_df)
}


CohenD_allConditions <- function(data, metadata, gene_sets, variable) {
  
  # Step 1: Check if variable exists in metadata
  if (!variable %in% colnames(metadata)) {
    stop(paste("Error: Variable", variable, "not found in metadata."))
  }
  
  # Step 2: Compute scores for all methods and signatures
  listScores <- CalculateScores(data = data, metadata = metadata, gene_sets = gene_sets, method = "all") 
  
  # Step 3: Flatten into a data frame with method & signature columns
  dfScores <- flatten_results(listScores)
  
  # Step 4: Initialize result list
  result_list <- list()
  
  # Step 5: Loop through each signature
  for (signature in unique(dfScores$signature)) {
    
    # Filter for the specific signature
    df_subset <- dfScores[dfScores$signature == signature, ]
    
    # Initialize storage for this signature
    cohen_d_results <- list()
    p_value_results <- list()
    
    # Step 6: Loop through each method
    for (method in unique(df_subset$method)) {
      
      # Subset data for this method
      df_method <- df_subset[df_subset$method == method, ]
      
      # Compute Cohen's d and p-values
      cohen_results <- compute_cohen_d(df_method, variable)
      
      # Convert to named vectors (column names = comparisons)
      cohen_d_results[[method]] <- setNames(cohen_results$CohenD, paste0(cohen_results$Group1, ":", cohen_results$Group2))
      p_value_results[[method]] <- setNames(cohen_results$PValue, paste0(cohen_results$Group1, ":", cohen_results$Group2))
    }
    
    # Convert lists to data frames
    cohen_d_df <- as.data.frame(do.call(rbind, cohen_d_results))
    p_value_df <- as.data.frame(do.call(rbind, p_value_results))
    
    # Store results in the named list
    result_list[[signature]] <- list(CohenD = cohen_d_df, PValue = p_value_df)
  }
  
  # Step 7: Return the list
  return(result_list)
}


```

## Running


```{r fig.height=20, fig.width=20, warning=FALSE}

original_signatures <- list(SG1=Signature1, SG2=Signature2)

#original_signatures <- Senescence
FDR_Simulation_alt <- function(data, metadata, original_signatures, gene_list, number_of_sims, title_for_plot) {
  
  # Combine original and simulated gene sets
  all_signatures <- original_signatures
  
  for (sig in names(original_signatures)) {
    cur_sig <- original_signatures[[sig]]
    
    # Convert vector to data frame if needed
    if (is.vector(cur_sig)) {
      cur_sig <- data.frame(Gene = cur_sig, Signal = 1)
    }
    
    # Generate simulated signatures
    for (sim in 1:number_of_sims) {
      cur_model_sig <- cur_sig
      cur_model_sig$Gene <- sample(gene_list, nrow(cur_sig))  # Randomly sample genes
      all_signatures[[paste0(sig, "_sim", sim)]] <- cur_model_sig
    }
  }
  
  # Run CalculateScores on ALL signatures (original + simulated)
  scores_df <- CalculateScores(data = data, metadata = metadata, 
                               gene_sets = all_signatures, method = "ranking")
  
  return(scores_df)
}



simulationlist <- list()

for (tissue in names(TCGAobj_tissue)) {
  print(tissue)
  counts <- 2^TCGAobj_tissue[[tissue]]$logCPM
  metadata <- data.frame(sample=colnames(counts), tumour = TCGAobj_tissue[[tissue]]$metadata)

simulationlist[[tissue]] <- FDR_Simulation_alt(data = counts,
               metadata = metadata,
               original_signatures = original_signatures,
               gene_list = row.names(TCGAobj_tissue[["LAML"]]$logCPM),
               number_of_sims = 100,
               title_for_plot = paste0(tissue))
}

#I save the simulated signatures for future use
saveRDS(simulationlist, "~/Projects/StressGranules/SGPAPER/Simulations/simulationlist.RDS")

```

# Validation Image

This is a simple annex where I designed the image where I compare the behaviour of both my signatures on the discovery and validation datasets. 

```{r eval=FALSE, echo=FALSE, fig.height=4, fig.width=10}
cond_cohend <- list(A=c("YES"),  
                    B=c("NO"))

SG1 <- SG1 %>%
  mutate(Signal = ifelse(Signal == "Pos", 1, -1))

#I directly obtained the data directly from each datasets analysis. Please refer to the corresponding markdowns to get this information.

#The information here is directly obtained from the discovery dataset
Counts_Disc <- Counts
Metadata_Disc <- Metadata
Metadata_Disc$SGPres <- factor(Metadata_Disc$SGPres, levels = c("NO", "YES"))

(discimage <- PlotScores(data = 2^Counts_Disc,metadata = Metadata_Disc, gene_sets = list(SG1=SG1, SG2=SG2),
                             ColorVariable = "SGPres",
                             ColorValues = c("#779ECB", "#FFA071"),
                             GroupingVariable="SGPres",  
                             method ="ranking",  
                             ConnectGroups=TRUE, 
                             ncol = NULL, 
                             nrow = NULL, 
                             widthTitle=20, 
                             legend_nrow = 1, 
                             pointSize=4,
                             cond_cohend=cond_cohend,
                             title=paste0("Discovery Dataset"),
                             labsize=7, 
                             titlesize = 10, pvalcalc = T))

#The information here is directly obtained from the validation dataset
Counts_Val <- Validation[-nrow(Validation),-1]
Counts_Val[,1:22] <- lapply(Counts_Val[,1:22], as.numeric)

Metadata_Val <- data.frame(sampleID=colnames(Counts_Val), SGPres=t(Validation[nrow(Validation),-1]))
Metadata_Val$SGPres <- factor(Metadata_Val$SGPres, levels = c("NO", "YES"))
(valimage <- PlotScores(data = 2^Counts_Val,metadata = Metadata_Val, gene_sets = list(SG1=SG1, SG2=SG2),
                             ColorVariable = "SGPres",
                             ColorValues = c("#779ECB", "#FFA071"),
                             GroupingVariable="SGPres",  
                             method ="ranking",  
                             ConnectGroups=TRUE, 
                             ncol = NULL, 
                             nrow = NULL, 
                             widthTitle=20, 
                             legend_nrow = 1, 
                             pointSize=4,
                             cond_cohend=cond_cohend,
                             title=paste0("Validation Dataset"),
                             labsize=7, 
                             titlesize = 10, pvalcalc = T))

(Newgid <- ggarrange(discimage,valimage, ncol = 2,labels = c("A","B"), font.label = list(size=25)))

tiff("~/Projects/StressGranules/SGPAPER/SignatureValidation.tiff", width = 12, height=4, units = "in", res=300)
Newgid
dev.off()
```

# Session Info

```{r}
sessionInfo()
```
