---
title: "Analyses of GSE138988 and GSE119977"
subtitle: "Analyses comparing SG signal on WC transcriptome"
author: 
  - name: "Alexandre Kaizeler^[Disease Transcriptomics Group, Gulbenkian Institute of Molecular Medicine]"
    email: alexandreafonso@gimm.pt
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  rmdformats::downcute:
    downcute_theme: "chaos"
---


# Introduction

In this document, I present all analyses pertaining to the datasets **GSE138988**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE138988)), and  **GSE119977**  
([NCBI GEO Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE119977))
originally published by **T. Matheny, 2019 and 2021**.

## Methodology

```{r, message=FALSE}
# These are all the libraries I used throughout the analysis. Not all of them will be required on this guidebook
library(Rsubread)
library(limma)
library(edgeR)
library(DESeq2)
library(magrittr)
library(ggrepel)
library(tidyverse)
library(vsn)
library(ggplot2)
library(NMF)
library(biomaRt)
library(ggpubr)
library(factoextra)
library(ggfortify)
library(plotly)
library(reshape2)
library(fgsea)
library(msigdbr)
library(data.table)
library(colorspace)
library(readxl)

set.seed(1906)
```

## Loading up the data

By merging these two datasets, I have enough sample number and variety to do a linear model to isolate the SG signal. This is because I need stressed cells with and without SG, with respective unstressed controls. 

```{r}
#From Sorbitol testing
bigstudy <- read.table("~/Projects/StressGranules/SG_Transcriptome/RP_Data_2021/TrimmedSamples/alignment_STAR/featureCounts_results.txt", header = T)[,c(1,10:12,15,16,20:28,32:34)]

#From KO experiments
PE <- read.table("~/Projects/StressGranules/AS.WC_Transcriptome/GSE138988/TrimmedSamples/alignment_STAR/featureCounts_results_2.txt", header = T)[,c(1,7:12)]
SE <- read.table("~/Projects/StressGranules/AS.WC_Transcriptome/GSE138988/TrimmedSamples/alignment_STAR/featureCounts_results.txt", header = T)[,c(1,7:12)]

#Now we merge everything up
raw.read.counts <- merge(PE, SE, by="Geneid")
raw.read.counts <- merge(raw.read.counts, bigstudy, by="Geneid")


biomart <- fread("~/biomart.txt")[,-1]


colnames(raw.read.counts)[1] <- "ensembl_gene_id_version"
colnames(biomart) <- c("ensembl_gene_id_version", "hgnc_symbol")

eltable <- merge(raw.read.counts,biomart,by = "ensembl_gene_id_version", all= TRUE) #this creates a table with both ensemble ids and HGNC codes.

eltable <- subset(eltable, hgnc_symbol!="")#this removes reads that dont have a corresponding hgnc symbol.
eltable$ensembl_gene_id_version <- NULL


#I sum counts from rows with the same HGNC symbol.
read.counts <- eltable %>%
  group_by(hgnc_symbol) %>%
  summarise(across(everything(), sum, na.rm = TRUE))

read.counts <- as.data.frame(read.counts)

row.names(read.counts) <- read.counts$hgnc_symbol #changes rownames into hgnc symbols
read.counts$hgnc_symbol <- NULL
```


## Extra Quality Control

Let's check for sample complexity. It is possible that one or more samples have different complexity. In specific, I am going to check if enough reads have been obtained from every sample. If not, we should have a great bias to housekeeping genes. Another way to put it, is to say that even with all reads, we won't obtain a great deal of different genes.

```{r, warning=FALSE, message=FALSE, results='hide'}

# Define color palette
custom_palette <- qualitative_hcl(10, palette = "Pastel")
custom_palette <- sample(custom_palette, 10)

# Use SampleColouring from metadata
SampleColouring <- c("PE.WTCT", "PE.WTCT", "PE.WTCT", "PE.WTAR", "PE.WTAR", "PE.WTAR", "SE.WTAR", "SE.WTAR", "SE.WTAR", "SE.KOAR", "SE.KOAR", "SE.KOAR", "WTAR", "WTAR", "WTAR", "WTSO", "WTSO", "WTSO", "KOSO", "KOSO", "GFUN", "GFUN", "GFUN", "KOUN", "KOUN", "KOUN", "GFSO", "GFSO", "GFSO")

# Build cumulative counts
final <- sapply(1:ncol(read.counts), function(i) {
  sort_vals <- read.counts[order(read.counts[, i], decreasing = TRUE), i]
  cumsum(sort_vals)
})
final <- as.data.frame(final)
colnames(final) <- colnames(read.counts)

# Melt and add gene position + sample info
finalmelt <- reshape2::melt(final, variable.name = "Sample")
finalmelt$genes <- rep(1:nrow(final), times = ncol(final))
finalmelt$SampleColouring <- rep(SampleColouring, each = nrow(final))

# Plot cumulative read coverage
ggplot(data = finalmelt, aes(x = genes, y = log10(value), group = Sample, color = SampleColouring)) +
  geom_line() +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  theme_bw() +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (log10)") +
  xlim(0, 20000) +
  ylim(6, 8) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

# Create percentage-based plot
orderfinal <- subset(finalmelt, genes == max(finalmelt$genes))
trial <- merge(orderfinal[, c("Sample", "value")], finalmelt, by = "Sample", suffixes = c(".total", ".current"))
trial$percent <- (trial$value.current / trial$value.total) * 100
trial$SampleColouring <- rep(SampleColouring, each = nrow(final))

ggplot(data = trial, aes(x = genes, y = percent, group = Sample, color = SampleColouring)) +
  geom_line() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  scale_color_manual(values = custom_palette, name = "Cell Line") +
  xlab("Number of Genes Covered") +
  ylab("Number of Reads (%)") +
  xlim(0, 20000) +
  theme_bw() +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()
  )

```

As we can see in these plots, even though there is a sample type with lower read depth, in this case there are no individual sample outliers in terms of complexity or read depth. Excellent!

# Differential Expression Analysis

## Filtration

### ***Before Filtration***

Let us begin with the filtration. Personally, I find that its easier to do visually. I am aware that my filtration is more violent than most. Usually, from RNA-Seq experiments, we commonly obtain bi-modal distributions, with the first mode corresponding to lowly expressed genes. I remove these by cutting just before the second mode. This time, I opted to also observe every sample type by itself.

```{r, warning=FALSE}
alternew.PE.WTCT <- log10(rowSums(read.counts[,c(1:3)]))
alternew.PE.WTCT <- as.data.frame(alternew.PE.WTCT)

alternew.PE.WTAR <- log10(rowSums(read.counts[,c(4:6)]))
alternew.PE.WTAR <- as.data.frame(alternew.PE.WTAR)

alternew.SE.WTAR <- log10(rowSums(read.counts[,c(7:9)]))
alternew.SE.WTAR <- as.data.frame(alternew.SE.WTAR)

alternew.SE.KOAR <- log10(rowSums(read.counts[,c(10:12)]))
alternew.SE.KOAR <- as.data.frame(alternew.SE.KOAR)

alternew.WTAR <- log10(rowSums(read.counts[,c(13:15)]))
alternew.WTAR <- as.data.frame(alternew.WTAR)

alternew.WTSO <- log10(rowSums(read.counts[,c(16:18)]))
alternew.WTSO <- as.data.frame(alternew.WTSO)

alternew.KOSO <- log10(rowSums(read.counts[,c(19:20)]))
alternew.KOSO <- as.data.frame(alternew.KOSO)

alternew.GFUN <- log10(rowSums(read.counts[,c(21:23)]))
alternew.GFUN <- as.data.frame(alternew.GFUN)

alternew.KOUN <- log10(rowSums(read.counts[,c(24:26)]))
alternew.KOUN <- as.data.frame(alternew.KOUN)

alternew.GFSO <- log10(rowSums(read.counts[,c(27:29)]))
alternew.GFSO <- as.data.frame(alternew.GFSO)

alternew <- log10(rowSums(read.counts))
alternew <- as.data.frame(alternew)

ggplot(alternew, aes(x=alternew)) + geom_density() + theme_bw() + geom_vline(xintercept = 4, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - All Samples")

ggplot(alternew.PE.WTCT, aes(x=alternew.PE.WTCT)) + geom_density(fill="#FFB6C1", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - PE.WTCT Samples")

ggplot(alternew.PE.WTAR, aes(x=alternew.PE.WTAR)) + geom_density(fill="#FFDAB9", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - PE.WTAR Samples")


ggplot(alternew.SE.WTAR, aes(x=alternew.SE.WTAR)) + geom_density(fill="#E6E6FA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - SE.WTAR Samples")

ggplot(alternew.SE.KOAR, aes(x=alternew.SE.KOAR)) + geom_density(fill="#98FB98", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.5, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - SE.KOAR Samples")

ggplot(alternew.WTAR, aes(x=alternew.WTAR)) + geom_density(fill="#87CEFA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - WTAR Samples")

ggplot(alternew.WTSO, aes(x=alternew.WTSO)) + geom_density(fill="#EEE8AA", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - WTSO Samples")

ggplot(alternew.KOSO, aes(x=alternew.KOSO)) + geom_density(fill="#F08080", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - KOSO Samples")

ggplot(alternew.GFUN, aes(x=alternew.GFUN)) + geom_density(fill="#B0E0E6", alpha=0.5) + theme_bw() + geom_vline(xintercept = 2.75, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - GFUN Samples")

ggplot(alternew.KOUN, aes(x=alternew.KOUN)) + geom_density(fill="#6495ED", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - KOUN Samples")

ggplot(alternew.GFSO, aes(x=alternew.GFSO)) + geom_density(fill="#D8BFD8", alpha=0.5) + theme_bw() + geom_vline(xintercept = 3.25, linetype = "dashed") + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) + ggtitle("Density of Reads - GFSO Samples")
```


By seeing the shape of the graph, we can see where we should cut. We have a **binomial distribution** (low reads and noise vs actual biological meaningful reads). The ideal place to cut them is right before the wedge.

### ***After Filtration***

I didn't do it this time, but I've maintained the code to filter by sample type. This time, I considered that it was more correct to filter all samples together.

```{r}
filterednew <- subset(read.counts, alternew>4)
filterednew.PE.WTCT <- subset(read.counts, alternew.PE.WTCT>3)
filterednew.PE.WTAR <- subset(read.counts, alternew.PE.WTAR>3)
filterednew.SE.WTAR <- subset(read.counts, alternew.SE.WTAR>3)
filterednew.SE.KOAR <- subset(read.counts, alternew.SE.KOAR>2.5)
filterednew.WTAR <- subset(read.counts, alternew.WTAR>3)
filterednew.WTSO <- subset(read.counts, alternew.WTSO>3)
filterednew.KOSO <- subset(read.counts, alternew.KOSO>2.75)
filterednew.GFUN <- subset(read.counts, alternew.GFUN>2.75)
filterednew.KOUN <- subset(read.counts, alternew.KOUN>3)
filterednew.GFSO <- subset(read.counts, alternew.GFSO>3.25)

#after cutting, its a good idea to visualize again
falternew <- log10(rowSums(filterednew))
falternew <- as.data.frame(falternew)
ggplot(falternew, aes(x=falternew)) + geom_density() + theme_bw() + ylab("Events") + xlab("Number of Reads (log10)") +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

#lets save the genes which we maintain
impgenes <- c(row.names(falternew))
impgenes.PE.WTCT <- c(row.names(filterednew.PE.WTCT))
impgenes.PE.WTAR <- c(row.names(filterednew.PE.WTAR))
impgenes.SE.WTAR <- c(row.names(filterednew.SE.WTAR))
impgenes.SE.KOAR <- c(row.names(filterednew.SE.KOAR))
impgenes.WTAR <- c(row.names(filterednew.WTAR))
impgenes.WTSO <- c(row.names(filterednew.WTSO))
impgenes.KOSO <- c(row.names(filterednew.KOSO))
impgenes.GFUN <- c(row.names(filterednew.GFUN))
impgenes.KOUN <- c(row.names(filterednew.KOUN))
impgenes.GFSO <- c(row.names(filterednew.GFSO))

#Now we need to select only the rows which have those genes.
finaltable <- read.counts[row.names(read.counts)%in%impgenes,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.PE.WTCT,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.PE.WTAR,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.SE.WTAR,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.SE.KOAR,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.WTAR,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.WTSO,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.KOSO,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.GFUN,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.KOUN,]
#finaltable <- finaltable[row.names(finaltable)%in%impgenes.GFSO,]
```


## Matrices

Good! Now we can proceed by designing our matrices. 

```{r}
#To be able to use Voom, we must first convert our read count table into a special DGE object.
D0 <- DGEList(finaltable)


#Then we calculate the normalisation factors for these new generated data frames
#Bear in mind this step calculates the normalisation factors, but does not normalise itself.
D0 <- calcNormFactors(D0)


#Lets create a dataframe with this information
#Just in case I end up needing it, this is the full name of every sample
sampletype <- c("WTUN", "WTUN", "WTUN", "WTAR", "WTAR", "WTAR", "WTAR", "WTAR", "WTAR", "KOAR", "KOAR", "KOAR", "WTAR", "WTAR", "WTAR", "WTSO", "WTSO", "WTSO", "KOSO", "KOSO", "GFUN", "GFUN", "GFUN", "KOUN", "KOUN", "KOUN", "GFSO", "GFSO", "GFSO")

studyorigin <- c(rep("Study1", 6), rep("Study3",6), rep("Study2", 17))
#In this version of stress, I descriminate the stress type
StressType <- c("NO", "NO", "NO", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Arsenite", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "Sorbitol", "NO", "NO", "NO", "NO", "NO", "NO","Sorbitol", "Sorbitol", "Sorbitol")
#In this version of stress, I do not
StressType_simple <- c("NO", "NO", "NO", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "NO", "NO", "NO", "NO", "NO", "NO","YES", "YES", "YES")

#the cell type refers to the g3bp1 state
CellType <- c(rep("WT",9),"KO", "KO", "KO", rep("WT",6),"KO", "KO", "GF", "GF", "GF", "KO", "KO", "KO", "GF", "GF", "GF")
SGPres <- c("NO", "NO", "NO", "YES", "YES", "YES", "YES", "YES", "YES", "NO", "NO", "NO", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "YES", "NO", "NO", "NO", "NO", "NO", "NO", "YES", "YES", "YES")

metadata <- data.frame(Study=studyorigin, Stress=StressType, Cell=CellType, SG=SGPres)


#Now we do the matrixes
mm <- model.matrix(~0+studyorigin+StressType+CellType+SGPres)


#and now we normalise. This step also turns our counts into log2-counts per million.
y <- voom(D0, mm, plot = T)
```

The Mean-Variance Trend graph from voom also allows to determine if the filtration was appropriate. In this case, it was.

## Normalisation

Now, we are almost ready to move on to the differential gene expression analysis. I will first check to see if any issues arose during the normalisation process, by looking at samples before and after this process.

```{r}
dfforboxy <- finaltable
orderforplots <- sampletype
colouring <- data.frame(Sample=colnames(dfforboxy), Types=orderforplots)
dfforboxy <- reshape2::melt(dfforboxy)
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
# Convert 'Types' to factor with custom order
dfforboxy$Types <- factor(dfforboxy$Types, levels = unique(sampletype))

prenorm <- ggplot(dfforboxy, aes(x=Sample, y=(log2(Expression+1)), fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c(custom_palette), labels=c(unique(orderforplots))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("Prior to Normalisation")

#Using the Standard Matrix

dfforboxy <- y$E
dfforboxy <- reshape2::melt(dfforboxy)
dfforboxy <- dfforboxy[,-1]
colnames(dfforboxy) <- c("Sample", "Expression")
dfforboxy <- merge(dfforboxy, colouring, by="Sample")
# Convert 'Types' to factor with custom order
dfforboxy$Types <- factor(dfforboxy$Types, levels = unique(sampletype))
postnorm <- ggplot(dfforboxy, aes(x=Sample, y=Expression, fill=Types)) + geom_boxplot(alpha=0.6) + theme_bw() + ylab("Read Counts (Log2CPM)") + xlab("Samples") + scale_fill_manual(name = "Sample Type", values = c(custom_palette), labels=c(unique(orderforplots))) +
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "None") + ggtitle("After Normalisation")

ggarrange(prenorm, postnorm, nrow=1)

```
In this case, from the boxplots, we can see that there no samples which can be considered outliers.

## PCA

I like to start with a Principal Component Analysis (PCA), which helps to identify what drives the largest variation in the data.

```{r}
#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- sampletype
dfforpca$StressType <- metadata$Stress
dfforpca$CellType <- metadata$Cell
dfforpca$SGPres <- metadata$SG
dfforpca$StudyNum <- metadata$Study


# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-5)], scale = FALSE)
p_sample<- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Sample") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_stress<- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StressType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StressType", values = custom_palette) +
  ggtitle("PCA - Coloured by StressType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_sg <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "SGPres", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "SGPres", values = custom_palette) +
  ggtitle("PCA - Coloured by SGPres") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_cell <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "CellType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "CellType", values = custom_palette) +
  ggtitle("PCA - Coloured by CellType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_study <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StudyNum", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StudyNum", values = custom_palette) +
  ggtitle("PCA - Coloured by StudyNum") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_sample
p_stress
p_sg
p_cell
p_study


#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])

```

As one can observe from the PCAs, it appears no single variable dominates the variance. However, we can see that samples which theoritically should be the same, PE.WTAR and SE.WTAR do not coalesce together. Therefore, I will treat most variables as "batch-effect", and will remove them. This way, we can compare samples from different cells, as they were the same.



## Batch Effect Correction

```{r}
batch_cols <- c("Study","Cell", "Stress")
metadata_cols <- c("SG")
mmbatch <- model.matrix(reformulate(paste0("~0 + ",paste0( batch_cols,collapse = " + "))), data=metadata)
mmkeep <- model.matrix(reformulate(paste0("~0 + ",paste0( metadata_cols,collapse =  "+" ))), data=metadata)
mm <- cbind(mmkeep,mmbatch)
D0 <- DGEList(2^y$E) #senescence_normcounts
D0 <- calcNormFactors(D0)
y <- voom(D0, mm, plot = F)
fit <- lmFit(y, mm)
beta <- fit$coefficients[,-(1:ncol(mmkeep)),drop=FALSE]
#beta <- fit$coefficients
beta[is.na(beta)] <- 0
corrcounts_v1 <- as.matrix(y$E) - beta %*% t(mmbatch)
offset <- apply(corrcounts_v1,1,min) - apply(y$E,1,min)
corrcounts_v1 <- as.data.frame(corrcounts_v1-offset)
corrcounts_v1 <- 2^corrcounts_v1

mm <- model.matrix(~0+SG, metadata)

y <- voom(corrcounts_v1, mm, plot = T)
```

## Batch Corrected PCA

```{r}
#This prepares my data for the PCA
dfforpca <- y$E
dfforpca <- t(dfforpca)
dfforpca <- as.data.frame(dfforpca) # needs to be a data.frame
dfforpca <- apply(dfforpca, 2, as.numeric) # Ensure numeric
dfforpca <- as.data.frame(dfforpca) # still needs to be a data.frame

# Add metadata, which we will then use to plot
dfforpca$Sample <- sampletype
dfforpca$StressType <- metadata$Stress
dfforpca$CellType <- metadata$Cell
dfforpca$SGPres <- metadata$SG
dfforpca$StudyNum <- metadata$Study


# PCA computation itself. I do one PCA for each variable I want to see.
pca_res <- prcomp(dfforpca[, 1:(ncol(dfforpca)-5)], scale = FALSE)
p_sample<- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "Sample", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_palette) +
  ggtitle("PCA - Coloured by Sample") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_stress<- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StressType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StressType", values = custom_palette) +
  ggtitle("PCA - Coloured by StressType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_sg <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "SGPres", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "SGPres", values = custom_palette) +
  ggtitle("PCA - Coloured by SGPres") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_cell <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "CellType", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "CellType", values = custom_palette) +
  ggtitle("PCA - Coloured by CellType") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_study <- autoplot(pca_res, data = dfforpca, x = 1, y = 2, colour = "StudyNum", size = 8) +
  theme_bw() +
  scale_colour_manual(name = "StudyNum", values = custom_palette) +
  ggtitle("PCA - Coloured by StudyNum") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())

p_sample
p_stress
p_sg
p_cell
p_study


#in case you want to see most important genes for PCA yourself
contribs1 <- pca_res$rotation[,1]
contribs2 <- pca_res$rotation[,2]

#These are the biggest contributors to the PCA1 (right side)
(pos1genes <- sort(contribs1, decreasing=T)[1:10])
#These are the biggest contributors to the PCA1 (left side)
(neg1genes <- sort(contribs1, decreasing=F)[1:10])

#These are the biggest contributors to the PCA2 (up side)
(pos2genes <- sort(contribs2, decreasing=T)[1:10])
#These are the biggest contributors to the PCA2 (down side)
(neg2genes <- sort(contribs2, decreasing=F)[1:10])


#These are my corrected normalised counts, which I then use as this serves as my validation dataset for the signature
write.table(t(dfforpca), "~/Projects/StressGranules/SGPAPER/DiscoveryDataset.txt")
```


## Linear Models

Finally, we move on to differential gene expression! In this chunk, I define the linear model I will actually use for my analyses. 

```{r}
#Now I simply create the linear model that isolates the SG signal
fitted <- lmFit(y$E, mm)

head(coef(fitted))

contr <- makeContrasts(SGYES - SGNO , levels = colnames(coef(fitted)))

tmp_contr <- contrasts.fit(fitted, contr)

tmp <- eBayes(tmp_contr)
top.table <- topTable(tmp, sort.by = "logFC", n=Inf)
DEGenes <- top.table
ndegs=20
```


# SG vs Non-SG

```{r, results='hide', warning=FALSE, message=FALSE}

upmarkers <- as.data.frame(DEGenes[DEGenes$logFC>1.5 & DEGenes$B>0,] )
downmarkers <- as.data.frame(DEGenes[DEGenes$logFC< -1.5 & DEGenes$B>0,] )

ggplot(data=DEGenes,
                 aes(x=logFC,
                     y=  B)) +
  geom_point(color="#CCCCCC")+ geom_vline(xintercept = -0.5, linetype="dashed")+geom_vline(xintercept = 0.5, linetype="dashed")+
  theme_bw() + ylab("B-Statistic")+xlab("Log Fold-Change")+
  theme(axis.line = element_line(colour = "black"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  ggtitle("SG vs No-SG")+
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  geom_label_repel(upmarkers[1:ndegs,] , mapping=aes(label=row.names(upmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#A10702",max.overlaps = 20, fontface="italic")+
  geom_label_repel(downmarkers[1:ndegs,] , mapping=aes(label=row.names(downmarkers)[1:ndegs]),size=4,point.padding=-0.5, color="#03256C",max.overlaps = 20, fontface="italic")+
  geom_point(data=upmarkers, aes(x=logFC, y=  B), color="#A10702",size=4)+
  geom_point(data=downmarkers, aes(x=logFC, y=  B), color="#03256C",size=4)+
  theme(text=element_text(size=15,  family="Arial MS"))

write.table(DEGenes, "~/Projects/StressGranules/SGPAPER/RealSGs/RealWC.txt", row.names=T, col.names = T)
```

From the volcano, we can see downregulation of FOS and IFIT transcripts, which are important in inflammation and stress signalling. 


## Machine Learning 

This is the process that I used to define my signature. This is the workaround that I decided on. Basicly, it keeps me the same number of samples, but randomly selects which samples are those. The idea is that this way I can do bootstrapping, which I consider more valuable in order to maintain robustness. The section is commented because it is slow to run.

```{r}
# expression_data <- t(y$E)
# library(glmnet)
# library(dplyr)
# 
# bootstrap_lasso <- function(data, labels, n_bootstrap = 100, alpha = 0.5) {
#   selected_genes_list <- vector("list", n_bootstrap)
# 
#   for (i in 1:n_bootstrap) {
#     # Create bootstrap sample (no separate test set)
#     boot_indices <- sample(seq_len(nrow(data)), size = nrow(data), replace = TRUE)
#     boot_data <- data[boot_indices, ]
#     boot_labels <- labels[boot_indices]
# 
#     # Fit lasso regression
#     x_boot <- as.matrix(boot_data)
#     y_boot <- as.factor(boot_labels)
# 
#     cv_fit <- cv.glmnet(x_boot, y_boot, family = "binomial", alpha = alpha)
#     best_lambda <- cv_fit$lambda.min
# 
#     lasso_model <- glmnet(x_boot, y_boot, family = "binomial", alpha = alpha, lambda = best_lambda)
#     selected_genes <- rownames(coef(lasso_model))[-1][coef(lasso_model)[-1] != 0]
# 
#     # Store selected genes
#     selected_genes_list[[i]] <- selected_genes
#   }
# 
#   # Combine results into a single data frame
#   all_genes <- unlist(selected_genes_list)
#   gene_counts <- as.data.frame(table(all_genes))
#   colnames(gene_counts) <- c("Gene", "Count")
# 
#   return(gene_counts)
# }
# 
# # Example usage
# gene_counts <- bootstrap_lasso(expression_data, SGPres, n_bootstrap = 1000, alpha=0.5)
# 
# # View genes selected across bootstraps
# print(gene_counts)
# 
# 
# Signature <- subset(gene_counts, Count>=900)
# 
# 
# dfforsig <- y$E
# signature_genes <- Signature$Gene
# Pos <- row.names(DEGenes[row.names(DEGenes) %in% signature_genes & DEGenes$t > 0 ,])
# Neg <- row.names(DEGenes[row.names(DEGenes) %in% signature_genes & DEGenes$t < 0 ,])
# 
# # Calculate scores using only the genes that are present
# scores <- colSums(dfforsig[Pos, , drop = FALSE]) - colSums(dfforsig[Neg, , drop = FALSE])
# 
# library(dendextend)
# 
# # Convert scores to a data frame
# score_df <- data.frame(Sample = names(scores), Score = scores, SG=SGStatus)
# 
# # Create the bar plot with samples ordered by clustering
# ggplot(score_df, aes(x = Sample, y = Score, fill=SGStatus)) +
#   geom_bar(stat = "identity") +
#   labs(title = "Hierarchically Clustered Sample Scores", x = "Sample", y = "Score") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1))
# 
# write.table(score_df, "~/Projects/StressGranules/SGPAPER/SG1.txt", row.names=T, col.names = T)

```

## Expression Graphs (Extra QC)

As an added control, lets plot these genes in regard to their expression.

```{r, fig.height=12, fig.width=12}
finagenes <- rbind(upmarkers, downmarkers)

dfforexpression <- y$E

samplecodes <- data.frame(samples=colnames(dfforexpression)[c(1:29)], purity=c(1:29))
sampletype <- data.frame(samples=colnames(dfforexpression), sampletype=sampletype)
genesforexp <- dfforexpression[row.names(dfforexpression)%in%row.names(finagenes),]
genesforexp <- reshape2::melt(genesforexp)
colnames(genesforexp) <- c("Gene", "samples", "expression")
genesforexp <- merge(genesforexp, samplecodes, by="samples")
genesforexp <- merge(genesforexp, sampletype, by="samples")
genesforexp <- genesforexp[,-1]
genesforexp[,2] <- as.numeric(genesforexp[,2])
genesforexp[,1] <- as.character(genesforexp[,1])
genesforexp <- genesforexp[order(genesforexp$Gene),]

codesforsgpres <- data.frame(sampletype=sampletype$sampletype, sgpresence=SGPres)
genesforexp <- merge(genesforexp, codesforsgpres, by="sampletype")

winnyplot <- ggplot(genesforexp, aes(x=sgpresence, y=expression)) + geom_violin(colour="black", alpha=0.5) + geom_point(aes(colour=sampletype)) + facet_wrap(.~ Gene, ncol=5, scales = "free")
winnyplot <- winnyplot + scale_fill_manual(name = "Sample Type", values = c(custom_palette), breaks = c(unique(sampletype))) + scale_colour_manual(name = "Sample Type", values = c(custom_palette), breaks = c(unique(sampletype))) 
winnyplot <- winnyplot + theme_bw() + xlab("Stress State") + ylab("Gene Expression (Log2 CPM)") + theme(
      strip.text.x = element_text(
        size = 12, face="italic"
        )) +
 theme(
   strip.background = element_rect(
     color="black", fill="#FFFFFF", size=1, linetype="solid"
     )
   ) + theme(
    #legend.position = c(.95, .40),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    ) + theme(legend.title = element_text(size = 12)) + ggtitle("Expression of DEA genes") + theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
winnyplot
```



## GSEA Enrichment

Now that we have our genes of interest, it would also be interesting to see what are the **altered gene sets** between upregulated and downregulated transcripts from the above comparison. Therefore, lets perform a *Gene Set Enrichment Analysis*

```{r, fig.height=20, fig.width=30, results=FALSE}
#for all genes
DEGenes <- DEGenes[order(DEGenes[,3]),]
gseatool <- DEGenes[,3]
gseatool <- as.data.frame(gseatool)
colnames(gseatool)[1] <- "t" 
gseatool[,2] <- row.names(DEGenes)
colnames(gseatool)[2] <- "Symbol" 
gseatool <- gseatool[,c(2,1)]

#for only the genes without the contamination


vectorranks <- gseatool$t
names(vectorranks) <- gseatool$Symbol


#to load our pathways
hallmarks.gs <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP")

hallmarks.gsets <- hallmarks.gs %>% split(x = .$gene_symbol, f = .$gs_name)

for (set in names(hallmarks.gsets)){
  hallmarks.gsets[[set]] <- toupper(hallmarks.gsets[[set]])
}

hallmarks.res <- fgsea(pathways = hallmarks.gsets, 
                       stats = vectorranks)

hallmarks.res.tidy <- hallmarks.res %>%
  as_tibble() %>%
  arrange(desc(NES))

hallmarks.res.tidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj)

hallmark.sg.t <- hallmarks.res.tidy[,c(1,6)]
hallmark.sg.t <- as.data.frame(hallmark.sg.t)

#this small part is generic
thresh <- 0.005

#this one is no longer generic
data_to_plot <- hallmarks.res.tidy %>% filter(padj < thresh)

statvec <- c()
for (i in 1:nrow(data_to_plot)){
  if (data_to_plot[i,6] > 0){
    status <- "Upregulated"
  }
 else {
   status <- "Downregulated"
 }   
statvec <- c(statvec, status)
}

data_to_plot$Status <- statvec

##-----HALMARK-----
gsea_plot_hall <- ggplot(data_to_plot, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= Status), alpha=0.75) + scale_fill_manual(values = c("#53A2BE","#BA3B46"), name = "Regulation Status", labels = c("Downregulated", "Upregulated"), breaks = c("Downregulated", "Upregulated")) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score (NES)",
       title= "Hallmark Pathways",
       caption = paste0("p-value < ", thresh)) + theme(plot.title = element_text(hjust = 1)) +
  theme_bw() + theme(text=element_text(family = "Quicksand")) + theme(legend.position = "none")

gsea_plot_hall+ 
  theme(text=element_text(size=25,  family="Arial MS"))
```

We have a downregulation  in inflammation pathways.

# Session Info

```{r}
sessionInfo()
```
